### 2017-07-30 mbt.tex
MI: Some changes in labels in mbt.tex. New experimental section with results for instances of size 50 from available datasets

% comment
% This is a general template file for the LaTeX package SVJour3mm
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
%\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
%gsave
%newpath
%  20 20 moveto
%  20 220 lineto
%  220 220 lineto
%  220 20 lineto
%closepath
%2 setlinewidth
%gsave
%  .4 setgray fill
%grestore
%stroke
%grestore
%\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
\usepackage[T1]{fontenc}
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
\usepackage{tkz-graph}
\usepackage{amsmath, amssymb}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{amsfonts}
\usepackage{subcaption}
\usepackage{hyperref}
%\usepackage{pdflscape} % Landscape pages
\usepackage{color}	% Different font colors
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{mathtools}
\usepackage{color}
\usepackage[linesnumbered]{algorithm2e}
\SetKwRepeat{Do}{do}{while}

\captionsetup{compatibility=false}

\newtheorem{observation}[theorem]{\textbf{Observation}}

\tikzset{
every edge/.style={fill=none} 
every node/.style={shape=circle,fill=gray!40,draw }}

%\newtheorem{proposition}{Proposition}
%\newtheorem{obsservation}{Observation}
%\newtheorem{corollary}{Corollary}
%newtheorem{lemma}{Lemma}
%\newtheorem{problem}{Problem}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%
\begin{document}

\title{Insert your title here%\thanks{Grants or other notes
%about the article that should go on the front page should be
%placed here. General acknowledgments should be placed at the end of the article.}
}
\subtitle{Do you have a subtitle?\\ If so, write it here}

%\titlerunning{Short form of title}        % if too long for running head

\author{First Author         \and
        Second Author %etc.
}

%\authorrunning{Short form of author list} % if too long for running head

\institute{F. Author \at
              first address \\
              Tel.: +123-45-678910\\
              Fax: +123-45-678910\\
              \email{fauthor@example.com}           %  \\
%             \emph{Present address:} of F. Author  %  if needed
           \and
           S. Author \at
              second address
}

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

\maketitle

\begin{abstract}
\keywords{}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\section{Introduction}
\label{intro}
The minimum broadcast time (MBT) problem consists of a set of communication nodes with a subset of source nodes. 
The task is to disseminate a signal to every node in a shortest possible time (delay), while abiding by communication rules.
An \emph{informed} node is a node that has received the signal.
Otherwise, a node is \emph{uninformed}.
At the beginning, the set of informed nodes is exactly the set of sources.
An informed node can send the signal to an uninformed node if the two nodes are located within a communication vicinity of each other.

The time is divided into a finite number of time steps.
Every informed node can, at each time step, forward the signal to at most one uninformed neighbor node.
Therefore, the number of informed nodes can at most be doubled from one step to the next.
%\subsection{Motivation and Related Work}
%This communication protocol differs from various wireless communication models where a signal can be relayed to all nodes within a visibility range of a sender.
This communication protocol appears in various practical application such as communication among computer processors or telephone networks.
%The applications are however not confined to wired networks.
Situations where the signals have to cover large distances typically assume sending the signal to one neighbor at a time.
This is common in satellite communication.

{Literature overview}

Deciding whether an instance of MBT has a solution with delay at most $t$ has been shown to be NP-complete \cite{slater81}. 
For bipartite planar graphs with maximum degree 3, NP-completeness persists even if $t=2$ or there is only one source \cite{jansen95}.
When $t=2$, the problem also remains NP-complete for cubic planar graphs \cite{middendorf93}, grid graphs with maximum degree 3,
complete grid graphs, chordal graphs, and for split graphs \cite{jansen95}. 
The single-source variant of the decision version of MBT is NP-complete for grid graphs with maximum degree 4, and for chordal graphs \cite{jansen95}.
The problem is known to be polynomial in trees \cite{slater81}.
Whether the problem is polynomial or NP-complete for split graphs with a single source was stated as an open questions in \cite{jansen95}, and has to the best of our knowledge not been answered yet.

A number of inexact methods, for both general and special graph classes, have been proposed in the literature during the last three decades.
One of the first works of this category \cite{scheuermann84} 
introduces an exact dynamic programming algorithm based on generating all maximum matchings in an induced bipartite graph.
Additional contributions of \cite{scheuermann84} are heuristic approaches for near optimal broadcasting.
From more recent works we mention \cite{hasson04}, which describes a meta heuristic algorithm for MBT, and provides a comparison with other existing methods.
The communication model is considered in an existing satellite navigation system in \cite{chu17}, where a greedy inexact methods is proposed together with a non-linear mathematical model.
Examples of additional efficient heuristics can be found e.g. in \cite{harutyunyan06,harutyunyan14,wang10,jimborean13}.

Approximation algorithms for MBT are studied in \cite{kortsarz95}. 
The authors argue that methods presented in \cite{scheuermann84} provide no guarantee on the performance, and show that the $n$-node wheel is an example of an unfavourable instance.
They introduce an $\mathcal{O}(\sqrt{n})$-additive approximation algorithm for broadcasting in general graphs.
They further provide approximation algorithms for several graph classes with small separators with approximation ratio proportional to the separator size times $\log n$.
Another algorithm with $\mathcal{O}\left(\frac{\log n}{\log \log n}\right)$-approximation\footnote{All logarithms in this paper are of base 2} ratio is given in~\cite{elkin03}.
Most of the works cited above consider a single source.

A related problem extensively studied in the literature is the minimum broadcast graph problem \cite{grigni91,mcgarvey16}. 
A broadcast graph supports a broadcast from any node to all other nodes in optimal time $\lceil\log n\rceil$.
For a given integer $n$, the problem is to find a broadcast graph of $n$ nodes such that e.g.\ the number of edges, or the maximum node degree, in the graph is minimized.
The authors of \cite{mcgarvey16} study ILP models for $c$-broadcast graphs, a generalization of the problem that allows transmission of the signal to at most $c$ neighbours in a single time step.

\section{Network Model and Definitions}

The communication network is represented by a connected graph $G=(V,E)$ and a subset $S\subseteq V$ referred to as the set of sources, where $|V|=n$. 
% Let $V\left[F\right]=\{v\in V:\text{ some edge in } F \text{ is incident to } v\}, F\subseteq E$.
% For disjoint node sets $V_1,V_S\subseteq V$, we denote by $G\left[V_1,V_2\right]$ the bipartite graph $(V_1\cup V_2,E')$, where  $E'=\{\{u,v\}\in E: u\in V_1,v\in V_2\}$.

\begin{definition} \label{def:broadcasttime}
The \emph{broadcast time} $\tau(G,S)$ of a node set $S\subseteq V$ in $G$ is defined as the smallest integer $t\geq 0$ for which there exist
a sequence $V_0\subseteq\dots\subseteq V_t$ of node sets and a function $\pi:V\setminus S\to V$, such that:
\begin{enumerate}
  \item $V_0=S$ and $V_t=V$, \label{def:boundary}
  \item for all $v\in V\setminus S$, $\{v,\pi(v)\}\in E$, \label{def:edge}
  \item for all $i=1,\ldots,t$ and all $v\in V_i$, $\pi(v)\in V_{i-1}$, and \label{def:parent}
  \item for all $u,v\in V_i\setminus V_{i-1}$, $\pi(u)=\pi(v)$ only if $u=v$. \label{def:unique}
\end{enumerate}
% % \begin{equation*}
% \tau(G,S)=
% \begin{cases}
% 	0, S=V,\\
% 	1+\min\{\tau(G,S\cup V\left[M\right]):M \text{ is a matching in } G\left[S, V\setminus S\right]\}, S\neq V.
% \end{cases}
% \label{eq:btime}
% \end{equation*}
\end{definition}
% %Broadcasting is defined as a sequence of sets $(S=V_0\subseteq\dots\subseteq V_k = V)$ where each $V_i$ represents the nodes informed after time step $i$, $0\leq i\leq t$.
% %For each node $v\in V_i\setminus V_{i-1}$, there exists a single node $\pi(v)\in V_{i-1}$ adjacent to $v$, which forwarded the signal to $v$.
% %Also, for every two nodes $u,v\in V_i\setminus V_{i-1}$ we have $\pi(u)=\pi(v)$.
% %The value $k$ is referred to as \emph{delay}.

Referring to Section \ref{intro}, the node set $V_i$ is the set of nodes that are informed in time step $i$.
Initially, only the sources are informed ($V_0=S$), whereas all nodes are informed after $t$ time steps ($V_t=V$),
and the set of informed nodes is monotonously non-decreasing ($V_{i_1}\subseteq V_i$ for $i=1,\ldots,t$).
The parent function $\pi$ maps each node to the node from which it received the signal.
Conditions \ref{def:edge}--\ref{def:parent} of Definition \ref{def:broadcasttime} thus reflect that the sender is a neighbor node in $G$,
and that it is informed at an earlier time step than the recipient node.
Because each node can send to at most one neighbor node in each time step,
condition \ref{def:unique} states that $\pi$ maps the set of nodes which are informed for the first time in step $i$ to distinct parent nodes.

The optimization problem in question is formulated as follows:
\begin{problem}[\textsc{Minimum Broadcast Time}]\label{prob:min}
Given $G=(V,E)$ and $S\subseteq V$, find $\tau(G,S)$.
% a sequence $(S=V_0\subseteq\dots\subseteq V_{\tau(G,S)}=V)$ and a function $\pi:V\setminus S\to V$, such that for each $v\in V\setminus S:\{v,\pi(v)\}\in E$, and for each $u,v\in V_i\setminus V_{i-1}: \pi(u)=\pi(v)\Leftrightarrow u=v$.
\end{problem}

\begin{definition} \label{def:broadcastgraph}
For any $V_0,\ldots,V_t$ and $\pi$ satisfying the conditions of Definition \ref{def:broadcasttime},
the corresponding \emph{communication forest} is
the digraph $D=(V,A)$, where $A=\left\{\left(v,\pi(v)\right): v\in V\right\}$.
Each connected component of $D$ is a \emph{communication tree}.
\end{definition}

\noindent
It is easily verified that the communication trees are indeed arborescences, rooted at distinct sources, with arcs pointing away from the source.
Let $T_s=\left(V_{T_s},A_{V_s}\right)$ denote the communication tree rooted at source $s\in S$, and define
the function $\alpha:V\to S$ such that $v\in V_{T_{\alpha(v)}}$ for all $v\in V$.
That is, $\alpha$ associates a node $v$ with the source in which the communication tree containing $v$ is rooted.
For any communication tree $T_s$, let $T^i_s$ be the subtree of $T$ induced by the set of nodes with distance at most $i$ edges from $s$.
Analogously, let $D^i=\left\{T_s^i: s\in S\right\}$.

% Given a feasible sequence of node sets and $\pi$, we derive the following broadcast trees \cite{grigni91}:
% For $s\in S$, a broadcast tree $T_s=(V_{T_s},E_{T_s})$ is a time-labeled directed subgraph of $G$ describing a broadcast originated in $s$ by the rules:
% \begin{enumerate}
% \item $T_s$ is rooted at $s$ with arcs directed towards the leaves.
% \item each node $v\in V_i$ is labeled with an integer $\tau(v)=i$,
% \item for $v\in V\setminus S, v\in V_{T_s}\Leftrightarrow \pi(v)\in V_{T_s}$, 
% \item $E_{T_s}=\{\{u,v\}: u,v\in V_{T_s}, \pi(v)=u\}$, and
% \item for two sources $s,s'\in S, s\neq s'$ we have that $V_{T_s}\cap V_{T_{s'}}=\emptyset$. 
% \end{enumerate}

% Any feasible sequence of node sets and $\pi$ yield a set of trees $T=\{T_s:s\in S\}$ that forms a partition of $G$ into trees referred to as \emph{broadcast forest}.

% Given a broadcast forest $T$, the delay is determined as $\max_{v\in V}\{\tau(v)\}$.
%A set of trees $T=\{T_s:\cup_{s\in S}V_{T_s}=V\}$ forms a partition of $G$ into trees referred to as \emph{broadcast forest}.
%$T$ can be derived from a given sequence of sets of nodes defining broadcasting and the mapping $\pi$.
% For any integer $i$, let $T^i_s$ be the subtree of $T_s$ obtained by pruning all nodes $v\in V_{T_s}$ with $\tau(v)>i$.
% Analogously, we define $T^i=\{T^i_s:s\in S\}$.

%We also define the set $A=\{(i,j),(j,i):\{i,j\}\in E\}$ that consists of all arcs that can be derived by directing edges in $E$.
The degree of node $v$ in graph $G$ is denoted by $\text{deg}_G(v)$.
For the diameter of $G$ we use the symbol $\Delta_G$.
The set of neighbors of $v\in V$ in $G$ is denoted by $N_G(v)$.
Whenever there is no risk of confusion, the subscript $G$ is omitted.
Likewise, $\text{deg}^-_D(v)$ and $\text{deg}^+_D(v)$ denote, respectively, the in-degree and the out-degree of digraph $D$,
and $N_D^+(v)=\left\{u\in V:(v,u)\in A\right\}$ and $N_D^-(v)=\left\{u\in V:(u,v)\in A\right\}$ denote the neighbor sets of node $v$ in $D$.
The node set $L(D)=\left\{v\in V: N_D^+(v)=\emptyset\right\}$ is referred to as the \emph{leaves} in $D$.

%For convenience, we consider the following definition of broadcast trees \cite{grigni91}:
%For $s\in S$, a broadcast tree $T_s$ with node set $V_{T_s}$ is a time-labeled directed subgraph of $G$ describing a broadcast originated by $s$ by the following rules:
%\begin{enumerate}
%\item $T_s$ is rooted at $s$ with arcs directed towards the leaves.
%\item Each node $v$ is labeled with an integer $\tau(v)$, where $\tau(s)=0$.
%\item Whenever $v$ is a parent of $u$ in $T_s$ $\tau(v)<\tau(u)$.
%\item Whenever $v$ and $u$ are siblings in $T_s$, $\tau(v)\neq \tau(u)$. 
%\end{enumerate}
%The number of leaves in $T_s$ is denoted by $L(T_s)$.
\input{sect3-exact-methods.tex}

\section{Lower bounds}
In this section, we study lower bounds on the delay for several restrictions of input graphs.
%An optimal solution is obtained by solving a sequence of decision problems with varying $t$. 
%It is therefore desirable to determine tight lower and upper bounds in order to arrive in the optimum after solving as few decision problems as possible.
Obvious bounds for a general graph instance are given by
\begin{observation}
For an instance $(G,S)$ of Problem \ref{prob:min},
\begin{equation}
\left\lceil\log\frac{n}{|S|}\right\rceil\leq t^* \leq n-|S|.
\label{eq:loglb}
\end{equation}
\label{obs:loglb}
\end{observation}

In the following, we use $m$-step Fibonacci numbers \cite{noe05}, a generalization of well known (2-step) Fibonacci numbers, defined by letting, 
$F^{(m)}_k=0$ for $k\leq 0$, $F^{(m)}_1=1$, and 
other terms according to linear recurrence relation 
\begin{align*}
F^{(m)}_k &=\sum\limits_{i=1}^m F^{(m)}_{k-i}, &\text{ for } k\geq 2.
\end{align*}

Consider a $d$-regular graph with a unique source $s$.
Any broadcast forest consists of a single tree $T_s$.
We investigate the number of leaves in $T_s$, and derive a lower bound on the delay for this graph class.

If the orientation of arcs in $T_s$ is disregarded, $L(T^1_s)=L(T^2_s)=2$.
For $i\geq 3$, $L(T^i_s)$ equals the number of nodes with degree $1,\dots,d-1$ in $T^{i-1}_s$, 
because in a $d$-regular graph, only nodes with degree lower than $d$ can inform new uninformed nodes.
It can also be interpreted as the sum of the number of leaves in $T^{i-d+1}_s,\dots,T^{i-1}_s$, which leads to %the following formula
\begin{equation*}
\label{eq:leafrec}
L(T^i_s)=\sum\limits_{j=i-d+1}^{i-1} L(T^j_s).
\end{equation*}  
This formula equals to the recursive definition of Fibonacci sequence of order $d-1$.
As each of the two base cases, $L(T^1_s)$ and $L(T^2_s)$, equals the double of the base cases of the Fibonacci sequence, the number of leaves in time step $i$ is calculated as
\begin{equation*}
\label{eq:fibleaf}
L(T^i_s)=2 F^{(d-1)}_i.
\end{equation*}  
The number of nodes in $T^i_s$ can be expressed as the sum of nodes newly informed in time steps $1,\dots,i$, that is, the sum of leaves in $T^1_s,\dots,T^i_s$. Thus,
\begin{equation}
\label{eq:fibcnt}
|V_{T^i_s}|=|V_i|=2\sum\limits_{j=1}^i F^{(d-1)}_j.
\end{equation}

\begin{proposition}
For a $d$-regular graph on $n$ nodes and $|S|$ sources, a lower bound on the delay is 
\begin{equation*}
\label{lem:lbreg1}
\underline{t}=\left\lceil\min\{k:2s\sum\limits_{j=1}^k F^{(d-1)}_j\geq n\}\right\rceil.
\end{equation*}
\label{prop:lbfib}
\end{proposition}
\begin{proof}
In order to inform $n$ nodes in the best possible scenario, the signal has to be relayed sufficiently many time steps so that the maximum possible number of informed nodes becomes $n$.
The maximum number of nodes informed within $i$ time steps is given by \eqref{eq:fibcnt}.
We therefore need to set the upper limit of the summation in \eqref{eq:fibcnt} so that the right-hand side exceeds $n$.
The reason why the result is divided by $|S|$ is that the best case scenario with several source nodes assumes that the signals initiated in individual sources are spread evenly.
\qed
\end{proof}

An additional knowledge of a degree sequence of $G$ can be exploited. 
\begin{lemma}
\label{lemma:degorder}
Let $(d_1,\dots,d_n)$ be a degree sequence and $t\leq \bar{t}-|S|$ a given delay.
Consider all possible instances of which graphs are described by this degree sequence.
The maximum number of nodes informed within $t$ time steps is achieved in an instance in which it is possible to inform  nodes in the order of their decreasing degree.

In other words, the instance with the shortest possible broadcast time is such that its nodes can be informed in the order of their decreasing degree.

\end{lemma}
\begin{proof}
A node with degree $d_i$ informed in $i$-th time step can inform at most $\min\{d_i-1,t-i\}$ nodes.
Consider $l,k$ such that $1\leq k < \ell\leq n$ and nodes $v_k$, $v_\ell$ with $\deg(v_k)=d_k$ and $\deg(v_\ell)=d_\ell$.
Then, $d_k-1>d_\ell-1$ and $t-k > t-\ell$.
If nodes are informed in the order of their decreasing degree, let $K_1$ and $L_1$ be the number of nodes that can be informed by $v_k$ and $v_l$, respectively:
$$
K_1=\min\{d_k-1,t-k\}, ~~~ L_1=\min\{d_\ell-1,t-\ell\}.
$$
If the order in which $v_k$ and $v_l$ are informed is switched, i.e. when $v_k$ and $v_\ell$ is informed in the $\ell$-th and $k$-th time step, respectively, 
we use $K_2$ and $L_2$ to denote the maximum number of nodes informed by $v_k$ and $v_\ell$:
$$
K_2=\min\{d_k-1,t-\ell\} ~~~ L_2=\min\{d_\ell-1,t-k\}.
$$
We now investigate what possible values can the expression 
\begin{equation}
\label{eq:degbounds}
(K_1+L_1)-(K_2+L_2)
\end{equation}
attain.
Observe that $K_1\geq L_2$ and $K_2\geq L_1$. 
If $K_2>L_1$, then $K_2=K_1=d_k-1$.
But in that case, $L_2=L_1=d_\ell-1$. 
We conclude that \eqref{eq:degbounds} always takes a non-negative value.
Therefore, if the nodes are informed in the order of their decreasing degrees,
the number of informed nodes within the give time step is no worse than when the nodes are informed in any other order.
\qed
\end{proof}

Alg.~\ref{alg:dreg} calculates a lower bound on the delay when given number of nodes, sources and a degree sequence in $G$ as an input.
The algorithm iteratively updates possible node degrees in $T^k$ in each time step $k$, and records the maximum possible number of nodes in $V_k$.
For the purpose of finding lower bounds, it is assumed that each node $v\in V_k$ with $\deg_{T^k}(v)<\deg_G(v)$ informs a new uninformed node.
The number of iterations is then the lowest possible delay for given input.
%The main idea is that in the most optimistic case, each node $v\in V_i$ with $\text{deg}_{F_i}(v)<\text{deg}_G(v)$ informs new not yet informed node.
%For the purpose of finding lower bounds, we can assume without loss of generality that $\text{deg}_{F_i}(u)\leq\text{deg}_{F_i}(v) \Leftrightarrow \text{deg}_G(u)\geq \text{deg}_G(v)$.
Once a node $v$ reaches its maximum degree, i.e., when $\text{deg}_{T^k}=\deg_G(v)$ for some $k$, $v$ does not inform any other node in the next time steps.

%For each $i=1,\dots d$, Alg. \ref{alg:dreg} keeps the  number of nodes with degree $i$.
%These values are updated iteratively using dynamic programming until the number of informed nodes reaches $n$.

\begin{algorithm}
\KwData{$n,m,d_1,\dots,d_n\in \mathbb{N}, m\leq n,\newline 1\leq d_n\leq\dots\leq d_2\leq d_1$}
$a_1\leftarrow\dots\leftarrow a_{m}\leftarrow 1$\;%\tcp{this is a comment}
$a_{m+1}\leftarrow\dots\leftarrow a_{n}\leftarrow 0$\;%\tcp{this is a comment}
$c\leftarrow m$;%\tcp{every source informs a new node}
$~k\leftarrow 0$\;
\While {$c<n$} {
$k\leftarrow k+1$\;
$c'\leftarrow 0$\;
\For{$i=1,\dots,c$} {
	\If {$a_i<d_i$} {
		$a_i\leftarrow a_i + 1$\;
		$c'\leftarrow c' + 1$\;
		\If {$c+c'\leq n$} {
			$a_{c+c'}\leftarrow 1$;\tcp{Newly informed node} 
		}
	}
}
$c\leftarrow c + c'$\;
}
\Return $k$\;
%\Return $\lceil k/s \rceil$\;
 \caption{Lower bound exploiting distribution of degrees}
\label{alg:dreg}
\end{algorithm}

The input is assumed to be correct in the sense that a graph with the given degree sequence exists, and by definition, the degree sequence is ordered non-increasingly.
For each iteration $k$, Alg.~\ref{alg:dreg} stores degrees of nodes in $T^k$ in variables $a_1,\dots,a_n$.
Note that the forest $T^k$ is not actually constructed. 
The algorithm operates merely with potential degrees of nodes in $T^k$.
Next, variable $c$ keeps the value $|V_k|$, i.e., the number nodes informed within $k$ steps.
Finally, $c'$ stores $|V_k\setminus V_{k-1}|$, thus the number of nodes newly informed in time step $k$.

\begin{proposition}
If $G$ is an arbitrary graph with $n$ nodes, $m$ sources and node degrees $d_i$, $1\leq i\leq n$, Alg.~\ref{alg:dreg} calculates a lower bound on the delay in $G$.
\label{prop:deg}
\end{proposition}
\begin{proof}
The assumption that nodes with higher degree in $G$ are informed first in the optimal solution is justified by Lemma \ref{lemma:degorder}.
In each iteration $k$, degrees in $T^k$ of maximum possible number of nodes are increased in the order of their decreasing degree.
The current maximum possible number of informed nodes is stored in the variable $c$, and is updated at the end of the outre loop.
So, when the condition $c<n$ is tested on line 7, $c$ contains the correct value of number of nodes in $V_k$.
As soon as it is possible that at least $n$ nodes are informed, the algorithm returns the number of necessary iterations.
\qed
\end{proof}

\input{sect5-upper-bounds.tex}
\input{sect6-experimental-results.tex}

%\begin{acknowledgements}
%If you'd like to thank anyone, place your comments here
%and remove the percent signs.
%\end{acknowledgements}

% BibTeX users please use one of
%\bibliographystyle{spbasic}      % basic style, author-year citations
%\bibliographystyle{spmpsci}      % mathematics and physical sciences
%\bibliographystyle{spphys}       % APS-like style for physics
%\bibliography{}   % name your BibTeX data base

% Non-BibTeX users please use
\begin{thebibliography}{}
%
% and use \bibitem to create references. Consult the Instructions
% for authors for reference list style.
%
\bibitem{chu17}
Chu, X., Chen, Y.,
Time division inter‐satellite link topology generation problem: Modeling and solution,
International Journal of Satellite Communications and Networking, 194 -- 206, 36 (2017)

\bibitem{cormen90}
Cormen, T. H., Leiserson, C. E., Rivest, R. L,
Introduction to Algorithms, 
MIT Press, 401 -- 402, 1990.

\bibitem{elkin03}
Elkin, M., Kortsarz, G.,
Sublogarithmic approximation for telephone multicast: path out of jungle,
Symposium on Discrete Algorithms, 76 -- 85 (2003)

\bibitem{farley81}
Farley, A. M., Proskurowski, A.,
Broadcasting in Trees with Multiple Originators,
SIAM Journal on Algebraic Discrete Methods, 381 -- 386, 2, 4 (1981)

\bibitem{grigni91}
Grigni, M., Peleg, D.,
Tight bounds on minimum broadcast networks
Networks, 207-222, 4 (1991)

\bibitem{hasson04} 
Hasson, Y., Sipper, M.,
A Novel Ant Algorithm for Solving the Minimum Broadcast Time Problem,
International Conference on Parallel Problem Solving from Nature, 775 -- 780 (2004)

\bibitem{harutyunyan06}
Harutyunyan, H. A., Shao, B.,
An efficient heuristic for broadcasting in networks,
Journal of Parallel and Distributed Computing, 68 -- 76, 66, 1 (2006)

\bibitem{harutyunyan14}
Harutyunyan, H. A., Jimborean, C.,
New Heuristic for Message Broadcasting in Network,
IEEE 28th International Conference on Advanced Information Networking and Application, 517 -- 524, (2014)

\bibitem{jansen95}
Jansen, K., M\"uller, H.,
The minimum broadcast time problem for several processor networks, 
Theoretical Computer Science, 69 -- 85, 147 (1995)

\bibitem{kortsarz95}
Kortsarz, G., Peleg, D.,
Approximation algorithms for minimum-time broadcast
SIAM Journal on Discrete Mathematics, 401 -- 427, 8, 3 (1995)

\bibitem{mcgarvey16}
McGarvey, R. G., Rieksts, B. Q., Ventura, J. A., Ahn, N.,
Binary linear programming models for robust broadcasting in communication networks,
Discrete Applied Mathematics, 173 -- 84, 204, (2016)

\bibitem{middendorf93}
Middendorf, M.,
Minimum broadcast time is NP-complete for 3-regular planar graphs and deadline 2,
Information Processing Letters, 281 -- 287, 46 (1993)

\bibitem{noe05}
Noe, T. D., Post, J. V., 
Primes in Fibonacci n-step and Lucas n-Step Sequences,
J. Integer Seq. 8, Article 05.4.4, 2005.

\bibitem{scheuermann84}
Scheuermann, P., Wu, G.,
Heuristic Algorithms for Broadcasting in Point-to-Point Computer Networks,
IEEE Transactions on Computers, 804 -- 811, 33, 9 (1984)

\bibitem{datasets}
Singh, A.,
http://dcis.uohyd.ernet.in/~alokcs/dtp.zip

\bibitem{slater81}
Slater, P. J., Cockayne, E. J., Hedetniemi, S.T.,
Information dissemination in Trees,
SIAM Journal on Computing, 692 -- 701, 10, 4 (1981)

\bibitem{wang10}
Wang, W.,
Heuristics for Message Broadcasting in Arbitrary Networks,
Master thesis, Concordia University, Montr\'eal, Qu\'ebec, 
Retrieved from http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.633.5827\&rep=rep1\&type=pdf (2010)

\bibitem{jimborean13}
Jimborean, C.,
New Heuristics for Message Broadcasting in Arbitrary Networks,
Master thesis, Concordia University, Montr\'eal, Qu\'ebec, 
Retrieved from https://spectrum.library.concordia.ca/977717/1/Jimborean\_MCompSc\_F2013.pdf (2013)

% Format for Journal Reference
%Author, Article title, Journal, Volume, page numbers (year)
% Format for books
%\bibitem{RefB}
%Author, Book title, page numbers. Publisher, place (year)
% etc
\end{thebibliography}

\end{document}
% end of file template.tex

### 2017-07-31 mbt.tex
MI: Some sections moved to separate files. New section with experimental results contains objective functin values for dataset instances of size 50 and 100. Concluding section introduced

@@
 \end{abstract}
-
-\section{Introduction}
-\label{intro}
-The minimum broadcast time (MBT) problem consists of a set of communication nodes with a subset of source nodes. 
-The task is to disseminate a signal to every node in a shortest possible time (delay), while abiding by communication rules.
-An \emph{informed} node is a node that has received the signal.
-Otherwise, a node is \emph{uninformed}.
-At the beginning, the set of informed nodes is exactly the set of sources.
-An informed node can send the signal to an uninformed node if the two nodes are located within a communication vicinity of each other.
-
-The time is divided into a finite number of time steps.
-Every informed node can, at each time step, forward the signal to at most one uninformed neighbor node.
-Therefore, the number of informed nodes can at most be doubled from one step to the next.
-%\subsection{Motivation and Related Work}
-%This communication protocol differs from various wireless communication models where a signal can be relayed to all nodes within a visibility range of a sender.
-This communication protocol appears in various practical application such as communication among computer processors or telephone networks.
-%The applications are however not confined to wired networks.
-Situations where the signals have to cover large distances typically assume sending the signal to one neighbor at a time.
-This is common in satellite communication.
-
-{Literature overview}
-
-Deciding whether an instance of MBT has a solution with delay at most $t$ has been shown to be NP-complete \cite{slater81}. 
-For bipartite planar graphs with maximum degree 3, NP-completeness persists even if $t=2$ or there is only one source \cite{jansen95}.
-When $t=2$, the problem also remains NP-complete for cubic planar graphs \cite{middendorf93}, grid graphs with maximum degree 3,
-complete grid graphs, chordal graphs, and for split graphs \cite{jansen95}. 
-The single-source variant of the decision version of MBT is NP-complete for grid graphs with maximum degree 4, and for chordal graphs \cite{jansen95}.
-The problem is known to be polynomial in trees \cite{slater81}.
-Whether the problem is polynomial or NP-complete for split graphs with a single source was stated as an open questions in \cite{jansen95}, and has to the best of our knowledge not been answered yet.
-
-A number of inexact methods, for both general and special graph classes, have been proposed in the literature during the last three decades.
-One of the first works of this category \cite{scheuermann84} 
-introduces an exact dynamic programming algorithm based on generating all maximum matchings in an induced bipartite graph.
-Additional contributions of \cite{scheuermann84} are heuristic approaches for near optimal broadcasting.
-From more recent works we mention \cite{hasson04}, which describes a meta heuristic algorithm for MBT, and provides a comparison with other existing methods.
-The communication model is considered in an existing satellite navigation system in \cite{chu17}, where a greedy inexact methods is proposed together with a non-linear mathematical model.
-Examples of additional efficient heuristics can be found e.g. in \cite{harutyunyan06,harutyunyan14,wang10,jimborean13}.
-
-Approximation algorithms for MBT are studied in \cite{kortsarz95}. 
-The authors argue that methods presented in \cite{scheuermann84} provide no guarantee on the performance, and show that the $n$-node wheel is an example of an unfavourable instance.
-They introduce an $\mathcal{O}(\sqrt{n})$-additive approximation algorithm for broadcasting in general graphs.
-They further provide approximation algorithms for several graph classes with small separators with approximation ratio proportional to the separator size times $\log n$.
-Another algorithm with $\mathcal{O}\left(\frac{\log n}{\log \log n}\right)$-approximation\footnote{All logarithms in this paper are of base 2} ratio is given in~\cite{elkin03}.
-Most of the works cited above consider a single source.
-
-A related problem extensively studied in the literature is the minimum broadcast graph problem \cite{grigni91,mcgarvey16}. 
-A broadcast graph supports a broadcast from any node to all other nodes in optimal time $\lceil\log n\rceil$.
-For a given integer $n$, the problem is to find a broadcast graph of $n$ nodes such that e.g.\ the number of edges, or the maximum node degree, in the graph is minimized.
-The authors of \cite{mcgarvey16} study ILP models for $c$-broadcast graphs, a generalization of the problem that allows transmission of the signal to at most $c$ neighbours in a single time step.
-
+\input{sect1-introduction.tex}
 \section{Network Model and Definitions}
@@
 
-\section{Lower bounds}
-In this section, we study lower bounds on the delay for several restrictions of input graphs.
-%An optimal solution is obtained by solving a sequence of decision problems with varying $t$. 
-%It is therefore desirable to determine tight lower and upper bounds in order to arrive in the optimum after solving as few decision problems as possible.
-Obvious bounds for a general graph instance are given by
-\begin{observation}
-For an instance $(G,S)$ of Problem \ref{prob:min},
-\begin{equation}
-\left\lceil\log\frac{n}{|S|}\right\rceil\leq t^* \leq n-|S|.
-\label{eq:loglb}
-\end{equation}
-\label{obs:loglb}
-\end{observation}
-
-In the following, we use $m$-step Fibonacci numbers \cite{noe05}, a generalization of well known (2-step) Fibonacci numbers, defined by letting, 
-$F^{(m)}_k=0$ for $k\leq 0$, $F^{(m)}_1=1$, and 
-other terms according to linear recurrence relation 
-\begin{align*}
-F^{(m)}_k &=\sum\limits_{i=1}^m F^{(m)}_{k-i}, &\text{ for } k\geq 2.
-\end{align*}
-
-Consider a $d$-regular graph with a unique source $s$.
-Any broadcast forest consists of a single tree $T_s$.
-We investigate the number of leaves in $T_s$, and derive a lower bound on the delay for this graph class.
-
-If the orientation of arcs in $T_s$ is disregarded, $L(T^1_s)=L(T^2_s)=2$.
-For $i\geq 3$, $L(T^i_s)$ equals the number of nodes with degree $1,\dots,d-1$ in $T^{i-1}_s$, 
-because in a $d$-regular graph, only nodes with degree lower than $d$ can inform new uninformed nodes.
-It can also be interpreted as the sum of the number of leaves in $T^{i-d+1}_s,\dots,T^{i-1}_s$, which leads to %the following formula
-\begin{equation*}
-\label{eq:leafrec}
-L(T^i_s)=\sum\limits_{j=i-d+1}^{i-1} L(T^j_s).
-\end{equation*}  
-This formula equals to the recursive definition of Fibonacci sequence of order $d-1$.
-As each of the two base cases, $L(T^1_s)$ and $L(T^2_s)$, equals the double of the base cases of the Fibonacci sequence, the number of leaves in time step $i$ is calculated as
-\begin{equation*}
-\label{eq:fibleaf}
-L(T^i_s)=2 F^{(d-1)}_i.
-\end{equation*}  
-The number of nodes in $T^i_s$ can be expressed as the sum of nodes newly informed in time steps $1,\dots,i$, that is, the sum of leaves in $T^1_s,\dots,T^i_s$. Thus,
-\begin{equation}
-\label{eq:fibcnt}
-|V_{T^i_s}|=|V_i|=2\sum\limits_{j=1}^i F^{(d-1)}_j.
-\end{equation}
-
-\begin{proposition}
-For a $d$-regular graph on $n$ nodes and $|S|$ sources, a lower bound on the delay is 
-\begin{equation*}
-\label{lem:lbreg1}
-\underline{t}=\left\lceil\min\{k:2s\sum\limits_{j=1}^k F^{(d-1)}_j\geq n\}\right\rceil.
-\end{equation*}
-\label{prop:lbfib}
-\end{proposition}
-\begin{proof}
-In order to inform $n$ nodes in the best possible scenario, the signal has to be relayed sufficiently many time steps so that the maximum possible number of informed nodes becomes $n$.
-The maximum number of nodes informed within $i$ time steps is given by \eqref{eq:fibcnt}.
-We therefore need to set the upper limit of the summation in \eqref{eq:fibcnt} so that the right-hand side exceeds $n$.
-The reason why the result is divided by $|S|$ is that the best case scenario with several source nodes assumes that the signals initiated in individual sources are spread evenly.
-\qed
-\end{proof}
-
-An additional knowledge of a degree sequence of $G$ can be exploited. 
-\begin{lemma}
-\label{lemma:degorder}
-Let $(d_1,\dots,d_n)$ be a degree sequence and $t\leq \bar{t}-|S|$ a given delay.
-Consider all possible instances of which graphs are described by this degree sequence.
-The maximum number of nodes informed within $t$ time steps is achieved in an instance in which it is possible to inform  nodes in the order of their decreasing degree.
-
-In other words, the instance with the shortest possible broadcast time is such that its nodes can be informed in the order of their decreasing degree.
-
-\end{lemma}
-\begin{proof}
-A node with degree $d_i$ informed in $i$-th time step can inform at most $\min\{d_i-1,t-i\}$ nodes.
-Consider $l,k$ such that $1\leq k < \ell\leq n$ and nodes $v_k$, $v_\ell$ with $\deg(v_k)=d_k$ and $\deg(v_\ell)=d_\ell$.
-Then, $d_k-1>d_\ell-1$ and $t-k > t-\ell$.
-If nodes are informed in the order of their decreasing degree, let $K_1$ and $L_1$ be the number of nodes that can be informed by $v_k$ and $v_l$, respectively:
-$$
-K_1=\min\{d_k-1,t-k\}, ~~~ L_1=\min\{d_\ell-1,t-\ell\}.
-$$
-If the order in which $v_k$ and $v_l$ are informed is switched, i.e. when $v_k$ and $v_\ell$ is informed in the $\ell$-th and $k$-th time step, respectively, 
-we use $K_2$ and $L_2$ to denote the maximum number of nodes informed by $v_k$ and $v_\ell$:
-$$
-K_2=\min\{d_k-1,t-\ell\} ~~~ L_2=\min\{d_\ell-1,t-k\}.
-$$
-We now investigate what possible values can the expression 
-\begin{equation}
-\label{eq:degbounds}
-(K_1+L_1)-(K_2+L_2)
-\end{equation}
-attain.
-Observe that $K_1\geq L_2$ and $K_2\geq L_1$. 
-If $K_2>L_1$, then $K_2=K_1=d_k-1$.
-But in that case, $L_2=L_1=d_\ell-1$. 
-We conclude that \eqref{eq:degbounds} always takes a non-negative value.
-Therefore, if the nodes are informed in the order of their decreasing degrees,
-the number of informed nodes within the give time step is no worse than when the nodes are informed in any other order.
-\qed
-\end{proof}
-
-Alg.~\ref{alg:dreg} calculates a lower bound on the delay when given number of nodes, sources and a degree sequence in $G$ as an input.
-The algorithm iteratively updates possible node degrees in $T^k$ in each time step $k$, and records the maximum possible number of nodes in $V_k$.
-For the purpose of finding lower bounds, it is assumed that each node $v\in V_k$ with $\deg_{T^k}(v)<\deg_G(v)$ informs a new uninformed node.
-The number of iterations is then the lowest possible delay for given input.
-%The main idea is that in the most optimistic case, each node $v\in V_i$ with $\text{deg}_{F_i}(v)<\text{deg}_G(v)$ informs new not yet informed node.
-%For the purpose of finding lower bounds, we can assume without loss of generality that $\text{deg}_{F_i}(u)\leq\text{deg}_{F_i}(v) \Leftrightarrow \text{deg}_G(u)\geq \text{deg}_G(v)$.
-Once a node $v$ reaches its maximum degree, i.e., when $\text{deg}_{T^k}=\deg_G(v)$ for some $k$, $v$ does not inform any other node in the next time steps.
-
-%For each $i=1,\dots d$, Alg. \ref{alg:dreg} keeps the  number of nodes with degree $i$.
-%These values are updated iteratively using dynamic programming until the number of informed nodes reaches $n$.
-
-\begin{algorithm}
-\KwData{$n,m,d_1,\dots,d_n\in \mathbb{N}, m\leq n,\newline 1\leq d_n\leq\dots\leq d_2\leq d_1$}
-$a_1\leftarrow\dots\leftarrow a_{m}\leftarrow 1$\;%\tcp{this is a comment}
-$a_{m+1}\leftarrow\dots\leftarrow a_{n}\leftarrow 0$\;%\tcp{this is a comment}
-$c\leftarrow m$;%\tcp{every source informs a new node}
-$~k\leftarrow 0$\;
-\While {$c<n$} {
-$k\leftarrow k+1$\;
-$c'\leftarrow 0$\;
-\For{$i=1,\dots,c$} {
-	\If {$a_i<d_i$} {
-		$a_i\leftarrow a_i + 1$\;
-		$c'\leftarrow c' + 1$\;
-		\If {$c+c'\leq n$} {
-			$a_{c+c'}\leftarrow 1$;\tcp{Newly informed node} 
-		}
-	}
-}
-$c\leftarrow c + c'$\;
-}
-\Return $k$\;
-%\Return $\lceil k/s \rceil$\;
- \caption{Lower bound exploiting distribution of degrees}
-\label{alg:dreg}
-\end{algorithm}
-
-The input is assumed to be correct in the sense that a graph with the given degree sequence exists, and by definition, the degree sequence is ordered non-increasingly.
-For each iteration $k$, Alg.~\ref{alg:dreg} stores degrees of nodes in $T^k$ in variables $a_1,\dots,a_n$.
-Note that the forest $T^k$ is not actually constructed. 
-The algorithm operates merely with potential degrees of nodes in $T^k$.
-Next, variable $c$ keeps the value $|V_k|$, i.e., the number nodes informed within $k$ steps.
-Finally, $c'$ stores $|V_k\setminus V_{k-1}|$, thus the number of nodes newly informed in time step $k$.
-
-\begin{proposition}
-If $G$ is an arbitrary graph with $n$ nodes, $m$ sources and node degrees $d_i$, $1\leq i\leq n$, Alg.~\ref{alg:dreg} calculates a lower bound on the delay in $G$.
-\label{prop:deg}
-\end{proposition}
-\begin{proof}
-The assumption that nodes with higher degree in $G$ are informed first in the optimal solution is justified by Lemma \ref{lemma:degorder}.
-In each iteration $k$, degrees in $T^k$ of maximum possible number of nodes are increased in the order of their decreasing degree.
-The current maximum possible number of informed nodes is stored in the variable $c$, and is updated at the end of the outre loop.
-So, when the condition $c<n$ is tested on line 7, $c$ contains the correct value of number of nodes in $V_k$.
-As soon as it is possible that at least $n$ nodes are informed, the algorithm returns the number of necessary iterations.
-\qed
-\end{proof}
-
+\input{sect4-lower-bounds.tex}
 \input{sect5-upper-bounds.tex}
 \input{sect6-experimental-results.tex}
+\input{sect7-conclusion.tex}
 

### 2017-07-31 sect1-introduction.tex
MI: Some sections moved to separate files. New section with experimental results contains objective functin values for dataset instances of size 50 and 100. Concluding section introduced

\section{Introduction}
\label{intro}
The minimum broadcast time (MBT) problem consists of a set of communication nodes with a subset of source nodes. 
The task is to disseminate a signal to every node in a shortest possible time (delay), while abiding by communication rules.
An \emph{informed} node is a node that has received the signal.
Otherwise, a node is \emph{uninformed}.
At the beginning, the set of informed nodes is exactly the set of sources.
An informed node can send the signal to an uninformed node if the two nodes are located within a communication vicinity of each other.

The time is divided into a finite number of time steps.
Every informed node can, at each time step, forward the signal to at most one uninformed neighbor node.
Therefore, the number of informed nodes can at most be doubled from one step to the next.
%\subsection{Motivation and Related Work}
%This communication protocol differs from various wireless communication models where a signal can be relayed to all nodes within a visibility range of a sender.
This communication protocol appears in various practical application such as communication among computer processors or telephone networks.
%The applications are however not confined to wired networks.
Situations where the signals have to cover large distances typically assume sending the signal to one neighbor at a time.
This is common in satellite communication.

{Literature overview}

Deciding whether an instance of MBT has a solution with delay at most $t$ has been shown to be NP-complete \cite{slater81}. 
For bipartite planar graphs with maximum degree 3, NP-completeness persists even if $t=2$ or there is only one source \cite{jansen95}.
When $t=2$, the problem also remains NP-complete for cubic planar graphs \cite{middendorf93}, grid graphs with maximum degree 3,
complete grid graphs, chordal graphs, and for split graphs \cite{jansen95}. 
The single-source variant of the decision version of MBT is NP-complete for grid graphs with maximum degree 4, and for chordal graphs \cite{jansen95}.
The problem is known to be polynomial in trees \cite{slater81}.
Whether the problem is polynomial or NP-complete for split graphs with a single source was stated as an open questions in \cite{jansen95}, and has to the best of our knowledge not been answered yet.

A number of inexact methods, for both general and special graph classes, have been proposed in the literature during the last three decades.
One of the first works of this category \cite{scheuermann84} 
introduces an exact dynamic programming algorithm based on generating all maximum matchings in an induced bipartite graph.
Additional contributions of \cite{scheuermann84} are heuristic approaches for near optimal broadcasting.
From more recent works we mention \cite{hasson04}, which describes a meta heuristic algorithm for MBT, and provides a comparison with other existing methods.
The communication model is considered in an existing satellite navigation system in \cite{chu17}, where a greedy inexact methods is proposed together with a non-linear mathematical model.
Examples of additional efficient heuristics can be found e.g. in \cite{harutyunyan06,harutyunyan14,wang10,jimborean13}.

Approximation algorithms for MBT are studied in \cite{kortsarz95}. 
The authors argue that methods presented in \cite{scheuermann84} provide no guarantee on the performance, and show that the $n$-node wheel is an example of an unfavourable instance.
They introduce an $\mathcal{O}(\sqrt{n})$-additive approximation algorithm for broadcasting in general graphs.
They further provide approximation algorithms for several graph classes with small separators with approximation ratio proportional to the separator size times $\log n$.
Another algorithm with $\mathcal{O}\left(\frac{\log n}{\log \log n}\right)$-approximation\footnote{All logarithms in this paper are of base 2} ratio is given in~\cite{elkin03}.
Most of the works cited above consider a single source.

A related problem extensively studied in the literature is the minimum broadcast graph problem \cite{grigni91,mcgarvey16}. 
A broadcast graph supports a broadcast from any node to all other nodes in optimal time $\lceil\log n\rceil$.
For a given integer $n$, the problem is to find a broadcast graph of $n$ nodes such that e.g.\ the number of edges, or the maximum node degree, in the graph is minimized.
The authors of \cite{mcgarvey16} study ILP models for $c$-broadcast graphs, a generalization of the problem that allows transmission of the signal to at most $c$ neighbours in a single time step.

### 2017-07-31 sect4-lower-bounds.tex
MI: Some sections moved to separate files. New section with experimental results contains objective functin values for dataset instances of size 50 and 100. Concluding section introduced

\section{Lower bounds}
In this section, we study lower bounds on the delay for several restrictions of input graphs.
%An optimal solution is obtained by solving a sequence of decision problems with varying $t$. 
%It is therefore desirable to determine tight lower and upper bounds in order to arrive in the optimum after solving as few decision problems as possible.
Obvious bounds for a general graph instance are given by
\begin{observation}
For an instance $(G,S)$ of Problem \ref{prob:min},
\begin{equation}
\left\lceil\log\frac{n}{|S|}\right\rceil\leq t^* \leq n-|S|.
\label{eq:loglb}
\end{equation}
\label{obs:loglb}
\end{observation}

In the following, we use $m$-step Fibonacci numbers \cite{noe05}, a generalization of well known (2-step) Fibonacci numbers, defined by letting, 
$F^{(m)}_k=0$ for $k\leq 0$, $F^{(m)}_1=1$, and 
other terms according to linear recurrence relation 
\begin{align*}
F^{(m)}_k &=\sum\limits_{i=1}^m F^{(m)}_{k-i}, &\text{ for } k\geq 2.
\end{align*}

Consider a $d$-regular graph with a unique source $s$.
Any broadcast forest consists of a single tree $T_s$.
We investigate the number of leaves in $T_s$, and derive a lower bound on the delay for this graph class.

If the orientation of arcs in $T_s$ is disregarded, $L(T^1_s)=L(T^2_s)=2$.
For $i\geq 3$, $L(T^i_s)$ equals the number of nodes with degree $1,\dots,d-1$ in $T^{i-1}_s$, 
because in a $d$-regular graph, only nodes with degree lower than $d$ can inform new uninformed nodes.
It can also be interpreted as the sum of the number of leaves in $T^{i-d+1}_s,\dots,T^{i-1}_s$, which leads to %the following formula
\begin{equation*}
\label{eq:leafrec}
L(T^i_s)=\sum\limits_{j=i-d+1}^{i-1} L(T^j_s).
\end{equation*}  
This formula equals to the recursive definition of Fibonacci sequence of order $d-1$.
As each of the two base cases, $L(T^1_s)$ and $L(T^2_s)$, equals the double of the base cases of the Fibonacci sequence, the number of leaves in time step $i$ is calculated as
\begin{equation*}
\label{eq:fibleaf}
L(T^i_s)=2 F^{(d-1)}_i.
\end{equation*}  
The number of nodes in $T^i_s$ can be expressed as the sum of nodes newly informed in time steps $1,\dots,i$, that is, the sum of leaves in $T^1_s,\dots,T^i_s$. Thus,
\begin{equation}
\label{eq:fibcnt}
|V_{T^i_s}|=|V_i|=2\sum\limits_{j=1}^i F^{(d-1)}_j.
\end{equation}

\begin{proposition}
For a $d$-regular graph on $n$ nodes and $|S|$ sources, a lower bound on the delay is 
\begin{equation*}
\label{lem:lbreg1}
\underline{t}=\left\lceil\min\{k:2s\sum\limits_{j=1}^k F^{(d-1)}_j\geq n\}\right\rceil.
\end{equation*}
\label{prop:lbfib}
\end{proposition}
\begin{proof}
In order to inform $n$ nodes in the best possible scenario, the signal has to be relayed sufficiently many time steps so that the maximum possible number of informed nodes becomes $n$.
The maximum number of nodes informed within $i$ time steps is given by \eqref{eq:fibcnt}.
We therefore need to set the upper limit of the summation in \eqref{eq:fibcnt} so that the right-hand side exceeds $n$.
The reason why the result is divided by $|S|$ is that the best case scenario with several source nodes assumes that the signals initiated in individual sources are spread evenly.
\qed
\end{proof}

An additional knowledge of a degree sequence of $G$ can be exploited. 
\begin{lemma}
\label{lemma:degorder}
Let $(d_1,\dots,d_n)$ be a degree sequence and $t\leq \bar{t}-|S|$ a given delay.
Consider all possible instances of which graphs are described by this degree sequence.
The maximum number of nodes informed within $t$ time steps is achieved in an instance in which it is possible to inform  nodes in the order of their decreasing degree.

In other words, the instance with the shortest possible broadcast time is such that its nodes can be informed in the order of their decreasing degree.

\end{lemma}
\begin{proof}
A node with degree $d_i$ informed in $i$-th time step can inform at most $\min\{d_i-1,t-i\}$ nodes.
Consider $l,k$ such that $1\leq k < \ell\leq n$ and nodes $v_k$, $v_\ell$ with $\deg(v_k)=d_k$ and $\deg(v_\ell)=d_\ell$.
Then, $d_k-1>d_\ell-1$ and $t-k > t-\ell$.
If nodes are informed in the order of their decreasing degree, let $K_1$ and $L_1$ be the number of nodes that can be informed by $v_k$ and $v_l$, respectively:
$$
K_1=\min\{d_k-1,t-k\}, ~~~ L_1=\min\{d_\ell-1,t-\ell\}.
$$
If the order in which $v_k$ and $v_l$ are informed is switched, i.e. when $v_k$ and $v_\ell$ is informed in the $\ell$-th and $k$-th time step, respectively, 
we use $K_2$ and $L_2$ to denote the maximum number of nodes informed by $v_k$ and $v_\ell$:
$$
K_2=\min\{d_k-1,t-\ell\} ~~~ L_2=\min\{d_\ell-1,t-k\}.
$$
We now investigate what possible values can the expression 
\begin{equation}
\label{eq:degbounds}
(K_1+L_1)-(K_2+L_2)
\end{equation}
attain.
Observe that $K_1\geq L_2$ and $K_2\geq L_1$. 
If $K_2>L_1$, then $K_2=K_1=d_k-1$.
But in that case, $L_2=L_1=d_\ell-1$. 
We conclude that \eqref{eq:degbounds} always takes a non-negative value.
Therefore, if the nodes are informed in the order of their decreasing degrees,
the number of informed nodes within the give time step is no worse than when the nodes are informed in any other order.
\qed
\end{proof}

Alg.~\ref{alg:dreg} calculates a lower bound on the delay when given number of nodes, sources and a degree sequence in $G$ as an input.
The algorithm iteratively updates possible node degrees in $T^k$ in each time step $k$, and records the maximum possible number of nodes in $V_k$.
For the purpose of finding lower bounds, it is assumed that each node $v\in V_k$ with $\deg_{T^k}(v)<\deg_G(v)$ informs a new uninformed node.
The number of iterations is then the lowest possible delay for given input.
%The main idea is that in the most optimistic case, each node $v\in V_i$ with $\text{deg}_{F_i}(v)<\text{deg}_G(v)$ informs new not yet informed node.
%For the purpose of finding lower bounds, we can assume without loss of generality that $\text{deg}_{F_i}(u)\leq\text{deg}_{F_i}(v) \Leftrightarrow \text{deg}_G(u)\geq \text{deg}_G(v)$.
Once a node $v$ reaches its maximum degree, i.e., when $\text{deg}_{T^k}=\deg_G(v)$ for some $k$, $v$ does not inform any other node in the next time steps.

%For each $i=1,\dots d$, Alg. \ref{alg:dreg} keeps the  number of nodes with degree $i$.
%These values are updated iteratively using dynamic programming until the number of informed nodes reaches $n$.

\begin{algorithm}
\KwData{$n,m,d_1,\dots,d_n\in \mathbb{N}, m\leq n,\newline 1\leq d_n\leq\dots\leq d_2\leq d_1$}
$a_1\leftarrow\dots\leftarrow a_{m}\leftarrow 1$\;%\tcp{this is a comment}
$a_{m+1}\leftarrow\dots\leftarrow a_{n}\leftarrow 0$\;%\tcp{this is a comment}
$c\leftarrow m$;%\tcp{every source informs a new node}
$~k\leftarrow 0$\;
\While {$c<n$} {
$k\leftarrow k+1$\;
$c'\leftarrow 0$\;
\For{$i=1,\dots,c$} {
	\If {$a_i<d_i$} {
		$a_i\leftarrow a_i + 1$\;
		$c'\leftarrow c' + 1$\;
		\If {$c+c'\leq n$} {
			$a_{c+c'}\leftarrow 1$;\tcp{Newly informed node} 
		}
	}
}
$c\leftarrow c + c'$\;
}
\Return $k$\;
%\Return $\lceil k/s \rceil$\;
 \caption{Lower bound exploiting distribution of degrees}
\label{alg:dreg}
\end{algorithm}

The input is assumed to be correct in the sense that a graph with the given degree sequence exists, and by definition, the degree sequence is ordered non-increasingly.
For each iteration $k$, Alg.~\ref{alg:dreg} stores degrees of nodes in $T^k$ in variables $a_1,\dots,a_n$.
Note that the forest $T^k$ is not actually constructed. 
The algorithm operates merely with potential degrees of nodes in $T^k$.
Next, variable $c$ keeps the value $|V_k|$, i.e., the number nodes informed within $k$ steps.
Finally, $c'$ stores $|V_k\setminus V_{k-1}|$, thus the number of nodes newly informed in time step $k$.

\begin{proposition}
If $G$ is an arbitrary graph with $n$ nodes, $m$ sources and node degrees $d_i$, $1\leq i\leq n$, Alg.~\ref{alg:dreg} calculates a lower bound on the delay in $G$.
\label{prop:deg}
\end{proposition}
\begin{proof}
The assumption that nodes with higher degree in $G$ are informed first in the optimal solution is justified by Lemma \ref{lemma:degorder}.
In each iteration $k$, degrees in $T^k$ of maximum possible number of nodes are increased in the order of their decreasing degree.
The current maximum possible number of informed nodes is stored in the variable $c$, and is updated at the end of the outre loop.
So, when the condition $c<n$ is tested on line 7, $c$ contains the correct value of number of nodes in $V_k$.
As soon as it is possible that at least $n$ nodes are informed, the algorithm returns the number of necessary iterations.
\qed
\end{proof}

### 2018-06-26 mbt.tex
Initial commit in the repository. Created a .gitignore file for not tracking latex output files.

--- 2017-07-30 mbt.tex
@@
-% comment
 % This is a general template file for the LaTeX package SVJour3mm
@@
 % your LaTeX will extract the file if required
-%\begin{filecontents*}{example.eps}
+\begin{filecontents*}{example.eps}
 %!PS-Adobe-3.0 EPSF-3.0
@@
 %%EndComments
-%gsave
-%newpath
-%  20 20 moveto
-%  20 220 lineto
-%  220 220 lineto
-%  220 20 lineto
-%closepath
-%2 setlinewidth
-%gsave
-%  .4 setgray fill
-%grestore
-%stroke
-%grestore
-%\end{filecontents*}
+gsave
+newpath
+  20 20 moveto
+  20 220 lineto
+  220 220 lineto
+  220 20 lineto
+closepath
+2 setlinewidth
+gsave
+  .4 setgray fill
+grestore
+stroke
+grestore
+\end{filecontents*}
 %
@@
 \documentclass[smallextended]{svjour3}       % onecolumn (second format)
-\usepackage[T1]{fontenc}
 %\documentclass[twocolumn]{svjour3}          % twocolumn
@@
 \usepackage{amsmath, amssymb}
-\usepackage{multirow}
 \usepackage{graphicx}
@@
 The minimum broadcast time (MBT) problem consists of a set of communication nodes with a subset of source nodes. 
-The task is to disseminate a signal to every node in a shortest possible time (delay), while abiding by communication rules.
+The task is to disseminate a signal to every node in a shortest possible time $t^*$ (delay), while abiding by communication rules.
 An \emph{informed} node is a node that has received the signal.
@@
 At the beginning, the set of informed nodes is exactly the set of sources.
-An informed node can send the signal to an uninformed node if the two nodes are located within a communication vicinity of each other.
+An informed node $u$ can send the signal to an uninformed node $v$ if $u$ and $v$ are located within a communication vicinity of each other.
 
-The time is divided into a finite number of time steps.
-Every informed node can, at each time step, forward the signal to at most one uninformed neighbor node.
-Therefore, the number of informed nodes can at most be doubled from one step to the next.
+The time is divided into discrete time steps.
+At each time step, every informed node can forward the signal to at most one uninformed neighbor.
+The number of informed nodes at some time step can therefore be no more than the double of the number of informed nodes at the previous time step.
 %\subsection{Motivation and Related Work}
@@
 
-{Literature overview}
+\subsection{Literature overview}
 
-Deciding whether an instance of MBT has a solution with delay at most $t$ has been shown to be NP-complete \cite{slater81}. 
-For bipartite planar graphs with maximum degree 3, NP-completeness persists even if $t=2$ or there is only one source \cite{jansen95}.
-When $t=2$, the problem also remains NP-complete for cubic planar graphs \cite{middendorf93}, grid graphs with maximum degree 3,
-complete grid graphs, chordal graphs, and for split graphs \cite{jansen95}. 
-The single-source variant of the decision version of MBT is NP-complete for grid graphs with maximum degree 4, and for chordal graphs \cite{jansen95}.
+MBT has been shown to be NP-complete on arbitrary graphs in \cite{slater81}. 
+The problem remains NP-complete even for several restricted communication networks \cite{jansen95} such as
+split graphs with delay $t^*=2$, chordal graphs,
+bipartite and planar graphs with maximum degree 3 and a single source,
+several variants of grid graphs,
+bipartite planar graphs with delay $t^*=2$ and maximum degree 3. 
+Yet another NP-completeness result for 3-regular planar graphs with delay 2 is shown in \cite{middendorf93}.
 The problem is known to be polynomial in trees \cite{slater81}.
-Whether the problem is polynomial or NP-complete for split graphs with a single source was stated as an open questions in \cite{jansen95}, and has to the best of our knowledge not been answered yet.
+Whether the problem is polynomial or NP-complete for split graphs with a single source was stated as an open questions in \cite{jansen95}, and hasn't been answered yet to our best knowledge.
 
-A number of inexact methods, for both general and special graph classes, have been proposed in the literature during the last three decades.
-One of the first works of this category \cite{scheuermann84} 
-introduces an exact dynamic programming algorithm based on generating all maximum matchings in an induced bipartite graph.
-Additional contributions of \cite{scheuermann84} are heuristic approaches for near optimal broadcasting.
+A number of inexact methods for both general and special graph classes has been proposed in the literature during the last three decades.
+One of the first works of this category is \cite{scheuermann84}, 
+where the authors introduce an exact dynamic programming algorithm based on generating all maximum matchings in an induced bipartite graph.
+Additional contribution of \cite{scheuermann84} are heuristic approaches for near optimal broadcasting.
 From more recent works we mention \cite{hasson04}, which describes a meta heuristic algorithm for MBT, and provides a comparison with other existing methods.
@@
 Approximation algorithms for MBT are studied in \cite{kortsarz95}. 
-The authors argue that methods presented in \cite{scheuermann84} provide no guarantee on the performance, and show that the $n$-node wheel is an example of an unfavourable instance.
+The authors argue that methods presented in \cite{scheuermann84} provide no guarantee on the performance and state an $n$-node wheel as an example of an unfavourable instance.
 They introduce an $\mathcal{O}(\sqrt{n})$-additive approximation algorithm for broadcasting in general graphs.
@@
 Another algorithm with $\mathcal{O}\left(\frac{\log n}{\log \log n}\right)$-approximation\footnote{All logarithms in this paper are of base 2} ratio is given in~\cite{elkin03}.
-Most of the works cited above consider a single source.
+Most of the works above consider a single source.
 
-A related problem extensively studied in the literature is the minimum broadcast graph problem \cite{grigni91,mcgarvey16}. 
+A related problem extensively studied in the literature is the minimum broadcast graph \cite{grigni91,mcgarvey16}. 
 A broadcast graph supports a broadcast from any node to all other nodes in optimal time $\lceil\log n\rceil$.
-For a given integer $n$, the problem is to find a broadcast graph of $n$ nodes such that e.g.\ the number of edges, or the maximum node degree, in the graph is minimized.
+For a given $n$, the effort is to determine properties such as the minimum number of edges and the minimum degree in broadcast graph on $n$ nodes.
 The authors of \cite{mcgarvey16} study ILP models for $c$-broadcast graphs, a generalization of the problem that allows transmission of the signal to at most $c$ neighbours in a single time step.
 
-\section{Network Model and Definitions}
+\section{Network Model and Notation}
 
 The communication network is represented by a connected graph $G=(V,E)$ and a subset $S\subseteq V$ referred to as the set of sources, where $|V|=n$. 
-% Let $V\left[F\right]=\{v\in V:\text{ some edge in } F \text{ is incident to } v\}, F\subseteq E$.
-% For disjoint node sets $V_1,V_S\subseteq V$, we denote by $G\left[V_1,V_2\right]$ the bipartite graph $(V_1\cup V_2,E')$, where  $E'=\{\{u,v\}\in E: u\in V_1,v\in V_2\}$.
+Let $V\left[F\right]=\{v\in V:\text{ some edge in } F \text{ is incident to } v\}, F\subseteq E$.
+For disjoint node sets $V_1,V_S\subseteq V$, we denote by $G\left[V_1,V_2\right]$ the bipartite graph $(V_1\cup V_2,E')$, where  $E'=\{\{u,v\}\in E: u\in V_1,v\in V_2\}$.
 
-\begin{definition} \label{def:broadcasttime}
-The \emph{broadcast time} $\tau(G,S)$ of a node set $S\subseteq V$ in $G$ is defined as the smallest integer $t\geq 0$ for which there exist
-a sequence $V_0\subseteq\dots\subseteq V_t$ of node sets and a function $\pi:V\setminus S\to V$, such that:
-\begin{enumerate}
-  \item $V_0=S$ and $V_t=V$, \label{def:boundary}
-  \item for all $v\in V\setminus S$, $\{v,\pi(v)\}\in E$, \label{def:edge}
-  \item for all $i=1,\ldots,t$ and all $v\in V_i$, $\pi(v)\in V_{i-1}$, and \label{def:parent}
-  \item for all $u,v\in V_i\setminus V_{i-1}$, $\pi(u)=\pi(v)$ only if $u=v$. \label{def:unique}
-\end{enumerate}
-% % \begin{equation*}
-% \tau(G,S)=
-% \begin{cases}
-% 	0, S=V,\\
-% 	1+\min\{\tau(G,S\cup V\left[M\right]):M \text{ is a matching in } G\left[S, V\setminus S\right]\}, S\neq V.
-% \end{cases}
-% \label{eq:btime}
-% \end{equation*}
+\begin{definition}
+The \emph{broadcast time} $\tau(G,S)$ of a node set $S\subseteq V$ in $G$ is defined as: 
+\begin{equation*}
+\tau(G,S)=
+\begin{cases}
+	0, S=V,\\
+	1+\min\{\tau(G,S\cup V\left[M\right]):M \text{ is a matching in } G\left[S, V\setminus S\right]\}, S\neq V.
+\end{cases}
+\label{eq:btime}
+\end{equation*}
 \end{definition}
-% %Broadcasting is defined as a sequence of sets $(S=V_0\subseteq\dots\subseteq V_k = V)$ where each $V_i$ represents the nodes informed after time step $i$, $0\leq i\leq t$.
-% %For each node $v\in V_i\setminus V_{i-1}$, there exists a single node $\pi(v)\in V_{i-1}$ adjacent to $v$, which forwarded the signal to $v$.
-% %Also, for every two nodes $u,v\in V_i\setminus V_{i-1}$ we have $\pi(u)=\pi(v)$.
-% %The value $k$ is referred to as \emph{delay}.
-
-Referring to Section \ref{intro}, the node set $V_i$ is the set of nodes that are informed in time step $i$.
-Initially, only the sources are informed ($V_0=S$), whereas all nodes are informed after $t$ time steps ($V_t=V$),
-and the set of informed nodes is monotonously non-decreasing ($V_{i_1}\subseteq V_i$ for $i=1,\ldots,t$).
-The parent function $\pi$ maps each node to the node from which it received the signal.
-Conditions \ref{def:edge}--\ref{def:parent} of Definition \ref{def:broadcasttime} thus reflect that the sender is a neighbor node in $G$,
-and that it is informed at an earlier time step than the recipient node.
-Because each node can send to at most one neighbor node in each time step,
-condition \ref{def:unique} states that $\pi$ maps the set of nodes which are informed for the first time in step $i$ to distinct parent nodes.
-
-The optimization problem in question is formulated as follows:
-\begin{problem}[\textsc{Minimum Broadcast Time}]\label{prob:min}
-Given $G=(V,E)$ and $S\subseteq V$, find $\tau(G,S)$.
-% a sequence $(S=V_0\subseteq\dots\subseteq V_{\tau(G,S)}=V)$ and a function $\pi:V\setminus S\to V$, such that for each $v\in V\setminus S:\{v,\pi(v)\}\in E$, and for each $u,v\in V_i\setminus V_{i-1}: \pi(u)=\pi(v)\Leftrightarrow u=v$.
+%Broadcasting is defined as a sequence of sets $(S=V_0\subseteq\dots\subseteq V_k = V)$ where each $V_i$ represents the nodes informed after time step $i$, $0\leq i\leq t$.
+%For each node $v\in V_i\setminus V_{i-1}$, there exists a single node $\pi(v)\in V_{i-1}$ adjacent to $v$, which forwarded the signal to $v$.
+%Also, for every two nodes $u,v\in V_i\setminus V_{i-1}$ we have $\pi(u)=\pi(v)$.
+%The value $k$ is referred to as \emph{delay}.
+The optimization problem in question is formulated as follows \cite{jansen95,middendorf93}:
+\begin{problem}\label{prob:min}
+Given $G=(V,E)$ and $S\subseteq V$, find a sequence $(S=V_0\subseteq\dots\subseteq V_{\tau(G,S)}=V)$ and a function $\pi:V\setminus S\to V$, such that for each $v\in V\setminus S:\{v,\pi(v)\}\in E$, and for each $u,v\in V_i\setminus V_{i-1}: \pi(u)=\pi(v)\Leftrightarrow u=v$.
 \end{problem}
 
-\begin{definition} \label{def:broadcastgraph}
-For any $V_0,\ldots,V_t$ and $\pi$ satisfying the conditions of Definition \ref{def:broadcasttime},
-the corresponding \emph{communication forest} is
-the digraph $D=(V,A)$, where $A=\left\{\left(v,\pi(v)\right): v\in V\right\}$.
-Each connected component of $D$ is a \emph{communication tree}.
-\end{definition}
+Given a feasible sequence of node sets and $\pi$, we derive the following broadcast trees \cite{grigni91}:
+For $s\in S$, a broadcast tree $T_s=(V_{T_s},E_{T_s})$ is a time-labeled directed subgraph of $G$ describing a broadcast originated in $s$ by the rules:
+\begin{enumerate}
+\item $T_s$ is rooted at $s$ with arcs directed towards the leaves.
+\item each node $v\in V_i$ is labeled with an integer $\tau(v)=i$,
+\item for $v\in V\setminus S, v\in V_{T_s}\Leftrightarrow \pi(v)\in V_{T_s}$, 
+\item $E_{T_s}=\{\{u,v\}: u,v\in V_{T_s}, \pi(v)=u\}$, and
+\item for two sources $s,s'\in S, s\neq s'$ we have that $V_{T_s}\cap V_{T_{s'}}=\emptyset$. 
+\end{enumerate}
 
-\noindent
-It is easily verified that the communication trees are indeed arborescences, rooted at distinct sources, with arcs pointing away from the source.
-Let $T_s=\left(V_{T_s},A_{V_s}\right)$ denote the communication tree rooted at source $s\in S$, and define
-the function $\alpha:V\to S$ such that $v\in V_{T_{\alpha(v)}}$ for all $v\in V$.
-That is, $\alpha$ associates a node $v$ with the source in which the communication tree containing $v$ is rooted.
-For any communication tree $T_s$, let $T^i_s$ be the subtree of $T$ induced by the set of nodes with distance at most $i$ edges from $s$.
-Analogously, let $D^i=\left\{T_s^i: s\in S\right\}$.
+Any feasible sequence of node sets and $\pi$ yield a set of trees $T=\{T_s:s\in S\}$ that forms a partition of $G$ into trees referred to as \emph{broadcast forest}.
+The number of leaves leaves in $T_s$ is denoted by $L(T_s)$.
+The function $\alpha:V\to S$ associates a node $v$ with the source in which the tree containing $v$ is rooted.
 
-% Given a feasible sequence of node sets and $\pi$, we derive the following broadcast trees \cite{grigni91}:
-% For $s\in S$, a broadcast tree $T_s=(V_{T_s},E_{T_s})$ is a time-labeled directed subgraph of $G$ describing a broadcast originated in $s$ by the rules:
-% \begin{enumerate}
-% \item $T_s$ is rooted at $s$ with arcs directed towards the leaves.
-% \item each node $v\in V_i$ is labeled with an integer $\tau(v)=i$,
-% \item for $v\in V\setminus S, v\in V_{T_s}\Leftrightarrow \pi(v)\in V_{T_s}$, 
-% \item $E_{T_s}=\{\{u,v\}: u,v\in V_{T_s}, \pi(v)=u\}$, and
-% \item for two sources $s,s'\in S, s\neq s'$ we have that $V_{T_s}\cap V_{T_{s'}}=\emptyset$. 
-% \end{enumerate}
-
-% Any feasible sequence of node sets and $\pi$ yield a set of trees $T=\{T_s:s\in S\}$ that forms a partition of $G$ into trees referred to as \emph{broadcast forest}.
-
-% Given a broadcast forest $T$, the delay is determined as $\max_{v\in V}\{\tau(v)\}$.
+Given a broadcast forest $T$, the delay is determined as $\max_{v\in V}\{\tau(v)\}$.
 %A set of trees $T=\{T_s:\cup_{s\in S}V_{T_s}=V\}$ forms a partition of $G$ into trees referred to as \emph{broadcast forest}.
 %$T$ can be derived from a given sequence of sets of nodes defining broadcasting and the mapping $\pi$.
-% For any integer $i$, let $T^i_s$ be the subtree of $T_s$ obtained by pruning all nodes $v\in V_{T_s}$ with $\tau(v)>i$.
-% Analogously, we define $T^i=\{T^i_s:s\in S\}$.
+For any integer $i$, let $T^i_s$ be the subtree of $T_s$ obtained by pruning all nodes $v\in V_{T_s}$ with $\tau(v)>i$.
+Analogously, we define $T^i=\{T^i_s:s\in S\}$.
 
 %We also define the set $A=\{(i,j),(j,i):\{i,j\}\in E\}$ that consists of all arcs that can be derived by directing edges in $E$.
-The degree of node $v$ in graph $G$ is denoted by $\text{deg}_G(v)$.
+The degree of node $v$ in $G$ is denoted by $\text{deg}_G(v)$.
+If $G$ is directed, $\text{deg}^-_G(v)$ and $\text{deg}^+_G(v)$ denote in-degree and out-degree, respectively.
 For the diameter of $G$ we use the symbol $\Delta_G$.
-The set of neighbors of $v\in V$ in $G$ is denoted by $N_G(v)$.
-Whenever there is no risk of confusion, the subscript $G$ is omitted.
-Likewise, $\text{deg}^-_D(v)$ and $\text{deg}^+_D(v)$ denote, respectively, the in-degree and the out-degree of digraph $D$,
-and $N_D^+(v)=\left\{u\in V:(v,u)\in A\right\}$ and $N_D^-(v)=\left\{u\in V:(u,v)\in A\right\}$ denote the neighbor sets of node $v$ in $D$.
-The node set $L(D)=\left\{v\in V: N_D^+(v)=\emptyset\right\}$ is referred to as the \emph{leaves} in $D$.
+Whenever there is no danger of confusion, the subscript $G$ is omitted.
+The set of neighbors of $v\in V$ in $G$ is denoted by $N(v)$ and $N^+(v)$ in directed graphs.
 
@@
 %The number of leaves in $T_s$ is denoted by $L(T_s)$.
-\input{sect3-exact-methods.tex}
+\input{models.tex}
 
@@
 For an instance $(G,S)$ of Problem \ref{prob:min},
-\begin{equation}
-\left\lceil\log\frac{n}{|S|}\right\rceil\leq t^* \leq n-|S|.
-\label{eq:loglb}
-\end{equation}
-\label{obs:loglb}
+$$\left\lceil\log\frac{n}{|S|}\right\rceil\leq t^* \leq n-|S|.$$
 \end{observation}
@@
 \label{lem:lbreg1}
-\underline{t}=\left\lceil\min\{k:2s\sum\limits_{j=1}^k F^{(d-1)}_j\geq n\}\right\rceil.
+\underline{t}=\left\lceil\min\{k:2\sum\limits_{j=1}^k F^{(d-1)}_j\geq n\}/|S|\right\rceil.
 \end{equation*}
-\label{prop:lbfib}
 \end{proposition}
@@
 Let $(d_1,\dots,d_n)$ be a degree sequence and $t\leq \bar{t}-|S|$ a given delay.
-Consider all possible instances of which graphs are described by this degree sequence.
-The maximum number of nodes informed within $t$ time steps is achieved in an instance in which it is possible to inform  nodes in the order of their decreasing degree.
-
-In other words, the instance with the shortest possible broadcast time is such that its nodes can be informed in the order of their decreasing degree.
-
+The maximum number of nodes informed within $t$ time steps is achieved if nodes become informed in the order of their decreasing degree.
 \end{lemma}
@@
 $k\leftarrow k+1$\;
-$c'\leftarrow 0$\;
+$c_n\leftarrow 0$\;
 \For{$i=1,\dots,c$} {
@@
 		$a_i\leftarrow a_i + 1$\;
-		$c'\leftarrow c' + 1$\;
-		\If {$c+c'\leq n$} {
-			$a_{c+c'}\leftarrow 1$;\tcp{Newly informed node} 
+		$c_n\leftarrow c_n + 1$\;
+		\If {$c+c_n\leq n$} {
+			$a_{c+c_n}\leftarrow 1$;\tcp{Newly informed node} 
 		}
@@
 }
-$c\leftarrow c + c'$\;
+$c\leftarrow c + c_n$\;
 }
@@
 Next, variable $c$ keeps the value $|V_k|$, i.e., the number nodes informed within $k$ steps.
-Finally, $c'$ stores $|V_k\setminus V_{k-1}|$, thus the number of nodes newly informed in time step $k$.
+Finally, $c_n$ stores $|V_k\setminus V_{k-1}|$, thus the number of nodes newly informed in time step $k$.
 
@@
 If $G$ is an arbitrary graph with $n$ nodes, $m$ sources and node degrees $d_i$, $1\leq i\leq n$, Alg.~\ref{alg:dreg} calculates a lower bound on the delay in $G$.
-\label{prop:deg}
 \end{proposition}
@@
 
-\input{sect5-upper-bounds.tex}
-\input{sect6-experimental-results.tex}
+\input{upper.tex}
 
@@
 
-\bibitem{datasets}
-Singh, A.,
-http://dcis.uohyd.ernet.in/~alokcs/dtp.zip
-
 \bibitem{slater81}

### 2018-06-26 sect3-exact-methods.tex
some sections in separate files. Added required files for compiling the sources.

\section{Exact methods}

Problem \ref{prob:min} can be formulated as an integer linear program. % and solved to optimality. 
In this section, we present two different modeling approaches.

\subsection{Broadcast time model}
The first studied model is a straightforward formulation of the problem.
Consider variables 
$$ x_{uv}^k=
\begin{cases} 
1, \text{ if } v\in V_k \text{ and } \pi(v)=u,\\ 
0, \text{ otherwise},
\end{cases}
z_{k}=\begin{cases}
1, \text{ if } k\leq\tau(G,S),\\
0, \text{ otherwise}.
\end{cases}
$$
The worst case scenario is when $G$ is a path $v_1,\dots,v_n$ with $S=\{v_1\}$. 
In such an instance, the necessary number of time steps is $n-1$, which gives a trivial upper bound $\bar{t}=n-s$ on the broadcast time.
Problem \ref{prob:min} is then formulated as follows: 
\begin{subequations}\label{mod:basic}
\begin{align}
\label{mod:basic:obj} \min \sum\limits_{k=1}^{\bar{t}}z_k \\ 
\notag \text{s. t. } \\
\label{mod:basic:onefromroot} \sum_{v \in N(s)}x^1_{sv} & \leq 1 & s\in S,\\
\label{mod:basic:singlein} \sum\limits_{k=1}^{\bar{t}}\sum\limits_{v\in N(u)}x_{vu}^k & = 1 & u\in V \setminus S,\\
%\label{mod:basic:uniqueTout} \sum\limits_{v\in N(u)}x_{uv}^k & \leq 1  & u\in V,k=1,\dots,\bar{t},\\
%\label{mod:basic:tIncreases} \sum\limits_{v\in N(u)}x_{uv}^k &\leq\sum\limits_{\ell=1}^{k-1}\sum\limits_{w\in N(u)\setminus\{v\}} x_{wu}^{\ell}  & u\in V\setminus S, k=2,\dots,\bar{t},\\
\label{mod:basic:tIncreases} x_{uv}^k &\leq\sum\limits_{\ell=1}^{k-1}\sum\limits_{w\in N(u)\setminus\{v\}} x_{wu}^{\ell}  & \{u,v\}\in E, u\not\in S, k=2,\dots,\bar{t},\\
%\label{mod:basic:tcrel} \sum\limits_{k=1}^{\bar{t}}k\cdot x_{uv}^k & \leq t^* &  (u,v)\in A,\\
\label{mod:basic:tcrel} \sum\limits_{v\in N(u)}x_{uv}^k & \leq z_k &  u\in V,k=1,\dots,\bar{t},\\
%\label{mod:basic:tcrel} \sum\limits_{t=1}^{n-1}t\sum\limits_{j\in N(i)}x_{ij}^k & \leq c &  i\in V,\\
\label{mod:basic:positiveCost}x_{uv}^1 & = 0 & (u,v)\in A, u \in V\setminus S,\\
\label{mod:basic:dim}&&x \in \{0,1\}^{A\times V},z\in\{0,1\}^{\bar{t}}.
\end{align}~
\end{subequations}
%Constraints \eqref{mod:basic:onefromroot} indicate that for each source node $s$, there is at most one adjacent node $u\in V_1$ such that $\pi(u)=s$.
Constraints \eqref{mod:basic:onefromroot} indicate that each source node $s$ forwards the signal to at most one adjacent node $v$ in the first time step. 
%By \eqref{mod:basic:singlein}, for every non-source node $u$, there is exactly one node $v$ such that $\pi(u)=v$.
By \eqref{mod:basic:singlein}, every non-source node $u$ receives the signal from exactly one adjacent node $v$ in some time step $k$.
%The requirement that a non-source node has a neighbor $v\in V_k$ such that $\pi(v)=u$ only if there exists a node $w\in V_{k-1}$ such that $\pi(u)=w$ is modeled by \eqref{mod:basic:tIncreases}. 
The requirement that a non-source node $u$ informs a neighbor $v$ in the $k$-th time step only if $u$ is informed by some adjacent node $w$ in an earlier time step is modeled by \eqref{mod:basic:tIncreases}. 
%Constraints \eqref{mod:basic:uniqueTout} enforce that for each node $u\in V$ and each subset $V_k$, there is at most one adjacent node $v\in V_k$ with $\pi(v)=u$.
Constraints \eqref{mod:basic:tcrel} enforce that each node $u\in V$ forwards the signal to at most one adjacent node $v$ in each time step.
It also sets a correct value to the variable $z$ that appears in the objective function.
%The requirement that only informed nodes can relay a signal is modeled by \eqref{mod:basic:tIncreases}. 
%The maximum time step at which any transmission takes place is captured by \eqref{mod:basic:tcrel}, and finally, \eqref{mod:basic:positiveCost} states that a node that is not a source never transmits in the first time step.
%The length of the sequence of subsets is captured by \eqref{mod:basic:tcrel}, and finally, \eqref{mod:basic:positiveCost} state that if $\pi(v)\not\in S$ for some $j\in V$, then $v\not\in V_1$.
Finally, \eqref{mod:basic:positiveCost} state that non-source nodes do not transmit in the first time step.
\subsection{Binomial tree model}

The binomial tree $B^k$ of order $k$ is an ordered tree defined recursively as follow \cite{cormen90}:
\begin{itemize}
\item The binomial tree $B^0$ consists of a single node.
\item The binomial tree $B^k$ has a root with $k$ children where the $i$-th child is the root of a binomial tree of order $k-i$, $i=1,\dots,k$.
\end{itemize}
An example of $B^3$ is depicted in Fig.\ref{fig:beta}.
%If a solution to MBT consists of broadcast trees that are binomial, the number of informed nodes doubles in each time step.
For a given time step $k$, the maximum number of informed nodes within $k$ steps is $|S|2^k$.
This occurs when the solution of MBT consists of broadcast trees that are binomial.
%Any broadcast tree can be regarded as pruned binomial tree.  
%Problem \ref{prob:min} can therefore be restated as finding a partition of $G$ into $m$ pruned binomial trees 
\begin{observation}
\label{obs:btspread}
if $r\in S$ is the root of $B^k$, then $\tau(B^k,\{r\})=k$.
\end{observation}
\begin{figure}
\centering
\begin{tikzpicture}[->,scale=.7,every node/.style = {scale=.6,draw,shape=circle, align=center, fill=gray!30}, level/.style={sibling distance=2.5cm/#1,level distance=1.0cm}]]
   \node[] {1}
   	   child[] { node {2} 
	   	   child {node {4}
		  	child {node {8}} 
		   }
		   child {node {6} }
	   }
   child[] { node {3}
   	   child { node {7} }
	   }
   child[] { node {5}
	}
 ;
\end{tikzpicture}
\caption{A binomial tree with nodes labeled by their positions}
\label{fig:beta}
\end{figure}

\subsubsection{Binomial trees over positive integers}

Let $k\in \mathbb{N}$ and $I=\{1,\dots,2^k\}$. 
A directed binomial tree $B^k=(V_{B^k},A_{B^k})$ with arcs oriented towards the leaves has a regular structure that allows to define a systematic numbering of nodes so that a node number determines unambiguously a position in $B^k$.
That is, we need an applicable bijective function $\beta:V_{B^k}\to I$.
A suitable bijection $\beta$ assigns values from $I$ to nodes increasingly with decreasing outgoing degree.
If there is an ambiguity, a node whose parent has a lower number is assigned a lower number.
This function is defined recursively as
\begin{equation*}
\label{eq:beta}
\beta(v)=\begin{cases}
1,\text{ if } v \text{ is the root of } B^k,\\
\beta(u) + 2^{k-deg^+(v)-1}, (u,v)\in A_{B^k},\text{ otherwise}.
\end{cases}
\end{equation*}
Nodes in Fig. \ref{fig:beta} are labeled with their $\beta$-values.
Pairs of integers that are assigned to adjacent nodes in a binomial tree are defined by relation
\begin{equation*}
\label{eq:betarel}
R=\{(i,j)\in\mathbb{N}\times\mathbb{N}:j=i+2^k,k\geq\log i,k\in \mathbb{Z}\}.
\end{equation*}

We further define the function $\pi':\mathbb{N}\to\mathbb{Z}_+$ that for a given integer $i$ associated with node $v$
determines $\beta$-value of parent of $v$: 
\begin{align*}
\label{eq:piprime}
\pi'(1)&=0,& \\
\pi'(j)&=j-2^{\lceil\log j\rceil -1}, &j > 1.
\end{align*}
Using functions $\alpha$, $\beta$, $\pi'$ and the relation $R$, we introduce the notion of binomial trees over positive integers.
\begin{definition}
A pair $(I,X)$ where $I=\{1,\dots,2^k\}$, $k\in\mathbb{Z}_+$, $(\alpha(u),\beta(u))=(\alpha(v),\beta(v))\Leftrightarrow u=v$ and
$X=\{(i,j)\in R, i,j\in I\}$ is an \emph{integer binomial tree of order $k$}.

$(I,X)$, where $I\subseteq\{1,\dots,2^k\}, k\in \mathbb{Z}_+,\pi(j)\in I$ for all $j\in I\setminus\{1\}$ and
$X=\{(i,j)\in R, i,j\in I\}$ is an \emph{integer binomial tree of order $k$ pruned at $\{1,\dots,2^k\}\setminus I$}.
\end{definition}
We now develop a method for modelling Problem \ref{prob:min} whose main principle is finding a partition of $G$ into pruned binomial trees.
For any integer $i$, let $I_i=\{2^{i-1}+1,\dots,2^i\}$.
%\begin{problem}
%\label{prob:dec}
%Given $G=(V,E)$, $S\subseteq V$ and $t\in \mathbb{N}$, is there a sequence $S=V_0\subseteq\dots\subseteq V_t=V$ 
%and a mapping $\pi:V\setminus S\to V$, such that for each $v\in V\setminus S:\{v,\pi(v)\}\in E$ and for each  $u,v\in V_i\setminus V_{i-1}: \pi(u)\neq \pi(v)\Leftrightarrow u\neq v$?
%\end{problem}
%For a delay $t$, at most $s\cdot 2^t$ nodes can be informed within $t$ steps. 
%Therefore, we assume $n\leq 2^ks.
%This can be achieved when the broadcast forest $T$ consists of binomial trees $B^t$ of order $t$ rooted at sources $s\in S$.
%Hence, if there is a partition of $G$ into $s$ pruned binomial trees of order at most $t$ rooted at sources, then $(G,S,t)$ is a YES instance of Problem \ref{prob:dec}.
%Finding a partition of $G$ into $s$ pruned binomial trees can be equivalently formulated as finding a partition of $G''$ into $s$ (complete) binomial trees, 
%where $G''$ is constructed from $G$ as follows:
%Let $\alpha\coloneqq s\cdot 2^k-|V|$, and let $K_\alpha=(V_\alpha,E_\alpha)$ be a complete graph on $\alpha$ nodes.
%Each node in $K_\alpha$ is connected to every node $v\in V$ in the original graph $G$.
%Thus, $G''=(V'',E'')$ with $V''=V\cup V_\alpha$ and $E''=E\cup E_\alpha\cup \{\{u,v\}: u\in V \wedge v\in V_\alpha\}$.
%The set of arcs $A''$ is constructed by creating two arcs of opposite orientation for each edge, but arcs with orientation from $V_\alpha$ to $V$ are excluded. 
%Formally, $A''=A\cup\{(u,v),(v,u): \{u,v\}\in E_\alpha\}\cup\{(u,v):u\in V \wedge v\in V_\alpha\}$.

%\begin{observation}\label{obs:deg}
%For each $i\in\{1,\dots,t\}$, the set $\{v\in V^t: 1\leq\beta(v)\leq2^i\}\setminus\{v\in V^t:1\leq\beta(v)\leq2^{i-1}\}$ contains nodes with out-degree $t-i$.
%\end{observation}
%\begin{observation}\label{obs:childdeg}
%Children of $v\in V^t$ with $\text{deg}^+(v)=\ell$ have out-degree $0,\dots,\ell-1$.
%\end{observation}
\begin{observation}\label{obs:eqbetalevel}
$\pi'$ is injective.
%For $i_1,i_2\in I_i$, $\pi'(i_1)=\pi'(i_2)\Leftrightarrow i_1=i_2$.
\end{observation}
\begin{proof}
We notice that $\lceil\log (2^{i-1}+1)\rceil=\dots =\lceil\log 2^i\rceil$, and thus $\pi'(2^{i-1}+1),\dots,\pi'(2^i)$ are pairwise different.\qed 
\end{proof}

\begin{proposition}\label{lem:probeq}
A graph $G=(V,E)$ and $S\subseteq V$ has $\tau(G,S)\leq k$ iff it is possible to assign integers to nodes
such that they form $m$ (pruned) integer binomial trees of order $k$ or smaller rooted at sources.
\end{proposition}
\begin{proof}
Assume nodes in $V$ can be labeled by integers from $I=\{1,\dots,2^k\}$ so that they form (pruned) integer integer binomial trees $B_\ell$ of order at most $k$, $1\leq\ell\leq m$.
For $r\in V_{B_\ell}$ with $\alpha(r)=r$ and $\beta(r)=1$, by Obs. \ref{obs:btspread}, $\tau(B_\ell,\{r\})\leq k$.
The sequence of node sets $V_0,\dots,V_k$ is constructed by setting $V_0=\{v\in V:\beta(v)=1\}$, and $V_{i}=V_{i-1}\cup\{v\in V: \beta(v)\in I_i\}$.
Moreover, $\pi(v)=u\Leftrightarrow \alpha(v)=\alpha(u)~\&~\pi'(\beta(v))=\beta(u)$.
Also, $\pi(u)=\pi(v)\Leftrightarrow u=v$ follows from Obs. \ref{obs:eqbetalevel}, because nodes in $V_i$ have $\beta$ values in $I_i$.

Conversely, suppose there is a sequence of subsets and a function $\pi$ in $G$ with the desired properties.
Nodes are associated with integers from $I$ assigned according to the following steps:
\begin{enumerate}
\item $\beta(s)=1$ for all $s\in V_0$,
\item For $v\in V_i$, set $\beta(v)=j$ such that $j\in I_i$ and $(\pi'(\beta(v)),j)\in R, 1\leq i\leq k$.  
\end{enumerate}
\qed
\end{proof}
\subsubsection{The formulation}
Consider a graph $G'=(V',E')$ constructed  by adding a universal node $v_0$ to $G$. 
The set of nodes and edges is then $V'=V\cup \{v_0\}$ and $E'=E\cup\{\{v_0,v\}:v\in V\}$.
Apart from $z$-variables introduced in model \eqref{mod:basic}, the ILP formulation based on partition into binomial trees uses 
$$
y_{is}^v=\begin{cases}
1, \text{ if } \beta(v)=i \text{ and } \alpha(v)=s,\\ 
0, \text{ otherwise},\\
\end{cases}
$$
where $v\in V'$, $i\in I$, $s\in S$ and $0\leq j\leq \bar{t}$. 
With the definition of $G'$ above, it is straightforward to specify constraints that enforce desired values for $y$-variables.
Whenever $y_{is}^{v_0}=1$, it indicates that the binomial tree rooted at $s$ is pruned at node with position $i$.
%An obvious weakness of this approach is  that the number of nodes increases to $|V''|=\mathcal{O}(ns)$, and the dimension of variables is thus $\mathcal{O}(n^2s^2)$.
%However, once a suitable partition is found, the arcs of binomial trees contained in $K_\alpha$ can be diversely shuffled while preserving the layour of binomial trees in $G$.
%Instead of adding the entire complete graph $K_\alpha$, a single node $v_0$ with a loop $(v_0,v_0)$ is connected as an apex to the original $G$.
%Let us denote this multigraph as $G'=(V',E')$, where $V'=V\cup\{v_0\}$, $E'=E\cup\{\{u,v_0\}:u\in V\}\cup\{\{v_0\}\}$. 
%The arc set is then analogously defined as $A'=A\cup\{(u,v_0): u\in V\}\cup\{(v_0,v_0)\}$.
%The requirement for partition into binomial trees has to be adjusted accordingly.
%The subtrees contained in $G$ remain unchanged, every arc $(u,v)\in A^k_i, i=1,\dots,s$ in $G''$ with $u\in V$ and $v\in V_\alpha$ becomes $(u,v_0)$ in $G'$,
%and every $(u,v)\in A_\alpha$ becomes $(v_0,v_0)$.
%So, $v_0$ acts as a universal node that can substitute several nodes in each binomial tree.
Let us define the set $C(i)$ of $\beta$-positions of children (direct descendants) of node $v$ with $\beta(v)=i$ in a binomial tree of order $k$:
\begin{equation}
\label{eq:c1}
C(i)=\{2^j+i:j=\lceil\log_2 i\rceil,\dots,k-1\}.
\end{equation}
The formulation based on binomial trees is the following:
\begin{subequations}\label{mod:partition}
\begin{align}
\notag &\min\sum\limits_{j=0}^{\bar{t}}z_j,\\
\notag \text{s. t. } \\
\label{mod:part:nodeBelongs} \sum\limits_{i\in I}\sum\limits_{s\in S}y^v_{is} & = 1 & v\in V,\\
\label{mod:part:treeHasIJ} \sum\limits_{v\in V'}y^v_{is} & = 1 & i\in I,s\in S,\\
\label{mod:part:source1} y_{1s}^s & = 1  & s\in S,\\
%\label{mod:part:noReturn} y^u_{ij}+y^v_{lj} &\leq 1 & i\in I,l\in C(i), j\in J, u\in V_\alpha,v\in V,\\
%\label{mod:part:followArcs} y^u_{is}+y^v_{\ell s} &\leq 1 & i\in I,\ell\in C(i), s\in S, u,v\in V',(u,v)\not\in A',\\
%\label{mod:part:followArcsA} y^u_{is}+y^v_{\ell s} &\leq 1 & i\in I,\ell\in C(i), s\in S, u,v\in V,\{u,v\}\not\in E,\\
%\label{mod:part:followArcsB} y^{v_0}_{is}+y^v_{\ell s} &\leq 1 & i\in I,\ell\in C(i), s\in S, v\in V,\\
\label{mod:part:followArcsA}y^{v_0}_{is}+y^u_{i s} + \sum\limits_{v\in V\setminus N(u)}y^v_{\ell s}&\leq 1 & u\in V,i\in I,\ell\in C(i), s\in S,  \\
\label{mod:part:followArcsB}y^{v_0}_{is}+y^u_{\ell s} + \sum\limits_{v\in V\setminus N(u)}y^v_{i s}&\leq 1 & u\in V,i\in I,\ell\in C(i), s\in S,\\ 
\label{mod:part:yzrel}\sum\limits_{v\in V}y^v_{is} & \leq z_{\lceil\log i\rceil} & i\in I,s\in S,\\
\label{mod:part:dim}&&y \in \{0,1\}^{I\times S\times V'}, z\in \{0,1\}^{\bar{t}}.
\end{align}~
\end{subequations}

%As the model represents the decision problem, it suffices to find any feasible solution, and no objective function is needed.
The interpretation of constraints \eqref{mod:part:nodeBelongs} is that every node in the original graph $G$ belongs to exactly one binomial tree.
Note that these constraints are quantified only over $V$ and not over $V'$.
In this way it is achieved that $v_0$ can be regarded as a part of several binomial trees.
By \eqref{mod:part:treeHasIJ} is ensured that exactly one node, possibly $v_0$, is allocated to position $i$ of each binomial tree.
By the summation over $V'$ is ensured, that pruned nodes are collectively represented by $v_0$.
Next, \eqref{mod:part:source1} enforce that source nodes are always the first nodes in corresponding binomial trees, in accordance with definition \eqref{eq:beta} of the function $\beta$.
The remaining two sets of constraints guarantee that the arcs of binomial trees follow edges in $E$.
In particular, it is enforced by \eqref{mod:part:followArcsA} that if $u$ and $v$ are not adjacent in $G$, then $v$ must not act as a child of $u$ in any binomial tree.
Constraints, \eqref{mod:part:followArcsB} forbid any node from $V$ to be a child of $v_0$ in any binomial tree. 
This reflects the obvious fact that if a tree is pruned at some node, all its descendants must also be excluded from the tree.
%The definition of $A'$ also prevents arcs of the binomial trees to be oriented from $V_\alpha$ to $V$.
%In other words, once the signal leaves the original graph $G$ and enters $v_0$, it cannot return back to $G$.
Without \eqref{mod:part:followArcsA} and \eqref{mod:part:followArcsB}, it could be possible to find a feasible solution, even when no partition of $G$ into pruned binomial trees exists.
Finally, the relation \eqref{mod:part:yzrel} between $y$ and $z$ variables follows from Obs. \ref{obs:btspread}. 
It says that whenever there is a node in a position $i$, then the delay is at least $\lceil\log i\rceil$.
%Constraints \eqref{mod:part:followArcsA} and \eqref{mod:part:followArcsB} can be replaced by stronger
%\begin{subequations}
%\begin{align}
%\label{mod:part:followArcsAStrongerA}
%y^{v_0}_{is}+y^u_{i s} + \sum\limits_{v\in V\setminus N(u)}y^v_{\ell s}&\leq 1 & u\in V,i\in I,\ell\in C(i), s\in S,  \\
%\label{mod:part:followArcsAStrongerB}
%y^{v_0}_{is}+y^u_{\ell s} + \sum\limits_{v\in V\setminus N(u)}y^v_{i s}&\leq 1 & u\in V,i\in I,\ell\in C(i), s\in S. 
%\end{align}
%\end{subequations}
%
\subsubsection{Valid inequalities}
Let $W$ be a maximal independent set in $G$.
Model \eqref{mod:partition} is strengthened by 
\begin{align}
\label{mod:part:vibasic}
y^{v_0}_{is}+ \sum\limits_{v\in W}(y^v_{is}+y^v_{\ell s})&\leq 1 & i\in I,\ell\in C(i), s\in S, 
\end{align}
which exploits the fact that no pair of nodes in $W$ is adjacent, and so there must be no two nodes with adjacent $\beta$-positions.

We now generalize this idea by using the notion of graph power $G^m=(V,E^m)$ commonly defined as a graph with the same set of nodes as $G$,
and an edge between two nodes in $G^m$ is present iff there is a path of length at most $m$ between them in $G$.
For our purposes, we use a slightly modified definition of the edge set
$$E^m=\{\{u,v\}:\text{there exists a path between $u$ and $v$ in $G$ of length $m$}\}.$$
Definition \eqref{eq:c1} can be generalized to descendants of an arbitrary distance $m$ from $v$ in $B^k$:
\begin{equation}
C^{m+1}(i)=\bigcup_{j\in C^1(i)}C^m(j).
\end{equation}
For a given $m$, let $W_m$ be a maximal independent set in $G^m$.
Further strengthening of model \eqref{mod:partition} is achieved by introducing valid inequalities
\begin{align}
\label{mod:part:vigeneral}
y^{v_0}_{is}+ \sum\limits_{v\in W_m}(y^v_{is}+y^v_{\ell s})&\leq 1 & i\in I,\ell\in C^m(i), s\in S,1\leq m\leq \Delta_G-1. 
\end{align}
Clearly, inequality \eqref{mod:part:vibasic} is included in \eqref{mod:part:vigeneral} for $m=1$.
The distance between positions $i$ and $\ell=C^m(i)$ in a binomial tree is $m$.
The maximal independent set $W_m$ contains nodes such that length of any path between any two nodes is different from $m$, 
and so there must not be two nodes in $W_m$ with positions $i$ and $\ell$ at the same time.

\subsubsection{Symmetry removal}
Another improvement of this model is achieved by a symmetry removal.
If a broadcast tree is identical to a binomial tree, we notice that nodes with labels from $C(i)$, i.e., children of some node $v$ with $\beta(v)=i$, are informed in increasing time steps.
For example in $B^3$, $C(2)=\{4,6\}$ and the corresponding nodes are informed in time step 2 and 3, respectively.
If a label $\ell\in C(i)$ corresponds to a node of a binomial tree that is pruned (if $y^{v_0}_{\ell s}=1$ for some $s\in S$), 
all labels $j\in C(i)$ such that $j>\ell$ can also be pruned. 
Thus, adding 
\begin{align}
\label{mod:part:sr}
y^{v_0}_{js}&\leq y^{v_0}_{\ell s}&i\in I,j,\ell\in C(i),j<\ell, s\in S
\end{align}
to the model reduces the set of feasible solutions.

%An objective function is naturally lacking in the formulation of the decision problem.
%It is nevertheless straightforward to create an ILP model for the corresponding optimization problem.
%Let $z_i=1 \Leftrightarrow t=i$ be a new variable and let $\bar{t}$ be an upper bound on the delay in $(G,S)$.
%Problem \ref{prob:min} is formulated as follows:
%\begin{subequations}
%\begin{align}
%\notag &\min\sum\limits_{j=0}^{\bar{t}}z_j,\\
%\notag \text{s. t. } \\
%\notag \eqref{mod:part:nodeBelongs} - \eqref{mod:part:source1},& \eqref{mod:part:followArcsAStrongerA} - \eqref{mod:part:followArcsAStrongerB}, \eqref{mod:part:vi}, \eqref{mod:part:sr},\\
%\notag\sum\limits_{v\in V}y^v_{is} & \leq z_{\lceil\log i\rceil} & i\in I,s\in S,\\
%\notag\label{mod:part:optdim}y &\in \{0,1\}^{I\times S\times V'}, z \in \{0,1\}^{\bar{t}}.
%\end{align}~
%\end{subequations}

### 2018-06-26 sect5-upper-bounds.tex
some sections in separate files. Added required files for compiling the sources.

\section{Upper bounds}

%The algorithms presented in this section iteratively construct broadcast foerest $T_i, i=1,2,\dots$.
The following inexact method is based on the idea of finding maximum cardinality matching in $G\left[V_{T_i},V\setminus V_{T_i}\right]$, and extending $T_i$ by this matching.
It means that in each iteration $i$, the maximum possible number of nodes are informed.
The maximum cardinality matching can be regarded as finding $|V_{T_i}|$ node-disjoint  binomial trees of order at most one with roots in $V_{T_i}$, maximizing the number of edges.
By generalizing this notion, we iteratively look for $|V_{T_i}|$ node-disjoint (pruned) binomial trees of an arbitrary given order $k$ valued between 1 and $n-|S|$.
Even though this problem is NP-hard for $k\geq 2$ \cite{jansen95}, it is expected that the computational time is sound in most practical instances.
After obtaining the set of binomial trees, first $p$ nodes in each tree are selected and added to the broadcast forest.
The parameter $p\in \{1,\dots,2^k\}$ is also a part of the input.
\begin{algorithm}[]
\KwData{$G=(V,E), S\subseteq V, k\in \{1,\dots,n-|S|\}, p\in \{1,\dots,2^k\}$}
\textbf{for }$s\in S\textbf{ do }T_s=(V_s,A_s), V_s\leftarrow \{s\}, E_s\leftarrow\emptyset$\;
$I=\{1,\dots,2^k\}$\\
$T=\{T_s:s\in S\}$\ ~~~\tcp{broadcast forest}
$t\leftarrow 0$\;
\While{$V_T\neq V$} {
	$S\leftarrow V_T$\;
	Find a set of pruned binomial trees $B=\{B_1,\dots,B_{|V_T|}\}$ of order at most $k$ with roots in $V_T$ by solving model \eqref{mod:genmatch}\;
	$V_T\leftarrow V_T\cup \{v:v\in V_B:\beta(v)\leq p\}$\;
	$E_T\leftarrow E_T\cup \{\{u,v\}\in E_B: \beta(u)\leq p,\beta(v)\leq p\}$\;
	$t\leftarrow t+1$\;
}
\Return t\;
%\Return $\lceil k/s \rceil$\;
 \caption{A method for determining an upper bound}
\label{alg:match}
\end{algorithm}

Alg. \ref{alg:match} describes the process formally.
Initially, the broadcast forest consists of isolated sources.
The binomial trees are determined by solving the ILP model \eqref{mod:genmatch}. 
This model is a modification of formulation \eqref{mod:partition}, and uses the same type of variables.
The objective function is to maximize the number of nodes involved in the binomial trees.
%The index set $I=\{1,\dots,2^k\}$ depends on the input parameter $k$.
The constraints \eqref{mod:genmatch:nodeBelongs} state that each node belongs to at most one binomial tree.
Compared to \eqref{mod:part:nodeBelongs}, \eqref{mod:genmatch:nodeBelongs} is an inequality, because the binomial trees do not necessarily form a partition of $G$, and so not all nodes have to be used.
The remaining constraints are taken from formulation \eqref{mod:partition}. 
Note that in every iteration, the model is solved with a different set $S$, as all nodes already included in the broadcast forest are roots of the binomial trees.
The model considers the entire set $V$, but it is also possible to restrict this set to nodes with distance at most $k$ from some node in $V_T$.
When this restriction is not imposed, nodes with larger distance are not a part of any binomial tree in the current iteration due to \eqref{mod:part:followArcsA} - \eqref{mod:part:followArcsB}.

\begin{subequations}\label{mod:genmatch}
\begin{align}
\notag \max\sum\limits_{v\in V}&\sum\limits_{i\in I}\sum\limits_{s\in S}   y_{is}^v,\\
\notag \text{s. t. } \\
\label{mod:genmatch:nodeBelongs} \sum\limits_{i\in I}\sum\limits_{s\in S}y^v_{is} & \leq 1 & v\in V,\\
\notag\eqref{mod:part:treeHasIJ} - \eqref{mod:part:followArcsB},\\
%\label{mod:genmatch:treeHasIJ} \sum\limits_{v\in V'}y^v_{is} & = 1 & i\in I,s\in V_T,\\
%\label{mod:genmatch:source1} y_{1s}^s & = 1  & s\in V_T,\\
%\label{mod:part:noReturn} y^u_{ij}+y^v_{lj} &\leq 1 & i\in I,l\in C(i), j\in J, u\in V_\alpha,v\in V,\\
%\label{mod:part:followArcs} y^u_{is}+y^v_{\ell s} &\leq 1 & i\in I,\ell\in C(i), s\in S, u,v\in V',(u,v)\not\in A',\\
%\label{mod:part:followArcsA} y^u_{is}+y^v_{\ell s} &\leq 1 & i\in I,\ell\in C(i), s\in S, u,v\in V,\{u,v\}\not\in E,\\
%\label{mod:part:followArcsB} y^{v_0}_{is}+y^v_{\ell s} &\leq 1 & i\in I,\ell\in C(i), s\in S, v\in V,\\
%\label{mod:genmatch:followArcsA}y^{v_0}_{is}+y^u_{i s} + \sum\limits_{v\in V\setminus N(u)}y^v_{\ell s}&\leq 1 & u\in V,i\in I,\ell\in C(i), s\in V_T,  \\
%\label{mod:genmatch:followArcsB}y^{v_0}_{is}+y^u_{\ell s} + \sum\limits_{v\in V\setminus N(u)}y^v_{i s}&\leq 1 & u\in V,i\in I,\ell\in C(i), s\in V_T,\\ 
\label{mod:genmatch:dim}&&y \in \{0,1\}^{I\times S\times V'}.
\end{align}~
\end{subequations}

%
%We present a simple general method for determining an upper bound on the optimal solution.
%The idea is to construct broadcast trees in parallel from their roots in sources.
%A new node is connected to a tree is it was not connected to any other tree in any of the previous iterations.
%The steps are summarized in Alg. \ref{alg:genub}.
%
%\begin{algorithm}[]
%\KwData{$G=(V,E), S\subseteq V$}
%$U\leftarrow S$;\tcp{Set of marked nodes}
%$T_s=(V_s,A_s), V_s\leftarrow \{s\}, E_s\leftarrow\emptyset$\;
%\While{$|U|<n$} {
%	\For{$s\in S$} {
%		Select $v\in V\setminus U:\exists u\in V_s \text{ s. t.} \{u,v\}\in E$ according to some strategy\; 	
%		$U\leftarrow U\cup\{v\}$\;
%		$V_s\leftarrow V_s\cup\{v\}$\;
%		$A_s\leftarrow A_s\cup\{u,v\}$\;
%	}
%}
%\For{$s\in S$} {
%$	
%\sigma(v)=\begin{cases}
%		0, \text{ if } v\in L(T_s),\\
%		\max\limits_{k\in 1,\dots,|N^+(v)|}\{k+\sigma(j_k):j_k\in N^+(v)\wedge \forall \ell<k:\sigma(j_\ell)>\sigma(j_k)\},\\
%		\hfill \text{ if }v\in V_s\setminus L(T_s).
%
%\end{cases}\;
%$
%}
%\Return $\max_{s\in S}\sigma(s)$\;
%%\Return $\lceil k/s \rceil$\;
% \caption{A method for determining an upper bound}
%\label{alg:genub}
%\end{algorithm}
%The algorithm consists of two phases. 
%In the first phase (lines 3 -- 10), $m$ broadcast trees are constructed.
%The second phase (lines 11 -- 13) uses a recursive formula that determines in linear time an optimal broadcast time for a subtree of $T_s$ rooted at $v\in V_s$.
%The value  $\sigma(s)$ than gives the optimal broadcast time for $T_s$.
%The complexity of Alg. \ref{alg:genub} depends on the strategy according to which the trees are constructed.
%For common strategies such as BFS and DFS it is $\mathcal{O}(n+|E|)$, for the shortest paths trees $\mathcal{O}(|E| + n\log n)$.
%
%

### 2018-07-16 mbt.tex
some sections in separate files. Added required files for compiling the sources.
proposition rephrased

@@
 % your LaTeX will extract the file if required
-\begin{filecontents*}{example.eps}
+%\begin{filecontents*}{example.eps}
 %!PS-Adobe-3.0 EPSF-3.0
@@
 %%EndComments
-gsave
-newpath
-  20 20 moveto
-  20 220 lineto
-  220 220 lineto
-  220 20 lineto
-closepath
-2 setlinewidth
-gsave
-  .4 setgray fill
-grestore
-stroke
-grestore
-\end{filecontents*}
+%gsave
+%newpath
+%  20 20 moveto
+%  20 220 lineto
+%  220 220 lineto
+%  220 20 lineto
+%closepath
+%2 setlinewidth
+%gsave
+%  .4 setgray fill
+%grestore
+%stroke
+%grestore
+%\end{filecontents*}
 %
@@
 %The number of leaves in $T_s$ is denoted by $L(T_s)$.
-\input{models.tex}
+\input{sect3-exact-methods.tex}
 
@@
 Let $(d_1,\dots,d_n)$ be a degree sequence and $t\leq \bar{t}-|S|$ a given delay.
-The maximum number of nodes informed within $t$ time steps is achieved if nodes become informed in the order of their decreasing degree.
+Consider all possible instances of which graphs are described by this degree sequence.
+The maximum number of nodes informed within $t$ time steps is achieved in an instance in which it is possible to inform  nodes in the order of their decreasing degree.
+
+In other words, the instance with the shortest possible broadcast time is such that its nodes can be informed in the order of their decreasing degree.
+
 \end{lemma}
@@
 
-\input{upper.tex}
+\input{sect5-upper-bounds.tex}
 

### 2018-07-25 mbt.tex
fixed fibonacci formula for LB
Notation in Alg. 1. changed from c_n to c'
label Network model nad Notation --> Definitions
change label
pokus
DH: Made minor correction just to see how git works.

@@
+% comment
 % This is a general template file for the LaTeX package SVJour3mm
@@
 The minimum broadcast time (MBT) problem consists of a set of communication nodes with a subset of source nodes. 
-The task is to disseminate a signal to every node in a shortest possible time $t^*$ (delay), while abiding by communication rules.
+The task is to disseminate a signal to every node in a shortest possible time (delay), while abiding by communication rules.
 An \emph{informed} node is a node that has received the signal.
@@
 
-\subsection{Literature overview}
+{Literature overview}
 
@@
 
-\section{Network Model and Notation}
+\section{Network Model and Definitions}
 
@@
 \label{lem:lbreg1}
-\underline{t}=\left\lceil\min\{k:2\sum\limits_{j=1}^k F^{(d-1)}_j\geq n\}/|S|\right\rceil.
+\underline{t}=\left\lceil\min\{k:2s\sum\limits_{j=1}^k F^{(d-1)}_j\geq n\}\right\rceil.
 \end{equation*}
@@
 $k\leftarrow k+1$\;
-$c_n\leftarrow 0$\;
+$c'\leftarrow 0$\;
 \For{$i=1,\dots,c$} {
@@
 		$a_i\leftarrow a_i + 1$\;
-		$c_n\leftarrow c_n + 1$\;
-		\If {$c+c_n\leq n$} {
-			$a_{c+c_n}\leftarrow 1$;\tcp{Newly informed node} 
+		$c'\leftarrow c' + 1$\;
+		\If {$c+c'\leq n$} {
+			$a_{c+c'}\leftarrow 1$;\tcp{Newly informed node} 
 		}
@@
 }
-$c\leftarrow c + c_n$\;
+$c\leftarrow c + c'$\;
 }
@@
 Next, variable $c$ keeps the value $|V_k|$, i.e., the number nodes informed within $k$ steps.
-Finally, $c_n$ stores $|V_k\setminus V_{k-1}|$, thus the number of nodes newly informed in time step $k$.
+Finally, $c'$ stores $|V_k\setminus V_{k-1}|$, thus the number of nodes newly informed in time step $k$.
 

### 2018-07-25 mbt.tex
DH: Sharpened definitions in Section 2.

@@
 \documentclass[smallextended]{svjour3}       % onecolumn (second format)
+\usepackage[T1]{fontenc}
 %\documentclass[twocolumn]{svjour3}          % twocolumn
@@
 At the beginning, the set of informed nodes is exactly the set of sources.
-An informed node $u$ can send the signal to an uninformed node $v$ if $u$ and $v$ are located within a communication vicinity of each other.
+An informed node can send the signal to an uninformed node if the two nodes are located within a communication vicinity of each other.
 
-The time is divided into discrete time steps.
-At each time step, every informed node can forward the signal to at most one uninformed neighbor.
-The number of informed nodes at some time step can therefore be no more than the double of the number of informed nodes at the previous time step.
+The time is divided into a finite number of time steps.
+Every informed node can, at each time step, forward the signal to at most one uninformed neighbor node.
+Therefore, the number of informed nodes can at most be doubled from one step to the next.
 %\subsection{Motivation and Related Work}
@@
 
-MBT has been shown to be NP-complete on arbitrary graphs in \cite{slater81}. 
-The problem remains NP-complete even for several restricted communication networks \cite{jansen95} such as
-split graphs with delay $t^*=2$, chordal graphs,
-bipartite and planar graphs with maximum degree 3 and a single source,
-several variants of grid graphs,
-bipartite planar graphs with delay $t^*=2$ and maximum degree 3. 
-Yet another NP-completeness result for 3-regular planar graphs with delay 2 is shown in \cite{middendorf93}.
+Deciding whether an instance of MBT has a solution with delay at most $t$ has been shown to be NP-complete \cite{slater81}. 
+For bipartite planar graphs with maximum degree 3, NP-completeness persists even if $t=2$ or there is only one source \cite{jansen95}.
+When $t=2$, the problem also remains NP-complete for cubic planar graphs \cite{middendorf93}, grid graphs with maximum degree 3,
+complete grid graphs, chordal graphs, and for split graphs \cite{jansen95}. 
+The single-source variant of the decision version of MBT is NP-complete for grid graphs with maximum degree 4, and for chordal graphs \cite{jansen95}.
 The problem is known to be polynomial in trees \cite{slater81}.
-Whether the problem is polynomial or NP-complete for split graphs with a single source was stated as an open questions in \cite{jansen95}, and hasn't been answered yet to our best knowledge.
+Whether the problem is polynomial or NP-complete for split graphs with a single source was stated as an open questions in \cite{jansen95}, and has to the best of our knowledge not been answered yet.
 
-A number of inexact methods for both general and special graph classes has been proposed in the literature during the last three decades.
-One of the first works of this category is \cite{scheuermann84}, 
-where the authors introduce an exact dynamic programming algorithm based on generating all maximum matchings in an induced bipartite graph.
-Additional contribution of \cite{scheuermann84} are heuristic approaches for near optimal broadcasting.
+A number of inexact methods, for both general and special graph classes, have been proposed in the literature during the last three decades.
+One of the first works of this category \cite{scheuermann84} 
+introduces an exact dynamic programming algorithm based on generating all maximum matchings in an induced bipartite graph.
+Additional contributions of \cite{scheuermann84} are heuristic approaches for near optimal broadcasting.
 From more recent works we mention \cite{hasson04}, which describes a meta heuristic algorithm for MBT, and provides a comparison with other existing methods.
@@
 Approximation algorithms for MBT are studied in \cite{kortsarz95}. 
-The authors argue that methods presented in \cite{scheuermann84} provide no guarantee on the performance and state an $n$-node wheel as an example of an unfavourable instance.
+The authors argue that methods presented in \cite{scheuermann84} provide no guarantee on the performance, and show that the $n$-node wheel is an example of an unfavourable instance.
 They introduce an $\mathcal{O}(\sqrt{n})$-additive approximation algorithm for broadcasting in general graphs.
@@
 Another algorithm with $\mathcal{O}\left(\frac{\log n}{\log \log n}\right)$-approximation\footnote{All logarithms in this paper are of base 2} ratio is given in~\cite{elkin03}.
-Most of the works above consider a single source.
+Most of the works cited above consider a single source.
 
-A related problem extensively studied in the literature is the minimum broadcast graph \cite{grigni91,mcgarvey16}. 
+A related problem extensively studied in the literature is the minimum broadcast graph problem \cite{grigni91,mcgarvey16}. 
 A broadcast graph supports a broadcast from any node to all other nodes in optimal time $\lceil\log n\rceil$.
-For a given $n$, the effort is to determine properties such as the minimum number of edges and the minimum degree in broadcast graph on $n$ nodes.
+For a given integer $n$, the problem is to find a broadcast graph of $n$ nodes such that e.g.\ the number of edges, or the maximum node degree, in the graph is minimized.
 The authors of \cite{mcgarvey16} study ILP models for $c$-broadcast graphs, a generalization of the problem that allows transmission of the signal to at most $c$ neighbours in a single time step.
@@
 The communication network is represented by a connected graph $G=(V,E)$ and a subset $S\subseteq V$ referred to as the set of sources, where $|V|=n$. 
-Let $V\left[F\right]=\{v\in V:\text{ some edge in } F \text{ is incident to } v\}, F\subseteq E$.
-For disjoint node sets $V_1,V_S\subseteq V$, we denote by $G\left[V_1,V_2\right]$ the bipartite graph $(V_1\cup V_2,E')$, where  $E'=\{\{u,v\}\in E: u\in V_1,v\in V_2\}$.
+% Let $V\left[F\right]=\{v\in V:\text{ some edge in } F \text{ is incident to } v\}, F\subseteq E$.
+% For disjoint node sets $V_1,V_S\subseteq V$, we denote by $G\left[V_1,V_2\right]$ the bipartite graph $(V_1\cup V_2,E')$, where  $E'=\{\{u,v\}\in E: u\in V_1,v\in V_2\}$.
 
-\begin{definition}
-The \emph{broadcast time} $\tau(G,S)$ of a node set $S\subseteq V$ in $G$ is defined as: 
-\begin{equation*}
-\tau(G,S)=
-\begin{cases}
-	0, S=V,\\
-	1+\min\{\tau(G,S\cup V\left[M\right]):M \text{ is a matching in } G\left[S, V\setminus S\right]\}, S\neq V.
-\end{cases}
-\label{eq:btime}
-\end{equation*}
+\begin{definition} \label{def:broadcasttime}
+The \emph{broadcast time} $\tau(G,S)$ of a node set $S\subseteq V$ in $G$ is defined as the smallest integer $t\geq 0$ for which there exist
+a sequence $V_0\subseteq\dots\subseteq V_t$ of node sets and a function $\pi:V\setminus S\to V$, such that:
+\begin{enumerate}
+  \item $V_0=S$ and $V_t=V$, \label{def:boundary}
+  \item for all $v\in V\setminus S$, $\{v,\pi(v)\}\in E$, \label{def:edge}
+  \item for all $i=1,\ldots,t$ and all $v\in V_i$, $\pi(v)\in V_{i-1}$, and \label{def:parent}
+  \item for all $u,v\in V_i\setminus V_{i-1}$, $\pi(u)=\pi(v)$ only if $u=v$. \label{def:unique}
+\end{enumerate}
+% % \begin{equation*}
+% \tau(G,S)=
+% \begin{cases}
+% 	0, S=V,\\
+% 	1+\min\{\tau(G,S\cup V\left[M\right]):M \text{ is a matching in } G\left[S, V\setminus S\right]\}, S\neq V.
+% \end{cases}
+% \label{eq:btime}
+% \end{equation*}
 \end{definition}
-%Broadcasting is defined as a sequence of sets $(S=V_0\subseteq\dots\subseteq V_k = V)$ where each $V_i$ represents the nodes informed after time step $i$, $0\leq i\leq t$.
-%For each node $v\in V_i\setminus V_{i-1}$, there exists a single node $\pi(v)\in V_{i-1}$ adjacent to $v$, which forwarded the signal to $v$.
-%Also, for every two nodes $u,v\in V_i\setminus V_{i-1}$ we have $\pi(u)=\pi(v)$.
-%The value $k$ is referred to as \emph{delay}.
-The optimization problem in question is formulated as follows \cite{jansen95,middendorf93}:
-\begin{problem}\label{prob:min}
-Given $G=(V,E)$ and $S\subseteq V$, find a sequence $(S=V_0\subseteq\dots\subseteq V_{\tau(G,S)}=V)$ and a function $\pi:V\setminus S\to V$, such that for each $v\in V\setminus S:\{v,\pi(v)\}\in E$, and for each $u,v\in V_i\setminus V_{i-1}: \pi(u)=\pi(v)\Leftrightarrow u=v$.
+% %Broadcasting is defined as a sequence of sets $(S=V_0\subseteq\dots\subseteq V_k = V)$ where each $V_i$ represents the nodes informed after time step $i$, $0\leq i\leq t$.
+% %For each node $v\in V_i\setminus V_{i-1}$, there exists a single node $\pi(v)\in V_{i-1}$ adjacent to $v$, which forwarded the signal to $v$.
+% %Also, for every two nodes $u,v\in V_i\setminus V_{i-1}$ we have $\pi(u)=\pi(v)$.
+% %The value $k$ is referred to as \emph{delay}.
+
+Referring to Section \ref{intro}, the node set $V_i$ is the set of nodes that are informed in time step $i$.
+Initially, only the sources are informed ($V_0=S$), whereas all nodes are informed after $t$ time steps ($V_t=V$),
+and the set of informed nodes is monotonously non-decreasing ($V_{i_1}\subseteq V_i$ for $i=1,\ldots,t$).
+The parent function $\pi$ maps each node to the node from which it received the signal.
+Conditions \ref{def:edge}--\ref{def:parent} of Definition \ref{def:broadcasttime} thus reflect that the sender is a neighbor node in $G$,
+and that it is informed at an earlier time step than the recipient node.
+Because each node can send to at most one neighbor node in each time step,
+condition \ref{def:unique} states that $\pi$ maps the set of nodes which are informed for the first time in step $i$ to distinct parent nodes.
+
+The optimization problem in question is formulated as follows:
+\begin{problem}[\textsc{Minimum Broadcast Time}]\label{prob:min}
+Given $G=(V,E)$ and $S\subseteq V$, find $\tau(G,S)$.
+% a sequence $(S=V_0\subseteq\dots\subseteq V_{\tau(G,S)}=V)$ and a function $\pi:V\setminus S\to V$, such that for each $v\in V\setminus S:\{v,\pi(v)\}\in E$, and for each $u,v\in V_i\setminus V_{i-1}: \pi(u)=\pi(v)\Leftrightarrow u=v$.
 \end{problem}
 
-Given a feasible sequence of node sets and $\pi$, we derive the following broadcast trees \cite{grigni91}:
-For $s\in S$, a broadcast tree $T_s=(V_{T_s},E_{T_s})$ is a time-labeled directed subgraph of $G$ describing a broadcast originated in $s$ by the rules:
-\begin{enumerate}
-\item $T_s$ is rooted at $s$ with arcs directed towards the leaves.
-\item each node $v\in V_i$ is labeled with an integer $\tau(v)=i$,
-\item for $v\in V\setminus S, v\in V_{T_s}\Leftrightarrow \pi(v)\in V_{T_s}$, 
-\item $E_{T_s}=\{\{u,v\}: u,v\in V_{T_s}, \pi(v)=u\}$, and
-\item for two sources $s,s'\in S, s\neq s'$ we have that $V_{T_s}\cap V_{T_{s'}}=\emptyset$. 
-\end{enumerate}
+\begin{definition} \label{def:broadcastgraph}
+For any $V_0,\ldots,V_t$ and $\pi$ satisfying the conditions of Definition \ref{def:broadcasttime},
+the corresponding \emph{communication forest} is
+the digraph $D=(V,A)$, where $A=\left\{\left(v,\pi(v)\right): v\in V\right\}$.
+Each connected component of $D$ is a \emph{communication tree}.
+\end{definition}
 
-Any feasible sequence of node sets and $\pi$ yield a set of trees $T=\{T_s:s\in S\}$ that forms a partition of $G$ into trees referred to as \emph{broadcast forest}.
-The number of leaves leaves in $T_s$ is denoted by $L(T_s)$.
-The function $\alpha:V\to S$ associates a node $v$ with the source in which the tree containing $v$ is rooted.
+\noindent
+It is easily verified that the communication trees are indeed arborescences, rooted at distinct sources, with arcs pointing away from the source.
+Let $T_s=\left(V_{T_s},A_{V_s}\right)$ denote the communication tree rooted at source $s\in S$, and define
+the function $\alpha:V\to S$ such that $v\in V_{T_{\alpha(v)}}$ for all $v\in V$.
+That is, $\alpha$ associates a node $v$ with the source in which the communication tree containing $v$ is rooted.
+For any communication tree $T_s$, let $T^i_s$ be the subtree of $T$ induced by the set of nodes with distance at most $i$ edges from $s$.
+Analogously, let $D^i=\left\{T_s^i: s\in S\right\}$.
 
-Given a broadcast forest $T$, the delay is determined as $\max_{v\in V}\{\tau(v)\}$.
+% Given a feasible sequence of node sets and $\pi$, we derive the following broadcast trees \cite{grigni91}:
+% For $s\in S$, a broadcast tree $T_s=(V_{T_s},E_{T_s})$ is a time-labeled directed subgraph of $G$ describing a broadcast originated in $s$ by the rules:
+% \begin{enumerate}
+% \item $T_s$ is rooted at $s$ with arcs directed towards the leaves.
+% \item each node $v\in V_i$ is labeled with an integer $\tau(v)=i$,
+% \item for $v\in V\setminus S, v\in V_{T_s}\Leftrightarrow \pi(v)\in V_{T_s}$, 
+% \item $E_{T_s}=\{\{u,v\}: u,v\in V_{T_s}, \pi(v)=u\}$, and
+% \item for two sources $s,s'\in S, s\neq s'$ we have that $V_{T_s}\cap V_{T_{s'}}=\emptyset$. 
+% \end{enumerate}
+
+% Any feasible sequence of node sets and $\pi$ yield a set of trees $T=\{T_s:s\in S\}$ that forms a partition of $G$ into trees referred to as \emph{broadcast forest}.
+
+% Given a broadcast forest $T$, the delay is determined as $\max_{v\in V}\{\tau(v)\}$.
 %A set of trees $T=\{T_s:\cup_{s\in S}V_{T_s}=V\}$ forms a partition of $G$ into trees referred to as \emph{broadcast forest}.
 %$T$ can be derived from a given sequence of sets of nodes defining broadcasting and the mapping $\pi$.
-For any integer $i$, let $T^i_s$ be the subtree of $T_s$ obtained by pruning all nodes $v\in V_{T_s}$ with $\tau(v)>i$.
-Analogously, we define $T^i=\{T^i_s:s\in S\}$.
+% For any integer $i$, let $T^i_s$ be the subtree of $T_s$ obtained by pruning all nodes $v\in V_{T_s}$ with $\tau(v)>i$.
+% Analogously, we define $T^i=\{T^i_s:s\in S\}$.
 
 %We also define the set $A=\{(i,j),(j,i):\{i,j\}\in E\}$ that consists of all arcs that can be derived by directing edges in $E$.
-The degree of node $v$ in $G$ is denoted by $\text{deg}_G(v)$.
-If $G$ is directed, $\text{deg}^-_G(v)$ and $\text{deg}^+_G(v)$ denote in-degree and out-degree, respectively.
+The degree of node $v$ in graph $G$ is denoted by $\text{deg}_G(v)$.
 For the diameter of $G$ we use the symbol $\Delta_G$.
-Whenever there is no danger of confusion, the subscript $G$ is omitted.
-The set of neighbors of $v\in V$ in $G$ is denoted by $N(v)$ and $N^+(v)$ in directed graphs.
+The set of neighbors of $v\in V$ in $G$ is denoted by $N_G(v)$.
+Whenever there is no risk of confusion, the subscript $G$ is omitted.
+Likewise, $\text{deg}^-_D(v)$ and $\text{deg}^+_D(v)$ denote, respectively, the in-degree and the out-degree of digraph $D$,
+and $N_D^+(v)=\left\{u\in V:(v,u)\in A\right\}$ and $N_D^-(v)=\left\{u\in V:(u,v)\in A\right\}$ denote the neighbor sets of node $v$ in $D$.
+The node set $L(D)=\left\{v\in V: N_D^+(v)=\emptyset\right\}$ is referred to as the \emph{leaves} in $D$.
 

### 2018-08-06 mbt.tex
DH: Corrected minor errors.

@@
 \usepackage{amsmath, amssymb}
+\usepackage{multirow}
 \usepackage{graphicx}
@@
 Initially, only the sources are informed ($V_0=S$), whereas all nodes are informed after $t$ time steps ($V_t=V$),
-and the set of informed nodes is monotonously non-decreasing ($V_{i_1}\subseteq V_i$ for $i=1,\ldots,t$).
+and the set of informed nodes is monotonously non-decreasing ($V_{i-1}\subseteq V_i$ for $i=1,\ldots,t$).
 The parent function $\pi$ maps each node to the node from which it received the signal.
@@
 the function $\alpha:V\to S$ such that $v\in V_{T_{\alpha(v)}}$ for all $v\in V$.
-That is, $\alpha$ associates a node $v$ with the source in which the communication tree containing $v$ is rooted.
+That is, $\alpha$ associates a node $v$ with the source at which the communication tree containing $v$ is rooted.
 For any communication tree $T_s$, let $T^i_s$ be the subtree of $T$ induced by the set of nodes with distance at most $i$ edges from $s$.
@@
 
-\section{Lower bounds}
-In this section, we study lower bounds on the delay for several restrictions of input graphs.
-%An optimal solution is obtained by solving a sequence of decision problems with varying $t$. 
-%It is therefore desirable to determine tight lower and upper bounds in order to arrive in the optimum after solving as few decision problems as possible.
-Obvious bounds for a general graph instance are given by
-\begin{observation}
-For an instance $(G,S)$ of Problem \ref{prob:min},
-$$\left\lceil\log\frac{n}{|S|}\right\rceil\leq t^* \leq n-|S|.$$
-\end{observation}
-
-In the following, we use $m$-step Fibonacci numbers \cite{noe05}, a generalization of well known (2-step) Fibonacci numbers, defined by letting, 
-$F^{(m)}_k=0$ for $k\leq 0$, $F^{(m)}_1=1$, and 
-other terms according to linear recurrence relation 
-\begin{align*}
-F^{(m)}_k &=\sum\limits_{i=1}^m F^{(m)}_{k-i}, &\text{ for } k\geq 2.
-\end{align*}
-
-Consider a $d$-regular graph with a unique source $s$.
-Any broadcast forest consists of a single tree $T_s$.
-We investigate the number of leaves in $T_s$, and derive a lower bound on the delay for this graph class.
-
-If the orientation of arcs in $T_s$ is disregarded, $L(T^1_s)=L(T^2_s)=2$.
-For $i\geq 3$, $L(T^i_s)$ equals the number of nodes with degree $1,\dots,d-1$ in $T^{i-1}_s$, 
-because in a $d$-regular graph, only nodes with degree lower than $d$ can inform new uninformed nodes.
-It can also be interpreted as the sum of the number of leaves in $T^{i-d+1}_s,\dots,T^{i-1}_s$, which leads to %the following formula
-\begin{equation*}
-\label{eq:leafrec}
-L(T^i_s)=\sum\limits_{j=i-d+1}^{i-1} L(T^j_s).
-\end{equation*}  
-This formula equals to the recursive definition of Fibonacci sequence of order $d-1$.
-As each of the two base cases, $L(T^1_s)$ and $L(T^2_s)$, equals the double of the base cases of the Fibonacci sequence, the number of leaves in time step $i$ is calculated as
-\begin{equation*}
-\label{eq:fibleaf}
-L(T^i_s)=2 F^{(d-1)}_i.
-\end{equation*}  
-The number of nodes in $T^i_s$ can be expressed as the sum of nodes newly informed in time steps $1,\dots,i$, that is, the sum of leaves in $T^1_s,\dots,T^i_s$. Thus,
-\begin{equation}
-\label{eq:fibcnt}
-|V_{T^i_s}|=|V_i|=2\sum\limits_{j=1}^i F^{(d-1)}_j.
-\end{equation}
-
-\begin{proposition}
-For a $d$-regular graph on $n$ nodes and $|S|$ sources, a lower bound on the delay is 
-\begin{equation*}
-\label{lem:lbreg1}
-\underline{t}=\left\lceil\min\{k:2s\sum\limits_{j=1}^k F^{(d-1)}_j\geq n\}\right\rceil.
-\end{equation*}
-\end{proposition}
-\begin{proof}
-In order to inform $n$ nodes in the best possible scenario, the signal has to be relayed sufficiently many time steps so that the maximum possible number of informed nodes becomes $n$.
-The maximum number of nodes informed within $i$ time steps is given by \eqref{eq:fibcnt}.
-We therefore need to set the upper limit of the summation in \eqref{eq:fibcnt} so that the right-hand side exceeds $n$.
-The reason why the result is divided by $|S|$ is that the best case scenario with several source nodes assumes that the signals initiated in individual sources are spread evenly.
-\qed
-\end{proof}
-
-An additional knowledge of a degree sequence of $G$ can be exploited. 
-\begin{lemma}
-\label{lemma:degorder}
-Let $(d_1,\dots,d_n)$ be a degree sequence and $t\leq \bar{t}-|S|$ a given delay.
-Consider all possible instances of which graphs are described by this degree sequence.
-The maximum number of nodes informed within $t$ time steps is achieved in an instance in which it is possible to inform  nodes in the order of their decreasing degree.
-
-In other words, the instance with the shortest possible broadcast time is such that its nodes can be informed in the order of their decreasing degree.
-
-\end{lemma}
-\begin{proof}
-A node with degree $d_i$ informed in $i$-th time step can inform at most $\min\{d_i-1,t-i\}$ nodes.
-Consider $l,k$ such that $1\leq k < \ell\leq n$ and nodes $v_k$, $v_\ell$ with $\deg(v_k)=d_k$ and $\deg(v_\ell)=d_\ell$.
-Then, $d_k-1>d_\ell-1$ and $t-k > t-\ell$.
-If nodes are informed in the order of their decreasing degree, let $K_1$ and $L_1$ be the number of nodes that can be informed by $v_k$ and $v_l$, respectively:
-$$
-K_1=\min\{d_k-1,t-k\}, ~~~ L_1=\min\{d_\ell-1,t-\ell\}.
-$$
-If the order in which $v_k$ and $v_l$ are informed is switched, i.e. when $v_k$ and $v_\ell$ is informed in the $\ell$-th and $k$-th time step, respectively, 
-we use $K_2$ and $L_2$ to denote the maximum number of nodes informed by $v_k$ and $v_\ell$:
-$$
-K_2=\min\{d_k-1,t-\ell\} ~~~ L_2=\min\{d_\ell-1,t-k\}.
-$$
-We now investigate what possible values can the expression 
-\begin{equation}
-\label{eq:degbounds}
-(K_1+L_1)-(K_2+L_2)
-\end{equation}
-attain.
-Observe that $K_1\geq L_2$ and $K_2\geq L_1$. 
-If $K_2>L_1$, then $K_2=K_1=d_k-1$.
-But in that case, $L_2=L_1=d_\ell-1$. 
-We conclude that \eqref{eq:degbounds} always takes a non-negative value.
-Therefore, if the nodes are informed in the order of their decreasing degrees,
-the number of informed nodes within the give time step is no worse than when the nodes are informed in any other order.
-\qed
-\end{proof}
-
-Alg.~\ref{alg:dreg} calculates a lower bound on the delay when given number of nodes, sources and a degree sequence in $G$ as an input.
-The algorithm iteratively updates possible node degrees in $T^k$ in each time step $k$, and records the maximum possible number of nodes in $V_k$.
-For the purpose of finding lower bounds, it is assumed that each node $v\in V_k$ with $\deg_{T^k}(v)<\deg_G(v)$ informs a new uninformed node.
-The number of iterations is then the lowest possible delay for given input.
-%The main idea is that in the most optimistic case, each node $v\in V_i$ with $\text{deg}_{F_i}(v)<\text{deg}_G(v)$ informs new not yet informed node.
-%For the purpose of finding lower bounds, we can assume without loss of generality that $\text{deg}_{F_i}(u)\leq\text{deg}_{F_i}(v) \Leftrightarrow \text{deg}_G(u)\geq \text{deg}_G(v)$.
-Once a node $v$ reaches its maximum degree, i.e., when $\text{deg}_{T^k}=\deg_G(v)$ for some $k$, $v$ does not inform any other node in the next time steps.
-
-%For each $i=1,\dots d$, Alg. \ref{alg:dreg} keeps the  number of nodes with degree $i$.
-%These values are updated iteratively using dynamic programming until the number of informed nodes reaches $n$.
-
-\begin{algorithm}
-\KwData{$n,m,d_1,\dots,d_n\in \mathbb{N}, m\leq n,\newline 1\leq d_n\leq\dots\leq d_2\leq d_1$}
-$a_1\leftarrow\dots\leftarrow a_{m}\leftarrow 1$\;%\tcp{this is a comment}
-$a_{m+1}\leftarrow\dots\leftarrow a_{n}\leftarrow 0$\;%\tcp{this is a comment}
-$c\leftarrow m$;%\tcp{every source informs a new node}
-$~k\leftarrow 0$\;
-\While {$c<n$} {
-$k\leftarrow k+1$\;
-$c'\leftarrow 0$\;
-\For{$i=1,\dots,c$} {
-	\If {$a_i<d_i$} {
-		$a_i\leftarrow a_i + 1$\;
-		$c'\leftarrow c' + 1$\;
-		\If {$c+c'\leq n$} {
-			$a_{c+c'}\leftarrow 1$;\tcp{Newly informed node} 
-		}
-	}
-}
-$c\leftarrow c + c'$\;
-}
-\Return $k$\;
-%\Return $\lceil k/s \rceil$\;
- \caption{Lower bound exploiting distribution of degrees}
-\label{alg:dreg}
-\end{algorithm}
-
-The input is assumed to be correct in the sense that a graph with the given degree sequence exists, and by definition, the degree sequence is ordered non-increasingly.
-For each iteration $k$, Alg.~\ref{alg:dreg} stores degrees of nodes in $T^k$ in variables $a_1,\dots,a_n$.
-Note that the forest $T^k$ is not actually constructed. 
-The algorithm operates merely with potential degrees of nodes in $T^k$.
-Next, variable $c$ keeps the value $|V_k|$, i.e., the number nodes informed within $k$ steps.
-Finally, $c'$ stores $|V_k\setminus V_{k-1}|$, thus the number of nodes newly informed in time step $k$.
-
-\begin{proposition}
-If $G$ is an arbitrary graph with $n$ nodes, $m$ sources and node degrees $d_i$, $1\leq i\leq n$, Alg.~\ref{alg:dreg} calculates a lower bound on the delay in $G$.
-\end{proposition}
-\begin{proof}
-The assumption that nodes with higher degree in $G$ are informed first in the optimal solution is justified by Lemma \ref{lemma:degorder}.
-In each iteration $k$, degrees in $T^k$ of maximum possible number of nodes are increased in the order of their decreasing degree.
-The current maximum possible number of informed nodes is stored in the variable $c$, and is updated at the end of the outre loop.
-So, when the condition $c<n$ is tested on line 7, $c$ contains the correct value of number of nodes in $V_k$.
-As soon as it is possible that at least $n$ nodes are informed, the algorithm returns the number of necessary iterations.
-\qed
-\end{proof}
-
+\input{sect4-lower-bounds.tex}
 \input{sect5-upper-bounds.tex}
+\input{sect6-experimental-results.tex}
+\input{sect7-conclusion.tex}
 
@@
 
+\bibitem{datasets}
+Singh, A.,
+http://dcis.uohyd.ernet.in/~alokcs/dtp.zip
+
 \bibitem{slater81}

### 2018-08-09 sect3-exact-methods.tex
broadcast time model updated according to what is actually implemented: constraints (1d) has a sum on the lhs
MI: Decision version of the straightfwd model has been stated and the idea behind it described

@@
 %\label{mod:basic:uniqueTout} \sum\limits_{v\in N(u)}x_{uv}^k & \leq 1  & u\in V,k=1,\dots,\bar{t},\\
-%\label{mod:basic:tIncreases} \sum\limits_{v\in N(u)}x_{uv}^k &\leq\sum\limits_{\ell=1}^{k-1}\sum\limits_{w\in N(u)\setminus\{v\}} x_{wu}^{\ell}  & u\in V\setminus S, k=2,\dots,\bar{t},\\
-\label{mod:basic:tIncreases} x_{uv}^k &\leq\sum\limits_{\ell=1}^{k-1}\sum\limits_{w\in N(u)\setminus\{v\}} x_{wu}^{\ell}  & \{u,v\}\in E, u\not\in S, k=2,\dots,\bar{t},\\
+\label{mod:basic:tIncreases} \sum\limits_{v\in N(u)}x_{uv}^k &\leq\sum\limits_{\ell=1}^{k-1}\sum\limits_{w\in N(u)\setminus\{v\}} x_{wu}^{\ell}  & u\in V\setminus S, k=2,\dots,\bar{t},\\
+%\label{mod:basic:tIncreases} x_{uv}^k &\leq\sum\limits_{\ell=1}^{k-1}\sum\limits_{w\in N(u)\setminus\{v\}} x_{wu}^{\ell}  & \{u,v\}\in E, u\not\in S, k=2,\dots,\bar{t},\\
 %\label{mod:basic:tcrel} \sum\limits_{k=1}^{\bar{t}}k\cdot x_{uv}^k & \leq t^* &  (u,v)\in A,\\
@@
 Finally, \eqref{mod:basic:positiveCost} state that non-source nodes do not transmit in the first time step.
+
+%In Model \ref{mod:basic}, superscript of $x$-variables represents a time step in which a signal is transmitted along corresponding arc.
+The nature of MBT suggests another modelling approach derived from Model \ref{mod:basic}. 
+For a given maximum broadcast time $t_{\text{max}}$, we maximize the number of arcs along which a signal is relayed within $t_{\text{max}}$.
+If the optimal value is $n-1$, it is clear that we found a spanning broadcast tree with broadcast time $t_{\text{max}}$.
+In case the optimal value is less than $n-1$, the gien $t_{\text{max}}$ is insufficient and we have to try solving the problem with increased time limit.
+Clearly, an additional knowledge of upper and lower bound spares some computations. 
+A lower bound is the initial value of $t_{\text{max}}$. 
+Similarly, if an upper bound $\bar{t}$ is known and $t_{\text{max}}=\bar{t}-1$, the process can be terminated when the objective function is less then $n-1$, because $\bar{t}$ is than the optimal value.
+
+The decision model \label{mod:basic:dec}  has the form
+\begin{subequations}\label{mod:basic:dec}
+\begin{align}
+\label{mod:basic:dec:obj} \max \sum\limits_{(u,v) \in A}\sum\limits_{k=1}^{t_{\text{max}}}x_{u,v}^k \\ 
+\notag \text{s. t. } \\
+\label{mod:basic:dec:atMost1in} \sum\limits_{k=1}^{t_{\text{max}}}\sum\limits_{v\in N(u)}x_{vu}^k & \leq 1 & u\in V \setminus S,\\
+\label{mod:basic:dec:0toSource} \sum\limits_{k=1}^{t_{\text{max}}}\sum\limits_{v\in N(u)}x_{vu}^k & = 0  & u\in S,\\
+\notag \eqref{mod:basic:onefromroot}, \eqref{mod:basic:tIncreases}, \eqref{mod:basic:positiveCost},\\
+\label{mod:basic:dec:dim}&&x \in \{0,1\}^{A\times V}.
+\end{align}~
+\end{subequations}
+
+Constraint \eqref{mod:basic:singlein} is replaced by \eqref{mod:basic:dec:atMost1in} and \eqref{mod:basic:dec:0toSource}. 
+The former is an inequality, because not all nodes are necessarily reached within the given time limit.
+The latter makes sure that sources are not reached by any signal. 
+This was filtered out by optimality in \eqref{mod:basic}, but it must be forbidden explicitly in \eqref{mod:basic:dec} due to the changed objective function.
+
 \subsection{Binomial tree model}

### 2018-08-10 mbt.tex
Started writing about combinatorial relaxations

@@
 \input{sect1-introduction.tex}
-\section{Network Model and Definitions}
+\section{Network Model and Definitions} \label{sec:def}
 
@@
 Because each node can send to at most one neighbor node in each time step,
-condition \ref{def:unique} states that $\pi$ maps the set of nodes which are informed for the first time in step $i$ to distinct parent nodes.
+condition \ref{def:unique} states that $\pi$ maps the set of nodes becoming informed in step $i$ to distinct parent nodes.
 
@@
 \end{problem}
+
+Directed versions of Defintion \ref{def:broadcasttime} and Problem \ref{prob:min} are defined analogously.
+For a digraph $D=(V,A)$, condition \ref{def:edge} is modified to state that $(\pi(v),v)\in A$ for all $v\in V\setminus S$.
+We denote by $\tau(D,S)$ the broadcast time of $S$ in $D$.
 
@@
 That is, $\alpha$ associates a node $v$ with the source at which the communication tree containing $v$ is rooted.
-For any communication tree $T_s$, let $T^i_s$ be the subtree of $T$ induced by the set of nodes with distance at most $i$ edges from $s$.
+For any communication tree $T_s$, let $T^i_s$ be the subtree of $T_s$ induced by the set of nodes with distance at most $i$ edges from $s$.
 Analogously, let $D^i=\left\{T_s^i: s\in S\right\}$.

### 2018-08-10 sect4-lower-bounds.tex
Started writing about combinatorial relaxations

@@
 \section{Lower bounds}
-In this section, we study lower bounds on the delay for several restrictions of input graphs.
+Strong lower bounds on the minimum objective function value are of vital importance to combinatorial optimization algorithms.
+In this section, we study three types of lower bounds on the broadcast time $\tau(G,S)$.
+
+\subsection{Analytical lower bounds} \label{sec:lbanalyt}
 %An optimal solution is obtained by solving a sequence of decision problems with varying $t$. 
@@
 
-An additional knowledge of a degree sequence of $G$ can be exploited. 
+\subsection{Continuous relaxations of integer programming models} \label{sec:lblprel}
+DH: \emph{To be written: Demonstrate why LP-bounds are expected to be weak (no stronger than the analytical bounds).}
+
+\subsection{Combinatorial relaxations} \label{sec:lbcombrel}
+
+DH: \emph{I keep the old content here so far, but a suggestion to a new version is given after the old one.}
+
+Monotonous degree sequences of $G$ can be exploited more extensively than what was demonstrated in Section \ref{sec:lbanalyt}. 
 \begin{lemma}
@@
 \end{proof}
+
+DH: \emph{A suggestion to a new version follows.}
+\bigskip
+
+Lower bounds on the broadcast time $\tau(G,S)$ are obtained by omitting one or more of the conditions imposed in Definition \ref{def:broadcasttime}.
+For the purpose of strongest possible bounds, the relaxations thus constructed can be supplied with conditions that are redundant in the problem definition.
+
+Recall from Section \ref{sec:def} that $D=(V,A)$ is the communication forest corresponding to $V_0,\ldots,V_t$, and $\pi$.
+Conditions \ref{def:boundary}--\ref{def:unique} of Definition \ref{def:broadcasttime} imply that
+\begin{enumerate}
+\setcounter{enumi}{4}
+  \item for all $v\in V$, $\delta_D^+(v)+\delta_D^-(v)\leq\delta_G(v)$. \label{def:degree}
+\end{enumerate}
+
+\noindent
+A lower bound on $\tau(G,S)$ is then given by the solution to:
+\begin{problem}[\textsc{Node Degree Relaxation}]\label{prob:degree}
+Find the smallest integer $t\geq 0$ for which there exist
+a sequence $V_0\subseteq\dots\subseteq V_t$ of node sets and a function $\pi:V\setminus S\to V$,
+satisfying conditions \ref{def:boundary} and \ref{def:parent}--\ref{def:degree}.
+\end{problem}
+
+Except from the node degrees, explicit information about $G$ is not part of the input to Problem \ref{prob:degree},
+and two nodes $u,v\in V\setminus S$, where $\delta_G(u)=\delta_G(v)$, are considered identical.
+
+In order to develop an algorithm for the \textsc{Node Degree Relaxation}, we derive some optimality conditions of its single-source version.
+Thus, let $S=\left\{v_1\right\}$ and $V\setminus S=\left\{v_2,\ldots,v_n\right\}$, where $\delta_G(v_2)\geq\delta_G(v_3)\geq\dots\geq\delta_G(v_n)$,
+and denote $d_i=\delta_G(v_i)$ ($i=1,\ldots,n$).
+Let $\tau_i=\tau\left(T[i],\{v_i\}\right)$ denote the broadcast time of node $v_i$ in the subtree $T[i]$ of $T$ rooted at $v_i$.
+The broadcast time $\tau\left(V,\{v_1\}\right)$ is hence no less than the sum of the time $\lceil\log\beta(v_i)\rceil$ it takes to inform $v_i$,
+and the subsequent time $\tau_i$ needed to inform all nodes in $T[i]$.
+Consequently, if $T$ is an optimal broadcast tree,
+\begin{equation}
+  \tau\left(V,\{v_1\}\right)=\max\left\{\lceil\log\beta(v_i)\rceil+\tau_i: i=2,\ldots,n\right\}. \label{eq:tau}
+\end{equation}
+
+\begin{lemma} \label{lem:degree}
+Problem \ref{prob:degree} has an optimal solution where, for all $i=2,\ldots,n-1$, $\delta_T^+(v_i)\geq\delta_T^+(v_{i+1})$ and $\tau_i\geq\tau_{i+1}$.
+\end{lemma}
+\begin{proof}
+Consider an optimal broadcast tree $T$, and let $j\in\{2,\ldots,n-1\}$ be the largest value of $i$ for which $\tau_i<\tau_{i+1}$.
+\emph{By utilizing (\ref{eq:tau}, go on to argue that a simple swap does not increase $\tau\left(\right)$}
+\emph{Slightly more tricky: Argue that moving subtrees can fix the first inequality. Swap again to restore the second}.
+\end{proof}

### 2018-08-11 sect3-exact-methods.tex
Decision version of the binomial tree model was stated with objective function maximizing number of assigned labels. Another simple upper bound based on Prim's alg. was described.

@@
 %In Model \ref{mod:basic}, superscript of $x$-variables represents a time step in which a signal is transmitted along corresponding arc.
+\subsubsection{Decision version}
+\label{sec:decbasic}
 The nature of MBT suggests another modelling approach derived from Model \ref{mod:basic}. 
@@
 
+The objective function \eqref{mod:basic:dec:obj} maximizes the number of arcs relaying a signal.
+The actual target function is the minimum time required for spreading the signal to all nodes, and so we are looking for the smallest value of $t_{text{max}}$.
 \subsection{Binomial tree model}
@@
 
+\subsubsection{Decision version}
+
+In a similar manner as we derived the decision version \eqref{mod:basic:dec} of model \eqref{mod:basic}, it is possible to construct a decision version of the binomial tree model \eqref{mod:partition} as follows:
+\begin{subequations}\label{mod:genmatch}
+\begin{align}
+\notag \max\sum\limits_{v\in V}&\sum\limits_{i\in I}\sum\limits_{s\in S}   y_{is}^v,\\
+\notag \text{s. t. } \\
+\label{mod:genmatch:nodeBelongs} \sum\limits_{i\in I}\sum\limits_{s\in S}y^v_{is} & \leq 1 & v\in V,\\
+\notag\eqref{mod:part:treeHasIJ} - \eqref{mod:part:followArcsB},\\
+%\label{mod:genmatch:treeHasIJ} \sum\limits_{v\in V'}y^v_{is} & = 1 & i\in I,s\in V_T,\\
+%\label{mod:genmatch:source1} y_{1s}^s & = 1  & s\in V_T,\\
+%\label{mod:part:noReturn} y^u_{ij}+y^v_{lj} &\leq 1 & i\in I,l\in C(i), j\in J, u\in V_\alpha,v\in V,\\
+%\label{mod:part:followArcs} y^u_{is}+y^v_{\ell s} &\leq 1 & i\in I,\ell\in C(i), s\in S, u,v\in V',(u,v)\not\in A',\\
+%\label{mod:part:followArcsA} y^u_{is}+y^v_{\ell s} &\leq 1 & i\in I,\ell\in C(i), s\in S, u,v\in V,\{u,v\}\not\in E,\\
+%\label{mod:part:followArcsB} y^{v_0}_{is}+y^v_{\ell s} &\leq 1 & i\in I,\ell\in C(i), s\in S, v\in V,\\
+%\label{mod:genmatch:followArcsA}y^{v_0}_{is}+y^u_{i s} + \sum\limits_{v\in V\setminus N(u)}y^v_{\ell s}&\leq 1 & u\in V,i\in I,\ell\in C(i), s\in V_T,  \\
+%\label{mod:genmatch:followArcsB}y^{v_0}_{is}+y^u_{\ell s} + \sum\limits_{v\in V\setminus N(u)}y^v_{i s}&\leq 1 & u\in V,i\in I,\ell\in C(i), s\in V_T,\\ 
+\label{mod:genmatch:dim}&&y \in \{0,1\}^{I\times S\times V'}.
+\end{align}~
+\end{subequations}
+This model is a modification of formulation \eqref{mod:partition}, and uses the same type of variables.
+The objective function is to maximize the number of nodes involved in the binomial trees.
+In order to determine the minimum broadcast time, we are looking for the minimum $k$ such that each node can be assigned a label from the set $I=\{1,\dots,2^k\}$ while satisfying given constraints.
+%The index set $I=\{1,\dots,2^k\}$ depends on the input parameter $k$.
+The constraints \eqref{mod:genmatch:nodeBelongs} state that each node belongs to at most one binomial tree.
+Compared to \eqref{mod:part:nodeBelongs}, \eqref{mod:genmatch:nodeBelongs} is an inequality, because the binomial trees do not necessarily form a partition of $G$, and so not all nodes have to be used.
+The remaining constraints are taken from formulation \eqref{mod:partition}.
+
+The approach is then analogous to the idea indicated in Section \ref{sec:decbasic}.
+Assume we are given a lower bound $\underline{t}$ and an upper bound $\bar{t}$ on the minimum broadcast time.
+Initially, we define the set $I=\{1,\dots,2^{\underline{t}}\}$ and iteratively solve the model while doubling the set $I$ until the objective function attains the value $n$.
+That indicates that it is the first iteration in which all nodes are assigned a label, and it can be concluded that the minimum broadcast time is $\log_2|I|$.
+
 %An objective function is naturally lacking in the formulation of the decision problem.

### 2018-08-11 sect5-upper-bounds.tex
Decision version of the binomial tree model was stated with objective function maximizing number of assigned labels. Another simple upper bound based on Prim's alg. was described.

@@
 
-%The algorithms presented in this section iteratively construct broadcast foerest $T_i, i=1,2,\dots$.
+A knowledge of an upper bound $\bar{t}$ affects the number of variables in all studied models. 
+Particularly in the decision versions, the iterative approach can be terminated once the solution is dissatisfactory for broadcast time limit $\bar{t}-1$.
+The algorithms presented in this section iteratively construct broadcast foerest $T_i, i=1,2,\dots$.
+\subsection{Prim based method}
+
+A fast straightforward algorithm determining an upper bound is inspired by Prim algorithm for minimum spanning tree problem.
+Initially, the constructed broadcast forest consists of all sources. 
+In each iteration, the nodes already in the forest are sorted according to the number of neighbors not yet added to the forest.
+For each node we then select a neighbor to be appended to the forest.
+\begin{algorithm}[]
+\KwData{$G=(V,E), S\subseteq V$}
+\textbf{for }$s\in S\textbf{ do }T_s=(V_s,A_s), V_s\leftarrow \{s\}, E_s\leftarrow\emptyset$\;
+$T=\{T_s:s\in S\}$\ ~~~\tcp{broadcast forest}
+$t\leftarrow 0$\;
+\While{$V_T\neq V$} {
+	Sort nodes in $V_T$ increasingly by the number of neighbors in $V\setminus V_T$\;
+	$W\leftarrow\emptyset$\;
+	\For{$u\in V_T$} {
+		Find $u\in V\setminus (V_T\cup W)$ according to some strategy\; 	
+	}
+	$E_T\leftarrow E_T\cup \{\{u,w\}: u\in V_T, w\in N(u)\cap W\}$\;
+	$V_T\leftarrow V_T\cup W$\;
+	$t\leftarrow t+1$\;
+}
+\Return t\;
+%\Return $\lceil k/s \rceil$\;
+ \caption{A greedy Prim-like method for determining an upper bound}
+\label{alg:match}
+\end{algorithm}
+
+\subsection{Restricted binomial tree method}
+
 The following inexact method is based on the idea of finding maximum cardinality matching in $G\left[V_{T_i},V\setminus V_{T_i}\right]$, and extending $T_i$ by this matching.
@@
 $I=\{1,\dots,2^k\}$\\
-$T=\{T_s:s\in S\}$\ ~~~\tcp{broadcast forest}
+$T=\{T_s:s\in S\}$\;
 $t\leftarrow 0$\;
@@
 %\Return $\lceil k/s \rceil$\;
- \caption{A method for determining an upper bound}
+ \caption{A method for determining an upper bound based on iterative search for pruned binomial trees}
 \label{alg:match}
@@
 The binomial trees are determined by solving the ILP model \eqref{mod:genmatch}. 
-This model is a modification of formulation \eqref{mod:partition}, and uses the same type of variables.
-The objective function is to maximize the number of nodes involved in the binomial trees.
-%The index set $I=\{1,\dots,2^k\}$ depends on the input parameter $k$.
-The constraints \eqref{mod:genmatch:nodeBelongs} state that each node belongs to at most one binomial tree.
-Compared to \eqref{mod:part:nodeBelongs}, \eqref{mod:genmatch:nodeBelongs} is an inequality, because the binomial trees do not necessarily form a partition of $G$, and so not all nodes have to be used.
-The remaining constraints are taken from formulation \eqref{mod:partition}. 
 Note that in every iteration, the model is solved with a different set $S$, as all nodes already included in the broadcast forest are roots of the binomial trees.
@@
 
-\begin{subequations}\label{mod:genmatch}
-\begin{align}
-\notag \max\sum\limits_{v\in V}&\sum\limits_{i\in I}\sum\limits_{s\in S}   y_{is}^v,\\
-\notag \text{s. t. } \\
-\label{mod:genmatch:nodeBelongs} \sum\limits_{i\in I}\sum\limits_{s\in S}y^v_{is} & \leq 1 & v\in V,\\
-\notag\eqref{mod:part:treeHasIJ} - \eqref{mod:part:followArcsB},\\
-%\label{mod:genmatch:treeHasIJ} \sum\limits_{v\in V'}y^v_{is} & = 1 & i\in I,s\in V_T,\\
-%\label{mod:genmatch:source1} y_{1s}^s & = 1  & s\in V_T,\\
-%\label{mod:part:noReturn} y^u_{ij}+y^v_{lj} &\leq 1 & i\in I,l\in C(i), j\in J, u\in V_\alpha,v\in V,\\
-%\label{mod:part:followArcs} y^u_{is}+y^v_{\ell s} &\leq 1 & i\in I,\ell\in C(i), s\in S, u,v\in V',(u,v)\not\in A',\\
-%\label{mod:part:followArcsA} y^u_{is}+y^v_{\ell s} &\leq 1 & i\in I,\ell\in C(i), s\in S, u,v\in V,\{u,v\}\not\in E,\\
-%\label{mod:part:followArcsB} y^{v_0}_{is}+y^v_{\ell s} &\leq 1 & i\in I,\ell\in C(i), s\in S, v\in V,\\
-%\label{mod:genmatch:followArcsA}y^{v_0}_{is}+y^u_{i s} + \sum\limits_{v\in V\setminus N(u)}y^v_{\ell s}&\leq 1 & u\in V,i\in I,\ell\in C(i), s\in V_T,  \\
-%\label{mod:genmatch:followArcsB}y^{v_0}_{is}+y^u_{\ell s} + \sum\limits_{v\in V\setminus N(u)}y^v_{i s}&\leq 1 & u\in V,i\in I,\ell\in C(i), s\in V_T,\\ 
-\label{mod:genmatch:dim}&&y \in \{0,1\}^{I\times S\times V'}.
-\end{align}~
-\end{subequations}
+Not only determine the algorithms the upper bound on the minimum broadcast time, but they also construct a solution with this broadcast time.
 

### 2018-08-12 sect3-exact-methods.tex
MI: the objective function in the decision version of the straightfwd model has been adjusted, becasue its original version did not give at most n-s, but in some instances gave something less than that. Now it is always n-s if the solution is feasible. The description has been updated accordingly.

@@
 The nature of MBT suggests another modelling approach derived from Model \ref{mod:basic}. 
-For a given maximum broadcast time $t_{\text{max}}$, we maximize the number of arcs along which a signal is relayed within $t_{\text{max}}$.
-If the optimal value is $n-1$, it is clear that we found a spanning broadcast tree with broadcast time $t_{\text{max}}$.
-In case the optimal value is less than $n-1$, the gien $t_{\text{max}}$ is insufficient and we have to try solving the problem with increased time limit.
+For a given maximum broadcast time $t_{\text{max}}$, we maximize the number of nodes $v$ that receive a signal from some neighbor $u$ within $t_{\text{max}}$ time steps.
+If the optimal value is $n-s$, it is clear that we found a spanning broadcast tree with broadcast time $t_{\text{max}}$.
+In case the optimal value is less than $n-s$, the given $t_{\text{max}}$ is insufficient, and we have to try solving the problem with increased time limit.
 Clearly, an additional knowledge of upper and lower bound spares some computations. 
 A lower bound is the initial value of $t_{\text{max}}$. 
-Similarly, if an upper bound $\bar{t}$ is known and $t_{\text{max}}=\bar{t}-1$, the process can be terminated when the objective function is less then $n-1$, because $\bar{t}$ is than the optimal value.
+Similarly, if an upper bound $\bar{t}$ is known and $t_{\text{max}}=\bar{t}-1$, the process can be terminated even when the objective function is less then $n-s$, because $\bar{t}$ is than the optimal value.
 
@@
 \begin{align}
-\label{mod:basic:dec:obj} \max \sum\limits_{(u,v) \in A}\sum\limits_{k=1}^{t_{\text{max}}}x_{u,v}^k \\ 
+\label{mod:basic:dec:obj} \max \sum\limits_{v \in V\setminus S}\sum\limits_{u \in N(v)} \sum\limits_{k=1}^{t_{\text{max}}}x_{uv}^k \\ 
 \notag \text{s. t. } \\
@@
 This was filtered out by optimality in \eqref{mod:basic}, but it must be forbidden explicitly in \eqref{mod:basic:dec} due to the changed objective function.
-
-The objective function \eqref{mod:basic:dec:obj} maximizes the number of arcs relaying a signal.
-The actual target function is the minimum time required for spreading the signal to all nodes, and so we are looking for the smallest value of $t_{text{max}}$.
+The actual target function is the minimum time required for spreading the signal to all nodes, and so we are looking for the smallest value of $t_{\text{max}}$.
 \subsection{Binomial tree model}

### 2018-08-13 sect4-lower-bounds.tex

@@
 and denote $d_i=\delta_G(v_i)$ ($i=1,\ldots,n$).
+
+Consider an optimal communication tree $T$. That is, $T$ is a subtree of $G$ rooted at the unique source $v_1$ such that
+$\tau\left(G,\{v_1\}\right)=\tau\left(T,\{v_1\}\right)$.
 Let $\tau_i=\tau\left(T[i],\{v_i\}\right)$ denote the broadcast time of node $v_i$ in the subtree $T[i]$ of $T$ rooted at $v_i$.
-The broadcast time $\tau\left(V,\{v_1\}\right)$ is hence no less than the sum of the time $\lceil\log\beta(v_i)\rceil$ it takes to inform $v_i$,
+The optimal broadcast time $\tau\left(T,\{v_1\}\right)$ is hence no less than the sum of the time $\lceil\log\beta(v_i)\rceil$ it takes to inform $v_i$,
 and the subsequent time $\tau_i$ needed to inform all nodes in $T[i]$.
-Consequently, if $T$ is an optimal broadcast tree,
+Taking the largest of all such lower bounds, we get:
 \begin{equation}
-  \tau\left(V,\{v_1\}\right)=\max\left\{\lceil\log\beta(v_i)\rceil+\tau_i: i=2,\ldots,n\right\}. \label{eq:tau}
+  \tau\left(T,\{v_1\}\right)=\max\left\{\lceil\log\beta(v_i)\rceil+\tau_i: i=2,\ldots,n\right\}. \label{eq:tau}
 \end{equation}

### 2018-08-13 sect5-upper-bounds.tex
MI: new experiments for size 50
MI: Alg. 2 and its description in Sect 5.1 fixed.

@@
 A knowledge of an upper bound $\bar{t}$ affects the number of variables in all studied models. 
-Particularly in the decision versions, the iterative approach can be terminated once the solution is dissatisfactory for broadcast time limit $\bar{t}-1$.
-The algorithms presented in this section iteratively construct broadcast foerest $T_i, i=1,2,\dots$.
-\subsection{Prim based method}
+Particularly in the decision versions, the iterative approach can be terminated once the solution is not satisfactory for broadcast time limit $\bar{t}-1$.
+The algorithms presented in this section iteratively construct broadcast forest $T_i, i=1,2,\dots$.
+\subsection{Matching heuristic}
 
-A fast straightforward algorithm determining an upper bound is inspired by Prim algorithm for minimum spanning tree problem.
-Initially, the constructed broadcast forest consists of all sources. 
+A fast straightforward algorithm determining an upper bound is based on repetitive construction of a matching between already connected nodes and their neighbors.
+Initially, the constructed broadcast forest consists of isolated sources. 
 In each iteration, the nodes already in the forest are sorted according to the number of neighbors not yet added to the forest.
 For each node we then select a neighbor to be appended to the forest.
+Let $T_s=(V_s,A_s)$ be the tree rooted at source $s\in S$, and let $T=(V_T, A_T)=\{T_s:s\in S\}$ be the broadcast forest.
+Further, nodest of the broadcast forest are numbered by integers $V_T=\{v_1,\dots,v_{|V_T|}\}$.
+Alg. \ref{alg:matchheur} formalizes the steps.
+%\begin{algorithm}[]
+%\KwData{$G=(V,E), S\subseteq V$}
+%\textbf{for }$s\in S\textbf{ do } V_s\leftarrow \{s\}, A_s\leftarrow\emptyset$\;
+%$t\leftarrow 0$\;
+%\While{$V_T\neq V$} {
+%	Sort nodes in $V_T$ increasingly by the number of neighbors in $V\setminus V_T$\;
+%	$W\leftarrow\emptyset$\;
+%	\For{$u\in V_T$} {
+%		Find $v\in V\setminus (V_T\cup W)$ such that $(u,v)\in A$\;
+%		$W\leftarrow W\cup\{v\}$\;
+%	}
+%	$A_T\leftarrow A_T\cup \{\{u,w\}: u\in V_T, w\in N(u)\cap W\}$\;
+%	$V_T\leftarrow V_T\cup W$\;
+%	$t\leftarrow t+1$\;
+%}
+%\Return t\;
+%%\Return $\lceil k/s \rceil$\;
+% \caption{A greedy matching heuristic method for determining an upper bound}
+%\label{alg:matchheur}
+%\end{algorithm}
+
 \begin{algorithm}[]
 \KwData{$G=(V,E), S\subseteq V$}
-\textbf{for }$s\in S\textbf{ do }T_s=(V_s,A_s), V_s\leftarrow \{s\}, E_s\leftarrow\emptyset$\;
-$T=\{T_s:s\in S\}$\ ~~~\tcp{broadcast forest}
+\textbf{for }$s\in S\textbf{ do } V_s\leftarrow \{s\}, A_s\leftarrow\emptyset$\;
 $t\leftarrow 0$\;
 \While{$V_T\neq V$} {
-	Sort nodes in $V_T$ increasingly by the number of neighbors in $V\setminus V_T$\;
+	Sort nodes in $V_T$ so that $i<j\Leftrightarrow |N(v_i)\cap(V\setminus V_T)|<|N(v_j)\cap (V\setminus V_T)|$\;
 	$W\leftarrow\emptyset$\;
-	\For{$u\in V_T$} {
-		Find $u\in V\setminus (V_T\cup W)$ according to some strategy\; 	
+	$D\leftarrow\emptyset$\;
+	\For{$i=1,\dots,|V_T|$} {
+		Find $v_j\in V\setminus (V_T\cup W)$ such that $(v_i,v_j)\in A$\;
+		$W\leftarrow W\cup\{v_j\}$\;
+		$D\leftarrow D\cup\{\left(v_i,v_j\right)\}$\;
 	}
-	$E_T\leftarrow E_T\cup \{\{u,w\}: u\in V_T, w\in N(u)\cap W\}$\;
 	$V_T\leftarrow V_T\cup W$\;
+	$A_T\leftarrow A_T\cup D$\;
 	$t\leftarrow t+1$\;
@@
 %\Return $\lceil k/s \rceil$\;
- \caption{A greedy Prim-like method for determining an upper bound}
-\label{alg:match}
+ \caption{A greedy matching heuristic method for determining an upper bound}
+\label{alg:matchheur}
 \end{algorithm}
+The selection of a neighbor (line 7) can be done arbitrarily, or according to some strategy such as neighbor with maximum degree is selected first.
+Note that for determining an upper bound, it is not necessary to store arcs sets $A_T$ and $D$, because we merely count the number of iterations.
+In order to construct an actual solution with broadcast time equal to the upper bound, the arcs sets are needed.
+Matchings created in each iteration are not necessarily maximum.
 

### 2018-08-14 sect3-exact-methods.tex
MI: Simplification of notation of Alg. 2 and a consistent update of Alg. 3. Fixed an error in constraint (1d). Constraints (4d) and (4e) updated to their stronger versions and the explanation in the text also changed. Although it is rather a description of what is written mathematically. Notation N(U) introduced in Sect. 2.

@@
 %\label{mod:basic:uniqueTout} \sum\limits_{v\in N(u)}x_{uv}^k & \leq 1  & u\in V,k=1,\dots,\bar{t},\\
-\label{mod:basic:tIncreases} \sum\limits_{v\in N(u)}x_{uv}^k &\leq\sum\limits_{\ell=1}^{k-1}\sum\limits_{w\in N(u)\setminus\{v\}} x_{wu}^{\ell}  & u\in V\setminus S, k=2,\dots,\bar{t},\\
+\label{mod:basic:tIncreases} \sum\limits_{v\in N(u)}x_{uv}^k &\leq\sum\limits_{\ell=1}^{k-1}\sum\limits_{w\in N(u)} x_{wu}^{\ell}  & u\in V\setminus S, k=2,\dots,\bar{t},\\
 %\label{mod:basic:tIncreases} x_{uv}^k &\leq\sum\limits_{\ell=1}^{k-1}\sum\limits_{w\in N(u)\setminus\{v\}} x_{wu}^{\ell}  & \{u,v\}\in E, u\not\in S, k=2,\dots,\bar{t},\\
@@
 %\label{mod:part:followArcsB} y^{v_0}_{is}+y^v_{\ell s} &\leq 1 & i\in I,\ell\in C(i), s\in S, v\in V,\\
-\label{mod:part:followArcsA}y^{v_0}_{is}+y^u_{i s} + \sum\limits_{v\in V\setminus N(u)}y^v_{\ell s}&\leq 1 & u\in V,i\in I,\ell\in C(i), s\in S,  \\
-\label{mod:part:followArcsB}y^{v_0}_{is}+y^u_{\ell s} + \sum\limits_{v\in V\setminus N(u)}y^v_{i s}&\leq 1 & u\in V,i\in I,\ell\in C(i), s\in S,\\ 
+%\label{mod:part:followArcsA}y^{v_0}_{is}+y^u_{i s} + \sum\limits_{v\in V\setminus N(u)}y^v_{\ell s}&\leq 1 & u\in V,i\in I,\ell\in C(i), s\in S,  \\
+%\label{mod:part:followArcsB}y^{v_0}_{is}+y^u_{\ell s} + \sum\limits_{v\in V\setminus N(u)}y^v_{i s}&\leq 1 & u\in V,i\in I,\ell\in C(i), s\in S,\\ 
+\label{mod:part:followArcsA}\sum\limits_{v\in \bar{N}(u)}y^v_{\ell s}&\leq \sum\limits_{v\in N(\bar{N}(u))}y_{is}^{v} & u\in V,i\in I,\ell\in C(i), s\in S,  \\
+\label{mod:part:followArcsB}\sum\limits_{v\in \bar{N}(\bar{N}(u))}y^v_{\ell s}&\leq \sum\limits_{v\in N(u)} y_{is}^{v} & u\in V,i\in I,\ell\in C(i), s\in S,\\ 
 \label{mod:part:yzrel}\sum\limits_{v\in V}y^v_{is} & \leq z_{\lceil\log i\rceil} & i\in I,s\in S,\\
@@
 The remaining two sets of constraints guarantee that the arcs of binomial trees follow edges in $E$.
-In particular, it is enforced by \eqref{mod:part:followArcsA} that if $u$ and $v$ are not adjacent in $G$, then $v$ must not act as a child of $u$ in any binomial tree.
-Constraints, \eqref{mod:part:followArcsB} forbid any node from $V$ to be a child of $v_0$ in any binomial tree. 
+%In particular, it is enforced by \eqref{mod:part:followArcsA} that if $u$ and $v$ are not adjacent in $G$, then $v$ must not act as a child of $u$ in any binomial tree.
+In particular, it is enforced by \eqref{mod:part:followArcsA} that if there is some  non-neighbor $v$ of $u$  with a label $\ell$ in a binomial tree rooted at $s$,
+then there must be a neighbor of some non-neighbor of $u$ with label $i$ in the same binomial tree.
+Similarly, Constraints \eqref{mod:part:followArcsB} state that if there is a node $v$ with label $\ell$ in a binomial tree rooted at $s$ in the complement of neighborhood of all non-neighbors of $u$,
+there must also be a neighbor of $u$ with label $i$ in this binomial tree. 
 This reflects the obvious fact that if a tree is pruned at some node, all its descendants must also be excluded from the tree.

### 2018-08-14 sect5-upper-bounds.tex
MI: Simplification of notation of Alg. 2 and a consistent update of Alg. 3. Fixed an error in constraint (1d). Constraints (4d) and (4e) updated to their stronger versions and the explanation in the text also changed. Although it is rather a description of what is written mathematically. Notation N(U) introduced in Sect. 2.

@@
 A knowledge of an upper bound $\bar{t}$ affects the number of variables in all studied models. 
-Particularly in the decision versions, the iterative approach can be terminated once the solution is not satisfactory for broadcast time limit $\bar{t}-1$.
-The algorithms presented in this section iteratively construct broadcast forest $T_i, i=1,2,\dots$.
+Particularly in the decision versions, the iterative approach can be terminated once the solution is found to be infeasible for broadcast time limit $\bar{t}-1$.
+The algorithms presented in this section iteratively construct broadcast forest $T=(V_T,A_T)$, where in the last iteration $V_T=V$.
 \subsection{Matching heuristic}
@@
 For each node we then select a neighbor to be appended to the forest.
-Let $T_s=(V_s,A_s)$ be the tree rooted at source $s\in S$, and let $T=(V_T, A_T)=\{T_s:s\in S\}$ be the broadcast forest.
-Further, nodest of the broadcast forest are numbered by integers $V_T=\{v_1,\dots,v_{|V_T|}\}$.
-Alg. \ref{alg:matchheur} formalizes the steps.
+Let $V_T=\{v_1,\dots,v_{|V_T|}\}$.
+The steps are formalized in Alg. \ref{alg:matchheur}.
 %\begin{algorithm}[]
@@
 \KwData{$G=(V,E), S\subseteq V$}
-\textbf{for }$s\in S\textbf{ do } V_s\leftarrow \{s\}, A_s\leftarrow\emptyset$\;
-$t\leftarrow 0$\;
+$V_T\leftarrow S, A_T \leftarrow \emptyset$\;
+$\bar{t}\leftarrow 0$\;
 \While{$V_T\neq V$} {
@@
 	$A_T\leftarrow A_T\cup D$\;
-	$t\leftarrow t+1$\;
+	$\bar{t}\leftarrow \bar{t}+1$\;
 }
-\Return t\;
+\Return $\bar{t}$\;
 %\Return $\lceil k/s \rceil$\;
@@
 
-The following inexact method is based on the idea of finding maximum cardinality matching in $G\left[V_{T_i},V\setminus V_{T_i}\right]$, and extending $T_i$ by this matching.
-It means that in each iteration $i$, the maximum possible number of nodes are informed.
-The maximum cardinality matching can be regarded as finding $|V_{T_i}|$ node-disjoint  binomial trees of order at most one with roots in $V_{T_i}$, maximizing the number of edges.
-By generalizing this notion, we iteratively look for $|V_{T_i}|$ node-disjoint (pruned) binomial trees of an arbitrary given order $k$ valued between 1 and $n-|S|$.
+The following inexact method is based on the idea of finding maximum cardinality matching in $G\left[V_T,V\setminus V_T\right]$, and extending $T$ by this matching.
+It means that in each iteration, the maximum possible number of nodes are informed.
+The maximum cardinality matching can be regarded as finding $|V_T|$ node-disjoint  binomial trees of order at most one with roots in $V_T$, maximizing the number of edges.
+By generalizing this notion, we iteratively look for $|V_T|$ node-disjoint (pruned) binomial trees of an arbitrary given order $k$ valued between 1 and $n-|S|$, but typically not more than 4.
 Even though this problem is NP-hard for $k\geq 2$ \cite{jansen95}, it is expected that the computational time is sound in most practical instances.
@@
 \KwData{$G=(V,E), S\subseteq V, k\in \{1,\dots,n-|S|\}, p\in \{1,\dots,2^k\}$}
-\textbf{for }$s\in S\textbf{ do }T_s=(V_s,A_s), V_s\leftarrow \{s\}, E_s\leftarrow\emptyset$\;
-$I=\{1,\dots,2^k\}$\\
-$T=\{T_s:s\in S\}$\;
-$t\leftarrow 0$\;
+$V_T\leftarrow S, A_T \leftarrow \emptyset$\;
+$\bar{t}\leftarrow 0$\;
 \While{$V_T\neq V$} {
@@
 	$V_T\leftarrow V_T\cup \{v:v\in V_B:\beta(v)\leq p\}$\;
-	$E_T\leftarrow E_T\cup \{\{u,v\}\in E_B: \beta(u)\leq p,\beta(v)\leq p\}$\;
-	$t\leftarrow t+1$\;
+	$A_T\leftarrow A_T\cup \{(u,v)\in A_B: \beta(u)\leq p,\beta(v)\leq p\}$\;
+	$\bar{t}\leftarrow \bar{t}+1$\;
 }
-\Return t\;
+\Return $\bar{t}$\;
 %\Return $\lceil k/s \rceil$\;
@@
 
-Not only determine the algorithms the upper bound on the minimum broadcast time, but they also construct a solution with this broadcast time.
+Like in Alg. \ref{alg:matchheur}, for calculating an upper bound, it is also not needed to store node set $A_T$ in Alg \ref{alg:match}, but it is essential when the actual broadcast tree is desirable.
 

### 2018-08-30 smt/ivanova.tex
MI: created folder with tex sources of revised version of SMT

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[natbib,smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
\bibliographystyle{aps-nameyear}
%
\usepackage{amsmath, amssymb}
\usepackage{graphicx}
\usepackage{amsfonts}
\usepackage{subcaption}
\usepackage{hyperref}
%\usepackage{pdflscape} % Landscape pages
\usepackage{color}	% Different font colors
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{mathtools}
\usepackage[]{algorithm2e}
\usepackage{color}
\usepackage{multirow} % Merging rows in tables
\usepackage{multicol} % Merging columns in tables
\usepackage{booktabs} % Horizontal rules in tables
\usepackage{mathrsfs} % Fancy math font for model labels
\usepackage{tikz}

\newcommand{\indentitem}{\setlength\itemindent{18pt}}
\renewcommand{\labelitemi}{$\bullet$}
\captionsetup{compatibility=false}
\newtheorem{prop}{Proposition}
\newtheorem{obs}{Observation}
\newtheorem{cor}{Corollary}
\newcommand\mycommfont[1]{\footnotesize\ttfamily\textcolor{blue}{#1}}
\SetCommentSty{mycommfont}
\SetKwRepeat{Do}{do}{while}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{} 
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%
% original amsmath definition
% \subequations:
% \long macro:->\refstepcounter {equation}\protected@edef \theparentequation {\theequation }\setcounter {parentequation}{\value {equation}}\setcounter {equation}{0}\def \theequation {\theparentequation \alph {equation}}\ignorespaces

% saved by hyperref in
% > \HyOrg@subequations=\long macro:

% hyperref-patched \subequations: (\endsubequations not modified)
% > \subequations=macro:
% ->\stepcounter {equation}\protected@edef \theHparentequation {\@ifundefined {th
% eHequation}\theequation \theHequation }\addtocounter {equation}{-1}\HyOrg@subeq
% uations \def \theHequation {\theHparentequation \alph {equation}}\ignorespaces 
% .

% extending the environment
%   1. with optional parameter: expected values resume or intermezzo or none.
%   2. while keeping the hyperref customization.

\makeatletter
\def\user@resume{resume}
\def\user@intermezzo{intermezzo}
%
\newcounter{previousequation}
\newcounter{lastsubequation}
\newcounter{savedparentequation}
\setcounter{savedparentequation}{1}
% 
% The following code causes that we are able to continue with numbering of a model in a next \align .
% For example:
%    a < b    (1a)
%    a < c    (1b)
% some text comes here and the numbering will continue:
%    a < d    (1c)
%
\renewenvironment{subequations}[1][]{%
      \def\user@decides{#1}%
      \setcounter{previousequation}{\value{equation}}%
      \ifx\user@decides\user@resume 
           \setcounter{equation}{\value{savedparentequation}}%
      \else  
      \ifx\user@decides\user@intermezzo
           \refstepcounter{equation}%
      \else
           \setcounter{lastsubequation}{0}%
           \refstepcounter{equation}%
      \fi\fi
      \protected@edef\theHparentequation{%
          \@ifundefined {theHequation}\theequation \theHequation}%
      \protected@edef\theparentequation{\theequation}%
      \setcounter{parentequation}{\value{equation}}%
      \ifx\user@decides\user@resume 
           \setcounter{equation}{\value{lastsubequation}}%
         \else
           \setcounter{equation}{0}%
      \fi
      \def\theequation  {\theparentequation  \alph{equation}}%
      \def\theHequation {\theHparentequation \alph{equation}}%
      \ignorespaces
}{%
%  \arabic{equation};\arabic{savedparentequation};\arabic{lastsubequation}
  \ifx\user@decides\user@resume
       \setcounter{lastsubequation}{\value{equation}}%
       \setcounter{equation}{\value{previousequation}}%
  \else
  \ifx\user@decides\user@intermezzo
       \setcounter{equation}{\value{parentequation}}%
  \else
       \setcounter{lastsubequation}{\value{equation}}%
       \setcounter{savedparentequation}{\value{parentequation}}%
       \setcounter{equation}{\value{parentequation}}%
  \fi\fi
%  \arabic{equation};\arabic{savedparentequation};\arabic{lastsubequation}
  \ignorespacesafterend
}
\makeatother

\begin{document}

\title{Integer Programming Formulations for the Shared Multicast Tree Problem
%\thanks{Grants or other notes
%about the article that should go on the front page should be
%placed here. General acknowledgments should be placed at the end of the article.}
}
% \subtitle{Tightening the LP bounds}
\titlerunning{IP formulations for SMT}        % if too long for running head

\author{Marika Ivanova        \and
         Dag Haugland
}
%\authorrunning{Short form of author list} % if too long for running head

\institute{M. Ivanova \at
              Department of Informatics, University of Bergen, Norway \\
              Tel.: +47 942 77 070\\
              \email{Marika.Ivanova@uib.no}         %  \\
%             \emph{Present address:} of F. Author  %  if needed
           \and
           D. Haugland \at
              Department of Informatics, University of Bergen, Norway \\
              Tel.: +47 55584033\\
              \email{Dag.Haugland@uib.no}         %  \\
}

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor
\maketitle

\begin{abstract}
We study the Shared Multicast Tree (SMT) problem in wireless networks.
To support a multicast session between a set of network nodes, SMT aims to establish a wireless 
connection between them, such that the total energy consumption is minimized.
All destinations of the multicast message must be connected, while non-destinations are optional nodes that can be used to relay messages. 
In the well-studied minimum energy multicast (MEM)
problem, the optimal tree depends on the source of the message.
In contrast, SMT asks for a source-independent tree, which substantially reduces the processing overhead at each node.
This means that SMT can be regarded as a Steiner tree problem.
The objective function reflecting power consumption, however, distinguishes SMT clearly from the traditional minimum Steiner tree problem.

In the present work, we develop two 
integer programming formulations for SMT.
Both models are subsequently extended and strengthened by redundant valid inequalities.
Further strengthening is achieved by extending the set of variables and corresponding constraints.
Theorems on relations between the LP bounds corresponding to the models are stated and proved.
As the number of variables in the strongest formulations is a polynomial of degree four in the number of network nodes,
the models are impractical for computing lower bounds in instances beyond a fairly small size.
To make them applicable to larger instances, a constraint generation scheme is developed.
Results from computational experiments with the models demonstrate good promise of the approaches taken.
\keywords{Wireless communication, multicast tree, Steiner tree, LP bound, valid inequalities}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\input{sec1-intro.tex}
\input{sec2-sbtsmt.tex}
\input{sec3-milp.tex}
\input{sec4-relations.tex}
\input{sec5-strong.tex}
\input{sec6-exp.tex}
\input{sec7-conclusion.tex}

%as required. Don't forget to give each section
%and subsection a unique label (see Sect.~\ref{sec:1}).
%\paragraph{Paragraph headings} Use paragraph headings as needed.
%\begin{equation}
%a^2+b^2=c^2
%\end{equation}

% For one-column wide figures use
%\begin{figure}
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
%  \includegraphics{example.eps}
% figure caption is below the figure
%\caption{Please write your figure caption here}
%\label{fig:1}       % Give a unique label
%\end{figure}
%
% For two-column wide figures use
%\begin{figure*}
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
%  \includegraphics[width=0.75\textwidth]{example.eps}
% figure caption is below the figure
%\caption{Please write your figure caption here}
%\label{fig:2}       % Give a unique label
%\end{figure*}
%
% For tables use
%\begin{table}
% table caption is above the table
%\caption{Please write your table caption here}
%\label{tab:1}       % Give a unique label
% For LaTeX tables use
%\begin{tabular}{lll}
%\hline\noalign{\smallskip}
%first & second & third  \\
%\noalign{\smallskip}\hline\noalign{\smallskip}
%number & number & number \\
%number & number & number \\
%\noalign{\smallskip}\hline
%\end{tabular}
%\end{table}

%\begin{acknowledgements}
%If you'd like to thank anyone, place your comments here
%and remove the percent signs.
%\end{acknowledgements}

% BibTeX users please use one of
%\bibliographystyle{spbasic}      % basic style, author-year citations
%\bibliographystyle{spmpsci}      % mathematics and physical sciences
%\bibliographystyle{spphys}       % APS-like style for physics
%\bibliography{}   % name your BibTeX data base

% Non-BibTeX users please use
\begin{thebibliography}{}

%
% and use \bibitem to create references. Consult the Instructions
% for authors for reference list style.
%
% \bibitem[Dian{\'e} and Plesn{\'i}k(1993)]{diane93ipf}
% Dian{\'e} M, Plesn{\'i}k J (1993)
% An integer programming formulation of the Steiner problem in graphs.
% Zeitschrift f{\"u}r Operations Research
% 37(1):107--111

\bibitem[Althaus et al.(2003)]{althaus03}
Althaus E, C\u{a}linescu G, M\u{a}ndoiu II, Prasad S, Tchervenski N, Zelikovsky A (2003)
Power efficient range assignment in ad-hoc wireless networks.
In: Proceedings IEEE WCNC’03, pp 1889--1894

\bibitem[Altinkemer et al.(2005)]{altinkemer05}
Altinkemer K, Salman FS, Bellur P (2005)
Solving the minimum energy broadcasting problem in ad hoc wireless networks by integer programming.
In: Proceedings INOC 2005, pp B2.635--B2.642

\bibitem[Bauer et al.(2008)]{bauer08}
Bauer J, Haugland D, Yuan D (2008)
Analysis and computational study of several integer programming formulations for minimum-energy multicasting in wireless ad hoc networks.
Networks 52(2):57–68

\bibitem[\v{C}agalj et al.(2002)]{cagalj02}
\v{C}agalj M, Hubaux J-P, Enz C (2002)
Minimum-energy broadcast in all-wireless networks: NP-completeness and distribution issues.
In: Proceedings ACM MOBICOM 2002 pp 172--182

\bibitem[Clementi et al.(1999)]{clementi99}
Clementi A, Penna P, Silvestri R (1999)
Hardness results for the power range assignment problem in packet radio networks.
In: Hochbaum D et al. (eds) RANDOM-APPROX’99,
LNCS 1671:197--208

\bibitem[Das et al.(2003)]{das03}
Das AK, Marks RJ, El-Sharkawi M, Arabshani P, Gray A (2003)
Minimum power broadcast trees for wireless networks: integer programming formulations.
In: Proceedings IEEE INFOCOM 2003 Vol.\ 2, pp 1001--1010

\bibitem[Das et al.(2005)]{das05}
Das AK, Marks RJ, El-Sharkawi E, Arabshahi P, Gray A (2005)
Optimization methods for minimum power bidirectional topology construction in wireless networks with sectored antennas.
In: Proceedings IEEE GLOBECOM ’05, pp 3962--3968

\bibitem[Goemans and Myung(1993)]{goemans93catalog}
Goemans MX, Myung Y-S (1993)
A catalog of Steiner tree formulations.
Networks
23(1):19--28

\bibitem[Haugland and Yuan(2011)]{Haugland11Compact}
Haugland D, Yuan D (2011)
Compact integer programming models for power-optimal trees in ad hoc wireless networks.
In: Kennington J, Olinick E, Rajan D (eds) Wireless network design - optimization models and solution procedure.
International series in operations research \& management science.
Springer, New York, pp 219--246

\bibitem[Hsiao et al.(2013)]{hsiao13}
Hsiao P-C, Chiang T-C, Fu L-C (2013)
Static and dynamic minimum energy broadcast problem in wireless ad-hoc networks: A PSO-based approach and analysis
Applied Soft Computing
13:786--4801

\bibitem[Ivanova(2016)]{ivanova16isco}
Ivanova M (2016)
Shared Multicast Trees in Ad Hoc Wireless Networks.
In: Cerulli R, Fujishige S, Mahjoub A (eds) Combinatorial Optimization ISCO 2016,
LNCS 9849:273--284

\bibitem[Kirousis et al.(1997)]{kirousis97}
Kirousis LM, Kranakis E, Krizanc D, Pelc A (1997)
Power consumption in packet radio networks.
In: Reischuk M (ed): STACS'97 Proceedings,
LNCS 1200:363--374

\bibitem[Leggieri et al.(2008)]{leggieri08}
Leggieri V, Nobili P, Triki C (2008)
Minimum power multicasting problem in wireless networks.
Math Meth Oper Res 68:295–311

\bibitem[Montemanni and Gambardella(2004)]{montemanni04}
Montemanni R, Gambardella LM (2004)
Exact algorithms for the minimum power symmetric connectivity problem in wireless networks.
Comput Oper Res 31(10):1667--1680

\bibitem[Montemanni and Leggieri(2011)]{montemanni11}
Montemanni R, Leggieri V (2011)
A branch and price algorithm for the minimum power multicasting problem in wireless sensor networks.
Math Meth Oper Res
74:327--342

\bibitem[Papadimitriou and Georgiadis(2006)]{Papadimitriou06SBT}
Papadimitriou I, Georgiadis L (2006)
Minimum-energy Broadcasting in Multi-hop Wireless Networks Using a Single Broadcast Tree.
Mobile Netw Appl
11(3):361--375

\bibitem[Polzin and Daneshmand(2001)]{Polzin}
Polzin T, Daneshmand SV (2001)
A comparison of Steiner tree relaxations.
Discrete Appl Math
112(1--3):241--261

\bibitem[Wieselthier et al.(2000)]{Wieseltier00onthe}
Wieselthier JE, Nguyen GD, Ephremides A (2000)
On the Construction of Energy-Efficient Broadcast and Multicast Trees in Wireless Networks,
In: Proceedings IEEE INFOCOM 2000 Vol.\ 2, pp 585--594

\bibitem[Yuan(2005)]{yuan05}
Yuan D (2005)
An integer programming approach for the minimum-energy broadcast problem in wireless networks.
In: Proceedings INOC 2005, pp B2.643–B2.650

\bibitem[Yuan et al.(2008)]{yuan08}
Yuan D, Bauer J, Haugland D (2008)
Minimum-energy broadcast and multicast in wireless networks: An
integer programming approach and improved heuristic algorithms.
Ad Hoc Netw
6(5):696–717

\bibitem[Yuan and Haugland(2012)]{Haugland12Dual}
Yuan D, Haugland D (2012)
Dual Decomposition for Computational Optimization of Minimum-Power Shared Broadcast Tree in Wireless Networks.
IEEE T Mobile Comput
12(11):2008--2019

%\bibitem{RefJ}
% Format for Journal Reference
%Author, Article title, Journal, Volume, page numbers (year)
% Format for books
%\bibitem{RefB}
%Author, Book title, page numbers. Publisher, place (year)
% etc
\end{thebibliography}

\end{document}
% end of file template.tex

### 2018-08-30 smt/sec1-intro.tex
MI: created folder with tex sources of revised version of SMT

\section{Introduction}
\label{intro}

Wireless ad hoc networks (WANETs) are suitable in various practical applications in both civil and military sector, where a communication infrastructure is either absent or cannot be relied on.
Owing to their quick deployment and simple configuration, WANETs are useful in emergency operations such as natural disaster relief and military command and control.
Recently, the WANET concept has been introduced in ad hoc smart lighting in homes and in offices as well as ad hoc street light networks, where the control includes adjusting dimmable lights.

The purpose of a multicast communication session in a WANET is to route information from a sending device to a set of receiving devices.
Connection links are established by assigning sufficient power to the sending devices.
Since the devices typically use batteries as power supply, they are heavily energy-constrained.
Therefore, power efficiency is an important measure in designing WANETs.
Individual devices work as transceivers, which means that they have the ability to both transmit and receive a signal.
Moreover, the power level of a device can be dynamically adjusted during a multicast session.
\citet{montemanni11} name several applications of wireless sensor networks where power saving is essential.

Unlike wired networks, where signal passing takes place along predefined links, nodes in WANETs use omnidirectional antennas, and hence a message reaches all nodes within the communication range of its sender.
This range is determined by the power assigned to the sender, which is the maximum rather than the sum of the powers necessary to reach all intended receivers.
This feature is often referred to as the \emph{wireless advantage} \citep{Wieseltier00onthe}.

A well-known and extensively studied problem in the network design literature is the \textsc{Minimum Energy Broadcast} (MEB) problem.
Given a set of wireless devices with one designated source node, the goal is to assign powers to individual nodes.
The power assignments determine the communication ranges, and induce a broadcast tree such that a signal initiated by the source reaches, directly or via other nodes,
all the remaining nodes.
In the MEB problem, the task is to find power assignments enabling such a signal broadcast while minimizing the total energy consumption.
A natural and frequently addressed extension of MEB, is the \textsc{Minimum Energy Multicast} (MEM) problem.
Analogous to the difference between the \textsc{Minimum Spanning Tree} and the \textsc{Minimum Steiner Tree} problems,
MEM is distinguished from MEB in that not all nodes in the network are destinations of the multicast message.
Hence, the more general problem amounts to finding a Steiner tree spanning the destination nodes such that the sum of the power assignments is minimized.
Non-destination nodes are allowed in the Steiner tree, and will be included to the extent that they contribute to reduced power consumption.

In certain practical settings, MEB fails to reflect reality adequately.
Consider an application where more than one node can act as a source.
% Every node may initiate a message intended for the remaining nodes.
The optimal broadcast trees corresponding to two different sources are not necessarily identical,
and the optimal broadcast trees must be calculated separately for every possible source node.
In order to relay a received signal correctly, the nodes must be able to recognize which node initiated the received signal, and therefore also determine which broadcast tree to use.
Further, which power level to be set must be assessed by the relaying node.
Such overhead calculations require additional energy, and enabling the communication devices to accomplish them may even be impractical.

The idea behind the \textsc{Minimum Shared Broadcast Tree} (SBT) problem is to maintain a single, shared broadcast tree, to be applied regardless of which node is the signal source.
Such a tree is not necessarily optimal for any of the individual sources, but routing at each node is considerably simplified.
Provided that a shared broadcast tree is used, the nodes are no longer required to identify the source of the message in order to set a correct power level.
Instead, only the immediate neighbour from which the signal was received must be recognized.
The objective function in SBT captures not only the power levels of the nodes, but depends also on how frequently a node actually transmits with a certain power level.

The combinatorial optimization literature is already rich on Integer Programming (IP) approaches to power minimization in wireless networks.
Since the work of \citet{Wieseltier00onthe}, in particular the MEB and MEM problems have been studied extensively.
Noteworthy contributions to this research include the IP models studied by \citet{das03},
\citet{altinkemer05}, \citet{leggieri08}, \citet{yuan05}, \citet{yuan08}, \citet{bauer08}, and \citet{montemanni11}.
\citet{cagalj02} proved that MEB, and thereby MEM, is NP-hard.
Numerous contributions to fast, but possible inexact methods for MEM have thus appeared, and the interested reader is referred to \citep{hsiao13} for an overview.

The SBT problem has also received some attention in the scientific literature.
It was introduced by \citet{Papadimitriou06SBT}, who also proved that the problem is NP-hard.
In the same article, an approximation algorithm for SBT was developed and analyzed experimentally.
Building on this work, \citet{Haugland12Dual} presented an integer programming model for SBT, along with an associated decomposition approach.

The SBT and SMT problems resemble the more frequently studied \textsc{Range Assignment Problem} (RAP) in the sense that they all ask for an undirected source-independent tree of minimum power.
However, RAP concerns cases where a connected topology must be established by means of \emph{bidirectional} links,
with the feature that power must be assigned to both the sending and the receiving device in each link.
This scenario induces a spanning tree problem, where costs are incurred only for edges that are the most expensive ones incident to any of their end nodes.
Simple instances \citep{Haugland12Dual} show that because of the difference in objective functions, the optimal solutions to RAP and SBT can differ substantially.
\citet{kirousis97} and \citet{clementi99} conduct
hardness results for instances of RAP embedded in three- and two-dimensional Euclidean space, respectively.
Later, IP models have been developed by e.g.\ \citet{althaus03}, \citet{montemanni04}, \citet{das05}, and \citet{Haugland11Compact}.

For the same reasons that MEB has been generalized to MEM, it is natural and desirable to extend SBT to its multicast equivalent.
The \textsc{Minimum Shared Multicast Tree} (SMT) problem is thus at the forefront of this paper,
and neither MEB, MEM, nor RAP will be considered further.
Focus is henceforth on instances in which some of the nodes neither initiate new messages nor need messages from others.
Utilizing such \emph{non-destination} nodes to forward messages can in many cases contribute to substantial reductions in the total power consumption.
Non-destination nodes thus have the potential to play the role of Steiner nodes in the multicast tree.
All devices that can initiate a transmission, referred to as \emph{destination} nodes, also have to receive every message.

Some preliminary work on SMT has recently been published.
\citet{ivanova16isco} introduces the problem, and develops an IP model.
Moreover, the author studies inexact construction and improvement methods, for which computational results are reported.
Contributions from the present work include two sets of IP models for SMT, presented in order of increasing lower bounding capabilities.
By rigorous analysis of the models, we prove relations between the bounds offered by the corresponding continuous relaxations.
Intended for instances beyond the size that can be solved to optimality, we also develop a constraint generation procedure by which strong bounds can be computed.

The remainder of this paper is organized as follows: Section \ref{sec:SBT} describes the SMT problem and gives detailed explanation of its objective function.
Integer linear programming formulations, valid inequalities and their analysis are presented in Section \ref{sec:ILP},
followed by Section \ref{sec:comp}, where the models under study are compared with respect to strength.
In Section \ref{sec:strong}, the models are further extended by variables in higher dimensional spaces,
and a constraint generation procedure is developed.
Computational results are reported in Section \ref{sec:exp},
and suggestions to future work are summarized in Section \ref{sec:conclusion}.

### 2018-08-30 smt/sec2-sbtsmt.tex
MI: created folder with tex sources of revised version of SMT

\section{The Minimum Shared Multicast Tree Problem}
\label{sec:SBT}

\subsection{Assumptions and notation} \label{sec:notation}

A WANET is modeled by a complete graph $G=(V,E)$, where the set $V$ of nodes represents the set of wireless devices and the set of edges $E=\{\{i,j\}:i,j\in V, i\neq j\}$ corresponds to the potential links between them.
The set $A=\{(i,j):i,j\in V,\{i,j\}\in E\}$ consists of all arcs that can be derived by directing the edges in $E$.
Nodes in the set $D\subseteq V$ are the destination nodes, also referred to as \emph{source} nodes.
For an arbitrary $i\in V$, sets $V\setminus \{i\}$ and $D\setminus\{i\}$ are denoted $V_i$ and $D_i$, respectively.

Sending a message directly from (to) node $i$ to (from) node $j$ requires a certain \emph{power} $p_{ij}=p_{ji}$.
Whenever appropriate, $p_{ij}$ is referred to as the \emph{cost} of $\{i,j\}$,
and edges with higher cost than others are said to be \emph{more expensive}.
When the nodes are embedded in the plane, $p_{ij}$ is often assumed \citep{Papadimitriou06SBT,Haugland12Dual} to be
proportional to the Euclidean distance between $i$ and $j$ raised to a power between 2 and 4.
For simplicity, we assume for all $i\in V$ that $p_{ij}\neq p_{ik}$ when $j\neq k$, and define
$W_{ij}=\left\{k\in V_i: p_{ik}\geq p_{ij}\right\}$
as the set of nodes requiring no less power at $i$ than $j$ does.
The notation $G'\subseteq G$ means that $G'=(V',E')$ is a subgraph of $G$, i.e., $E'\subseteq E$ consists of the edges in $E$ with both end nodes in $V'\subseteq V$.

Let $y \in \{0,1\}^E$ and $X\in\{0,1\}^{A\times D}$ be binary vectors with components corresponding to
edges and pairs of arcs and destination nodes, respectively.
The vector $X^s$ then consists of all components of $X$ corresponding to a pair of which $s$ is the destination.
The subgraph of $G$ \emph{induced by} $y$ is defined as $G_y=(V,E_y)$, where $E_y=\left\{\{i,j\}\in E: y_{ij}=1\right\}$.
Directed graphs induced by binary vectors over $A$ are defined analogously.

Several integer programming models, all of which exclusively have binary variables, will be considered in this text.
If $\mathcal{M}$ is such a model, $\text{LP}(\mathcal{M})$ denotes the corresponding continuous relaxation
where the domain of each variable is replaced by the interval $[0,1]$.
Further, $z\left(\mathcal{M}\right)$ and $z\left(\text{LP}(\mathcal{M})\right)$ denote the optimal objective function values of the respective models.

\subsection{Problem definition} \label{sec:probdef}

To illustrate the problem under study, consider the tree depicted in Fig. \ref{fig:objexp}.
Six destination nodes ($D=\{i_1,i_2,a,b,c,d\}$) are spanned by use of three Steiner nodes ($V\setminus D = \{e,i,i_3\}$).
Consider node $i$ and its three neighbour nodes $i_1$, $i_2$ and $i_3$, ordered by decreasing power requirement at $i$.
That is, $p_{ii_1}>p_{ii_2}>p_{ii_3}$.
If the transmitting node is outside the shaded area ($i_2$, $c$, or $d$), then node $i$ relays the signal to two of its neighbors, \emph{including $i_1$}.
Exploiting the wireless advantage, node $i$ reaches both neighbors if it is assigned power level $p_{ii_1}$.
Otherwise, if the signal source is either $i_1$, $a$, or $b$, node $i$ passes the signal on to neighbors $i_2$ and $i_3$.
This is accomplished by power $p_{ii_2}$.

\begin{figure}[h!]
\centering
\begin{tikzpicture}[scale=.6]
 \draw[rounded corners=5mm, fill=gray!20, draw=white] (-.7,-.2) -- (3.5,0.7) -- (0.2,4.2) -- cycle;
 \begin{scope}[every node/.style={circle,draw=black,fill=gray!60,minimum size=1.5em, inner sep=2, scale=.8}]
  \node (a) at (0,0.5) {$a$};
  \node (b) at (0.5,3) {$b$};
  \node (c) at (9,2.5) {$c$};
  \node (d) at (9.5,0.9) {$d$};
  \node[scale=.85] (i1) at (2,1) {$i_1$};
  \node[scale=.85] (i2) at (4,3.1) {$i_2$};
  \node[scale=.85,fill=none] (i3) at (5.8,1.2) {$i_3$};
  \node[fill=none] (i) at (4.5,1.5) {$i$};
  \node[fill=none] (e) at (8,1.5) {$e$};
 \end{scope}
 \draw (a) -- (i1) -- (b);
 \draw (i1) -- (i) -- (i3) -- (e) -- (c);
 \draw (e) -- (d);
 \draw (i) -- (i2);
\end{tikzpicture}
\caption{A multicast tree illustrating the cost contribution from node $i$ to the objective function.
	 Destinations and Steiner nodes are denoted by filled and empty circles, respectively.}
\label{fig:objexp}
\end{figure}

In more general terms, let $T$ be a multicast tree, and let $i\in V_T$ be a node with at least two neighbors in $T$.
For multicasting a message, the power to be assigned to node $i$ is either $p_{ii_1}$ or $p_{ii_2}$, where $i_1$ and $i_2$ are neighbors of $i$ such that
$p_{ii_1}>p_{ii_2}>p_{ij}$ for all other neighbors $j\in V_T\setminus\{i_1,i_2\}$.
If the path from the message source to $i$ intersects $i_1$, then power $p_{ii_2}$ is sufficient.
Otherwise, the power $p_{ii_1}$ is required.
At a leaf $i$ with neighbor $i_1$ in $T$, the power assignment is $p_{ii_1}$ if $i$ also is the message source, and zero otherwise.

All nodes in $D$ are assumed to be equally likely to initiate a message.
The frequency at which power levels $p_{ii_1}$ and $p_{ii_2}$ (let $p_{ii_2}=0$ if $i$ is a leaf) are required,
is thus proportional to the number of sources in $D$ that utilize the respective arc for message forwarding.
In the example in Fig.\ \ref{fig:objexp}, the expected power consumption at node $i$ is thus proportional to $4p_{ii_1} + 3p_{ii_2}$.
Summing over all nodes yields the total cost of the tree.
% The objective in the problem addressed in the current work is to minimize this cost.

To express the cost of $T$ more rigorously, let $T^s=(V_T,A^s)$ denote the directed tree (arborescence) obtained by directing all edges in $E_T$ such that they point away from the \emph{root} at $s$.
For each arc $(i,j)\in A$, we let $n_{ij}(T)$ denote the number of destinations $s$ such that $(i,j)$ is the most expensive arc leaving node $i$ in $T^s$.
That is,
\[
  n_{ij}(T) = \left|\left\{s\in D: p_{ij}=\max_{k\in V_i: (i,k)\in A^s}p_{ik}\right\}\right|,
\]

\noindent
and the total cost of tree $T$ becomes:
$$
  c(T) = \sum_{i\in V_T}\sum_{j\in V_T:\{i,j\}\in E_T}p_{ij}n_{ij}(T).
$$ 
The problem under study is thus defined in network optimization terms:

\begin{problem}
\label{def:problem}
\textsc{Minimum Shared Multicast Tree}: Find a tree $T\subseteq G$ spanning $D$ such that $c(T)$ is minimized.
\end{problem}

While the feasible domains of the \textsc{Minimum Steiner Tree} problem and SMT are identical,
the cost functions differ.
In the former problem, all edges present in the tree contribute to the objective function by a cost that is given by the input data.
The input cost factor $p_{ij}$ is, in the case of SMT, scaled by a topology dependent factor $n_{ij}(T)$.
Only edges that are the most or second-most expensive edge incident to either of their end nodes get a positive scaling factor.
Some edges in $T$ can thus be scaled down to zero cost, while the most expensive ones are scaled up.

Problem \ref{def:problem} also has close resemblance with the SBT problem, which is the special case of SMT where $D=V$.
Like many other network design problems, \textsc{Minimum Shared Multicast Tree} is NP-hard.
This follows directly from the NP-hardness of SBT \citep{Papadimitriou06SBT}.

In the forthcoming sections, we develop and analyze IP models for SMT.
All formulations under study are extensions of models previously developed for the related problems mentioned above.

### 2018-08-30 smt/sec5-strong.tex
MI: created folder with tex sources of revised version of SMT

\section{Stronger Formulations}
\label{sec:strong}

Aiming for stronger LP bounds than those provided by the relaxations of the models introduced in Section \ref{sec:ILP},
this section is devoted to extensions of $\mathcal{X}_2$ and $\mathcal{F}_2$.
Variables with four node indices are added to both models.
Constraints connecting new and original variables then contribute to the improved bounds.

\subsection{Extension of $\mathcal{X}_2$ by destination-to-destination variables [$\mathcal{X}_3$]}
\label{sec:smtx3}

Let $A(D)=A\cap(D\times D)$ be the set of ordered pairs of distinct destination nodes.
For a tree $T$ solving Problem \ref{def:problem}, define the variables $x\in\{0,1\}^{A\times A(D)}$, where
$x_{ij}^{st}=1$ iff $(i,j) \in A$ is an arc on the path from $s$ to $t$ in arborescence $T^s$.
% \newline\newline
% $x_{ij}^{st}=
% \begin{cases}
%     1 & \text{if $(i,j) \in A$ is an arc on the path from $s$ to $t$ in $T$, $(s,t)\in A(D)$},\\
%     0 & \text{otherwise}.
% \end{cases}$
% \newline\newline
Model $\mathcal{X}_2$ is strengthened by constraints on $x$, which yields model $\mathcal{X}_3$: 
\newline
\newline
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                     % 
%             MODEL X3                %
%             (4 idx vars)            %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{subequations}[resume]\label{eq:smt-x3}
\begin{align}
\notag\makebox[0pt][l]{$\displaystyle{} \min \sum\limits_{(i,j) \in A} p_{ij} \sum\limits_{s \in D} \pi^s_{ij} $}\\
\text{s.t.} \notag \\
(\ref{con:dd:arrowFromDest}) - (\ref{con:dd:vardim}) &\text{ and }(\ref{con:dd:yvar}) - (\ref{con:vi:sumYImpSumX}) \notag\\
\label{con:mf:flowNormal} \sum\limits_{\substack{ j\in V_i}}x^{st}_{ij}-\sum\limits_{\substack{j\in V_i }}x^{st}_{ji} & = 0 & (s,t)\in A(D), i\in V\setminus\{s,t\},\\
\label{con:mf:flowDest} \sum\limits_{\substack{j\in V_t}}x^{st}_{jt}    & = 1  & (s,t)\in A(D),\\
\label{con:mf:fcap} x^{st}_{ij} &\leq X^{s}_{ij}& (i,j)\in A, (s,t)\in A(D),\\
\label{con:mf:fsym} x^{st}_{ij} &=  x^{ts}_{ji}& (i,j)\in A, (s,t)\in A(D),\\ 
\label{con:vi:sumFImpSumY} \sum_{k\in W_{ij}}x^{st}_{ik} & \leq \sum_{k\in W_{ij}} \pi^{s}_{jk} & (i,j)\in A, (s,t)\in A(D),\\
\label{con:mf:xydim} x&\in\{0,1\}^{A\times A(D)}. 
\end{align}~
\end{subequations}

Variable $x_{ij}^{st}$ is also interpreted as the \emph{flow of commodity $(s,t)$} along arc $(i,j)$.
Correspondingly, (\ref{con:mf:flowNormal})--(\ref{con:mf:flowDest}) can be viewed as flow conservation constraints,
stating that there exists a path from $s$ to $t$ in $T^s$.
Next, constraints (\ref{con:mf:fcap}) express the obvious relation between $X$ and $x$, that $(i,j)$ is an arc on the path from $s$ to $t$ only if it exists in $T^s$.
The flow symmetry equations (\ref{con:mf:fsym}) state that the path from $s$ to $t$ contains $(i,j)$ iff the reverse path contains the reverse arc $(j,i)$.
Finally, constraints (\ref{con:vi:sumFImpSumY}) reflect that if the path from $s$ to $t$ intersects $(i,j)$,
then the most expensive arc leaving $i$ in $T^s$ has cost $p_{ij}$ or more.

% In terms of multicast trees, $x_{ij}^{st}=1$ if and only if $(i,j)$ carries a signal that is initiated in $s$ and destined for $t$.
% Consider nodes $i,j\in V$ and a pair $(s,t)$ of destinations.
% If a message from $s$ to $t$ is sent through $(j,k)$ such that $p_{jk}\geq p_{ji}$, then the message from $s$ must be relayed by $j$ using power level at least $p_{ji}$.
% This is expressed by valid inequality (\ref{con:vi:sumFImpSumY}).

\subsection{Extension of $\mathcal{F}_2$ by variables of joint flow [$\mathcal{F}_3$]}

Similarly to the extension $\mathcal{X}_3$ of $\mathcal{X}_2$ by $x$-variables, the model $\mathcal{F}_2$ is extended by variables with four node indices.
Analogously to $A(D)$, let $E(D_0)=E\cap (D_0\times D_0)$ be the set of pairs of \emph{unordered} destination nodes other than $s_0$.
Define variables $f\in\{0,1\}^{A\times E(D)}$, where
$f_{ij}^{st}=f_{ji}^{ts}$ takes the value 1 iff the path in $T^0$ from $s_0$ to $s$ intersects with the path from $s_0$ to $t$ at arc $(i,j)$.
% \newline\newline
% $f^{st}_{ij}=f^{ts}_{ij}=
% \begin{cases}
%     1 & \text{if the paths in $T^0$ from $s_0$ to respectively $s$ and $t$ intersect} \\
%       & \text{at $(i,j)$,} \\
%     0 & \text{otherwise.}
% \end{cases}$
% \newline\newline
%
Variables $f_{ij}^{st}$ can also be regarded as a the \emph{joint flow} from $s_0$ to $s$ and $t$ \citep{Polzin}.
In the cited work, they are successfully used to strengthen formulations for the \textsc{Minimum Steiner Tree} problem.
The new variables enable formulation of an extended model denoted $\mathcal{F}_3$:

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                     %
%             MODEL F3                %
%                                     %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{subequations}[resume]
\begin{align}
\notag\min & \sum\limits_{(i,j) \in A} p_{ij} \sum\limits_{s \in D} \pi^s_{ij}\\ \notag
\text{s.t.}& && \\
\notag(\ref{con:pf1:flow})-(\ref{con:vi:sumYImpSumXTrans}) & ~\text{and} ~(\ref{con:vi:Y1}) \\
\label{con:pf2:flowHook} \sum\limits_{\substack{ j \in V_i }}\left(f^{st}_{ij}-f^{st}_{ji}\right) & \leq 
  \begin{cases}
    1, & i = s_0, \\
    0, & i\in V_0,
  \end{cases} \quad \{s,t\}\in E(D_0), \\
\label{con:pf2:startInSource}  f^{st}_{ij} & \leq F^s_{ij} \quad (i,j)\in A, \{s,t\}\in E(D_0),\\
\label{con:pf2:stopInDest}  f^{st}_{ij} & \leq F^t_{ij} \quad (i,j)\in A, \{s,t\}\in E(D_0),\\
\label{con:pf2:stronger}  F^{s}_{ij}+F^{t}_{ij}-f^{st}_{ij} &\leq g_{ij} \quad (i,j)\in A, \{s,t\}\in E(D_0),\\ 
\label{con:pf2:sumFImpSumY} \sum\limits_{k\in W_{ij}}\left(F^{t}_{ik} + F^{s}_{ki} - f^{st}_{ik} - f^{st}_{ki}\right) & \leq \sum\limits_{k\in W_{ij}}  \pi^{s}_{ik} \quad (i,j)\in A,\{s,t\}\in E(D_0),\\
\label{con:pf2:sumFImpSumY2} \sum\limits_{k\in W_{ij}}F^{t}_{ik} & \leq \sum\limits_{k\in W_{ij}}  \pi^{0}_{ik} \quad (i,j)\in A, t\in D_0,\\
\label{con:pf2:dim} f&\in\{0,1\}^{A\times E(D_0)}.
\end{align}~
\end{subequations}

By (\ref{con:pf2:flowHook}), it is ensured that the joint flow is non-increasing by increasing distance from the root $s_0$.
While constraints (\ref{con:pf2:startInSource})--(\ref{con:pf2:stopInDest}) ensure $f_{ij}^{st}=0$ if $F_{ij}^s=0$ or $F_{ij}^t=0$,
inequalities (\ref{con:pf2:stronger}) impose $f_{ij}^{st}=1$ if $F_{ij}^s=F_{ij}^t=1$.
By virtue of (\ref{con:pf2:stopInDest}), (\ref{con:pf2:stronger}) replace the weaker capacity constraints (\ref{con:pf1:xfrel}).
Finally, (\ref{con:pf2:sumFImpSumY})--(\ref{con:pf2:sumFImpSumY2}) are valid inequalities justified by:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                     %
%         Proposition  		      %
%   (transformation x_{ij}^{st})      %
%                                     %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{prop}\label{prop:transX}
Assume $(F,f,g,\pi)$ satisfies (\ref{con:pf1:flow})--(\ref{con:vi:sumYImpSumXTrans}), (\ref{con:vi:Y1}), (\ref{con:pf2:flowHook})--(\ref{con:pf2:stronger}), and (\ref{con:pf2:dim}).
Let $T^0\subseteq G_g$ be the arborescence spanning $D$ (see Lemma \ref{lem:modelcorrect}),
and let $T^s$ be the arborescence obtained by redirecting the arcs in $T^0$ such that they point away from $s\in D$. Then, for all nodes $i$ and $j$ ($i\neq j$) spanned by $T^0$,
$$
F^t_{ij}+F^s_{ji}-f^{st}_{ij}-f^{st}_{ji} = 
	\begin{cases}
		1, ~\text{if $(i,j)$ is an arc on the path from~$s$ to $t$ in $T^s$,} \\
		0, ~\text{otherwise.}
	\end{cases}
$$
\end{prop}
%
\begin{proof}
Because $g_{ij}+g_{ji}\leq 1$, we have $F^t_{ij}+F^s_{ji}\leq 1$.
Constraints (\ref{con:pf2:startInSource})--(\ref{con:pf2:stopInDest}) thus yield $f^{st}_{ij}+f^{st}_{ji}\leq 1$, and
$F^t_{ij}+F^s_{ji}-f^{st}_{ij}-f^{st}_{ji}\in \{0,1\}$ follows.
Inequalities (\ref{con:pf2:stronger}) in combination with (\ref{con:pf2:startInSource})--(\ref{con:pf2:stopInDest}) also imply $f^{st}_{ij}=F^t_{ij}F^s_{ji}$.

The path from $s$ to $t$ in $T^s$ contains $(i,j)$ iff $(i,j)\in A_{F^t}\setminus A_{F^s}$ or $(j,i)\in A_{F^s}\setminus A_{F^t}$.
This condition is equivalent to $F_{ij}^t=1-F_{ij}^s=1$ or $F_{ji}^s=1-F_{ji}^t=1$,
which holds iff $F_{ij}^t(1-F_{ij}^s)+F_{ji}^s(1-F_{ji}^t)=F_{ij}^t-f_{ij}^{st}+F_{ji}^s-f_{ji}^{st}=1$.
\qed
\end{proof}

Constraints (\ref{con:pf2:sumFImpSumY}) reflect the observation that if $(i,j)$ is on the path from $s$ to $t$ in $T^s$,
then the cost of the most expensive arc leaving $i$ in $T^s$ is at least $p_{ij}$.
Validity of (\ref{con:pf2:sumFImpSumY2}) is explained analogously, as $F_{ij}^t=1$ iff $(i,j)$ is on the path from $s_0$ to $t$ in $T^0$.

\subsection{Constraint Generation Scheme}
\label{sec:cg}
Owing to the large number of constraints in the stronger models, $\text{LP}(\mathcal{X}_3)$ and $\text{LP}(\mathcal{F}_3)$ are impractical for computing lower bounds, even in fairly small instances.
To enable computation of strong lower bounds, we first solve a reduced version of the LP, where some of the constraints are omitted.
Some of the relaxed constraints that are violated in the solution are next added, and the process is repeated until no violations exist.
This approach is known as a \emph{constraint generation} (CG) scheme.

\subsubsection{Implementation}%

Preliminary computational experiments indicate that solving $\text{LP}(\mathcal{F}_k)$ ($k=1,2$)
in order to compute a stronger bound than the one output from $\text{LP}(\mathcal{X}_k)$ is also more time-consuming.
The disproportion in running time is even more considerable in a comparison between $\text{LP}(\mathcal{F}_3)$ and $\text{LP}(\mathcal{X}_3)$ (see Section \ref{sec:expsmall}).
For these practical reasons, we apply the CG idea to $\text{LP}(\mathcal{X}_3)$.

Consider an optimal solution $(X,y,\pi)$ to $\text{LP}(\mathcal{X}_2)$, and let $X$ and $\pi$ be fixed.
Whether there exists an $x\in\left[0,1\right]^{A\times A(D)}$ satisfying (\ref{con:mf:flowNormal})-(\ref{con:mf:fcap}) can be checked by solving a maximum flow problem for each $(s,t)\in A(D)$.
The constraints state that it should be feasible to send one unit of flow from source $s\in D$ to destination $t\in D_s$,
such that the flow on arc $(i,j)$ is no more than $X_{ij}^s$.
Further, the symmetry constraints (\ref{con:mf:fsym}) implicitly impose the upper bound $X_{ji}^t$ on the same flow.
Finally, the side constraints (\ref{con:vi:sumFImpSumY}) impose upper bounds on the total flow on subsets of arcs leaving a given node.
By applying the symmetry $x_{ij}^{st}=x_{ji}^{ts}$ also to (\ref{con:vi:sumFImpSumY}), the extended maximum flow model, denoted $\mathcal{MF}$, is completed as
(the $x$-variables do not have the superscripts $s$ and $t$ as $\{s,t\}$ is fixed within each model instance):

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                     %
%          Model 2MF                  %
%        (13a) - (13f)                %
%                                     %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{subequations}\label{eq:mf}
\begin{align}
\label{objective:maxf} &\max\sum\limits_{i \in V_s} x_{si} \\ 
\text{s.t.}  \notag   \\
\label{con:maxf:flowNormal}  \sum\limits_{\substack{ j\in V_i}}\left(x_{ij}-x_{ji}\right) & = 0 & i\in V\setminus\{s,t\},\\
\label{con:maxf:fcapst}   x_{ij} &\leq \min\{X_{ij}^s, X_{ji}^t\}     &  (i,j)\in A,  \\ 	 
\label{con:maxf:sumFImpSumYst} \sum_{k\in W_{ij}}x_{ik} & \leq \sum_{k\in W_{ij}} \pi^{s}_{ik} & (i,j)\in A,\\
\label{con:maxf:sumFImpSumYts} \sum_{k\in W_{ij}}x_{ki} & \leq \sum_{k\in W_{ij}} \pi^{t}_{ik} & (i,j)\in A,\\
\label{con:maxf:fdim}x&\in\left[0,1\right]^{A}. 
\end{align}~
\end{subequations}

For $Q\subseteq E(D)=\left\{\{s,t\}\in E: s,t\in D\right\}$, let $\mathcal{X}_2+Q$ denote the extension of $\mathcal{X}_2$ by constraints (\ref{con:mf:flowNormal})--(\ref{con:vi:sumFImpSumY}) for all $\{s,t\}\in Q$.
In the CG procedure described in Algorithm \ref{alg:cg},
we first solve $\text{LP}(\mathcal{X}_2)$ to obtain solution vectors $X$ and $\pi$.
For each $\{s,t\}\in E(D)$, we then check whether the relaxed constraints (\ref{con:mf:flowNormal})--(\ref{con:vi:sumFImpSumY}) can be satisfied,
given the values of $X$ and $\pi$.
This question is answered by solving $\mathcal{MF}$
since the objective function in $\mathcal{MF}$ takes optimal value 1 iff the constraints are satisfiable.
In each iteration,
relaxed constraints corresponding to a selection of pairs $\{s,t\}$ for which the maximum flow is below 1, are added to the original model.
How to make this selection is discussed next.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                     %
%         Algorithm 1                 %
%                                     %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{algorithm}
$Q\leftarrow\emptyset$\;
\Do {$\Delta Q\neq\emptyset$}{
	$(X,y,\pi)\leftarrow\text{optimal solution to $\text{LP}(\mathcal{X}_2+Q)$}$\;
	\For{$\{s,t\}\in E(D)\setminus Q$} {
		$v_{st}\leftarrow \text{optimal objective function value in $\mathcal{MF}$}$\;
	}
	$Q'\leftarrow\left\{\{s,t\}\in E(D)\setminus Q: v_{st}<1\right\}$\;
	Select some $\Delta Q\subseteq Q'$ such that $\Delta Q\neq\emptyset$ if $Q'\neq\emptyset$\;
	$Q\leftarrow Q\cup\Delta Q$\;
}
 \caption{Constraint generation}
\label{alg:cg}
\end{algorithm}

\subsubsection{Constraint selection}

Preliminary experiments have demonstrated that in many instances, the optimal values of $X$ and $\pi$ in $\text{LP}(\mathcal{X}_2)$ are too restrictive to allow for unit maximum flow.
For most $\{s,t\}\in E(D)$, the objective function of $\mathcal{MF}$ takes a value below 1 in optimum.
Adding the constraints corresponding to all $\{s,t\}$ for which this occurs thus involves computations almost as extensive as solving $\text{LP}(\mathcal{X}_3)$ directly.

Selection of pairs $\{s,t\}$ to be included in $\Delta Q$ in Algorithm \ref{alg:cg} is based on the following intuitive considerations:
\begin{description}
  \item[\textbf{Adjacency matters}:] Extending $Q$ by $\{s,t\}$ is more likely to imply $v_{uw}=1$ in the following iteration if $\{s,t\}\cap\{u,w\}\neq\emptyset$ ($\{u,w\}\in E(D)\setminus Q$).
  \item[\textbf{Extent of violation matters}:] The growth of $z\left(\text{LP}(\mathcal{X}_2+Q)\right)$ increases by decreasing value of $v_{st}$.
\end{description}

Balancing the needs for rapid growth in $z\left(\text{LP}(\mathcal{X}_2+Q)\right)$ and small value of $|Q|$,
we pursue the CG scheme as follows:
In the graph $(D,Q')$, where edges correspond to pairs $\{s,t\}$ at which $v_{st}<1$, let the edge weights be $1-v_{st}$.
Let $\Delta Q$ be a matching of maximum weight in $(D,Q')$.
Thus, for each $\{u,w\}\in E(D)\setminus Q$ where (\ref{con:mf:flowNormal})--(\ref{con:vi:sumFImpSumY}) are not satisfiable given the current values of $X$ and $\pi$,
$Q$ is extended by an adjacent edge.

### 2018-08-30 smt/sec6-exp.tex
MI: created folder with tex sources of revised version of SMT

\section{Experimental Evaluation}
\label{sec:exp}

The practical part of this work focuses on comparison of models presented in Sections \ref{sec:ILP} and \ref{sec:strong}.
Instances of given number of nodes and destinations are generated with random coordinates uniformly distributed on the square with corners at [0, 0] and [100, 100].
A time limit of 20 CPU-minutes is imposed on each run.
All experiments are run on an Intel Core 2 Quad CPU at 2.83 GHz and 8 GB RAM.
The models are implemented in Java with Concert technology and solved using CPLEX 12.5.1 optimizer.

\subsection{Comparing the computational burden and the lower bound of the LP relaxations}
\label{sec:expcomplp}

The first experiments give an overview of the LP relaxations of the models with respect to their strength and computational time. 
Tab.\ \ref{tab:small_inst_cost} summarizes optimal objective function values of all LP relaxations.
Each table entry is the value of $z(\text{LP}(\mathcal{M}))/z^*$ for the model $\mathcal{M}$ corresponding to the column,
averaged over 25 instances of the size corresponding to the row.
Here, $z^*$ denotes the minimum power.
Entries labeled by an asterisk correspond to runs that, in at least one of the 25 instances, are interrupted because the time bound was reached.
The dual simplex method is used to solve the LPs, such that a lower bound on $z(\text{LP}(\mathcal{M}))$ is available upon interruption.
The rows are clustered by $|V|/|D|$ ratio.

Confirming Props.\ \ref{prop:f1strx1} and \ref{prop:f2strx2}, model $\text{LP}(\mathcal{F}_k), (k=1,2)$ yields a consistently tighter lower bound than does $\text{LP}(\mathcal{X}_k)$. 
The difference is more prominent for $k=1$, where the bound obtained by $\text{LP}(\mathcal{F}_1)$ is 13\% tighter in average, while $\text{LP}(\mathcal{F}_2)$ gives a 5\% tighter bound in average.
A comparison between $\text{LP}(\mathcal{X}_3)$ and $\text{LP}(\mathcal{F}_3)$ shows that they give identical bounds in all instances that are solved to optimality within the time limit by use of
$\text{LP}(\mathcal{F}_3)$.
Using $\text{LP}(\mathcal{F}_3)$, this is possible in all instances.
In the instances in question, the bounds provided by the stronger models are significantly tighter than those of $\text{LP}(\mathcal{F}_2)$, and close or equal to the integer optimum.

\begin{table}[h!]
\centering
\setlength{\tabcolsep}{6pt} % Default value: 6pt
\renewcommand{\arraystretch}{1.4} % Default value: 1
\begin{tabular}{rrrrrrrr}
$|V|$ & $|D|$ & $\text{LP}(\mathcal{X}_1)$ & $\text{LP}(\mathcal{F}_1)$ & $\text{LP}(\mathcal{X}_2)$ & $\text{LP}(\mathcal{F}_2)$ & $\text{LP}(\mathcal{X}_3)$ &$\text{LP}(\mathcal{F}_3)$\\\hline
  12 & 8       & 78.37  & 82.69  & 85.40    & 86.95    & 99.92  &  99.92\textcolor{white}{$^*$} \\
  15 & 10      & 78.19  & 82.35  & 85.04    & 86.89    & 99.88  &  99.88\textcolor{white}{$^*$} \\
  18 & 12      & 74.05  & 79.78  & 80.65    & 83.47    & 94.39  &  55.27$^*$                    \\\hline
  14 & 7       & 74.83  & 80.65  & 83.13    & 84.86    & 99.92  &  99.92\textcolor{white}{$^*$} \\
  16 & 8       & 72.25  & 80.39  & 82.20    & 85.97    & 99.78  &  99.78\textcolor{white}{$^*$} \\
  18 & 9       & 65.68  & 77.17  & 74.48    & 79.84    & 99.70  &  74.17$^*$                    \\ \hline
  15 & 5       & 65.99  & 78.08  & 80.87    & 86.11    & 100.00 & 100.00\textcolor{white}{$^*$} \\ 
  18 & 6       & 66.08  & 77.13  & 79.71    & 84.23    & 99.94  &  99.94\textcolor{white}{$^*$} \\ 
  21 & 7       & 62.26  & 76.05  & 75.89    & 83.33    & 99.96  &  99.27$^*$
\end{tabular}
\caption{Average LP bounds as percentages of integer optimum}
\label{tab:small_inst_cost}
\end{table}

Average running times are reported in Tab. \ref{tab:small_inst_time}.
The 20 CPU-minutes time restriction manifests itself when applying the strongest LP relaxations to some instances.
Solving $\text{LP}(\mathcal{F}_k), (k=1,2,3)$ is consistently more time-consuming than solving $\text{LP}(\mathcal{X}_k)$, which demonstrates a trade-off between strength and running time.  
Moreover, the running times of $\text{LP}(\mathcal{X}_3)$ and $\text{LP}(\mathcal{F}_3)$ are, in instances of the actual size, considerably longer than the time it takes to solve $\mathcal{F}_1$ to optimality using branch-and-bound (B\&B).
This implies that both $\text{LP}(\mathcal{X}_3)$ and $\text{LP}(\mathcal{F}_3)$ are impractical in their basic form.
Bounds corresponding to the other models are computed in a few seconds.

\begin{table}[h!]
\centering
\setlength{\tabcolsep}{6pt} % Default value: 6pt
\renewcommand{\arraystretch}{1.4} % Default value: 1
\begin{tabular}{rrrrrrrrr}
 $|V|$ & $|D|$ & $\text{LP}(\mathcal{X}_1)$ & $\text{LP}(\mathcal{F}_1)$ & $\text{LP}(\mathcal{X}_2)$ & $\text{LP}(\mathcal{F}_2)$ & $\text{LP}(\mathcal{X}_3)$ & $\text{LP}(\mathcal{F}_3)$ & $\mathcal{F}_1$\\ \hline
  12 & 8       & 0   & 0   & 0     & 0     & 8    & 26\textcolor{white}{$^*$}   & 2   \\
  15 & 10      & 1   & 1   & 1     & 1     & 146  & 566\textcolor{white}{$^*$}  & 16  \\
  18 & 12      & 3   & 4   & 3     & 5     & 1161 & 1171$^*$                    & 115 \\\hline
  14 & 7       & 0   & 0   & 0     & 0     & 5    & 46\textcolor{white}{$^*$}   & 5   \\ 
  16 & 8       & 1   & 1   & 1     & 1     & 72   & 304\textcolor{white}{$^*$}  & 18  \\
  18 & 9       & 1   & 2   & 2     & 2     & 326  & 1140$^*$                    & 65  \\ \hline
  15 & 5       & 0   & 0   & 0     & 0     & 1    & 8\textcolor{white}{$^*$}    & 3   \\ 
  18 & 6       & 0   & 0   & 0     & 0     & 9    & 109\textcolor{white}{$^*$}  & 12  \\ 
  21 & 7       & 1   & 2   & 2     & 3     & 107  & 884$^*$                     & 67
\end{tabular}
\caption{Comparison of average running time in CPU-seconds}
\label{tab:small_inst_time}
\end{table}

\subsection{Computing optimal solutions}
\label{sec:expsmall}

Small instances are often solved to optimality relatively quickly with B\&B.
We investigate what is the maximum size of instances that can be solved to optimality within 20 minutes using models $\mathcal{X}_1$ and $\mathcal{F}_1$.
The experiments compare sets of instances with different $|V|/|D|$ ratios.
For instances of the size given in columns 1--2, Table \ref{tab:small_inst} shows the average running time in CPU-seconds (columns 3--4),
the number of instances solved to optimality (columns 5--6),
and the average ratio between the upper and lower bounds on the minimum cost (columns 7--8) upon interruption of the solver.
Each table value is obtained after solving 25 instances of given size.

Unlike the case of LP relaxation, the running time needed to solve $\mathcal{F}_1$ by B\&B is shorter than the time needed to solve $\mathcal{X}_1$.
This is observed for all instance sizes.
As a consequence, the number of instances solved to optimality within the given time limit is smaller for $\mathcal{X}_1$ than for $\mathcal{F}_1$.
B\&B applied to $\mathcal{F}_1$ solves to optimality all the instances with no more than 20 nodes, as well as those where $|V|=21$ and $|D|=7$.
When the weaker model $\mathcal{X}_1$ is used, more time is needed in some of these instances.
Furthermore, when instances are not solved to optimality, $\mathcal{F}_1$ leaves a tighter optimality gap than does $\mathcal{X}_1$.
Note that instances with lower $|V|/|D|$ ratio are more difficult to solve than those with higher $|V|/|D|$.
This applies to both $\mathcal{F}_1$ and $\mathcal{X}_1$.
\begin{table}[h!]
\centering
\setlength{\tabcolsep}{12pt} % Default value: 6pt
\renewcommand{\arraystretch}{1.4} % Default value: 1
\begin{tabular}{rrrrrrrr}
  ~ & ~ & \multicolumn{2}{c}{average time [s]} &\multicolumn{2}{c}{\# solved} &\multicolumn{2}{c}{\shortstack{average \\remaining gap[\%]}}\\ \hline
 $|V|$ & $|D|$ & $\mathcal{X}_1$   & $\mathcal{F}_1$   & $\mathcal{X}_1$ & $\mathcal{F}_1$ & $\mathcal{X}_1$ & $\mathcal{F}_1$\\ \hline
  18 & 12      & 139  & 84   & 24 & 25 & 0.52  & 0.00  \\
  21 & 14      & 1006 & 789  & 10 & 18 & 9.64  & 5.08  \\ 
  24 & 16      & 1161 & 1137 & 3  & 4  & 21.04 & 20.52 \\ \hline 
  20 & 10      & 387  & 190  & 21 & 25 & 3.00  & 0.00  \\
  22 & 11      & 880  & 637  & 14 & 22 & 7.16  & 0.64  \\
  24 & 12      & 1169 & 997  & 3  & 10 & 16.44 & 10.44 \\ \hline
  21 & 7       & 282  & 113  & 21 & 25 & 2.56  & 0.00  \\ 
  24 & 8       & 650  & 442  & 16 & 21 & 8.64  & 2.72  \\ 
  27 & 9       & 1146 & 1013 & 4  & 8  & 22.80 & 14.64
\end{tabular}
\caption{Results obtained from B\&B applied to $\mathcal{X}_1$ and $\mathcal{F}_1$}
\label{tab:small_inst}
\end{table}

\subsection{Experiments on strong lower bounds}

As the instance size increases, exact solution of the SMT problem becomes unrealistic.
To assess the quality of fast and possible inexact methods \citep{ivanova16isco}, access to tightest possible lower bounds is crucial.
In the next set of experiments, we compare two different methods for computing the bounds.
Section \ref{sec:expcomplp} indicates that $z\left(\text{LP}(\mathcal{X}_3)\right)$ is a strong bound,
but solving $\text{LP}(\mathcal{X}_3)$ requires excessive computations.
At this point, the CG scheme (Alg.\ \ref{alg:cg}) comes into play.
Because $\text{LP}(\mathcal{F}_1)$ is solved quickly, lower bounds can alternatively be computed 
by applying B\&B to $\mathcal{F}_1$ for a restricted amount of time.
While the initial bound $z\left(\text{LP}(\mathcal{F}_1)\right)$ might be weak, fast solution of $\text{LP}(\mathcal{F}_1)$ enables rapid growth in the B\&B-tree,
and, possibly, also in the corresponding lower bound.

We compare the progress of the lower bounds obtained by the CG method with the one obtained by B\&B applied to $\mathcal{F}_1$.
 as a function of running time.
Instances with $|V|=2|D|\in\{24, 26,\ldots, 34\}$ are investigated, with 5 instances for each value of $|V|$.
The results are illustrated in Fig. \ref{fig:cggr}.
Each curve represents the progress of the lower bound within each instance set.
The horizontal axes represent running times ranging from 0 to 20 CPU-minutes.
The vertical axes correspond to the lower bounds as a percentage of $z\left(\text{LP}(\mathcal{F}_1)\right)$, averaged over all instances in the set.

\begin{figure}[!htb]
    \centering
    \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{lower-bound-24-12}
        \caption{$|V|=24, |D|=12$}
        \label{fig:cggr24-12}
    \end{subfigure}
    \hfill %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{lower-bound-26-13}
        \caption{$|V|=26, |D|=13$}
        \label{fig:cggr26-13}
    \end{subfigure}

    \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{lower-bound-28-14}
        \caption{$|V|=28, |D|=14$}
        \label{fig:cggr28-14}
    \end{subfigure}
    \hfill %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{lower-bound-30-15}
        \caption{$|V|=30, |D|=15$}
        \label{fig:cggr30-15}
    \end{subfigure}

    \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{lower-bound-32-16}
        \caption{$|V|=32, |D|=16$}
        \label{fig:cggr32-16}
    \end{subfigure}
    \hfill %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{lower-bound-34-17}
        \caption{$|V|=34, |D|=17$}
        \label{fig:cggr34-17}
    \end{subfigure}
  \caption{Comparison of lower bounds obtained by CG and B\&B on model $\mathcal{F}_1$. }
  \label{fig:cggr}
\end{figure}

Reflecting the progress of the dual simplex method applied to $\text{LP}(\mathcal{X}_2)$ and $\text{LP}(\mathcal{F}_1)$, respectively,
both CG and B\&B increase the lower bound sharply at the beginning. 
While solution of $\text{LP}(\mathcal{F}_1)$ is still in progress, the growth of the lower bound is faster than what is observed for CG (solution of $\text{LP}(\mathcal{F}_1)$). 
This is particularly apparent for instances of size 28 and larger (Fig. \ref{fig:cggr28-14} - \ref{fig:cggr34-17}). %and can be explained as a consequence of Prop. \ref{prop:f1strx1}.
Once $\text{LP}(\mathcal{F}_1)$ is solved, the increase slows down substantially, and remains modest until the time limit is reached. 
An analogous, but more moderate, gradual slow-down is also observed for CG.
In each instance set, the two curves intersect at some point in time.
From this moment on, CG provides a tighter lower bound.
Experiments reported in Fig.\ \ref{fig:cggr} thus indicate that CG is the more suitable method for obtaining tight lower bounds within a given time.

### 2018-08-30 smt/sec7-conclusion.tex
MI: created folder with tex sources of revised version of SMT

\section{Conclusion and Future Work}
\label{sec:conclusion}
We have presented and analyzed two approaches of modelling the SMT problem as an integer linear program.
The first formulation, $\mathcal{X}_1$, is based on a formulation of the broadcast version of the problem, while the second one, $\mathcal{F}_1$, is an extension of network flow formulation of the minimum Steiner tree problem.
Both formulations are improved by valid inequalities resulting in models $\mathcal{X}_2$ and $\mathcal{F}_2$, respectively. 
Further strengthening is achieved by introducing variables of higher dimension and associated constraints, which leads to models $\mathcal{X}_3$ and $\mathcal{F}_3$.
A theoretical analysis of the formulations shows that  $\mathcal{F}_1$ and $\mathcal{F}_2$ are at least as strong as $\mathcal{X}_1$ and $\mathcal{X}_2$, respectively. 
It is conjectured that an analogous relation holds between $\mathcal{F}_3$ and $\mathcal{X}_3$.
Experimental evaluation reveals that instances with up to approximately 20 nodes are practically solvable to optimality.
Subsequent investigation suggests that the LP relaxations of $\mathcal{F}_3$ and $\mathcal{X}_3$ are fairly strong, but the running time prohibits direct solution.
Nevertheless, experiments also show that a constraint generation procedure applied to $\mathcal{X}_3$, developed in the current work, provides tight lower bounds.
The procedure also demonstrates that when the number of nodes is no more than 30, many instances of the LP relaxation of $\mathcal{X}_3$ have an integer optimum.

Follow-up research should uncover whether the conjecture on the relation between the LP relaxations of $\mathcal{F}_3$ and $\mathcal{X}_3$ holds.
A search for facet-defining valid inequalities, or even convex hull formulations, is an interesting and ambitious direction of future work.
Owing to the limited size of solvable instances, there is a potential in studying inexact methods, such as approximation algorithms and domain specific heuristics.
Instance preprocessing combined with variable fixing can possibly increase the size of instances solvable to optimality with the current models.

%We have presented two approaches of modelling SMT as an integer linear program.
%The first formulation is based on broadcast trees, while the second  formulation is an extension of network flow model of the minimum Steiner tree problem.
%Both formulations are improved by valid inequalities and further strengthened by variables of higher dimension and associated constraints.
%A~theoretical analysis reveals that the models based on network flows are at least as strong as the broadcast tree models, with one exception of the strongest extensions of the models, where this claim remains a conjecture.
%Experimental study discovers that practically solvable to optimality are instances with up to approximately 20 nodes.
%Subsequent investigation of LP relaxations suggests that constraint generation applied to the strongest broadcast tree based models is a promising technique for obtaining lower bounds.
%
%%The follow-up research should prove the conjecture that also the strongest network flow based models are at least as strong as the corresponding broadcast tree based models.
%The follow-up research should prove the remaining conjecture regarding relations between the models.
%Additional investigation of facet defining valid inequalities, or even convex hull formulations is a possible direction of future work.
%Owing to the limited size of solvable instances, there is a potential in studying inexact methods such as approximation algorithms and domain specific heuristics.
%Instance preprocessing together with variable fixing could possibly increase the size of instances solvable to optimality with existing models.

### 2018-09-05 smt/sec2-sbtsmt.tex
MI: Added a reference Halgamuge et. al. (2009), and a paragraph (2nd last paragraph in Sect. 2) that attempts to explain the difference between our energy moddel and the one in the cited paper, and why we do what we do. Label sizes of graphs increased even further

@@
 
+Different energy models can be found in the literature. 
+\cite{halgamuge} present a comprehensive energy model for wireless sensor networks, 
+where the sources of energy consumption consist of sensing, logging, processing, transimission, actuation, transient and cluster formation.
+Our energy model focuses on the energy consumed by transmission, which is also tho most energy consuming aspect of wireless networks (51\% according to \cite{halgamuge}).
+Some of the measures such as sensing and cluster formation are not relevant in our model.
+Another distinction is that nodes in our model are all equivalent and the data is not aggregated in the nodes.
+For that reason we disregard the potential energy consumption sources occuring in sensor networks.
+However, it is easy to incorporate some of the disregarded measures by adding a constant for every used Steiner node.
+
 In the forthcoming sections, we develop and analyze IP models for SMT.

### 2018-09-05 smt/response.tex
MI: (SMT) Suggestions from Reviewer 2 - Abstract shortened from 22 to 15 lines. Introduced Section 2 Literature overview, although nothing has been changed yet in this section. Section 3 divided into 3 subsections - Assumptions and notation, Network model, Problem definition. The last paragraph in Section 1 (Introduction) is updated according to the new structure of Section 2 and 3.

\documentclass[12pt]{article}
\usepackage{a4wide,url}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{color}

\begin{document}
\section*{Response Letter}
We are grateful to the reviewers for their time, effort, and constructive remarks.
All remarks have been accounted for in the revised version, as detailed in our response below.

\section*{Reviewer 2}
Our response in \textcolor{blue}{blue} follows the corresponding remark from the reviewer in \textcolor{red}{red}.

\begin{itemize}
\item \textcolor{red}{The abstract should be shorter.} \textcolor{blue}{
We have revised the abstract, and reduced its length from 22 to 15 lines.
}
\item \textcolor{red}{The literature review has not been organized in a proper format. The references are so old. New related works need to be explained. The authors should use a separate Section for this purpose and compare the novelty of the present approach with them.} \textcolor{blue}{
Following the recommendation of the reviewer, we have reorganized the literature review, and devoted a new section (Section 2 in the revised manuscript)
to this topic. We have also included a review of recent and relevant literature (references ??? in the revision).
In paragraph ??? of Section 2, we now state explicitly the contributions from the present work, and explain how they relate to the current literature.
}
\item \textcolor{red}{They also should consider a separate section on the network model and assumptions.} \textcolor{blue}{
We have restructured the former Section 2 (now Section 3) into three subsections.
The first is devoted to mathematical notation and underlying assumptions, the second addresses the network model,
while the problem definition is given in the third subsection.
}
\item \textcolor{red}{The diagrams of figures are not distinguishable in black and white print. The legends should be corrected.} \textcolor{blue}{
While paying particular attention to their appearance in black and white print, we have given all figures,
including their legends, a complete overhaul.
}
\item \textcolor{red}{All equations should be terminated with "." or "," as well as sentences.} \textcolor{blue}{
Done.
}
\item \textcolor{red}{The legends, markers and other details of the figures are so small.} \textcolor{blue}{
Legends, markers, and other details are now given in larger size.
}
\item \textcolor{red}{A fundamental weak point is that there are no numerical results comparing the proposed scheme with the other new works;} \textcolor{blue}{
We agree that a comparison with alternative contemporary methods largely benefits the paper.
To that end, we have included a comparison with ??? \cite{???}.
Results from the experiments are documented in the new Section 7.?
}
\item \textcolor{red}{The simulation graphs need more expiation and justification.} \textcolor{blue}{
For better justification of the numerical results illustrated graphically in the former Section 6 (now Section 7),
we have added more explanations.
}
\item \textcolor{red}{ The main aim of this paper is the energy, How is the energy issue modeled? Why is not the real models used in the paper?} \textcolor{blue}{
In the new Section 3.2 on the network model, we now explain more detailed how the energy is modeled.
Justification is added in terms of references to relevant and frequently cited works (see [?], ....).
}
\item \textcolor{red}{Overall the novelty of the proposed approach in unclear and the paper is hard to follow
The authors have not considers the wireless problems in the paper.} \textcolor{blue}{
The novelty of and the contributions from the paper are now explicitly stated in Section 2.
We also explain how the paper relates to other works on energy minimization in wireless networks.
}
\end{itemize}

\section*{Reviewer 3}
Our response in \textcolor{blue}{blue} follows the corresponding remark from the reviewer in \textcolor{red}{red}.

\begin{itemize}
\item \textcolor{red}{The paper provides two new ILP formulations for the Shared Multicast Tree (SMT) problem in wireless ad hoc networks.
To address the problem of ILP scalability, the paper also provides a study and proofs of related theorems that shed light on the quality of
solutions achievable by  LP  bounds.
As the number and also provides a new constraint generation scheme.
Numerical results demonstrate the usefulness of the approach. The paper is well written.} \textcolor{blue}{
We are grateful to the reviewer for her/his encouraging remarks, and we appreciate the (s)he finds the paper well written.
}
\item \textcolor{red}{Overall, the methodology makes contribution in the field of optimization as the approach can be used to address other similar problems
that can be modeled as ILP which is not scalable.
However, this paper aims to address the problem of SMT in wireless ad hoc networks and it is
really motivated by the important problem of saving energy in sensor networks.
There I see a certain minor weakness.
The model considered ignores details of energy consumption is sensor network
(see for example Halgamuge et al., Progress in Electromagnetics Research, 12, 2009, or other sources).
It will be good to my opinion to point out this weakness and provide certain qualitative comments to benefit the practitioners,
on how the work can be used in practice, or what is still needed to be done so the work can be useful.} \textcolor{blue}{
We thank the reviewer for suggesting this very useful reference, a citation of which has been included in the revised version of the paper.
We fully agree that there are gaps between the coarse energy model applied in our work and more detailed energy models for energy consumption is sensor networks,
e.g.\ those analyzed by Halgamuge et al.\ (2009).
These gaps, and their possible implications on practical usability of our approaches, are now discussed in the new Section 3.2.
}
\end{itemize}

\noindent
Final comment: We thank the reviewers again for their constructive remarks, which largely helped us to improve the presentation of our findings.

\end{document}

### 2018-09-05 smt/ivanova.tex
MI: Added a reference Halgamuge et. al. (2009), and a paragraph (2nd last paragraph in Sect. 2) that attempts to explain the difference between our energy moddel and the one in the cited paper, and why we do what we do. Label sizes of graphs increased even further
MI: (SMT) Suggestions from Reviewer 2 - Abstract shortened from 22 to 15 lines. Introduced Section 2 Literature overview, although nothing has been changed yet in this section. Section 3 divided into 3 subsections - Assumptions and notation, Network model, Problem definition. The last paragraph in Section 1 (Introduction) is updated

@@
 All destinations of the multicast message must be connected, while non-destinations are optional nodes that can be used to relay messages. 
-In the well-studied minimum energy multicast (MEM)
-problem, the optimal tree depends on the source of the message.
-In contrast, SMT asks for a source-independent tree, which substantially reduces the processing overhead at each node.
-This means that SMT can be regarded as a Steiner tree problem.
-The objective function reflecting power consumption, however, distinguishes SMT clearly from the traditional minimum Steiner tree problem.
+% MI: I believe these sentences can be disregarded, because the related problems are discussed in detail in the Introduction.
+%In the well-studied minimum energy multicast (MEM)
+%problem, the optimal tree depends on the source of the message.
+%In contrast, SMT asks for a source-independent tree, which substantially reduces the processing overhead at each node.
+%This means that SMT can be regarded as a Steiner tree problem.
+%The objective function reflecting power consumption, however, distinguishes SMT clearly from the traditional minimum Steiner tree problem.
+The objective function reflecting power consumption distinguishes SMT clearly from the traditional minimum Steiner tree problem.
 
-In the present work, we develop two 
-integer programming formulations for SMT.
-Both models are subsequently extended and strengthened by redundant valid inequalities.
-Further strengthening is achieved by extending the set of variables and corresponding constraints.
+%In the present work,
+We develop two integer programming formulations for SMT.
+Both models are subsequently extended and strengthened. % by redundant valid inequalities.
+%Further strengthening is achieved by extending the set of variables and corresponding constraints.
 Theorems on relations between the LP bounds corresponding to the models are stated and proved.
 As the number of variables in the strongest formulations is a polynomial of degree four in the number of network nodes,
-the models are impractical for computing lower bounds in instances beyond a fairly small size.
-To make them applicable to larger instances, a constraint generation scheme is developed.
+the models are impractical for computing lower bounds in instances beyond a fairly small size, 
+%To make them applicable to larger instances, a constraint generation scheme is developed.
+and therefore a constraint generation scheme is developed.
 Results from computational experiments with the models demonstrate good promise of the approaches taken.
@@
 \input{sec1-intro.tex}
+\input{sec2-literature.tex}
 \input{sec2-sbtsmt.tex}
@@
 
+\bibitem[Halgamuge et al.(2009)]{halgamuge}
+Halgamuge MN, Zukerman M, Ramamohanarao K (2009)
+An estimation of sensor energy consumption.
+Progress In Electromagnetics Research B 12:259--295
+
 \bibitem[Haugland and Yuan(2011)]{Haugland11Compact}

### 2018-09-05 smt/sec1-intro.tex
MI: (SMT) Suggestions from Reviewer 2 - Abstract shortened from 22 to 15 lines. Introduced Section 2 Literature overview, although nothing has been changed yet in this section. Section 3 divided into 3 subsections - Assumptions and notation, Network model, Problem definition. The last paragraph in Section 1 (Introduction) is updated according to the new structure of Section 2 and 3.

@@
 
-The combinatorial optimization literature is already rich on Integer Programming (IP) approaches to power minimization in wireless networks.
-Since the work of \citet{Wieseltier00onthe}, in particular the MEB and MEM problems have been studied extensively.
-Noteworthy contributions to this research include the IP models studied by \citet{das03},
-\citet{altinkemer05}, \citet{leggieri08}, \citet{yuan05}, \citet{yuan08}, \citet{bauer08}, and \citet{montemanni11}.
-\citet{cagalj02} proved that MEB, and thereby MEM, is NP-hard.
-Numerous contributions to fast, but possible inexact methods for MEM have thus appeared, and the interested reader is referred to \citep{hsiao13} for an overview.
-
-The SBT problem has also received some attention in the scientific literature.
-It was introduced by \citet{Papadimitriou06SBT}, who also proved that the problem is NP-hard.
-In the same article, an approximation algorithm for SBT was developed and analyzed experimentally.
-Building on this work, \citet{Haugland12Dual} presented an integer programming model for SBT, along with an associated decomposition approach.
-
-The SBT and SMT problems resemble the more frequently studied \textsc{Range Assignment Problem} (RAP) in the sense that they all ask for an undirected source-independent tree of minimum power.
-However, RAP concerns cases where a connected topology must be established by means of \emph{bidirectional} links,
-with the feature that power must be assigned to both the sending and the receiving device in each link.
-This scenario induces a spanning tree problem, where costs are incurred only for edges that are the most expensive ones incident to any of their end nodes.
-Simple instances \citep{Haugland12Dual} show that because of the difference in objective functions, the optimal solutions to RAP and SBT can differ substantially.
-\citet{kirousis97} and \citet{clementi99} conduct
-hardness results for instances of RAP embedded in three- and two-dimensional Euclidean space, respectively.
-Later, IP models have been developed by e.g.\ \citet{althaus03}, \citet{montemanni04}, \citet{das05}, and \citet{Haugland11Compact}.
-
-For the same reasons that MEB has been generalized to MEM, it is natural and desirable to extend SBT to its multicast equivalent.
-The \textsc{Minimum Shared Multicast Tree} (SMT) problem is thus at the forefront of this paper,
-and neither MEB, MEM, nor RAP will be considered further.
-Focus is henceforth on instances in which some of the nodes neither initiate new messages nor need messages from others.
-Utilizing such \emph{non-destination} nodes to forward messages can in many cases contribute to substantial reductions in the total power consumption.
-Non-destination nodes thus have the potential to play the role of Steiner nodes in the multicast tree.
-All devices that can initiate a transmission, referred to as \emph{destination} nodes, also have to receive every message.
-
-Some preliminary work on SMT has recently been published.
-\citet{ivanova16isco} introduces the problem, and develops an IP model.
-Moreover, the author studies inexact construction and improvement methods, for which computational results are reported.
 Contributions from the present work include two sets of IP models for SMT, presented in order of increasing lower bounding capabilities.
@@
 
-The remainder of this paper is organized as follows: Section \ref{sec:SBT} describes the SMT problem and gives detailed explanation of its objective function.
+The remainder of this paper is organized as follows: 
+Section \ref{sec:literature} summarizes relevant literature and previous works.
+%Section \ref{sec:SBT} describes the SMT problem and gives detailed explanation of its objective function.
+Section \ref{sec:SBT} divided into three subsections introduces the notation, explains the network model, and gives formal definition of the problem.
 Integer linear programming formulations, valid inequalities and their analysis are presented in Section \ref{sec:ILP},

### 2018-09-05 smt/sec2-literature.tex
MI: (SMT) Suggestions from Reviewer 2 - Abstract shortened from 22 to 15 lines. Introduced Section 2 Literature overview, although nothing has been changed yet in this section. Section 3 divided into 3 subsections - Assumptions and notation, Network model, Problem definition. The last paragraph in Section 1 (Introduction) is updated according to the new structure of Section 2 and 3.

\section{Literature Overview}
\label{sec:literature}
The combinatorial optimization literature is already rich on Integer Programming (IP) approaches to power minimization in wireless networks.
Since the work of \citet{Wieseltier00onthe}, in particular the MEB and MEM problems have been studied extensively.
Noteworthy contributions to this research include the IP models studied by \citet{das03},
\citet{altinkemer05}, \citet{leggieri08}, \citet{yuan05}, \citet{yuan08}, \citet{bauer08}, and \citet{montemanni11}.
\citet{cagalj02} proved that MEB, and thereby MEM, is NP-hard.
Numerous contributions to fast, but possible inexact methods for MEM have thus appeared, and the interested reader is referred to \citep{hsiao13} for an overview.

The SBT problem has also received some attention in the scientific literature.
It was introduced by \citet{Papadimitriou06SBT}, who also proved that the problem is NP-hard.
In the same article, an approximation algorithm for SBT was developed and analyzed experimentally.
Building on this work, \citet{Haugland12Dual} presented an integer programming model for SBT, along with an associated decomposition approach.

The SBT and SMT problems resemble the more frequently studied \textsc{Range Assignment Problem} (RAP) in the sense that they all ask for an undirected source-independent tree of minimum power.
However, RAP concerns cases where a connected topology must be established by means of \emph{bidirectional} links,
with the feature that power must be assigned to both the sending and the receiving device in each link.
This scenario induces a spanning tree problem, where costs are incurred only for edges that are the most expensive ones incident to any of their end nodes.
Simple instances \citep{Haugland12Dual} show that because of the difference in objective functions, the optimal solutions to RAP and SBT can differ substantially.
\citet{kirousis97} and \citet{clementi99} conduct
hardness results for instances of RAP embedded in three- and two-dimensional Euclidean space, respectively.
Later, IP models have been developed by e.g.\ \citet{althaus03}, \citet{montemanni04}, \citet{das05}, and \citet{Haugland11Compact}.

For the same reasons that MEB has been generalized to MEM, it is natural and desirable to extend SBT to its multicast equivalent.
The \textsc{Minimum Shared Multicast Tree} (SMT) problem is thus at the forefront of this paper,
and neither MEB, MEM, nor RAP will be considered further.
Focus is henceforth on instances in which some of the nodes neither initiate new messages nor need messages from others.
Utilizing such \emph{non-destination} nodes to forward messages can in many cases contribute to substantial reductions in the total power consumption.
Non-destination nodes thus have the potential to play the role of Steiner nodes in the multicast tree.
All devices that can initiate a transmission, referred to as \emph{destination} nodes, also have to receive every message.

Some preliminary work on SMT has recently been published.
\citet{ivanova16isco} introduces the problem, and develops an IP model.
Moreover, the author studies inexact construction and improvement methods, for which computational results are reported.

### 2018-09-06 smt/sec1-intro.tex
MI: (SMT) Added a reference (Guo and Yang), newly written text is now in a blue font. Old text that is moved to different sections/paragraphs remains black

@@
 This feature is often referred to as the \emph{wireless advantage} \citep{Wieseltier00onthe}.
+
+\textcolor{blue}{\subsection{Related problems}}
 
@@
 
+The SBT and SMT problems resemble the more frequently studied \textsc{Range Assignment Problem} (RAP) in the sense that they all ask for an undirected source-independent tree of minimum power.
+However, RAP concerns cases where a connected topology must be established by means of \emph{bidirectional} links,
+with the feature that power must be assigned to both the sending and the receiving device in each link.
+This scenario induces a spanning tree problem, where costs are incurred only for edges that are the most expensive ones incident to any of their end nodes.
+
+\textcolor{blue}{\subsection{Contributions}}
+
 Contributions from the present work include two sets of IP models for SMT, presented in order of increasing lower bounding capabilities.
@@
 
+For the same reasons that MEB has been generalized to MEM, it is natural and desirable to extend SBT to its multicast equivalent.
+The \textsc{Minimum Shared Multicast Tree} (SMT) problem is thus at the forefront of this paper,
+and neither MEB, MEM, nor RAP will be considered further.
+Focus is henceforth on instances in which some of the nodes neither initiate new messages nor need messages from others.
+Utilizing such \emph{non-destination} nodes to forward messages can in many cases contribute to substantial reductions in the total power consumption.
+Non-destination nodes thus have the potential to play the role of Steiner nodes in the multicast tree.
+All devices that can initiate a transmission, referred to as \emph{destination} nodes, also have to receive every message.
+
+\textcolor{blue}{\subsection{Structure of the paper}}
+
 The remainder of this paper is organized as follows: 
-Section \ref{sec:literature} summarizes relevant literature and previous works.
+\textcolor{blue}{Section \ref{sec:literature} summarizes relevant literature and previous works.}
 %Section \ref{sec:SBT} describes the SMT problem and gives detailed explanation of its objective function.
-Section \ref{sec:SBT} divided into three subsections introduces the notation, explains the network model, and gives formal definition of the problem.
+\textcolor{blue}{Section \ref{sec:SBT} divided into three subsections introduces the notation, explains the network model, and gives formal definition of the problem.}
 Integer linear programming formulations, valid inequalities and their analysis are presented in Section \ref{sec:ILP},

### 2018-09-06 smt/sec2-literature.tex
MI: (SMT) Added a reference (Guo and Yang), newly written text is now in a blue font. Old text that is moved to different sections/paragraphs remains black

@@
 Noteworthy contributions to this research include the IP models studied by \citet{das03},
-\citet{altinkemer05}, \citet{leggieri08}, \citet{yuan05}, \citet{yuan08}, \citet{bauer08}, and \citet{montemanni11}.
+\citet{altinkemer05}, \citet{yuan05}, \citet{yuan08}, \citet{bauer08}, \citet{montemanni11} and \citet{leggieri08}. 
+\textcolor{blue}{
+The latter paper proposes an IP formulation of MEM based on a Set Covering model together with a preprocessing procedure reducing the number of constraints.}
 \citet{cagalj02} proved that MEB, and thereby MEM, is NP-hard.
 Numerous contributions to fast, but possible inexact methods for MEM have thus appeared, and the interested reader is referred to \citep{hsiao13} for an overview. 
+\textcolor{blue}{Another study concerning MEM is \cite{guo}, where the authors investigate energy conservation while using adaptive antennas. 
+Unlike omnidirectional antennas, adaptive antennas can concentrate their transmission power in a smaller angle.}
 
@@
 
-The SBT and SMT problems resemble the more frequently studied \textsc{Range Assignment Problem} (RAP) in the sense that they all ask for an undirected source-independent tree of minimum power.
-However, RAP concerns cases where a connected topology must be established by means of \emph{bidirectional} links,
-with the feature that power must be assigned to both the sending and the receiving device in each link.
-This scenario induces a spanning tree problem, where costs are incurred only for edges that are the most expensive ones incident to any of their end nodes.
 Simple instances \citep{Haugland12Dual} show that because of the difference in objective functions, the optimal solutions to RAP and SBT can differ substantially.
@@
 
-For the same reasons that MEB has been generalized to MEM, it is natural and desirable to extend SBT to its multicast equivalent.
-The \textsc{Minimum Shared Multicast Tree} (SMT) problem is thus at the forefront of this paper,
-and neither MEB, MEM, nor RAP will be considered further.
-Focus is henceforth on instances in which some of the nodes neither initiate new messages nor need messages from others.
-Utilizing such \emph{non-destination} nodes to forward messages can in many cases contribute to substantial reductions in the total power consumption.
-Non-destination nodes thus have the potential to play the role of Steiner nodes in the multicast tree.
-All devices that can initiate a transmission, referred to as \emph{destination} nodes, also have to receive every message.
-
 Some preliminary work on SMT has recently been published.

### 2018-09-06 smt/sec2-sbtsmt.tex
MI: (SMT) Suggestions from Reviewer 2 - Abstract shortened from 22 to 15 lines. Introduced Section 2 Literature overview, although nothing has been changed yet in this section. Section 3 divided into 3 subsections - Assumptions and notation, Network model, Problem definition. The last paragraph in Section 1 (Introduction) is updated according to the new structure of Section 2 and 3.
MI: (SMT) Added a reference (Guo and Yang), newly written text is now in a blue font. Old text that is moved to different sections/paragraphs remains black

@@
 
-\subsection{Problem definition} \label{sec:probdef}
+\textcolor{blue}{\subsection{Network model} \label{sec:netmod}}
 
@@
 
+\textcolor{blue}{
+Different energy models can be found in the literature. 
+\cite{halgamuge} present a comprehensive energy model for wireless sensor networks, 
+where the sources of energy consumption consist of sensing, logging, processing, transimission, actuation, transient and cluster formation.
+Our energy model focuses on the energy consumed by transmission, which is also tho most energy consuming aspect of wireless networks (51\% according to \cite{halgamuge}).
+Some of the measures such as sensing and cluster formation are not relevant in our model.
+Another distinction is that nodes in our model are all equivalent and the data is not aggregated in the nodes.
+On top of that, all the related works menioned in Section \ref{sec:literature} assume transmission as the sole energy consumption source. 
+For these reasons we disregard the potential energy consumption sources occuring in sensor networks.}
+
+\textcolor{blue}{
+However, it is easy to incorporate some of the disregarded measures by adding a constant for every used Steiner node.}
+
+\subsection{Problem definition} \label{sec:probdef}
+
 To express the cost of $T$ more rigorously, let $T^s=(V_T,A^s)$ denote the directed tree (arborescence) obtained by directing all edges in $E_T$ such that they point away from the \emph{root} at $s$.
@@
 
-Different energy models can be found in the literature. 
-\cite{halgamuge} present a comprehensive energy model for wireless sensor networks, 
-where the sources of energy consumption consist of sensing, logging, processing, transimission, actuation, transient and cluster formation.
-Our energy model focuses on the energy consumed by transmission, which is also tho most energy consuming aspect of wireless networks (51\% according to \cite{halgamuge}).
-Some of the measures such as sensing and cluster formation are not relevant in our model.
-Another distinction is that nodes in our model are all equivalent and the data is not aggregated in the nodes.
-For that reason we disregard the potential energy consumption sources occuring in sensor networks.
-However, it is easy to incorporate some of the disregarded measures by adding a constant for every used Steiner node.
-
 In the forthcoming sections, we develop and analyze IP models for SMT.

### 2018-09-11 smt/response.tex
MI (SMT): Addressed the criticism of Reviewer 2 and added more explanation to the graphs in Fig. 3. Also updated the response letter accordingly.

@@
 For better justification of the numerical results illustrated graphically in the former Section 6 (now Section 7),
-we have added more explanations.
+we have added more explanations in the caption of Fig. 3.
 }
@@
 In the new Section 3.2 on the network model, we now explain more detailed how the energy is modeled.
-Justification is added in terms of references to relevant and frequently cited works (see [?], ....).
+At the end of the section, we also describe how our model differs from another models in the literature.
+Justification is added in terms of references to relevant and frequently cited works mentioned in Section 2 (see Wieselthier et al. (2000), Das et al. (2003), Altinkemer et al. (2005),
+Yuan (2005), Yuan et al. (2008), Bauer et al. (2008), Montemanni and Leggieri (2011) and Leggieri et al. (2008), etc.).
 }

### 2018-09-11 smt/sec6-exp.tex
MI (SMT): Addressed the criticism of Reviewer 2 and added more explanation to the graphs in Fig. 3. Also updated the response letter accordingly.

@@
 
-We compare the progress of the lower bounds obtained by the CG method with the one obtained by B\&B applied to $\mathcal{F}_1$.
- as a function of running time.
+We compare the progress of the lower bounds obtained by the CG method with the one obtained by B\&B applied to $\mathcal{F}_1$ as a function of running time.
 Instances with $|V|=2|D|\in\{24, 26,\ldots, 34\}$ are investigated, with 5 instances for each value of $|V|$.
@@
     \end{subfigure}
-  \caption{Comparison of lower bounds obtained by CG and B\&B on model $\mathcal{F}_1$. }
+  \caption{Comparison of lower bounds obtained by CG and B\&B on model $\mathcal{F}_1$.
+	\textcolor{blue}{Each figure depicts a progress of lower bounds for instances with varying number of destination and non-destination nodes.
+	The lower bounds are obtained by solving CG scheme applied on model $\mathcal{F}_1$ (black curves), and produced during the course of classic B\&B algorithm (grey curves)
+	The tightness of the bounds increases with time, and it can be observed from the graphs that the CG bound becomes tighter at some point.
+	We express the quality of lower bounds with respect to the value of simple continuous relaxatin, as indicated by the vertical axes.
+	It can also be noted that the lower bound obtained from B\&B grows sharply during the computation of the simplex method in the root of the B\&B tree, 
+	which is happening until the curve reaches 100\%.
+	After that point the bound grows rather slowly.
+	CG on the other hand produces bound that grows faster even after the value of first relaxation in the root is reached.
+	}}
   \label{fig:cggr}

### 2018-09-12 smt/response.tex
MI (SMT): Added two references that might address the complain 'too old references'. There is a new Section 7.3 devoted to the description of the meta heuristic algorithm from [Pajor et Al. (2018)] and prepared writing about the experiments planned with this method. The response letter is updated accordingly, and now it seems almost complete, the only thing that I really cannot figure out is how to respond to the complaint about the absent contribution. So far I only created a separate section with a clear title, but I don't know how to explain the contribution more

--- 2018-09-05 smt/response.tex
@@
 Following the recommendation of the reviewer, we have reorganized the literature review, and devoted a new section (Section 2 in the revised manuscript)
-to this topic. We have also included a review of recent and relevant literature (references ??? in the revision).
-In paragraph ??? of Section 2, we now state explicitly the contributions from the present work, and explain how they relate to the current literature.
+to this topic. We have also included a review of more recent and relevant literature (references \cite{bein10, bhukya14}).
+In Section 1.2, we now state explicitly the contributions from the present work, and explain how they relate to the current literature.
 }
@@
 We agree that a comparison with alternative contemporary methods largely benefits the paper.
-To that end, we have included a comparison with ??? \cite{???}.
-Results from the experiments are documented in the new Section 7.?
+To that end, we have included a comparison with an adapted metaheuristic algorithm based on iterative combination of elite solutions \cite{pajor18}.
+The method is described along with presentatin of results from the experiments in the new Section 7.3.
 }
@@
 We fully agree that there are gaps between the coarse energy model applied in our work and more detailed energy models for energy consumption is sensor networks,
-e.g.\ those analyzed by Halgamuge et al.\ (2009).
+e.g.\ those analyzed in \cite{halgamuge}.
 These gaps, and their possible implications on practical usability of our approaches, are now discussed in the new Section 3.2.
@@
 
+\begin{thebibliography}{}
+
+\bibitem{bein10}
+Bein D, Zheng SQ (2010)
+Energy efficient all-to-all broadcast in all-wireless networks.
+Information Sciences 180(10):1781--1792
+
+\bibitem{bhukya14}
+Bhukya WN, Singh A (2014)
+\emph{p}-shrink: A Heuristic for Improving Minimum All-to-All Power Broadcast Trees in Wireless Networks.
+In: Proceedings of Ninth International Conference on Wireless Communication and Sensor Networks 2014 
+Vol.\ 299, pp 61--69
+
+\bibitem{halgamuge}
+Halgamuge MN, Zukerman M, Ramamohanarao K (2009)
+An estimation of sensor energy consumption.
+Progress In Electromagnetics Research B 12:259--295
+
+\bibitem{pajor18}
+Pajor T, Uchoa E, Werneck RF (2018)
+A robust and scalable algorithm for the Steiner problem in graphs.
+Math Prog Comp 10:69--118
+
+\end{thebibliography}
 \end{document}

### 2018-09-12 smt/ivanova.tex
MI: (SMT) Added a reference (Guo and Yang), newly written text is now in a blue font. Old text that is moved to different sections/paragraphs remains black
MI (SMT): Addressed the criticism of Reviewer 2 and added more explanation to the graphs in Fig. 3. Also updated the response letter accordingly.
MI (SMT): Added two references that might address the complain 'too old references'. There is a new Section 7.3 devoted to the description of the meta heuristic algorithm from [Pajor et Al. (2018)] and prepared writing about the experiments planned with this method. The response letter is updated

@@
 Networks 52(2):57–68
+\textcolor{blue}{\bibitem[Bein and Zheng(2010)]{bein10}
+Bein D, Zheng SQ (2010)
+Energy efficient all-to-all broadcast in all-wireless networks.
+Information Sciences 180(10):1781--1792}
+\textcolor{blue}{\bibitem[Bhukya and Singh(2014)]{bhukya14}
+Bhukya WN, Singh A (2014)
+\emph{p}-shrink: A Heuristic for Improving Minimum All-to-All Power Broadcast Trees in Wireless Networks.
+In: Proceedings of Ninth International Conference on Wireless Communication and Sensor Networks 2014 
+Vol.\ 299, pp 61--69}
 
@@
 23(1):19--28
+
+\bibitem[Guo and Yang(2004)]{guo}
+Guo S, Yang O (2004)
+Minimum Energy Multicast Routing for Wireless Ad-hoc Networks with Adaptive Antennas 
+In: Proceedings of the 12th IEEE International Conference on Network Protocols ICNP 2004,
+pp 151--160
 
@@
 74:327--342
+\textcolor{blue}{\bibitem[Pajor et al.(2018)]{pajor18}
+Pajor T, Uchoa E, Werneck RF (2018)
+A robust and scalable algorithm for the Steiner problem in graphs.
+Math Prog Comp 10:69--118
+}
 

### 2018-09-12 smt/sec2-literature.tex
MI (SMT): Added two references that might address the complain 'too old references'. There is a new Section 7.3 devoted to the description of the meta heuristic algorithm from [Pajor et Al. (2018)] and prepared writing about the experiments planned with this method. The response letter is updated accordingly, and now it seems almost complete, the only thing that I really cannot figure out is how to respond to the complaint about the absent contribution. So far I only created a separate section with a clear title, but I don't know how to explain the contribution more

@@
 Unlike omnidirectional antennas, adaptive antennas can concentrate their transmission power in a smaller angle.}
+
+\textcolor{blue}{\citet{bein10} study energy efficiency in wireless networks where all nodes can initiate a message, and the communicatin takes place in a single all-to-all broadcast tree.
+A heuristic algorithm generating all-to-all broadcast trees is proposed in \citet{bhukya14}, along with an experimental evaluation confirming its merit.}
 

### 2018-09-12 smt/sec6-exp.tex
MI (SMT): Added two references that might address the complain 'too old references'. There is a new Section 7.3 devoted to the description of the meta heuristic algorithm from [Pajor et Al. (2018)] and prepared writing about the experiments planned with this method. The response letter is updated accordingly, and now it seems almost complete, the only thing that I really cannot figure out is how to respond to the complaint about the absent contribution. So far I only created a separate section with a clear title, but I don't know how to explain the contribution more

--- 2018-08-30 smt/sec6-exp.tex
@@
 
+\textcolor{blue}{\subsection{Comparison with alternative methods}}
+
+\textcolor{blue}{
+To the best of our knowledge, SMT has not been studied extensively in terms of inexact methods.
+However, there is a manifold of a thorough research of the minimum Steiner tree problem, with the possibility of adapting the methods on SMT.
+We turn our attention to recently pusblished paper \citet{pajor18} which notains a metaheuristic algorithm for a fast generation of good solution to the Steiner tree problem.
+The method is general, and is therefore very suitable for adaptation to our purposes.
+The high level description is as follows:}
+\begin{algorithm}
+\textcolor{blue}{
+Create an empty pool $P$ of elite solutions with capacity $\gamma=\left\lceil\sqrt{M/2}\right\rceil$\;
+\For {$i\in \{1,\dots,M$\}} {
+Generate a new solution $S$ using a constructive algorithm, local search, and randomization\;
+\If {$i<\gamma$} {$P\leftarrow P\cup \{S\}$}
+\Else {
+$\hat{S}\leftarrow\arg\max_{S'\in P} c(S')$\;
+\If {$c(\hat{S})> c(S)$}{$P\leftarrow P\cup \{S\}\setminus\{\hat{S}\}$}
+$\hat{S}\leftarrow\arg\max_{S'\in P} c(S')$\;
+Generate a solution $\tilde{S}$ by combining $S$ with solutions in $P$\;
+\If {$c(\hat{S})> c(\tilde{S})$}{$P\leftarrow P\cup \{\tilde{S}\}\setminus\{\hat{S}\}$}
+}
+}
+\Return $\arg\min_{S'\in P} c(S')$\;
+ \caption{Outline of the metaheuristic algorithm}
+}
+\label{alg:meta}
+\end{algorithm}
+\textcolor{blue}{The construction procedure and local search is the adapted BIP algorithm together with refinements presented in \citet{ivanova16isco}.
+The remaining parts of the algorithm are carried out as discussed in \citet{pajor18}.
+The randomization is achieved by perturbing the edge costs in the input graph.
+The obtained solution $S$ is evaluated with respect to the real (original) edge costs and added to the pool $P$ if the pool capacity is not filled.
+If the pool is full and its worst solution $\hat{S}$ is worse than $S$, $\hat{S}$ is replaced by $S$ in $P$.
+Then a new solution $\tilde{S}$ is constructed by combining some solution in $P$ with $S$. 
+First, we adjust edge costs resulting in a modified input instance, 
+and then run the construction and local search yielding $\tilde{S}$.
+This new solution is again added to the pool if it is better than some solution already in $P$.
+The adjustment of the edges tries to promote edges that appear in both $S$ and a selected solution in $P$. 
+In that case, the original cost is preserved.
+If an edge is present in only one of the two solutions, its cost is multiplied by a random integer between 100 and 500.
+If an edge appears in neither of the solutions, its edge is multiplied by 1000, which makes it very unlikely to appear in the result $\tilde{S}$ of the combination.
+}
+\begin{table}[]
+\centering
+\textcolor{blue}{
+\begin{tabular}{rrrrrrrrrrr}
+\multicolumn{1}{l}{}      & \multicolumn{1}{l}{}      & \multicolumn{3}{c}{5 min}  & \multicolumn{3}{c}{10 min} & \multicolumn{3}{c}{20 min}     \\\hline
+\multicolumn{1}{l}{$|V|$} & \multicolumn{1}{l}{$|D|$} & \multicolumn{1}{l}{$\mathcal{X}_1$} & \multicolumn{1}{l}{$\mathcal{F}_1$} & \multicolumn{1}{l}{$\mathcal{M}$} & \multicolumn{1}{l}{$\mathcal{X}_1$} & \multicolumn{1}{l}{$\mathcal{F}_1$} & \multicolumn{1}{l}{$\mathcal{M}$} & \multicolumn{1}{l}{$\mathcal{X}_1$} & \multicolumn{1}{l}{$\mathcal{F}_1$} & \multicolumn{1}{l}{$\mathcal{M}$} \\\hline
+20  & 10   &     &   &   &   &    &    &    &    &       \\
+22  & 11   &     &   &   &   &    &    &    &    &       \\
+24  & 12   &     &   &   &   &    &    &    &    &                                  
+\end{tabular}
+}
+\caption{\textcolor{blue}{Comparison of the metaheuristic algorithm and solving IP models with imposed time-limit.}}
+\label{tab:metamod}
+\end{table}
+
+\textcolor{blue}{
+Tab.\ref{tab:metamod} summarizes the comparison of the metaheuristic algorithm with the ability of IP models to produce feasible solutions when restricted time limit.
+Each column represents the average percentage of objective function of the best method in a given instance.
+The averages are calculated over all instances of given $V$ and $D$.
+}
+
+\textcolor{blue}{
+The resulting values reveal that...
+}
+
 \subsection{Experiments on strong lower bounds}

### 2018-09-12 smt/ivanova.tex

@@
 
-\bibitem[Guo and Yang(2004)]{guo}
+\textcolor{blue}{\bibitem[Guo and Yang(2004)]{guo}
 Guo S, Yang O (2004)
@@
 In: Proceedings of the 12th IEEE International Conference on Network Protocols ICNP 2004,
-pp 151--160
+pp 151--160}
 

### 2018-09-16 smt/sec6-exp.tex
some results for the additional experiments (meta heuristics) are now in Tab 4.

@@
 \centering
+\setlength{\tabcolsep}{8pt} % Default value: 6pt
+\renewcommand{\arraystretch}{1.4} % Default value: 1
 \textcolor{blue}{
@@
 \multicolumn{1}{l}{}      & \multicolumn{1}{l}{}      & \multicolumn{3}{c}{5 min}  & \multicolumn{3}{c}{10 min} & \multicolumn{3}{c}{20 min}     \\\hline
-\multicolumn{1}{l}{$|V|$} & \multicolumn{1}{l}{$|D|$} & \multicolumn{1}{l}{$\mathcal{X}_1$} & \multicolumn{1}{l}{$\mathcal{F}_1$} & \multicolumn{1}{l}{$\mathcal{M}$} & \multicolumn{1}{l}{$\mathcal{X}_1$} & \multicolumn{1}{l}{$\mathcal{F}_1$} & \multicolumn{1}{l}{$\mathcal{M}$} & \multicolumn{1}{l}{$\mathcal{X}_1$} & \multicolumn{1}{l}{$\mathcal{F}_1$} & \multicolumn{1}{l}{$\mathcal{M}$} \\\hline
-20  & 10   &     &   &   &   &    &    &    &    &       \\
+\multicolumn{1}{l}{$|V|$} & \multicolumn{1}{l}{$|D|$} & \multicolumn{1}{c}{$\mathcal{X}_1$} & \multicolumn{1}{c}{$\mathcal{F}_1$} & \multicolumn{1}{c}{$\mathcal{M}$} & \multicolumn{1}{c}{$\mathcal{X}_1$} & \multicolumn{1}{c}{$\mathcal{F}_1$} & \multicolumn{1}{c}{$\mathcal{M}$} & \multicolumn{1}{c}{$\mathcal{X}_1$} & \multicolumn{1}{c}{$\mathcal{F}_1$} & \multicolumn{1}{c}{$\mathcal{M}$} \\
+\hline
+18  & 12   &100.6&100.0&101.4&   &    &    &    &    &       \\
+21  & 14   &     &   &   &   &    &    &    &    &       \\
+24  & 16   &     &   &   &   &    &    &    &    &   \\     
+\hline
+20  & 10   &100.6&100.0&101.4&   &    &    &    &    &       \\
 22  & 11   &     &   &   &   &    &    &    &    &       \\
-24  & 12   &     &   &   &   &    &    &    &    &                                  
+24  & 12   &     &   &   &   &    &    &    &    &   \\                               
+\hline
+21  & 7    &100.6&100.0&101.4&   &    &    &    &    &       \\
+24  & 8    &     &   &   &   &    &    &    &    &       \\
+27  & 9    &     &   &   &   &    &    &    &    &                                  
 \end{tabular}

### 2018-09-21 smt/sec7-heur.tex
DH: Added a section on the heuristic in a separate file. File names should be updated to become consistent with section numbers.

\textcolor{blue}{\section{Computing near-optimal solutions}} \label{sec:heur}

\textcolor{blue}{
Instances of size that prohibits computation of the optimal solution should be approached by some fast, but possibly inexact method.
Imposing a time bound on an IP-solver, and applying it to any of the models introduced in Section \ref{sec:ILP}, readily gives such a method.
Alternatively, a heuristic method dedicated to SMT can be developed.}

\textcolor{blue}{
To the best of our knowledge, the scientific literature offers few heuristic methods for SMT besides those studied by \citet{ivanova16isco}.
However, recent progress by \citet{pajor18} on heuristic methods for the related minimum Steiner tree problem can smoothly be exploited in order to generate analogous methods for SMT.
The main algorithmic components in the most basic heuristic are a construction algorithm, a local improvement algorithm,
and an algorithm for merging two Steiner trees in order to create a better one.
Construction and local improvement are applied in order to generate Steiner trees of presumably low cost.
If the cost is sufficiently low, the tree is stored in an ordered set $\mathcal{T}$ of at most $\gamma$ distinct trees.
Diversity among the trees is achieved by perturbing the edge costs randomly before construction and local improvement are applied.
After each tree generation, the new tree is merged with a randomly chosen tree from $\mathcal{T}$, and the resulting tree is kept in $\mathcal{T}$ if its cost is small.}

\textcolor{blue}{
Adapting the method by \citet{pajor18} to our problem is straightforward,
given the recently published construction and local improvement algorithms for SMT \citep{ivanova16isco}.
Let \textit{SMTHeur}$(G,D,p)$ denote the output produced when $(G,D,p)$ is input to the construction algorithm,
and the constructed tree is input to the local improvement method.
The adapted method is then described as follows, where $\mathcal{T}=\left(T_1,\ldots,T_{|\mathcal{T}|}\right)$
and $c\left(T_1\right)\leq\dots\leq c\left(T_{|\mathcal{T}|}\right)$:}

\begin{algorithm}
\textcolor{blue}{
\While{the time assigned to the heuristic is not expired} {
  \textbf{for} $\{i,j\}\in E$ \textbf{do} $\hat{p}_{ij}\leftarrow\text{random number uniformly distributed on
  $\left[\frac{1}{2}p_{ij},\frac{3}{2}p_{ij}\right]$}$\;
  $T\leftarrow\text{\textit{SMTHeur}$(G,D,\hat{p})$}$\;
  \textbf{if} $|\mathcal{T}|<\gamma$ \textbf{then} $\mathcal{T}\leftarrow \mathcal{T}\cup \{T\}$\;
  \Else {
    \textbf{if} $c(T)<c(T_{\gamma})$ \textbf{then} $\mathcal{T}\leftarrow \mathcal{T}\setminus\{T_{\gamma}\}$, $\mathcal{T}\leftarrow \mathcal{T}\cup \{T\}$\;
    $r\leftarrow\text{random number uniformly distributed on $\{1,\ldots,\gamma\}$}$\;
    \textbf{for} $\{i,j\}\in E_T\cap E_{T_r}$ \textbf{do} $\tilde{p}_{ij}\leftarrow p_{ij}$\;
    \For{$\{i,j\}\in \left(E_T\cup E_{T_r}\right)\setminus \left(E_T\cap E_{T_r}\right)$}{
      $\nu\leftarrow\text{random number uniformly distributed on $\{100,\ldots,500\}$}$\;
      $\tilde{p}_{ij}\leftarrow \nu p_{ij}$\;
    }
    \textbf{for} $\{i,j\}\in E\setminus E_T\setminus E_{T_r}$ \textbf{do} $\tilde{p}_{ij}\leftarrow 1000p_{ij}$\;
    $\tilde{T}\leftarrow\text{\textit{SMTHeur}$(G,D,\tilde{p})$}$\;
    \textbf{if} $c(\tilde{T})<c(T_{\gamma})$ \textbf{then} $\mathcal{T}\leftarrow \mathcal{T}\setminus\{T_{\gamma}\}$, $\mathcal{T}\leftarrow \mathcal{T}\cup \{\tilde{T}\}$\;
  }
}
\Return $T_1$
}

\textcolor{blue}{
\caption{Outline of the heuristic method}
\label{alg:meta}
}
\end{algorithm}

\textcolor{blue}{
Following \citet{pajor18}, the perturbed cost $\hat{p}_{ij}$ of edge $\{i,j\}$ is put equal to a random number with expected value $p_{ij}$.
Further, in the merge process, the adjusted weights $\tilde{p}_{ij}$ are determined in order to 
promote edges that appear in both $T$ and the randomly selected tree $T_{\gamma}$ in $\mathcal{T}$. 
The cost of edges appearing in exactly one of the trees are thus magnified.
For the purpose of diversification, the factor by which the cost is multiplied is chosen randomly.
To discourage the appearence of edges outside both $T$ and $T_r$ in the merge $\tilde{T}$,
their cost is magnified by a factor dominating the factors drawn for edges that occur in one of the trees.
}

\textcolor{blue}{
A few minor simplification over the approach taken by \citet{pajor18} are made in Alg.\ \ref{alg:meta}:
Rather than selecting the tree to leave $\mathcal{T}$ randomly, we consistently remove $T_{\gamma}$ when a new tree is added.
Also, for each new tre $T$ constructed, Alg.\ \ref{alg:meta} makes only one attempt to merge it with some $T_r$ stored in $\mathcal{T}$.
This contrasts the method in the cited work, where $\tilde{T}$ is chosen as the best tree found in a sequence of merges.
}

### 2018-09-23 smt/sec6-exp.tex
Talbe comparing meta algorithm and B&B filled. Added the description of the results.

--- 2018-09-12 smt/sec6-exp.tex
@@
 However, there is a manifold of a thorough research of the minimum Steiner tree problem, with the possibility of adapting the methods on SMT.
-We turn our attention to recently pusblished paper \citet{pajor18} which notains a metaheuristic algorithm for a fast generation of good solution to the Steiner tree problem.
-The method is general, and is therefore very suitable for adaptation to our purposes.
-The high level description is as follows:}
+We turn our attention to recently published paper \citet{pajor18} which contains a metaheuristic algorithm for a fast generation of good solution to the Steiner tree problem.
+The method is general, and is therefore very suitable for adaptation to our purposes.  The high level description is as follows:}
+
 \begin{algorithm}
@@
 \centering
+\setlength{\tabcolsep}{6pt} % Default value: 6pt
+\renewcommand{\arraystretch}{1.4} % Default value: 1
 \textcolor{blue}{
@@
 \multicolumn{1}{l}{}      & \multicolumn{1}{l}{}      & \multicolumn{3}{c}{5 min}  & \multicolumn{3}{c}{10 min} & \multicolumn{3}{c}{20 min}     \\\hline
-\multicolumn{1}{l}{$|V|$} & \multicolumn{1}{l}{$|D|$} & \multicolumn{1}{l}{$\mathcal{X}_1$} & \multicolumn{1}{l}{$\mathcal{F}_1$} & \multicolumn{1}{l}{$\mathcal{M}$} & \multicolumn{1}{l}{$\mathcal{X}_1$} & \multicolumn{1}{l}{$\mathcal{F}_1$} & \multicolumn{1}{l}{$\mathcal{M}$} & \multicolumn{1}{l}{$\mathcal{X}_1$} & \multicolumn{1}{l}{$\mathcal{F}_1$} & \multicolumn{1}{l}{$\mathcal{M}$} \\\hline
-20  & 10   &     &   &   &   &    &    &    &    &       \\
-22  & 11   &     &   &   &   &    &    &    &    &       \\
-24  & 12   &     &   &   &   &    &    &    &    &                                  
+\multicolumn{1}{l}{$|V|$} & \multicolumn{1}{l}{$|D|$} & \multicolumn{1}{c}{$\mathcal{X}_1$} & \multicolumn{1}{c}{$\mathcal{F}_1$} & \multicolumn{1}{c}{$\mathcal{M}$} & \multicolumn{1}{c}{$\mathcal{X}_1$} & \multicolumn{1}{c}{$\mathcal{F}_1$} & \multicolumn{1}{c}{$\mathcal{M}$} & \multicolumn{1}{c}{$\mathcal{X}_1$} & \multicolumn{1}{c}{$\mathcal{F}_1$} & \multicolumn{1}{c}{$\mathcal{M}$} \\
+\hline
+18  & 12   &100.0&100.0&101.8&100.0&100.0&100.0&100.0&100.0&100.0\\
+21  & 14   &106.5&107.4&101.2&102.7&103.1&100.1&100.9&100.3&100.0\\
+24  & 16   &144.2&112.2&101.3&124.0&114.0&100.1&104.3&112.7&100.0\\     
+\hline
+20  & 10   &100.6&100.0&101.4&100.1&100.0&100.0&100.1&100.0&100.0\\
+22  & 11   &102.1&103.2&101.7&101.3&100.7&100.1&100.7&100.2&100.1\\
+24  & 12   &106.6&111.7&101.7&105.0&105.5&100.0&103.1&106.2&100.0\\                               
+\hline
+21  & 7    &100.1&100.0&100.0&100.0&100.0&100.0&100.0&100.0&100.0\\
+24  & 8    &102.3&101.4&102.0&102.1&100.7&100.1&101.1&100.1&100.0\\
+27  & 9    &127.4&114.0&101.6&111.2&109.0&100.1&108.2&104.1&100.0                                  
 \end{tabular}
@@
 Tab.\ref{tab:metamod} summarizes the comparison of the metaheuristic algorithm with the ability of IP models to produce feasible solutions when restricted time limit.
-Each column represents the average percentage of objective function of the best method in a given instance.
+Each entry represents the average percentage  of the objective value yielded by the most successful method in a given instance.
+For example, if all methods find the optimum in every instance, the three corresponding values are all 100.0.
+Thus, the lowest value in a given instance type indicates that the corresponding method performed the most successfully. 
 The averages are calculated over all instances of given $V$ and $D$.
+Experiments with results in Tab. \ref{tab:small_inst} and Tab. \ref{tab:metamod} are carried out on the same sets of instances.
 }
@@
 \textcolor{blue}{
-The resulting values reveal that...
+The resulting values reveal that the metaheuristic algorithm converges quickly, and is often able to find optimal solutions faster than solving the IP models.
+The solutions are, however, not guaranteed to be optimal.
+The optimality is proven only on smaller instances, where the B\&B finds the optimum before the time limit is exceeded.
+It is likely that the inexact method run on large instances does not find the optimum, but it very clearly outperforms B\&B, 
+where the solutions found within the given time limit are far from the optimum.
 }
@@
 The vertical axes correspond to the lower bounds as a percentage of $z\left(\text{LP}(\mathcal{F}_1)\right)$, averaged over all instances in the set.
+\textcolor{blue}{The values can be retrieved from the output of Cplex during the course of simplex and B\&B algorithm.}
 
@@
 	The tightness of the bounds increases with time, and it can be observed from the graphs that the CG bound becomes tighter at some point.
-	We express the quality of lower bounds with respect to the value of simple continuous relaxatin, as indicated by the vertical axes.
+	We express the quality of lower bounds with respect to the value of simple continuous relaxation, as indicated by the vertical axes.
 	It can also be noted that the lower bound obtained from B\&B grows sharply during the computation of the simplex method in the root of the B\&B tree, 
@@
 	CG on the other hand produces bound that grows faster even after the value of first relaxation in the root is reached.
+%	The values can be retrieved from the output of CPLEX during the course of simplex and B\&B algorithm.
 	}}

### 2018-09-24 smt/response.tex
DH: Made some changes to the response letter.

@@
 We agree that a comparison with alternative contemporary methods largely benefits the paper.
-To that end, we have included a comparison with an adapted metaheuristic algorithm based on iterative combination of elite solutions \cite{pajor18}.
-The method is described along with presentatin of results from the experiments in the new Section 7.3.
+To that end, we have included a comparison with a recently published metaheuristic algorithm, based on iterative combination of elite solutions \cite{pajor18}.
+The method is described in the new Section 7, and results from experiments with the method are presented in the new Section 8.3.
 }
 \item \textcolor{red}{The simulation graphs need more expiation and justification.} \textcolor{blue}{
-For better justification of the numerical results illustrated graphically in the former Section 6 (now Section 7),
+For better justification of the numerical results illustrated graphically in the former Section 6 (now Section 8),
 we have added more explanations in the caption of Fig. 3.
@@
 In the new Section 3.2 on the network model, we now explain more detailed how the energy is modeled.
-At the end of the section, we also describe how our model differs from another models in the literature.
-Justification is added in terms of references to relevant and frequently cited works mentioned in Section 2 (see Wieselthier et al. (2000), Das et al. (2003), Altinkemer et al. (2005),
+At the end of the section, we also describe how our energy model relates to others in the literature.
+Justification is added in terms of references to \cite{halgamuge}, and to other relevant and frequently cited works mentioned in Section 2 (Wieselthier et al.
+(2000), Das et al.
+(2003), Altinkemer et al.
+(2005),
 Yuan (2005), Yuan et al. (2008), Bauer et al. (2008), Montemanni and Leggieri (2011) and Leggieri et al. (2008), etc.).
+See also our response to the second remark of Reviewer 3.
 }
@@
 We fully agree that there are gaps between the coarse energy model applied in our work and more detailed energy models for energy consumption is sensor networks,
-e.g.\ those analyzed in \cite{halgamuge}.
+e.g.\ those analyzed by Halgamuge et al.\ \cite{halgamuge}.
 These gaps, and their possible implications on practical usability of our approaches, are now discussed in the new Section 3.2.

### 2018-09-24 smt/sec2-literature.tex
DH: Summarizing previous changes.

@@
 \label{sec:literature}
+
 The combinatorial optimization literature is already rich on Integer Programming (IP) approaches to power minimization in wireless networks.

### 2018-09-24 smt/sec2-literature.tex
DH: Minor revision of most files.

--- 2018-09-06 smt/sec2-literature.tex
@@
 \textcolor{blue}{
-The latter paper proposes an IP formulation of MEM based on a Set Covering model together with a preprocessing procedure reducing the number of constraints.}
-\citet{cagalj02} proved that MEB, and thereby MEM, is NP-hard.
+The latter paper proposes an IP formulation of MEM based on a set covering model together with a preprocessing procedure reducing the number of constraints.}
+\citet{cagalj02} prove that MEB, and thereby MEM, is NP-hard.
 Numerous contributions to fast, but possible inexact methods for MEM have thus appeared, and the interested reader is referred to \citep{hsiao13} for an overview. 
-\textcolor{blue}{Another study concerning MEM is \cite{guo}, where the authors investigate energy conservation while using adaptive antennas. 
+\textcolor{blue}{Another study concerning MEM is the work by \cite{guo}, who investigate energy conservation while using adaptive antennas. 
 Unlike omnidirectional antennas, adaptive antennas can concentrate their transmission power in a smaller angle.}
 
+\textcolor{blue}{\citet{bein10} study energy efficiency in wireless networks where all nodes can initiate a message, and the communication takes place in a single all-to-all broadcast tree.
+A heuristic algorithm generating all-to-all broadcast trees is proposed by \citet{bhukya14}, along with an experimental evaluation confirming its merit.}
+
 The SBT problem has also received some attention in the scientific literature.
-It was introduced by \citet{Papadimitriou06SBT}, who also proved that the problem is NP-hard.
-In the same article, an approximation algorithm for SBT was developed and analyzed experimentally.
-Building on this work, \citet{Haugland12Dual} presented an integer programming model for SBT, along with an associated decomposition approach.
+It was first introduced by \citet{Papadimitriou06SBT}, who also proved that the problem is NP-hard.
+In the same article, an approximation algorithm for SBT is developed and analyzed experimentally.
+Building on this work, \citet{Haugland12Dual} present an integer programming model for SBT, along with an associated decomposition approach.
 

### 2018-09-24 smt/sec3-sbtsmt.tex
DH: Minor revision of most files.

--- 2018-08-30 smt/sec2-sbtsmt.tex
@@
 
-A WANET is modeled by a complete graph $G=(V,E)$, where the set $V$ of nodes represents the set of wireless devices and the set of edges $E=\{\{i,j\}:i,j\in V, i\neq j\}$ corresponds to the potential links between them.
+A WANET is modeled by a complete graph $G=(V,E)$, where the set $V$ of nodes represents the set of wireless devices,
+and the set of edges $E=\{\{i,j\}:i,j\in V, i\neq j\}$ corresponds to the potential links between them.
 The set $A=\{(i,j):i,j\in V,\{i,j\}\in E\}$ consists of all arcs that can be derived by directing the edges in $E$.
@@
 
-Sending a message directly from (to) node $i$ to (from) node $j$ requires a certain \emph{power} $p_{ij}=p_{ji}$.
+Sending a message directly from (to) node $i$ to (from) node $j$ requires a certain \emph{transmission power} $p_{ij}=p_{ji}$.
 Whenever appropriate, $p_{ij}$ is referred to as the \emph{cost} of $\{i,j\}$,
 and edges with higher cost than others are said to be \emph{more expensive}.
-When the nodes are embedded in the plane, $p_{ij}$ is often assumed \citep{Papadimitriou06SBT,Haugland12Dual} to be
+When the nodes are embedded in the plane, $p_{ij}$ is often assumed \citep{Papadimitriou06SBT,Haugland12Dual,halgamuge} to be
 proportional to the Euclidean distance between $i$ and $j$ raised to a power between 2 and 4.
@@
 as the set of nodes requiring no less power at $i$ than $j$ does.
-The notation $G'\subseteq G$ means that $G'=(V',E')$ is a subgraph of $G$, i.e., $E'\subseteq E$ consists of the edges in $E$ with both end nodes in $V'\subseteq V$.
+The notation $G'\subseteq G$ means that $G'=(V_{G'},E_{G'})$ is a subgraph of $G$, i.e., $E_{G'}\subseteq E$ consists of the edges in $E$ with both end nodes in $V_{G'}\subseteq V$.
 
@@
 
-In more general terms, let $T$ be a multicast tree, and let $i\in V_T$ be a node with at least two neighbors in $T$.
+In more general terms, let $T\subseteq G$ be a multicast tree, and let $i\in V_T$ be a node with at least two neighbors in $T$.
 For multicasting a message, the power to be assigned to node $i$ is either $p_{ii_1}$ or $p_{ii_2}$, where $i_1$ and $i_2$ are neighbors of $i$ such that
@@
 Summing over all nodes yields the total cost of the tree.
-% The objective in the problem addressed in the current work is to minimize this cost.
 
-To express the cost of $T$ more rigorously, let $T^s=(V_T,A^s)$ denote the directed tree (arborescence) obtained by directing all edges in $E_T$ such that they point away from the \emph{root} at $s$.
+\textcolor{blue}{
+The cost model outlined above aims for minimization of the transmission energy.
+As pointed out by \citet{halgamuge}, transmission constitutes a substantial part of the total energy consumption in wireless sensor networks.
+The energy consumed by other operations, such as sensing, logging, processing, actuation, and cluster formation is less sensitive to the topology
+of the multicast tree, and is neglected in the current work.
+This is in line with the approach taken in the wide range of literature on optimization models for wireless networks mentioned in Section \ref{sec:literature}.
+For a detailed energy model for wireless sensor networks, the reader is referred to the article by \citet{halgamuge}.}
+
+\subsection{Problem definition} \label{sec:probdef}
+
+To express the cost of the multicast tree more rigorously, let $T^s=(V_T,A^s)$ denote the directed tree (arborescence) obtained by directing all edges in $T$ such that they point away from the \emph{root} at $s$.
 For each arc $(i,j)\in A$, we let $n_{ij}(T)$ denote the number of destinations $s$ such that $(i,j)$ is the most expensive arc leaving node $i$ in $T^s$.

### 2018-09-24 smt/sec4-milp.tex
DH: Minor revision of most files.

\section{IP Formulations}
\label{sec:ILP}

A basic element of every IP formulation for SMT is a set of constraints modelling a Steiner tree.
We investigate two such constraint sets formulated in terms of variables with up to three node indices, and make necessary extensions required by the objective function of SMT.
The resulting models are subsequently strengthened by valid inequalities.

\subsection{Formulation based on broadcast trees}

The first model extends the SBT formulation introduced by \citet{Haugland12Dual} by non-destination nodes, in order to formulate the multicast version of the problem \citep{ivanova16isco}.

\subsubsection{Underlying Steiner tree formulation}

Feasible solutions to SMT are characterized by a Steiner tree $T$ covering $D$.
Define for each edge $\{i,j\}\in E$ and each destination node $s\in D$, the binary variables $y\in\{0,1\}^E$ and $X\in\{0,1\}^{A\times D}$,
where $y_{ij}=1$ iff $\{i,j\} \in E_T$,
and $X^{s}_{ij}=1$ iff $(i,j)\in A^s$.
% \[y_{ij}=
% \begin{cases}
%     1 & \text{if $\{i,j\} \in E_T$},\\
%     0 & \text{otherwise},
%   \end{cases}
% \]
% 
% \[X^{s}_{ij}=
% \begin{cases}
%     1 & \text{if $(i,j)$ is an arc in the arborescence $T^s$},\\
%     0 & \text{otherwise}.
%   \end{cases}
% \]
To induce a graph with an acyclic connected component spanning $D$, the following constraints are imposed:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                     % 
%     STEINER TREE MODEL  X0          %
%                                     %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{subequations}\label{mod:x0}
\begin{align}
\label{con:dd:arrowFromDest} \sum\limits_{j\in V_i}X^s_{ji} & = 1 & i,s\in D,i\neq s,\\
\label{con:dd:arrowFromNonDestB} \sum\limits_{j\in V_{i}}X^s_{ji} & \leq 1 & i\in V \setminus D, s\in D,\\
\label{con:dd:arrowFromNonDestA} X^s_{ij} & \leq \sum\limits_{k\in V_{i}\setminus \{j\}}X^s_{ki} & i\in V \setminus D,(i,j)\in A, s\in D,\\
\label{con:dd:oneDir} X^s_{ij} + X^s_{ji} & = y_{ij} & \{i,j\}\in E, s\in D,\\
\label{con:dd:startInSource} X^s_{is} & = 0 &  s\in D, (i,s)\in A,&\\
\label{con:dd:vardim}y \in \{0,1\}^{E}, X & \in \{0,1\}^{A\times D}.
\end{align}~
\end{subequations}

Constraints (\ref{con:dd:arrowFromDest}) ensure that all destination nodes but $s$ have a unique entering arc in $T^s$.
Inequalities (\ref{con:dd:arrowFromNonDestB})--(\ref{con:dd:arrowFromNonDestA})
say that non-destination nodes do not have more than one entering arc, and that they do have one if they have leaving arcs.
Constraints (\ref{con:dd:oneDir}) state that the arc set of $T^s$ consists of directed versions of the edge set of $T$,
and (\ref{con:dd:startInSource}) ensure that $s$ is the root of $T^s$.
More formally, the effect of the constraints can be stated as follows:

\begin{lemma} \label{lem:x0valid}
If $(X,y)$ satisfies (\ref{con:dd:arrowFromDest})--(\ref{con:dd:vardim}), then $G_y$ has a tree spanning $D$ as one of its connected components.
Conversely, for any tree $T\subseteq G$ spanning $D$, there exists a pair $(X,y)$ satisfying (\ref{con:dd:arrowFromDest})--(\ref{con:dd:vardim}) such that $T$
is a connected component of $G_y$.
\end{lemma}
\begin{proof}
By (\ref{con:dd:startInSource}) and (\ref{con:dd:arrowFromDest})--(\ref{con:dd:arrowFromNonDestB}),
the in-degree of $s\in D$ is 0 in $G_{X^s}$, whereas it is at most one at all other nodes.
Hence, the connected components of $G_{X^s}$ are directed cycles, arborescences, and isolated nodes,
and one of the components is an arborescence rooted at $s$.
Constraints (\ref{con:dd:oneDir}) imply that $G_y$ consists of undirected versions of the same components,
showing that no cycle in $G_y$ contains destination nodes.
Let $t\in D$, and let $P$
be the maximal path in $G_{X^s}$ that terminates in $t$.
Existence and uniqueness of this path is assured since $t$ belongs to an acyclic connected component.
According to (\ref{con:dd:arrowFromNonDestA}), each non-destination in $P$ has an entering arc in $P$,
while (\ref{con:dd:arrowFromDest}) ensures that the same is true for every destination node, except $s$.
It follows that the first node of $P$ is $s$.
Thus, destinations $s$ and $t$ belong to a joint, acyclic connected component of $G_y$.

The second claim of the lemma is obvious.\qed
\end{proof}

\subsubsection{SMT model [$\mathcal{X}_1$]} \label{sec:x1}

To reflect that only expensive edges in $T$ contribute to the objective function of Problem \ref{def:problem}, we introduce the variables:
\newline\newline
$\pi^s_{ij}=
\begin{cases}
    1 & \text{if $(i,j)$ is the most expensive arc leaving node $i$ in $T^s$},\\
    0 & \text{otherwise},
  \end{cases}$
\newline\newline
leading to the following formulation $\mathcal{X}_1$ of SMT:

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                     % 
%             MODEL X1                %
%           (3 idx vars)              %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{subequations}[resume]\label{eq:smt-x2}
\begin{align}
\label{objective:mf} \makebox[0pt][l]{$\displaystyle{} \min \sum\limits_{(i,j) \in A} p_{ij} \sum\limits_{s \in D} \pi^s_{ij} $}\\
\notag\text{s.t.}\\
\notag(\ref{con:dd:arrowFromDest}) &- (\ref{con:dd:vardim})\\
\label{con:dd:yvar} X^s_{ij} & \leq \sum_{k\in W_{ij}}\pi^s_{ik} & s\in D, (i,j)\in A,\\
\label{con:dd:vardimy} \pi & \in \{0,1\}^{A\times D}.
\end{align}~
\end{subequations}

Constraints (\ref{con:dd:yvar}) say that if $(i,j)\in A_{T^s}$, then the cost of the most expensive arc in $T^s$ leaving $i$ is at least $p_{ij}$.
Since $\sum_{s\in D}\pi_{ij}^s=n_{ij}(T)$ is the number of arborescences in which $(i,j)$ is the most expensive arc leaving $i$,
(\ref{objective:mf}) thus agrees with the objective of Problem \ref{def:problem}.

\begin{prop}
\label{prop:x1valid}
If $(X,y,\pi)$ is an optimal solution to $\mathcal{X}_1$, then $G_y$ consists of isolated non-destination nodes and a tree that
solves Problem \ref{def:problem}.
\end{prop}
\begin{proof}
Follows directly from $p\geq 0$ and Lemma \ref{lem:x0valid}.
\qed
\end{proof}

Model $\mathcal{X}_1$ is a slightly modified version of the SMT model introduced by \citet{ivanova16isco},
which contains a set of constraints disallowing non-destination leaves, and a weaker version of constraints (\ref{con:dd:arrowFromNonDestA}).
Formulation (\ref{objective:mf})--(\ref{con:dd:vardimy}) is \emph{minimal},
in the sense that each set of constraints is necessary in order to give a valid formulation.
In the next section, we strengthen the formulation by adding redundant valid inequalities.

\subsubsection{Valid inequalities [$\mathcal{X}_2$]} \label{sec:x2}

Adding the following valid inequalities to $\mathcal{X}_1$ gives another SMT-model, denoted $\mathcal{X}_2$:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                     % 
%             MODEL X2                %
%      (3 idx vars + VIs )            %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{subequations}[resume]
\begin{align}
\label{con:dd:extraCon} \sum\limits_{j\in V_{i}}X^s_{ji} & \leq \sum\limits_{j\in V_{i}}X^s_{ij}  & 	i\in V\setminus D, s\in D,\\
\label{con:vi:Y1} \sum\limits_{j\in V_s}  \pi^{s}_{sj} & =1 & s\in D,\\
\label{con:vi:sumYImpSumX} \sum\limits_{j\in V_i\setminus\{s\} }\pi^{s}_{ij} & = \sum\limits_{j\in V_i}  X^{s}_{ji} & i\in V\setminus D, s\in D.
\end{align}
\end{subequations}
Constraints (\ref{con:dd:extraCon}) ensure that the number of arcs leaving a non-destination in $T^s$ is no less than the number of entering arcs.
Their effect is to disallow non-destination leaves.
Equations (\ref{con:vi:Y1}) say that in arborescence $T^s$, exactly one arc $(s,j)$ leaving the root $s$ is declared to be largest.
Validity of the constraints follows immediately because the root has at least one leaving arc.
As constraints (\ref{con:vi:sumYImpSumX}) state, if an arc enters non-destination node $i$ in $T^s$, then $i$ also has at least one out-going arc,
exactly one of which is the most expensive.
The root $s$ is excluded from the summation set on the left of (\ref{con:vi:sumYImpSumX}) since no arcs in $T^s$ enter $s$.

\begin{remark} \label{rem:x2}
The sets (\ref{con:dd:extraCon})--(\ref{con:vi:sumYImpSumX}) of valid inequalities are \emph{independent} in the following sense:
There exist instances where the removal of either set implies that $z(\text{LP}(\mathcal{X}_2))$ takes a smaller value.
This includes the instance where $|V|=9$, $|D|=5$, all power requirements (arc costs) $p_{ij}$ equal the square of the Euclidean distance between nodes $i$ and $j$,
and the nodes are located in the plane with respective coordinates (destinations mentioned first)
(47, 65), (8, 65), (26, 12), (78, 27), (8, 8), (73, 66), (13, 55), (91, 44), and (76, 74).
\end{remark}

\subsection{Formulation based on network flows}

There are many formulations for the \textsc{Minimum Steiner Tree} problem \citep{goemans93catalog} that can serve as a basis for modelling SMT.
We consider the multi-commodity network flow model developed by \citet{Polzin} (denoted $P_{F}$ in their work).
A given destination $s_0\in D$ plays a particular role in the models.
For the sake of simplified notation, $s_0$ as subscript (superscript) is henceforth replaced by subscript (superscript) 0.

\subsubsection{Underlying Steiner tree formulation}

To identify a Steiner arborescence $T^0=(V^0,A^0)$ rooted at $s_0$, we define variables $g\in\{0,1\}^A$ and $F\in\{0,1\}^{A\times D_0}$,
where $g_{ij}=1$ iff $(i,j)\in A$, and $F_{ij}^s=1$ iff  the path in $T^0$ from $s_0$ to destination $s$ contains arc $(i,j)$.
% the following variables are introduced for each arc $(i,j)\in A$ and destination $s\in D_0$:
% \newline\newline
%   $g_{ij}=
% \begin{cases}
%     1 & \text{if $(i,j) \in A^0$},\\
%     0 & \text{otherwise},
% \end{cases}$
% \newline\newline
%   $F^{s}_{ij}=
% \begin{cases}
%     1 & \text{if the path from $s_0$ to $s$ in $T^0$ contains $(i,j)$},\\
%     0 & \text{otherwise}.
% \end{cases}$
% \newline\newline
%
Variables $g_{ij}$ and $F_{ij}^s$ are traditionally referred to as the \emph{design variable} of arc $(i,j)$ and the
\emph{flow of commodity $s$} along $(i,j)$, respectively.
The constraints of the \emph{multi-commodity flow formulation}
of the \textsc{Minimum Steiner Tree} problem are:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                     % 
%     STEINER ARBORESCENCE F0         %
%        (from Polzin)                %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{subequations}
\begin{align}
\label{con:pf1:xfrel} F^{s}_{ij} & \leq g_{ij} & s\in D_0, (i,j)\in A, \\
\label{con:pf1:flow} \sum\limits_{\substack{ j \in V_i }}F^{s}_{ji}-\sum\limits_{\substack{j\in V_i}}F^{s}_{ij} &= 
  \begin{cases}
    1, & i=s,\\
    0, & i\in V\setminus \{s_0, s\},
  \end{cases} & s\in D_0,\\
\label{con:pf1:B}  \sum_{j\in V_i}g_{ji}&\leq 1 & i\in V\setminus D,\\
\label{con:pf1:noflowFromT} F_{si}^s&=0 & s\in D_0, i\in V_s,\\
\label{con:pf1:fitt=xit} F_{is}^s&=g_{is} & s\in D_0, i\in V_s,\\
\label{con:pf1:xi0=0} g_{i0}&=0 & i\in V_0,\\
\label{con:pf1:dim}g \in \{0,1\}^{A},F&\in\{0,1\}^{A \times D_0}.
\end{align}~
\end{subequations}

The \emph{flow conservation} constraints (\ref{con:pf1:flow}) ensure that $T^0$ is connected,
and the \emph{capacity} constraints (\ref{con:pf1:xfrel}) state that flow of any commodity is allowed only if $(i,j)\in A^0$.
Multi-commodity flow formulations appear abundantly in the network design literature, and proof of the following claim is therefore omitted:

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                     %
%         Proposition  		      %
% (F1 models an arborescence)         %
%                                     %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{lemma}
\label{lem:modelcorrect}
Assume $(F,g)$ satisfies (\ref{con:pf1:xfrel})--(\ref{con:pf1:dim}).
Then some connected component $T^0\subseteq G_{g}$ is an $s_0$-rooted arborescence spanning $D$,
and some connected component of $G_{F^s}\subseteq T^0$ is a path from $s_0$ to $s\in D_0$.
Conversely, for any $s_0$-rooted arborescence $T^0$ spanning $D$, there exists a pair $(F,g)$ satisfying (\ref{con:pf1:xfrel})--(\ref{con:pf1:dim}) such that $T^0$
is a connected component of $G_g$.
\end{lemma}

To model the objective function of SMT, the following result is useful:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                     %
%         Proposition  		      %
%   (transformation X_{ij}^s)         %
%                                     %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{prop}\label{prop:transx}
Assume $(F,g)$ satisfies (\ref{con:pf1:xfrel})--(\ref{con:pf1:dim}),
and let $T^s=(V^s,A^s)$ be the arborescence obtained by redirecting the arcs in $T^0$ such that they point away from $s\in D_0$. Then
$$
g_{ij} - F^s_{ij}+F^s_{ji} = 
	\begin{cases}
		1, ~\text{if $(i,j)\in A^s$}, \\
		0, ~\text{otherwise.}
	\end{cases}
$$
\end{prop}
%
\begin{proof}
Because $T^0$ is acyclic, it follows from the capacity constraints (\ref{con:pf1:xfrel}) that $g_{ij}+F^s_{ji}\leq 1$.
The same constraints also imply $g_{ij} - F^s_{ij}\geq 0$.
Hence, $g_{ij} - F^s_{ij}+F^s_{ji}\in\{0,1\}$.

Arborescence $T^s$ contains arc $(i,j)$ iff $(i,j)$ is in $A^0$ but not on a path to $s$, or its reverse arc $(j,i)$ is on such a path.
By Lemma \ref{lem:modelcorrect}, this disjunction is equivalent to $g_{ij}-F_{ij}^s=1$ or $F_{ji}=1$.
The result then follows since the two conditions are mutually exclusive.
\qed
\end{proof}

% $\mathcal{F}_1$ based on $\mathcal{F}_0$.
% For this purpose, it is necessary to find a way to represent the constraint (\ref{con:dd:yvar}) in the $\mathcal{F}_1$ space.
% The $\pi$-variables from $\mathcal{X}_1$ have to be used in $\mathcal{F}_1$, because they appear in the objective function which remains unchanged.

\subsubsection{SMT model [$\mathcal{F}_1$]}

Recall from Section \ref{sec:x1} that $\pi_{ij}^s$ determines whether $(i,j)$ is the most expensive arc leaving node $i$ in arborescence $T^s$.
Taking advantage of Prop.\ \ref{prop:transx}, we can thus formulate Problem \ref{def:problem} in terms of variables $F$, $g$, and $\pi$,
resulting in model $\mathcal{F}_1$:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                     % 
%             MODEL F1                %
%        (3 idx vars)                 %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{subequations}[resume]
\begin{align}
\notag\min & \sum\limits_{(i,j) \in A} p_{ij} \sum\limits_{s \in D} \pi^s_{ij}\\ 
\notag\text{s.t.} && \\
\notag(\ref{con:pf1:xfrel}) - (\ref{con:pf1:dim})\\
\label{con:pf1:yvar} g_{ij} - F^s_{ij}+F^s_{ji} & \leq \sum_{k\in W_{ij}}\pi^s_{ik}  & s\in D_0, (i,j)\in A,\\
\label{con:pf1:yvar0} g_{ij} & \leq \sum_{k\in W_{ij}}\pi^0_{ik}   &  (i,j)\in A,\\
\label{con:pf1:dimy} \pi&\in \{0,1\}^{A\times D}.
\end{align}~
\end{subequations}

Like (\ref{con:dd:yvar}), constraints (\ref{con:pf1:yvar}) ensure that if $(i,j)\in A^s$, then $\pi_{ik}^s=1$ for some
arc $(i,k)$ at least as expensive as $(i,j)$.
As $g_{ij}$ determines whether $(i,j)\in A^0$, the corresponding constraint for destination $s_0$ is simplified to (\ref{con:pf1:yvar0}).
Validity of $\mathcal{F}_1$ is thus stated as a corollary to Prop.\ \ref{prop:transx}:

\begin{cor}
If $(F,g,\pi)$ is an optimal solution to $\mathcal{F}_1$, then $G_g$ consists of isolated non-destination nodes and an arborescence $T^0$.
The tree obtained by disregarding arc directions in $T^0$ is an optimal solution to Problem \ref{def:problem}.
\end{cor}

\begin{remark} \label{rem:f1a}
Like $\mathcal{X}_1$, also $\mathcal{F}_1$ is a minimal formulation.
Constraints (\ref{con:pf1:B}) prevent non-destinations from having multiple entering arcs.
In a \textsc{Minimum Steiner Tree} formulation, where a linear function of $g$ is minimized subject to constraints
(\ref{con:pf1:xfrel})--(\ref{con:pf1:flow}) and (\ref{con:pf1:noflowFromT})--(\ref{con:pf1:dim}),
constraints (\ref{con:pf1:B}) are implied by optimality.
The necessity of (\ref{con:pf1:B}) in formulation $\mathcal{F}_1$ of SMT is demonstrated by the optimal solution to $\mathcal{F}_1$ after removal of (\ref{con:pf1:B})
in the following instance: $V=\{s_0,a,\ldots,i\}$ with coordinates
(3, 61),
(79, 80),
(32, 54),
(93, 21),
(12, 70),
(93, 60),
(72, 78),
(76, 80),
(39, 84), and
(82, 80), respectively, edge costs equal to the square of the Euclidean distance between end points, and $D=\{s_0,a,\ldots,f\}$.
Figure \ref{fig:BProof} depicts the optimal solution to SMT, and the solution to $\mathcal{F}_1$ after removal of (\ref{con:pf1:B}).
%The optimal solution with objective value 25156 to the depicted instance obtained by solving SMT-$\mathcal{X}_1$ is shown in Fig. \ref{fig:BorigSMT}.
%The solution in Fig \ref{fig:Bpf2} yielded by solving SMT-$\mathcal{F}_1$ without the constraint (\ref{con:pf1:B}) has objective value 25148, but is not a feasible solution to Problem \ref{def:problem}, because of the cycle $(g,h,i,d,g)$.
%The non-existence of such a cycle in a solution given by model $\mathcal{X}_1$ is ensured by constraints (\ref{con:dd:arrowFromDest}), (\ref{con:dd:arrowFromNonDestB}) and (\ref{con:dd:oneDir}).
%A detailed proof of this claim can be found in \cite{ivanova16isco}.
%A transmission commenced in node $c$ is sent via arc $(g,f)$.
%As a consequence of the link between $g$ and $d$ in Fig. \ref{fig:Bpf2}, the node $d$ also receives the message.
%This link is absent in Fig. \ref{fig:BorigSMT}, and so $i$ has to relay the signal using arc $(i,d)$, causing the higher total objective value.
\end{remark}
\begin{remark} \label{rem:f1b}
The obviously valid inequalities (\ref{con:pf1:noflowFromT})--(\ref{con:pf1:fitt=xit}) are not necessary in the \textsc{Minimum Steiner Tree} formulation,
but have to be included in the formulation $\mathcal{F}_1$ of SMT in order to disallow nodes in $D_0$ to have multiple entering arcs.
Removal of (\ref{con:pf1:noflowFromT}) or (\ref{con:pf1:fitt=xit}) implies that the optimal solution to the instance where
$V=\{s_0,a,\ldots,k\}$ with coordinates
(43, 15),
(89, 88),
(56, 12),
(44, 66),
(60, 42),
(91, 48),
(64, 8),
(64, 34),
(68, 9),
(41, 53),
(85, 67), and
(93, 49), respectively,
edge costs defined as in Remark \ref{rem:f1a}, and $D=\{s_0,a,\ldots,h\}$,
is not a feasible solution to SMT.
\end{remark}

% \begin{figure}[!htb]
%     \centering
%     \begin{subfigure}[b]{0.4\textwidth}
%         \includegraphics[width=\textwidth]{conBNec}
%         \caption{Optimal SMT-solution with total cost 25156.\newline~}
%         \label{fig:BorigSMT}
%     \end{subfigure}
%     \hfill
%     \begin{subfigure}[b]{0.4\textwidth}
%         \includegraphics[width=\textwidth]{conBNec2}
%         \caption{Optimal solution $G_g$ to $\mathcal{F}_1$ after omission of constraints (\ref{con:pf1:B}).
%                  Its total cost is 25148, but as it is acyclic, it is infeasible in SMT.}
%         \label{fig:Bpf2}
%     \end{subfigure}
%     \caption{Instance with $D=\{s_0,a,\ldots,e\}$ demonstrating necessity of constraints (\ref{con:pf1:B}) in $\mathcal{F}_1$.
% 		The edge labels denote costs.
% 		For better legibility, some edges are contracted in the illustration.} 
%     \label{fig:BProof}
% \end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%      Figure                                        %
%    explaining why some constraints are necessary   %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\drawnodes}{
\begin{scope}[every node/.style={circle,draw=black,fill=gray!40,minimum size=1.5em, inner sep=2, scale=.8}]
 \node (s0) at (11,10) {$s_0$};
 \node (a) at (9,11) {$a$};
 \node (b) at (8,8) {$b$};
 \node[fill=none] (f) at (5,11) {$f$};
 \node[fill=none] (g) at (4.5,5) {$g$};
 \node[fill=none] (h) at (4,3) {$h$};
 \node[fill=none] (i) at (4.5,0.8) {$i$};
 \node (c) at (5,2) {$c$};
 \node (d) at (8,1) {$d$};
 \node (e) at (12,0.5) {$e$};
\end{scope}
}
\begin{figure}[!htb]
\centering
\begin{subfigure}[b]{0.48\textwidth}
\begin{tikzpicture}[scale=.6]
 \drawnodes
 \begin{scope}[above=5pt, every edge/.style={draw=black,thick}, every node/.style={scale=.8}]
 \draw  (s0) edge node[right=5pt,above=-1pt]{162} (a);
 \draw  (a) edge node{925} (f);
 \draw  (f) edge node[right=2pt]{949} (b);
 \draw  (f) edge node[right=-1pt]{1125} (g);
 \draw  (g) edge node[left=-1pt]{20} (h);
 \draw  (h) edge node[right=3pt,above=-2pt]{9} (c);
 \draw  (c) edge node[left=3pt,above=-3pt]{9} (i);
 \draw  (i) edge node{521} (d);
 \draw  (d) edge node{1521} (e);
 \end{scope}
\end{tikzpicture}
\caption{Optimal SMT-solution with total cost 25156.\newline~}
\label{fig:BorigSMT}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.48\textwidth}
\begin{tikzpicture}[scale=.6]
\drawnodes
 \begin{scope}[->,above=5pt, every edge/.style={draw=black,thick}, every node/.style={scale=.8}]
 \draw  (s0) edge node[right=5pt,above=-1pt]{162} (a);
 \draw  (a) edge node{925} (f);
 \draw  (f) edge node[right=2pt]{949} (b);
 \draw  (f) edge node[right=-1pt]{1125} (g);
 \draw  (g) edge node[left=-1pt]{20} (h);
 \draw  (g) edge node[right=-1pt]{53} (c);
 \draw  (d) edge node[right=2pt]{765} (g);
 \draw  (h) edge node[left=6pt,above=-6pt]{36} (i);
 \draw  (i) edge node{521} (d);
 \draw  (d) edge node{1521} (e);
 \end{scope}
\end{tikzpicture}
\caption{Optimal solution $G_g$ to $\mathcal{F}_1$ after omission of constraints (\ref{con:pf1:B}).
         Its total cost is 25148, but the cycle shows that it is infeasible in SMT.}
\label{fig:Bpf2}
\end{subfigure}
\caption{Instance demonstrating necessity of constraints (\ref{con:pf1:B}) in $\mathcal{F}_1$.
         Edge labels denote costs.
         For better legibility, some edges are contracted in the illustration.} 
\label{fig:BProof}
\end{figure}

\subsubsection{Valid inequalities [$\mathcal{F}_2$]}

Valid inequalities analogous to those leading to model $\mathcal{X}_2$ in Section \ref{sec:x2} are added to $\mathcal{F}_1$, resulting in a model denoted $\mathcal{F}_2$.
Analogously to (\ref{con:dd:extraCon}), to enforce non-destination nodes included in the spanning arborescence to have at least one child node,
the following constraints are introduced \citep{Polzin}:
\begin{subequations}[resume]
\begin{flalign}
\label{con:pf1:flowX}  \sum\limits_{\substack{ j\in V_i }}g_{ji}-\sum\limits_{\substack{j\in V_i}}g_{ij}    & \leq 0    \qquad\qquad i\in V\setminus D.
\end{flalign}
\end{subequations}

According to Prop.\ \ref{prop:transx}, if $i\in V\setminus D$ and $g_{ji} - F^s_{ji}+F^s_{ij}=1$ for some $j\in V_i$ and $s\in D_0$,
then $i$ is a Steiner node used to establish connection between $s$ and some other destination.
Since $i$ also has a leaving arc in $T^s$, we have $\pi_{ik}^s=1$ for some $k\neq i,s$.
This proves validity of the constraint $\sum_{j\in V_i}\left(g_{ji} - F^s_{ji}+F^s_{ij}\right)=\sum_{j\in V_i\setminus\{s\}}\pi_{ij}^s$
Application of flow conservation (\ref{con:pf1:flow}) yields the valid constraints
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                     %
%             MODEL F2                %
%                                     %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{subequations}[resume]
\begin{flalign}
\label{con:vi:sumYImpSumXTrans} \sum\limits_{j\in V_i\setminus\{s\}}\pi^{s}_{ij} & = \sum\limits_{j\in V_i}  g_{ji}  \quad\quad   i\in V\setminus D, s\in D. 
\end{flalign}
\end{subequations}

\noindent
Finally, constraints (\ref{con:vi:Y1}) are valid,
completing formulation $\mathcal{F}_2$ as model $\mathcal{F}_1$ extended by (\ref{con:pf1:flowX})--(\ref{con:vi:sumYImpSumXTrans}) and (\ref{con:vi:Y1}).
%
\begin{remark} \label{rem:f2}
With $s_0$ equal to the destination node in position (47, 65), the instance given in Remark \ref{rem:x2} also proves independence of the three sets
(\ref{con:vi:Y1}), (\ref{con:vi:sumYImpSumXTrans})--(\ref{con:pf1:flowX}) of valid inequalities.
\end{remark}

### 2018-09-24 smt/sec8-exp.tex
DH: Minor revision of most files.

--- 2018-09-23 smt/sec6-exp.tex
@@
 
-\begin{algorithm}
-\textcolor{blue}{
-Create an empty pool $P$ of elite solutions with capacity $\gamma=\left\lceil\sqrt{M/2}\right\rceil$\;
-\For {$i\in \{1,\dots,M$\}} {
-Generate a new solution $S$ using a constructive algorithm, local search, and randomization\;
-\If {$i<\gamma$} {$P\leftarrow P\cup \{S\}$}
-\Else {
-$\hat{S}\leftarrow\arg\max_{S'\in P} c(S')$\;
-\If {$c(\hat{S})> c(S)$}{$P\leftarrow P\cup \{S\}\setminus\{\hat{S}\}$}
-$\hat{S}\leftarrow\arg\max_{S'\in P} c(S')$\;
-Generate a solution $\tilde{S}$ by combining $S$ with solutions in $P$\;
-\If {$c(\hat{S})> c(\tilde{S})$}{$P\leftarrow P\cup \{\tilde{S}\}\setminus\{\hat{S}\}$}
-}
-}
-\Return $\arg\min_{S'\in P} c(S')$\;
- \caption{Outline of the metaheuristic algorithm}
-}
-\label{alg:meta}
-\end{algorithm}
 \textcolor{blue}{The construction procedure and local search is the adapted BIP algorithm together with refinements presented in \citet{ivanova16isco}.

### 2018-09-24 smt/response.tex
DH: Summarizing previous changes.
MI (SMT): Updated response letter - bibliography section removed, references unified to Name (year) format, comment about blue and black thext in the final version of the paper.

@@
 Following the recommendation of the reviewer, we have reorganized the literature review, and devoted a new section (Section 2 in the revised manuscript)
-to this topic. We have also included a review of more recent and relevant literature (references \cite{bein10, bhukya14}).
+to this topic. We have also included a review of more recent and relevant literature (references Bein and Zheng (2010), Bhukya and Singh (2014)).
 In Section 1.2, we now state explicitly the contributions from the present work, and explain how they relate to the current literature.
@@
 We agree that a comparison with alternative contemporary methods largely benefits the paper.
-To that end, we have included a comparison with a recently published metaheuristic algorithm, based on iterative combination of elite solutions \cite{pajor18}.
+To that end, we have included a comparison with a recently published metaheuristic algorithm, based on iterative combination of elite solutions Pajor et al. (2018).
 The method is described in the new Section 7, and results from experiments with the method are presented in the new Section 8.3.
@@
 At the end of the section, we also describe how our energy model relates to others in the literature.
-Justification is added in terms of references to \cite{halgamuge}, and to other relevant and frequently cited works mentioned in Section 2 (Wieselthier et al.
-(2000), Das et al.
-(2003), Altinkemer et al.
-(2005),
+Justification is added in terms of references to  Halgamuge et al. (2009), 
+and to other relevant and frequently cited works mentioned in Section 2 (Wieselthier et al. (2000), Das et al. (2003), Altinkemer et al. (2005),
 Yuan (2005), Yuan et al. (2008), Bauer et al. (2008), Montemanni and Leggieri (2011) and Leggieri et al. (2008), etc.).
@@
 \noindent
-Final comment: We thank the reviewers again for their constructive remarks, which largely helped us to improve the presentation of our findings.
+Final comment: 
+Newly added parts of the paper are also in \textcolor{blue}{blue} font.
+In several instances the text is restructured, but the sentences are not altered. In these cases the font color remains black.
+Such updates should be apparent due to new highlighted sections.
 
-\begin{thebibliography}{}
+We thank the reviewers again for their constructive remarks, which largely helped us to improve the presentation of our findings.
 
-\bibitem{bein10}
-Bein D, Zheng SQ (2010)
-Energy efficient all-to-all broadcast in all-wireless networks.
-Information Sciences 180(10):1781--1792
-
-\bibitem{bhukya14}
-Bhukya WN, Singh A (2014)
-\emph{p}-shrink: A Heuristic for Improving Minimum All-to-All Power Broadcast Trees in Wireless Networks.
-In: Proceedings of Ninth International Conference on Wireless Communication and Sensor Networks 2014 
-Vol.\ 299, pp 61--69
-
-\bibitem{halgamuge}
-Halgamuge MN, Zukerman M, Ramamohanarao K (2009)
-An estimation of sensor energy consumption.
-Progress In Electromagnetics Research B 12:259--295
-
-\bibitem{pajor18}
-Pajor T, Uchoa E, Werneck RF (2018)
-A robust and scalable algorithm for the Steiner problem in graphs.
-Math Prog Comp 10:69--118
-
-\end{thebibliography}
 \end{document}

### 2018-09-24 smt/sec1-intro.tex

@@
 
+\textcolor{blue}{When proven optimality is out of reach, effective methods for computing near-optimal solutions are crucial.
+A heuristic method tailored for SMT is another contribution from the current text,
+which thus demonstrates how recent progress on the \textsc{Minimum Steiner Tree} problem can be transferred to the problem under study.}
+
+\textcolor{blue}{\subsection{Scope and structure of the paper}}
+
 For the same reasons that MEB has been generalized to MEM, it is natural and desirable to extend SBT to its multicast equivalent.
 The \textsc{Minimum Shared Multicast Tree} (SMT) problem is thus at the forefront of this paper,
-and neither MEB, MEM, nor RAP will be considered further.
+and neither MEB, MEM, nor RAP will be \textcolor{blue}{given much attention}.
 Focus is henceforth on instances in which some of the nodes neither initiate new messages nor need messages from others.
@@
 
-\textcolor{blue}{\subsection{Structure of the paper}}
-
 The remainder of this paper is organized as follows: 
 \textcolor{blue}{Section \ref{sec:literature} summarizes relevant literature and previous works.}
-%Section \ref{sec:SBT} describes the SMT problem and gives detailed explanation of its objective function.
-\textcolor{blue}{Section \ref{sec:SBT} divided into three subsections introduces the notation, explains the network model, and gives formal definition of the problem.}
+\textcolor{blue}{Section \ref{sec:SBT} introduces the notation, explains the network model, and gives a formal definition of the problem.}
 Integer linear programming formulations, valid inequalities and their analysis are presented in Section \ref{sec:ILP},
@@
 and a constraint generation procedure is developed.
+\textcolor{blue}{A fast method for computing good, but not necessarily optimal solutions is given in Section \ref{sec:heur}.}
 Computational results are reported in Section \ref{sec:exp},

### 2018-09-24 smt/sec5-relations.tex

\section{Relations Between the Models}
\label{sec:comp}

In this section, we analyze and compare the lower bounds provided by optimal solutions with the continuous relaxations of the models formulated in Section \ref{sec:ILP}.
For any integer programming models $\mathcal{M}$ and $\mathcal{M}'$ of SMT, we say that $\text{LP}(\mathcal{M})$ is \emph{at least as strong as} $\text{LP}(\mathcal{M}')$ if
$z\left(\text{LP}(\mathcal{M})\right)\geq z\left(\text{LP}(\mathcal{M}')\right)$ for all SMT-instances,
and that it is \emph{stronger} if there also exists an instance in which the inequality is strict.

\subsection{Comparing $\mathcal{X}_1$ and $\mathcal{F}_1$}
For all $(i,j)\in A$, define $F_{ij}^*=\max_{s\in D_0}F_{ij}^s$.
To prove that $\text{LP}(\mathcal{F}_1)$ is at least as strong as $\text{LP}(\mathcal{X}_1)$, the following claim is instrumental.

\begin{lemma} \label{lem:lpf1feas}
All feasible solutions $(F,g,\pi)$ to $\text{LP}(\mathcal{F}_1)$ satisfy $\sum_{k\in V_i}g_{ki}=1$ ($i\in D_0$) and $\sum_{k\in V_i}g_{ki}\leq 1$ ($i\in V\setminus D$).
\end{lemma}
\begin{proof}
Let $i\in D_0$. Utilizing first (\ref{con:pf1:fitt=xit}), next flow conservation (\ref{con:pf1:flow}) at $i=s$, and finally (\ref{con:pf1:noflowFromT}), we get
$\sum_{j\in V_i}g_{ji}=\sum_{j\in V_i}F_{ji}^i = 1+\sum_{j\in V_i}F_{ij}^i=1.$
For $i\in V\setminus D$, the claim is identical to (\ref{con:pf1:B}). \qed
\end{proof}

\begin{lemma} \label{lem:lpf1opt}
In all SMT-instances, there exists an optimal solution $(F,g,\pi)$ to $\text{LP}(\mathcal{F}_1)$ satisfying
\begin{subequations}
\begin{align}
 & \min\{F_{ij}^s, F_{ji}^s\} = 0 & \{i,j\}\in E, s\in D_0, \text{and} \label{eq:lemlpf1a} \\
 & g_{ij} = F_{ij}^* & (i,j)\in A. \label{eq:lemlpf1b}
\end{align}
\end{subequations}
\end{lemma}

\begin{proof}
Obviously, an optimal solution to $\text{LP}(\mathcal{F}_1)$ exists, so let $(\bar{F},\bar{g},\pi)$ denote one such solution.
For all $(i,j)\in A$ and $s\in D_0$, let $F_{ij}^s=\bar{F}_{ij}^s-\min\{\bar{F}_{ij}^s, \bar{F}_{ji}^s\}$, and let $g$ be chosen according to (\ref{eq:lemlpf1b}).
Consequently, (\ref{eq:lemlpf1a})--(\ref{eq:lemlpf1b}) are satisfied, and the objective function values of $(F,g,\pi)$ and $(\bar{F},\bar{g},\pi)$ are equal.

To prove optimality of $(F,g,\pi)$, it suffices to prove feasibility.
The capacity constraints (\ref{con:pf1:xfrel}) follow directly from (\ref{eq:lemlpf1b}).
Because $\bar{F}$ satisfies the flow conservation constraints (\ref{con:pf1:flow}), also $F$ does, since
$F_{ji}^s-F_{ij}^s=\bar{F}_{ji}^s-\bar{F}_{ij}^s$.
Further, $g_{ij}=F_{ij}^*\leq\bar{F}_{ij}^*\leq\bar{g}_{ij}$, proving that also (\ref{con:pf1:B}) and (\ref{con:pf1:yvar})--(\ref{con:pf1:yvar0}) are satisfied.
Finally, (\ref{con:pf1:noflowFromT})--(\ref{con:pf1:fitt=xit}) follow directly from $\bar{F}_{si}^s=0$ and $\bar{F}_{is}^s=\bar{g}_{is}$,
and (\ref{con:pf1:xi0=0}) follow from $g_{i0}\leq \bar{g}_{i0}=0$.
Thus, $(F,g,\pi)$ is feasible in $\text{LP}(\mathcal{F}_1)$.\qed
\end{proof}

\begin{lemma} \label{lem:lpf1gijgji}
If $(F,g,\pi)$ is feasible in $\text{LP}(\mathcal{F}_1)$ with (\ref{eq:lemlpf1a})--(\ref{eq:lemlpf1b}) satisfied, then
\begin{equation}
\begin{aligned}
 & g_{ij}+g_{ji} \leq 1 & \{i,j\}\in E. \label{eq:lemlpf1c}
\end{aligned}
\end{equation}
\end{lemma}
\begin{proof}
Assume $g_{ij}+g_{ji}>1$ for some $\{i,j\}\in E$.
Because $g_{i0}=0$, this implies $i\neq s_0\neq j$.
By (\ref{eq:lemlpf1b}), there exist $s,t\in D_0$ such that $F_{ij}^s+F_{ji}^t>1$.
According to (\ref{eq:lemlpf1a}), $F_{ij}^s>1-F_{ji}^t\geq 0$ shows that $F_{ji}^s=0$.
Flow conservation at node $i$ thus implies $1<F_{ij}^s+F_{ji}^t\leq\sum_{k\in V_i\setminus\{j\}}F_{ki}^s+F_{ji}^t
\leq\sum_{k\in V_i\setminus\{j\}}g_{ki}+g_{ji}$, contradicting Lemma \ref{lem:lpf1feas}.\qed
\end{proof}

\begin{prop}
\label{prop:f1strx1}
$\text{LP}(\mathcal{F}_1)$ is at least as strong as $\text{LP}(\mathcal{X}_1)$. 
\end{prop}
\begin{proof}
Consider an optimal solution $(F,g,\pi)$ to $\text{LP}(\mathcal{F}_1)$ satisfying the conditions of Lemma \ref{lem:lpf1opt}.
We prove that there exist corresponding $X\in[0,1]^{A\times D}$ and $y\in[0,1]^E$ such that $(X,y,\pi)$ is a feasible solution to $\text{LP}(\mathcal{X}_1)$.
Let
\begin{subequations}
\begin{align}
 & X_{ij}^0=g_{ij} &  (i,j)\in A, \label{eq:trangX0}\\
 & X_{ij}^s=g_{ij}+F_{ji}^s-F_{ij}^s &  (i,j)\in A, s\in D_0, \label{eq:trangX}\\
 & y_{ij}=g_{ij}+g_{ji} &  \{i,j\}\in E. \label{eq:trangy}
\end{align}
\end{subequations}

\noindent
For distinct destinations $i,s\in D_0$, Lemma \ref{lem:lpf1feas} yields
$$\sum_{j\in V_i}X_{ji}^s=\sum_{j\in V_i}\left(g_{ji}+F_{ij}^s-F_{ji}^s\right)=\sum_{j\in V_i}g_{ji}=1,$$
proving that (\ref{con:dd:arrowFromDest}) is satisfied for $i\neq s_0\neq s$. 
By applying (\ref{con:pf1:xi0=0}), we get for $i=s_0\neq s$, that
$\sum_{j\in V_i}X_{ji}^s$ $=\sum_{j\in V_0}\left(g_{j0}+F_{0j}^s-F_{j0}^s\right)$ $=\sum_{j\in V_0}F_{0j}^s=1$,
where the latter identity is obtained by summing all flow conservation constraints (\ref{con:pf1:flow}) in which variables are superscripted by destination $s$.
Likewise, for $i\neq s_0=s$, we get 
$\sum_{j\in V_i}X_{ji}^s=\sum_{j\in V_i}g_{ji}=1$.
Hence, $X$ satisfies (\ref{con:dd:arrowFromDest}).

Also for $i\in V\setminus D$, we have
$\sum_{j\in V_i}X_{ji}^s=\sum_{j\in V_i}g_{ji}$, and constraints (\ref{con:dd:arrowFromNonDestB}) follow from (\ref{con:pf1:B}).
Constraints (\ref{con:dd:oneDir}) follow directly from the choice of values of $X$ and $y$,
and (\ref{con:dd:startInSource}) follows from (\ref{con:pf1:noflowFromT})--(\ref{con:pf1:fitt=xit}).
Further, $y_{ij}\in[0,1]$ holds according to Lemma \ref{lem:lpf1gijgji}, and $X_{ij}^s\in[0,1]$ follows from the capacity constraint (\ref{con:pf1:xfrel}).

It remains to prove (\ref{con:dd:arrowFromNonDestA}).
First, flow conservation at node $i$ is used to show
\[
  X_{ij}^s - \sum_{k\in V_{i}\setminus\{j\}}X^s_{ki} = g_{ij}+F_{ji}^s-F_{ij}^s - \sum_{k\in V_{i}\setminus\{j\}}\left(g_{ki}+F_{ik}^s-F_{ki}^s\right)
  = g_{ij}-\sum_{k\in V_{i}\setminus\{j\}}g_{ki}.
\]
It follows from condition (\ref{eq:lemlpf1b}) in Lemma \ref{lem:lpf1opt} that 
for some $t\in D_0$,
\[
  X_{ij}^s - \sum_{k\in V_{i}\setminus\{j\}}X^s_{ki}
  = F_{ij}^{t}-\sum_{k\in V_{i}\setminus\{j\}}g_{ki}
  \leq F_{ij}^{t}-\sum_{k\in V_{i}\setminus\{j\}}F_{ki}^{t}.
\]
If $F_{ij}^{t}=0$, the right hand side of the inequality is obviously non-positive.
To prove the same when $F_{ij}^{t}>0$, observe that (\ref{eq:lemlpf1a}) implies $F_{ji}^{t}=0$, which yields
\[
  F_{ij}^{t}-\sum_{k\in V_{i}\setminus\{j\}}F_{ki}^{t}
  = F_{ij}^{t}-\sum_{k\in V_{i}}F_{ki}^{t}
  \leq \sum_{k\in V_i}F_{ik}^{t}-\sum_{k\in V_{i}}F_{ki}^{t} = 0.
\]

Since the objective function values of $(F,g,\pi)$ and $(X,y,\pi)$ are identical in their respective relaxations, the proof is complete.
\qed\end{proof}

\begin{remark}
\label{rem:f1strx1}
Section \ref{sec:expcomplp} reports instances proving that $\text{LP}(\mathcal{F}_1)$ is \emph{stronger than} $\text{LP}(\mathcal{X}_1)$. 
\end{remark}

\subsection{Comparing $\mathcal{X}_2$ and $\mathcal{F}_2$}
For all $(i,j)\in A$ where $i,j\in V\setminus D$, let
$$\sigma_{ij}(F,g) = \min\left\{g_{ij} - F^*_{ij}, \sum_{k\in V_{i}}\left(g_{ik}-g_{ki}\right)\right\}.$$
That is, $\sigma_{ij}(F,g)$ is the smallest among slacks in constraints (\ref{con:pf1:xfrel}) at arc $(i,j)$ and constraint (\ref{con:pf1:flowX}) at node $i$.

\begin{lemma} \label{lem:lpf2}
In all SMT-instances, there exists an optimal solution $(F,g,\pi)$ to $\text{LP}(\mathcal{F}_2)$ satisfying (\ref{eq:lemlpf1a}),
\begin{subequations}
\begin{align}
 & \min\left\{g_{ij} - F_{ij}^*, g_{ji} - F_{ji}^*\right\} = 0 & \{i,j\}\in E, & \label{eq:lemlpf2a} \\
 & g_{ij} = F_{ij}^* & (i,j)\in A: \{i,j\}\cap D\neq\emptyset, & ~\text{and} \label{eq:lemlpf2b} \\
 & \sigma_{ij}(F,g)=0 & (i,j)\in A: \{i,j\}\subseteq V\setminus D. & \label{eq:lemlpf2c}
\end{align}
\end{subequations}
\end{lemma}

\begin{proof}
By following the initial step of the proof of Lemma \ref{lem:lpf1opt},
it is shown that $\text{LP}(\mathcal{F}_2)$ has an optimal solution $(F,\bar{g},\pi)$ satisfying (\ref{eq:lemlpf1a}).
For all $(i,j)\in A$, let $\delta_{ij}=\min\left\{\bar{g}_{ij}-F_{ij}^*, \bar{g}_{ji}-F_{ji}^*\right\}$,
and let $g_{ij}=\bar{g}_{ij}-\delta_{ij}$.
Hence, $(F,g)$ satisfies (\ref{eq:lemlpf2a}).
By exploiting $g\leq\bar{g}$, $g_{ij}-g_{ji}=\bar{g}_{ij}-\bar{g}_{ji}$, and feasibility of $(F,\bar{g},\pi)$,
it is seen straightforwardly that also $(F,g,\pi)$ is feasible.

For $j\in D_0$ (for $j=s_0$), (\ref{eq:lemlpf2b}) follows directly from (\ref{con:pf1:fitt=xit}) (from (\ref{con:pf1:xi0=0})).
Consider an $i\in D$ and $j\in V_i$ such that $g_{ij}>F_{ij}^*$.
Reducing $g_{ij}$ to $F_{ij}^*$, while keeping all other variables unchanged, preserves feasibility and the objective function value.
Thus, $(F,g)$ is henceforth assumed to satisfy (\ref{eq:lemlpf2b}).

To prove that also (\ref{eq:lemlpf2c}) can be satisfied,
consider the \emph{slack-induced digraph} $H[F,g]=(V\setminus D,A[F,g])$ of non-destination nodes, where
% $A_H=\left\{(i,j)\in A:g_{ij} = g_{ji} + \sum_{k\in V_{i}\setminus\{j\}}\left(g_{ki}-g_{ik}\right) > F_{ij}^*\right\}$.
$A[F,g]=\left\{(i,j)\in A: i,j\in V\setminus D, F_{ij}^* < g_{ij}\right\}$.
Assume $C=(i_0,\ldots,i_{c-1}, i_c)$ is a cycle of length $c$ in $H$ ($i_c=i_0$),
and let $\varepsilon = \min\left\{g_{i_li_{l+1}}-F_{i_li_{l+1}}^*: l=0,\ldots,c-1\right\}>0$.
Define $g_{i_li_{l+1}}'=g_{i_li_{l+1}}-\varepsilon$ ($l=0,\ldots,c-1$), and $g_{ij}'=g_{ij}$ for arcs $(i,j)$ not in $C$.
Because $\sum_{j\in V_i}g_{ji}'-\sum_{j\in V_i}g_{ij}' = \sum_{j\in V_i}g_{ji}-\sum_{j\in V_i}g_{ij}$,
$g'$ satisfies (\ref{con:pf1:flowX}).
Since $g'\leq g$, it follows that $(F,g',\pi)$ is feasible in $\text{LP}(\mathcal{F}_2)$, with the same objective function value as $(F,g,\pi)$.
% Let $H'=(V,A_{H'})$ be defined analogously to $H$, but with $g$ replaced by $g'$.
Then, $A[F,g']\subset A[F,g]$, and for some $l=0,\ldots,c-1$, $(i_l,i_{l+1})\in A[F,g]\setminus A[F,g']$.
An induction argument hence proves the existence of an optimal solution to $\text{LP}(\mathcal{F}_2)$ for which the corresponding slack-induced digraph is acyclic.
It is henceforth assumed that $(F,g,\pi)$ is such a solution.

Assume arc $(j_0,j_1)$ violates (\ref{eq:lemlpf2c}).
We show that the capacity can be reduced to $\tilde{g}\leq g$ such that
\begin{itemize}
  \item $(F,\tilde{g},\pi)$ is feasible in $\text{LP}(\mathcal{F}_2)$,
  \item $\sigma_{ij}(F,\tilde{g})>0$ only for arcs $(i,j)$ where $\sigma_{ij}(F,g)>0$, and
  \item either $\sigma_{j_0j_1}(F,\tilde{g})=0$ or $A[F,\tilde{g}]\subsetneq A[F,g]$.
\end{itemize}
\noindent
Finitely many reductions, and replacements of $g$ by $\tilde{g}$, are then sufficient to achieve $\sigma_{j_0j_1}(F,g)=0$, which completes the proof.

Let $(j_1,\ldots,j_p)$ be a path with $p\geq 1$ nodes in $H[F,g]$, such that $j_p$ has no out-neighbors in $H[F,g]$.
Node $j_p$ exists since the digraph is acyclic.
Define the maximum capacity reduction as
\[ \sigma^*=\min\left\{\sigma_{j_0j_1}(F,g),\min\left\{g_{j_l,j_{l+1}}-F^*_{j_l,j_{l+1}}: l=1,\ldots,p-1\right\}\right\}>0.\]
Let $\tilde{g}_{j_lj_{l+1}}=g_{j_lj_{l+1}}-\sigma^*$ for $l=0,1,\ldots,p-1$, and let $\tilde{g}_{ij}=g_{ij}$ for all other arcs $(i,j)\in A$.
Then $(F,\tilde{g},\pi)$ is feasible since, in comparison to $(F,g,\pi)$,  the left hand side of inequality (\ref{con:pf1:flowX}) is
increased by no more than the current slack at node $i=j_0$, reduced at node $i=j_p$, and unchanged at all other nodes $i\in V\setminus D$.
For all arcs $(i,j)$ between non-destination nodes, $\sigma_{ij}(F,\tilde{g})\leq\sigma_{ij}(F,g)$, except from arcs for which $i=j_p$.
But since $(j_p,j)\not\in A[F,g]$, $\sigma_{j_pj}(F,g)>0$.
% and the capacity reduction introduces no new violations of (\ref{eq:lemlpf2a})--(\ref{eq:lemlpf2c}).
It is finally observed that $\sigma_{j_0j_1}(F,\tilde{g})=0$ if $\sigma^*=\sigma_{j_0j_1}(F,g)$,
and that $\left(j_l,j_{l+1}\right)\in A[F,g]\setminus A[F,\tilde{g}]$ if $\sigma^*=g_{j_l,j_{l+1}}-F^*_{j_l,j_{l+1}}$ for some $l=1,\ldots,p-1$.
\qed\end{proof}

\begin{lemma} \label{lem:lpf2gijgji}
If $(F,g,\pi)$ is a feasible solution to $\text{LP}(\mathcal{F}_2)$ satisfying the conditions in Lemma \ref{lem:lpf2}, then (\ref{eq:lemlpf1c}) is satisfied.
Moreover,
\begin{equation}
\begin{aligned}
 & g_{ij} \leq \sum_{k\in V_i\setminus\{j\}}g_{ki} & (i,j)\in A. \label{eq:lemlpf2d}
\end{aligned}
\end{equation}
\end{lemma}

\begin{proof}
To prove (\ref{eq:lemlpf2d}), we only consider arcs $(i,j)$ where $g_{ij}>0$, as the inequality to be proved obviously holds otherwise.

Assume first that $g_{ij}=F_{ij}^*$.
Then there exists a destination node $t\in D_0\setminus\{i\}$ such that $F_{ij}^t=g_{ij}$.
Because (\ref{eq:lemlpf1a}) gives $F_{ji}^t=0$, flow conservation (\ref{con:pf1:flow}) at node $i$ implies 
$g_{ij}=F_{ij}^t\leq \sum_{k\in V_i\setminus\{j\}}F_{ki}^t\leq \sum_{k\in V_i\setminus\{j\}}g_{ki}$.

Assume next that $g_{ij}>F_{ij}^*$.
Conditions (\ref{eq:lemlpf2a}) and (\ref{eq:lemlpf2b}) of Lemma \ref{lem:lpf2} imply, respectively,
$g_{ji}=F_{ji}^*$ and $i,j\in V\setminus D$.
Because $g_{ij}>F_{ij}^*$, 
condition (\ref{eq:lemlpf2c}) of Lemma \ref{lem:lpf2} gives
$0=\sigma_{ij}(F,g) = \sum_{k\in V_{i}}\left(g_{ik}-g_{ki}\right)$. Thus,
\begin{equation}
  g_{ij} = g_{ji} + \sum_{k\in V_{i}\setminus\{j\}}\left(g_{ki}-g_{ik}\right). \label{eq:lemlpf2f}
\end{equation}
If $g_{ji}=0$, (\ref{eq:lemlpf2d}) follows directly from (\ref{eq:lemlpf2f}).
Otherwise, there exists a $t\in D_0$ such that $g_{ji}=F_{ji}^t$ and $F_{ij}^t=0$.
Flow conservation at $i$ then shows $g_{ji}=F_{ji}^t\leq\sum_{k\in V_{i}\setminus\{j\}}F_{ik}^t\leq\sum_{k\in V_{i}}g_{ik}$.
Adding this inequality to (\ref{eq:lemlpf2f}) proves (\ref{eq:lemlpf2d}).
% $g_{ij} \leq \sum_{k\in V_i\setminus\{j\}}g_{ik} + \sum_{k\in V_{i}\setminus\{j\}}\left(g_{ki}-g_{ik}\right) = \sum_{k\in V_{i}\setminus\{j\}}g_{ki}$.

To prove (\ref{eq:lemlpf1c}), assume $g_{ij}+g_{ji}>1$ for some $\{i,j\}\in E$.
By (\ref{eq:lemlpf2d}), this implies $1<g_{ij}+g_{ji}\leq\sum_{k\in V_{i}}g_{ki}$, contradicting Lemma \ref{lem:lpf1feas}.
\qed\end{proof}

\begin{prop}
\label{prop:f2strx2}
$\text{LP}(\mathcal{F}_2)$ is at least as strong as $\text{LP}(\mathcal{X}_2)$. 
\end{prop}
\begin{proof}
Consider an optimal solution $(F,g,\pi)$ to $\text{LP}(\mathcal{F}_2)$ satisfying the conditions of Lemma \ref{lem:lpf2}.
We prove that $(X,y,\pi)$ as defined by (\ref{eq:trangX0})--(\ref{eq:trangy}) is a feasible solution to $\text{LP}(\mathcal{X}_2)$.

The proof that $(X,y,\pi)$ satisfies constraints (\ref{con:dd:arrowFromDest})--(\ref{con:dd:arrowFromNonDestB}), (\ref{con:dd:oneDir})--(\ref{con:dd:startInSource}),
and (\ref{con:dd:yvar}),
follows the proof of Prop.\ \ref{prop:f1strx1}.
To prove (\ref{con:dd:arrowFromNonDestA}), let $i\in V\setminus D$, $j\in V_i$, and $s\in D$.
Then transformations (\ref{eq:trangX0})--(\ref{eq:trangX}) yield
$X_{ij}^s - \sum_{k\in V_{i}\setminus \{j\}}X_{ki}^s
= g_{ij} - \sum_{k\in V_{i}\setminus \{j\}}g_{ki}$, and (\ref{con:dd:arrowFromNonDestA}) follows from (\ref{eq:lemlpf2d}) in Lemma \ref{lem:lpf2gijgji}.
Because Lemma \ref{lem:lpf2gijgji} also states that $g_{ij}+g_{ji}\leq 1$, we further have that $X_{ij}^s\in[0,1]$ and $y_{ij}\in[0,1]$.

Constraints (\ref{con:dd:extraCon}) hold because
$\sum_{j\in V_i}\left(X_{ji}^s-X_{ij}^s\right)=\sum_{j\in V_i}\left(g_{ji}-g_{ij}\right)\leq 0$
according to constraints (\ref{con:pf1:flowX}).
Finally, (\ref{con:vi:sumYImpSumXTrans}) implies
$\sum_{j\in V_i}X_{ji}^s-\sum_{j\in V_i\setminus\{s\}}\pi_{ij}^s=\sum_{j\in V_i}g_{ji}-\sum_{j\in V_i\setminus\{s\}}\pi_{ij}^s=0$,
which proves that $(X,\pi)$ satisfies (\ref{con:vi:sumYImpSumX}).
\qed\end{proof}

\begin{remark}
\label{rem:f2strx2}
Section \ref{sec:expcomplp} reports instances proving that $\text{LP}(\mathcal{F}_2)$ is \emph{stronger than} $\text{LP}(\mathcal{X}_2)$. 
\end{remark}

### 2018-09-24 smt/sec7-heur.tex

@@
-\textcolor{blue}{\section{Computing near-optimal solutions}} \label{sec:heur}
+\textcolor{blue}{\section{Computing near-optimal solutions} \label{sec:heur}}
 
@@
 For the purpose of diversification, the factor by which the cost is multiplied is chosen randomly.
-To discourage the appearence of edges outside both $T$ and $T_r$ in the merge $\tilde{T}$,
+To discourage the appearance of edges outside both $T$ and $T_r$ in the merge $\tilde{T}$,
 their cost is magnified by a factor dominating the factors drawn for edges that occur in one of the trees.
@@
 Rather than selecting the tree to leave $\mathcal{T}$ randomly, we consistently remove $T_{\gamma}$ when a new tree is added.
-Also, for each new tre $T$ constructed, Alg.\ \ref{alg:meta} makes only one attempt to merge it with some $T_r$ stored in $\mathcal{T}$.
+Also, for each new tree $T$ constructed, Alg.\ \ref{alg:meta} makes only one attempt to merge it with some $T_r$ stored in $\mathcal{T}$.
 This contrasts the method in the cited work, where $\tilde{T}$ is chosen as the best tree found in a sequence of merges.

### 2018-09-26 smt/response.tex
MI (SMT): Minor changes in several files, mainly in Sect 8 - description of the results.

@@
 We fully agree that there are gaps between the coarse energy model applied in our work and more detailed energy models for energy consumption is sensor networks,
-e.g.\ those analyzed by Halgamuge et al.\ \cite{halgamuge}.
+e.g.\ those analyzed by Halgamuge et al. (2009).
 These gaps, and their possible implications on practical usability of our approaches, are now discussed in the new Section 3.2.
@@
 Newly added parts of the paper are also in \textcolor{blue}{blue} font.
-In several instances the text is restructured, but the sentences are not altered. In these cases the font color remains black.
+In cases when the text is restructured, but the sentences and expressions are not altered, the font color remains black.
 Such updates should be apparent due to new highlighted sections.

### 2018-09-26 smt/sec7-heur.tex
MI (SMT): Minor changes in several files, mainly in Sect 8 - description of the results.

@@
 Diversity among the trees is achieved by perturbing the edge costs randomly before construction and local improvement are applied.
-After each tree generation, the new tree is merged with a randomly chosen tree from $\mathcal{T}$, and the resulting tree is kept in $\mathcal{T}$ if its cost is small.}
+After each tree generation, the new tree is merged with a randomly chosen tree from $\mathcal{T}$, and the resulting tree is kept in $\mathcal{T}$ if its cost is small.
+%After each tree generation, the new tree is merged with a randomly chosen tree from $\mathcal{T}$, 
+%and the resulting tree is kept in $\mathcal{T}$ if its cost is smaller than some other tree already in $\mathcal{T}$.
+%If there is no empty capacity in $\mathcal{T}$, its most expensive tree is discarded.
+}
 

### 2018-09-26 smt/sec8-exp.tex
MI (SMT): Minor changes in several files, mainly in Sect 8 - description of the results.

@@
 All experiments are run on an Intel Core 2 Quad CPU at 2.83 GHz and 8 GB RAM.
-The models are implemented in Java with Concert technology and solved using CPLEX 12.5.1 optimizer.
+The models are implemented in Java with Concert technology and solved using Cplex 12.5.1 optimizer.
 
@@
 Small instances are often solved to optimality relatively quickly with B\&B.
-We investigate what is the maximum size of instances that can be solved to optimality within 20 minutes using models $\mathcal{X}_1$ and $\mathcal{F}_1$.
+We investigate what is the maximum size of instances that can be solved to optimality within 20 CPU-minutes using models $\mathcal{X}_1$ and $\mathcal{F}_1$.
 The experiments compare sets of instances with different $|V|/|D|$ ratios.
@@
 \textcolor{blue}{\subsection{Comparison with alternative methods}}
-
 \textcolor{blue}{
-To the best of our knowledge, SMT has not been studied extensively in terms of inexact methods.
-However, there is a manifold of a thorough research of the minimum Steiner tree problem, with the possibility of adapting the methods on SMT.
-We turn our attention to recently published paper \citet{pajor18} which contains a metaheuristic algorithm for a fast generation of good solution to the Steiner tree problem.
-The method is general, and is therefore very suitable for adaptation to our purposes.  The high level description is as follows:}
-
-\textcolor{blue}{The construction procedure and local search is the adapted BIP algorithm together with refinements presented in \citet{ivanova16isco}.
-The remaining parts of the algorithm are carried out as discussed in \citet{pajor18}.
-The randomization is achieved by perturbing the edge costs in the input graph.
-The obtained solution $S$ is evaluated with respect to the real (original) edge costs and added to the pool $P$ if the pool capacity is not filled.
-If the pool is full and its worst solution $\hat{S}$ is worse than $S$, $\hat{S}$ is replaced by $S$ in $P$.
-Then a new solution $\tilde{S}$ is constructed by combining some solution in $P$ with $S$. 
-First, we adjust edge costs resulting in a modified input instance, 
-and then run the construction and local search yielding $\tilde{S}$.
-This new solution is again added to the pool if it is better than some solution already in $P$.
-The adjustment of the edges tries to promote edges that appear in both $S$ and a selected solution in $P$. 
-In that case, the original cost is preserved.
-If an edge is present in only one of the two solutions, its cost is multiplied by a random integer between 100 and 500.
-If an edge appears in neither of the solutions, its edge is multiplied by 1000, which makes it very unlikely to appear in the result $\tilde{S}$ of the combination.
+We now investigate the ability of IP models $\mathcal{X}_1$ and $\mathcal{F}_1$ to produce a feasible solution when a time limit is imposed.
+We compare the quality of obtained solutions with those calculated by Alg. \ref{alg:meta}.
+The results for time restrictions 5, 10 and 20 CPU-minutes are reported in Tab. \ref{tab:metamod}.
+These experiments are carried out on the same instance set that is used in experiments presented in Tab. \ref{tab:small_inst}. 
 }
@@
 \hline
-18  & 12   &100.0&100.0&101.8&100.0&100.0&100.0&100.0&100.0&100.0\\
-21  & 14   &106.5&107.4&101.2&102.7&103.1&100.1&100.9&100.3&100.0\\
-24  & 16   &144.2&112.2&101.3&124.0&114.0&100.1&104.3&112.7&100.0\\     
+18  & 12   &100.1&100.1&100.0&100.0&100.0&100.0&100.0&100.0&100.0\\
+21  & 14   &107.9&108.8&100.0&102.7&103.2&100.0&101.2&100.7&100.0\\
+24  & 16   &153.4&115.6&100.0&126.2&114.2&100.0&104.3&112.7&100.0\\     
 \hline
-20  & 10   &100.6&100.0&101.4&100.1&100.0&100.0&100.1&100.0&100.0\\
-22  & 11   &102.1&103.2&101.7&101.3&100.7&100.1&100.7&100.2&100.1\\
-24  & 12   &106.6&111.7&101.7&105.0&105.5&100.0&103.1&106.2&100.0\\                               
+20  & 10   &100.6&100.1&100.0&100.1&100.0&100.0&100.1&100.0&100.0\\
+22  & 11   &102.8&103.9&100.0&101.3&100.7&100.1&100.7&100.2&100.1\\
+24  & 12   &108.2&113.4&100.0&105.0&105.5&100.0&103.1&106.2&100.0\\                               
 \hline
 21  & 7    &100.1&100.0&100.0&100.0&100.0&100.0&100.0&100.0&100.0\\
-24  & 8    &102.3&101.4&102.0&102.1&100.7&100.1&101.1&100.1&100.0\\
-27  & 9    &127.4&114.0&101.6&111.2&109.0&100.1&108.2&104.1&100.0                                  
+24  & 8    &102.8&101.9&100.0&102.1&100.7&100.0&101.1&100.1&100.0\\
+27  & 9    &133.2&118.7&100.0&111.3&109.1&100.0&108.2&104.1&100.0                                  
 \end{tabular}
@@
 \textcolor{blue}{
-Tab.\ref{tab:metamod} summarizes the comparison of the metaheuristic algorithm with the ability of IP models to produce feasible solutions when restricted time limit.
-Each entry represents the average percentage  of the objective value yielded by the most successful method in a given instance.
+Each entry contains the average percentage  of the objective value yielded by the most successful method in a given instance type.
 For example, if all methods find the optimum in every instance, the three corresponding values are all 100.0.
-Thus, the lowest value in a given instance type indicates that the corresponding method performed the most successfully. 
-The averages are calculated over all instances of given $V$ and $D$.
-Experiments with results in Tab. \ref{tab:small_inst} and Tab. \ref{tab:metamod} are carried out on the same sets of instances.
+Thus, the lowest value in a given instance type indicates that the corresponding method performs the most successfully. 
+The averages are calculated over all instances of given $|V|$ and $|D|$ (25 instances for each type).
 }
@@
 \textcolor{blue}{
-The resulting values reveal that the metaheuristic algorithm converges quickly, and is often able to find optimal solutions faster than solving the IP models.
+The resulting values reveal that the metaheuristic algorithm converges quickly, and is often able to find optimal solutions faster than Cplex applied to the IP models.
+The smallest 5 CPU-minutes time limit suffices to find a good quality solutions in our test scenarios, and increasing the time limit rarely improves the outcome.
 The solutions are, however, not guaranteed to be optimal.
-The optimality is proven only on smaller instances, where the B\&B finds the optimum before the time limit is exceeded.
-It is likely that the inexact method run on large instances does not find the optimum, but it very clearly outperforms B\&B, 
-where the solutions found within the given time limit are far from the optimum.
+The optimality is proven only for smaller instances, where the B\&B finds the optimum before the time limit is exceeded.
+It is likely that the inexact method executed on large instances fails to find the optimum, but it very clearly outperforms B\&B applied to the IP models, 
+where the solutions found within the given time limit have far larger objective function value.
 }
+\newline
 
@@
 As the instance size increases, exact solution of the SMT problem becomes unrealistic.
-To assess the quality of fast and possible inexact methods \citep{ivanova16isco}, access to tightest possible lower bounds is crucial.
+To assess the quality of fast and possible inexact methods, \textcolor{blue}{such as metaheuristic Alg. \ref{alg:meta}, 
+or construction methods by \citep{ivanova16isco}},  access to tightest possible lower bounds is crucial.
 In the next set of experiments, we compare two different methods for computing the bounds.
@@
   \caption{Comparison of lower bounds obtained by CG and B\&B on model $\mathcal{F}_1$.
-	\textcolor{blue}{Each figure depicts a progress of lower bounds for instances with varying number of destination and non-destination nodes.
+	\textcolor{blue}{Each figure depicts the progress of lower bounds for instances with varying number of destination and non-destination nodes.
 	The lower bounds are obtained by solving CG scheme applied on model $\mathcal{F}_1$ (black curves), and produced during the course of classic B\&B algorithm (grey curves)
@@
 	which is happening until the curve reaches 100\%.
-	After that point the bound grows rather slowly.
+	After that point, the bound grows rather slowly.
 	CG on the other hand produces bound that grows faster even after the value of first relaxation in the root is reached.

### 2018-09-26 smt/response.tex
DH: Minor revision of the bibliography and the response letter.

--- 2018-09-24 smt/response.tex
@@
 For better justification of the numerical results illustrated graphically in the former Section 6 (now Section 8),
-we have added more explanations in the caption of Fig. 3.
+we have added more explanations both in the caption of Fig.\ 3 and in the text referring to the figure.
 }

### 2018-09-26 smt/ivanova.tex
DH: Added a section on the heuristic in a separate file. File names should be updated to become consistent with section numbers.
MI (SMT): Minor changes in several files, mainly in Sect 8 - description of the results.
DH: Minor revision of the bibliography and the response letter.

--- 2018-09-12 smt/ivanova.tex
@@
 \input{sec2-literature.tex}
-\input{sec2-sbtsmt.tex}
-\input{sec3-milp.tex}
-\input{sec4-relations.tex}
-\input{sec5-strong.tex}
-\input{sec6-exp.tex}
-\input{sec7-conclusion.tex}
+\input{sec3-sbtsmt.tex}
+\input{sec4-milp.tex}
+\input{sec5-relations.tex}
+\input{sec6-strong.tex}
+\input{sec7-heur.tex}
+\input{sec8-exp.tex}
+\input{sec9-conclusion.tex}
 
@@
 \bibitem[Guo and Yang(2004)]{guo}
-Guo S, Yang O (2004)
+\textcolor{blue}{Guo S, Yang O (2004)
 Minimum Energy Multicast Routing for Wireless Ad-hoc Networks with Adaptive Antennas 
 In: Proceedings of the 12th IEEE International Conference on Network Protocols ICNP 2004,
-pp 151--160
+pp 151--160}
 

### 2018-09-26 smt/response.tex

--- 2018-09-26 smt/response.tex
@@
 Following the recommendation of the reviewer, we have reorganized the literature review, and devoted a new section (Section 2 in the revised manuscript)
-to this topic. We have also included a review of more recent and relevant literature (references Bein and Zheng (2010), Bhukya and Singh (2014)).
+to this topic. We have also included a review of more recent and relevant literature (Bein and Zheng 2010; Bhukya and Singh 2014).
 In Section 1.2, we now state explicitly the contributions from the present work, and explain how they relate to the current literature.
@@
 We agree that a comparison with alternative contemporary methods largely benefits the paper.
-To that end, we have included a comparison with a recently published metaheuristic algorithm, based on iterative combination of elite solutions Pajor et al. (2018).
-The method is described in the new Section 7, and results from experiments with the method are presented in the new Section 8.3.
+To that end, we have included a comparison with a recently published metaheuristic algorithm,
+originally designed for the \textsc{Minimum Steiner Tree} problem (Pajor et al.\ 2018).
+An adaptation to our problem is described in the new Section 7, and results from experiments with the method are presented in the new Section 8.3.
 }
@@
 Justification is added in terms of references to  Halgamuge et al. (2009), 
-and to other relevant and frequently cited works mentioned in Section 2 (Wieselthier et al. (2000), Das et al. (2003), Altinkemer et al. (2005),
-Yuan (2005), Yuan et al. (2008), Bauer et al. (2008), Montemanni and Leggieri (2011) and Leggieri et al. (2008), etc.).
+and to other relevant and frequently cited works mentioned in Section 2 (Wieselthier et al.\ 2000; Das et al.\ 2003; Altinkemer et al.\ 2005;
+Yuan 2005; Yuan et al.\ 2008; Bauer et al.\ 2008; Montemanni and Leggieri 2011; Leggieri et al.\ 2008).
 See also our response to the second remark of Reviewer 3.
@@
 We fully agree that there are gaps between the coarse energy model applied in our work and more detailed energy models for energy consumption is sensor networks,
-e.g.\ those analyzed by Halgamuge et al. (2009).
+e.g.\ those analyzed by Halgamuge et al.\ (2009).
 These gaps, and their possible implications on practical usability of our approaches, are now discussed in the new Section 3.2.

### 2018-09-26 smt/ivanova.tex

@@
 \textcolor{blue}{Guo S, Yang O (2004)
+Guo S, Yang O (2004)
 Minimum Energy Multicast Routing for Wireless Ad-hoc Networks with Adaptive Antennas 

### 2018-09-26 smt/sec7-heur.tex

--- 2018-09-24 smt/sec7-heur.tex
@@
 Diversity among the trees is achieved by perturbing the edge costs randomly before construction and local improvement are applied.
-After each tree generation, the new tree is merged with a randomly chosen tree from $\mathcal{T}$, and the resulting tree is kept in $\mathcal{T}$ if its cost is small.}
+After each tree generation, the new tree is merged with a randomly chosen tree from $\mathcal{T}$, 
+and the resulting tree is kept in $\mathcal{T}$ if its cost is smaller than some other tree already in $\mathcal{T}$.
+If also $|\mathcal{T}|=\gamma$, the most expensive tree $T_{\gamma}$ is removed from $\mathcal{T}$.
+}
 
@@
 This contrasts the method in the cited work, where $\tilde{T}$ is chosen as the best tree found in a sequence of merges.
-}
+} \\

### 2018-09-26 smt/sec8-exp.tex

@@
 \textcolor{blue}{
-We now investigate the ability of IP models $\mathcal{X}_1$ and $\mathcal{F}_1$ to produce a feasible solution when a time limit is imposed.
+We now investigate the ability of IP models $\mathcal{X}_1$ and $\mathcal{F}_1$ to produce good feasible solutions when a time limit is imposed.
 We compare the quality of obtained solutions with those calculated by Alg. \ref{alg:meta}.
-The results for time restrictions 5, 10 and 20 CPU-minutes are reported in Tab. \ref{tab:metamod}.
-These experiments are carried out on the same instance set that is used in experiments presented in Tab. \ref{tab:small_inst}. 
+The results for time restrictions 5, 10 and 20 CPU-minutes, respectively, are reported in Tab. \ref{tab:metamod}.
+These experiments are carried out on the instance sets that are investigated in the experiments reported in Section \ref{sec:expsmall}.
 }
+
 \begin{table}[]
@@
 \multicolumn{1}{l}{}      & \multicolumn{1}{l}{}      & \multicolumn{3}{c}{5 min}  & \multicolumn{3}{c}{10 min} & \multicolumn{3}{c}{20 min}     \\\hline
-\multicolumn{1}{l}{$|V|$} & \multicolumn{1}{l}{$|D|$} & \multicolumn{1}{c}{$\mathcal{X}_1$} & \multicolumn{1}{c}{$\mathcal{F}_1$} & \multicolumn{1}{c}{$\mathcal{M}$} & \multicolumn{1}{c}{$\mathcal{X}_1$} & \multicolumn{1}{c}{$\mathcal{F}_1$} & \multicolumn{1}{c}{$\mathcal{M}$} & \multicolumn{1}{c}{$\mathcal{X}_1$} & \multicolumn{1}{c}{$\mathcal{F}_1$} & \multicolumn{1}{c}{$\mathcal{M}$} \\
+\multicolumn{1}{l}{$|V|$} & \multicolumn{1}{l}{$|D|$} & \multicolumn{1}{c}{$\mathcal{X}_1$} & \multicolumn{1}{c}{$\mathcal{F}_1$} & \multicolumn{1}{c}{Alg.\ \ref{alg:meta}} & \multicolumn{1}{c}{$\mathcal{X}_1$} & \multicolumn{1}{c}{$\mathcal{F}_1$} & \multicolumn{1}{c}{Alg.\ \ref{alg:meta}} & \multicolumn{1}{c}{$\mathcal{X}_1$} & \multicolumn{1}{c}{$\mathcal{F}_1$} & \multicolumn{1}{c}{Alg.\ \ref{alg:meta}} \\
 \hline
@@
 }
-\caption{\textcolor{blue}{Comparison of the metaheuristic algorithm and solving IP models with imposed time-limit.}}
+\caption{\textcolor{blue}{Comparison of the metaheuristic algorithm (Alg.\ \ref{alg:meta}) and interrupted B\&B applied to IP models $\mathcal{X}_1$ and $\mathcal{F}_1$}}
 \label{tab:metamod}
@@
 \textcolor{blue}{
-Each entry contains the average percentage  of the objective value yielded by the most successful method in a given instance type.
-For example, if all methods find the optimum in every instance, the three corresponding values are all 100.0.
-Thus, the lowest value in a given instance type indicates that the corresponding method performs the most successfully. 
-The averages are calculated over all instances of given $|V|$ and $|D|$ (25 instances for each type).
+For a given instance, we define the relative performance of a method as the objective function value obtained by the method relative to the best performing method in that instance.
+For each time bound imposed, and for each instance set,
+we report the relative performances of $\mathcal{X}_1$, $\mathcal{F}_1$, and Alg.\ \ref{alg:meta}, averaged over all instances in the set.
+For example, if all methods find the optimum in every instance, the three corresponding values are all 100.0\%.
+Thus, the lowest value in a given instance set indicates the best performing method.
 }
@@
 \textcolor{blue}{
-The resulting values reveal that the metaheuristic algorithm converges quickly, and is often able to find optimal solutions faster than Cplex applied to the IP models.
-The smallest 5 CPU-minutes time limit suffices to find a good quality solutions in our test scenarios, and increasing the time limit rarely improves the outcome.
-The solutions are, however, not guaranteed to be optimal.
-The optimality is proven only for smaller instances, where the B\&B finds the optimum before the time limit is exceeded.
-It is likely that the inexact method executed on large instances fails to find the optimum, but it very clearly outperforms B\&B applied to the IP models, 
-where the solutions found within the given time limit have far larger objective function value.
+Table \ref{tab:metamod} reveals that Alg.\ \ref{alg:meta} gives the best average results for all combinations of time limits and most instance set.
+Even when the time bound is set beyond what is needed to compute the optimal solution by submitting $\mathcal{X}_1$ or $\mathcal{F}_1$ to B\&B (see Tab.\ \ref{tab:small_inst}),
+this observation applies.
+Although Alg.\ \ref{alg:meta} fails to prove optimality, it appears to be a better procedure than B\&B for computing near-optimal solutions within a given time bound.
 }
@@
 As the instance size increases, exact solution of the SMT problem becomes unrealistic.
-To assess the quality of fast and possible inexact methods, \textcolor{blue}{such as metaheuristic Alg. \ref{alg:meta}, 
-or construction methods by \citep{ivanova16isco}},  access to tightest possible lower bounds is crucial.
+To assess the quality of fast and possible inexact methods, \textcolor{blue}{such as Alg.\ \ref{alg:meta}, 
+or the construction methods by \citep{ivanova16isco}},  access to tightest possible lower bounds is crucial.
 In the next set of experiments, we compare two different methods for computing the bounds.
@@
 The vertical axes correspond to the lower bounds as a percentage of $z\left(\text{LP}(\mathcal{F}_1)\right)$, averaged over all instances in the set.
-\textcolor{blue}{The values can be retrieved from the output of Cplex during the course of simplex and B\&B algorithm.}
+\textcolor{blue}{
+	The tightness of the bounds increases with time, and it can be observed from the graphs that the CG bound becomes tighter at some point.
+	We express the lower bounds relative to $z\left(\text{LP}(\mathcal{F}_1)\right)$, as indicated by the vertical axes.
+	The lower bound obtained from B\&B grows sharply while the dual simplex method is solving the root problem of the B\&B tree, 
+	which goes on until the curve reaches 100\%.
+	After that point, the bound grows rather slowly.
+	In contrast, CG produces a bound that grows faster even after it has reached the value $z\left(\text{LP}(\mathcal{X}_1)\right)$.}
 
@@
 	\textcolor{blue}{Each figure depicts the progress of lower bounds for instances with varying number of destination and non-destination nodes.
-	The lower bounds are obtained by solving CG scheme applied on model $\mathcal{F}_1$ (black curves), and produced during the course of classic B\&B algorithm (grey curves)
-	The tightness of the bounds increases with time, and it can be observed from the graphs that the CG bound becomes tighter at some point.
-	We express the quality of lower bounds with respect to the value of simple continuous relaxation, as indicated by the vertical axes.
-	It can also be noted that the lower bound obtained from B\&B grows sharply during the computation of the simplex method in the root of the B\&B tree, 
-	which is happening until the curve reaches 100\%.
-	After that point, the bound grows rather slowly.
-	CG on the other hand produces bound that grows faster even after the value of first relaxation in the root is reached.
-%	The values can be retrieved from the output of CPLEX during the course of simplex and B\&B algorithm.
+	The lower bounds are obtained by applying the CG scheme to model $\mathcal{F}_1$ (black curves), and by running the B\&B algorithm (grey curves), respectively.
 	}}

### 2018-09-26 smt/sec9-conclusion.tex

--- 2018-08-30 smt/sec7-conclusion.tex
@@
 A search for facet-defining valid inequalities, or even convex hull formulations, is an interesting and ambitious direction of future work.
-Owing to the limited size of solvable instances, there is a potential in studying inexact methods, such as approximation algorithms and domain specific heuristics.
+Owing to the limited size of solvable instances, there is a potential in studying inexact methods, such as approximation algorithms and problem specific heuristics.
+\textcolor{blue}{In particular, experiments on a simple metaheuristic documented in the current work, suggest that fast and inexact methods have a potential to compute solutions close to optimality in large instances.}
 Instance preprocessing combined with variable fixing can possibly increase the size of instances solvable to optimality with the current models.

### 2018-09-26 smt/sec6-strong.tex
MI: some typos fixed

--- 2018-08-30 smt/sec5-strong.tex
@@
 
-Aiming for stronger LP bounds than those provided by the relaxations of the models introduced in Section \ref{sec:ILP},
+Aiming for stronger LP bounds than those provided by the relaxations of the models introduced in Section\ \ref{sec:ILP},
 this section is devoted to extensions of $\mathcal{X}_2$ and $\mathcal{F}_2$.
@@
 %
-Variables $f_{ij}^{st}$ can also be regarded as a the \emph{joint flow} from $s_0$ to $s$ and $t$ \citep{Polzin}.
+Variables $f_{ij}^{st}$ can also be regarded as the \emph{joint flow} from $s_0$ to $s$ and $t$ \citep{Polzin}.
 In the cited work, they are successfully used to strengthen formulations for the \textsc{Minimum Steiner Tree} problem.

### 2018-10-11 smt/response.tex
DH: Submitted version of the first revision.

@@
 Newly added parts of the paper are also in \textcolor{blue}{blue} font.
-In cases when the text is restructured, but the sentences and expressions are not altered, the font color remains black.
-Such updates should be apparent due to new highlighted sections.
+In cases where the text is restructured, but the sentences and expressions are not altered, only the section headings are in blue.
 

### 2018-10-11 smt/ivanova.tex
MI: some typos fixed
DH: Submitted version of the first revision.

@@
               Department of Informatics, University of Bergen, Norway \\
-              Tel.: +47 942 77 070\\
+              Tel.: +47 94277070\\
               \email{Marika.Ivanova@uib.no}         %  \\
@@
 \bibitem[Halgamuge et al.(2009)]{halgamuge}
-Halgamuge MN, Zukerman M, Ramamohanarao K (2009)
+\textcolor{blue}{Halgamuge MN, Zukerman M, Ramamohanarao K (2009)
 An estimation of sensor energy consumption.
-Progress In Electromagnetics Research B 12:259--295
+Progress In Electromagnetics Research B 12:259--295}
 

### 2018-12-27 smt/ivanova.tex
MI: Remarks from minor revision incorporated - mostly typos and grammar.

--- 2018-09-26 smt/ivanova.tex
@@
 Networks 52(2):57–68
-\textcolor{blue}{\bibitem[Bein and Zheng(2010)]{bein10}
+
+\bibitem[Bein and Zheng(2010)]{bein10}
 Bein D, Zheng SQ (2010)
 Energy efficient all-to-all broadcast in all-wireless networks.
-Information Sciences 180(10):1781--1792}
-\textcolor{blue}{\bibitem[Bhukya and Singh(2014)]{bhukya14}
+Information Sciences 180(10):1781--1792
+
+\bibitem[Bhukya and Singh(2014)]{bhukya14}
 Bhukya WN, Singh A (2014)
@@
 In: Proceedings of Ninth International Conference on Wireless Communication and Sensor Networks 2014 
-Vol.\ 299, pp 61--69}
+Vol.\ 299, pp 61--69
 
@@
 \bibitem[Guo and Yang(2004)]{guo}
-\textcolor{blue}{Guo S, Yang O (2004)
+Guo S, Yang O (2004)
 Minimum Energy Multicast Routing for Wireless Ad-hoc Networks with Adaptive Antennas 
 In: Proceedings of the 12th IEEE International Conference on Network Protocols ICNP 2004,
-pp 151--160}
+pp 151--160
 
@@
 74:327--342
-\textcolor{blue}{\bibitem[Pajor et al.(2018)]{pajor18}
+
+\bibitem[Pajor et al.(2018)]{pajor18}
 Pajor T, Uchoa E, Werneck RF (2018)
@@
 Math Prog Comp 10:69--118
-}
 

### 2018-12-27 smt/sec1-intro.tex
MI: Remarks from minor revision incorporated - mostly typos and grammar.

--- 2018-09-06 smt/sec1-intro.tex
@@
 
-Wireless ad hoc networks (WANETs) are suitable in various practical applications in both civil and military sector, where a communication infrastructure is either absent or cannot be relied on.
+Wireless ad hoc networks (WANETs) are suitable in various practical applications in both civil and military sector, 
+where a communication infrastructure is either absent or cannot be relied on.
 Owing to their quick deployment and simple configuration, WANETs are useful in emergency operations such as natural disaster relief and military command and control.
-Recently, the WANET concept has been introduced in ad hoc smart lighting in homes and in offices as well as ad hoc street light networks, where the control includes adjusting dimmable lights.
+Recently, the WANET concept has been introduced in ad hoc smart lighting in homes and in offices, 
+as well as ad hoc street light networks, where the control includes adjusting dimmable lights.
 
@@
 
-Unlike wired networks, where signal passing takes place along predefined links, nodes in WANETs use omnidirectional antennas, and hence a message reaches all nodes within the communication range of its sender.
+Unlike wired networks, where signal passing takes place along predefined links, nodes in WANETs use omnidirectional antennas, 
+and hence a message reaches all nodes within the communication range of its sender.
 This range is determined by the power assigned to the sender, which is the maximum rather than the sum of the powers necessary to reach all intended receivers.
@@
 
-\textcolor{blue}{\subsection{Related problems}}
+\subsection{Related problems}
 
 A well-known and extensively studied problem in the network design literature is the \textsc{Minimum Energy Broadcast} (MEB) problem.
-Given a set of wireless devices with one designated source node, the goal is to assign powers to individual nodes.
+Given a set of wireless devices with one designated source node, the goal is to assign power to individual nodes.
 The power assignments determine the communication ranges, and induce a broadcast tree such that a signal initiated by the source reaches, directly or via other nodes,
@@
 Further, which power level to be set must be assessed by the relaying node.
-Such overhead calculations require additional energy, and enabling the communication devices to accomplish them may even be impractical.
+Such overhead calculations require additional energy, and enforcing the communication devices to accomplish them may even be impractical.
 
@@
 
-The SBT and SMT problems resemble the more frequently studied \textsc{Range Assignment Problem} (RAP) in the sense that they all ask for an undirected source-independent tree of minimum power.
+The SBT and SMT problems resemble the more frequently studied \textsc{Range Assignment Problem} (RAP) 
+in the sense that they all ask for an undirected source-independent tree of minimum power.
 However, RAP concerns cases where a connected topology must be established by means of \emph{bidirectional} links,
@@
 
-\textcolor{blue}{\subsection{Contributions}}
+\subsection{Contributions}
 
-Contributions from the present work include two sets of IP models for SMT, presented in order of increasing lower bounding capabilities.
+Contributions from the present work include two sets of Integer Programming (IP) models for SMT, presented in order of increasing lower bounding capabilities.
 By rigorous analysis of the models, we prove relations between the bounds offered by the corresponding continuous relaxations.
@@
 
+When proven optimality is out of reach, effective methods for computing near-optimal solutions are crucial.
+A heuristic method tailored for SMT is another contribution from the current text,
+which thus demonstrates how recent progress on the \textsc{Minimum Steiner Tree} problem can be transferred to the problem under study.
+
+\subsection{Scope and structure of the paper}
+
 For the same reasons that MEB has been generalized to MEM, it is natural and desirable to extend SBT to its multicast equivalent.
 The \textsc{Minimum Shared Multicast Tree} (SMT) problem is thus at the forefront of this paper,
-and neither MEB, MEM, nor RAP will be considered further.
-Focus is henceforth on instances in which some of the nodes neither initiate new messages nor need messages from others.
+and neither MEB, MEM, nor RAP will be given much attention.
+Focus is henceforth on instances in which some of the nodes neither initiate new messages nor need to receive messages from others.
 Utilizing such \emph{non-destination} nodes to forward messages can in many cases contribute to substantial reductions in the total power consumption.
@@
 
-\textcolor{blue}{\subsection{Structure of the paper}}
-
 The remainder of this paper is organized as follows: 
-\textcolor{blue}{Section \ref{sec:literature} summarizes relevant literature and previous works.}
-%Section \ref{sec:SBT} describes the SMT problem and gives detailed explanation of its objective function.
-\textcolor{blue}{Section \ref{sec:SBT} divided into three subsections introduces the notation, explains the network model, and gives formal definition of the problem.}
+Section \ref{sec:literature} summarizes relevant literature and previous works.
+Section \ref{sec:SBT} introduces the notation, explains the network model, and gives a formal definition of the problem.
 Integer linear programming formulations, valid inequalities and their analysis are presented in Section \ref{sec:ILP},
-followed by Section \ref{sec:comp}, where the models under study are compared with respect to strength.
+followed by Section \ref{sec:comp}, where the models under study are compared with respect to their strength.
 In Section \ref{sec:strong}, the models are further extended by variables in higher dimensional spaces,
 and a constraint generation procedure is developed.
+A fast method for computing good, but not necessarily optimal solutions is given in Section \ref{sec:heur}.
 Computational results are reported in Section \ref{sec:exp},
 and suggestions to future work are summarized in Section \ref{sec:conclusion}.
+\newline

### 2018-12-27 smt/sec2-literature.tex
MI: Remarks from minor revision incorporated - mostly typos and grammar.

@@
 
-The combinatorial optimization literature is already rich on Integer Programming (IP) approaches to power minimization in wireless networks.
+The combinatorial optimization literature is already rich on IP approaches to power minimization in wireless networks.
 Since the work of \citet{Wieseltier00onthe}, in particular the MEB and MEM problems have been studied extensively.
@@
 \citet{altinkemer05}, \citet{yuan05}, \citet{yuan08}, \citet{bauer08}, \citet{montemanni11} and \citet{leggieri08}. 
-\textcolor{blue}{
-The latter paper proposes an IP formulation of MEM based on a set covering model together with a preprocessing procedure reducing the number of constraints.}
+The latter paper proposes an IP formulation of MEM based on a set covering model, together with a preprocessing procedure reducing the number of constraints.
 \citet{cagalj02} prove that MEB, and thereby MEM, is NP-hard.
-Numerous contributions to fast, but possible inexact methods for MEM have thus appeared, and the interested reader is referred to \citep{hsiao13} for an overview. 
-\textcolor{blue}{Another study concerning MEM is the work by \cite{guo}, who investigate energy conservation while using adaptive antennas. 
-Unlike omnidirectional antennas, adaptive antennas can concentrate their transmission power in a smaller angle.}
+Numerous contributions to fast, but possibly inexact methods for MEM have thus appeared, and the interested reader is referred to \citep{hsiao13} for an overview. 
+Another study concerning MEM is the work by \cite{guo}, who investigate energy conservation while using adaptive antennas. 
+Unlike omnidirectional antennas, adaptive antennas can concentrate their transmission power in a smaller angle.
 
-\textcolor{blue}{\citet{bein10} study energy efficiency in wireless networks where all nodes can initiate a message, and the communication takes place in a single all-to-all broadcast tree.
-A heuristic algorithm generating all-to-all broadcast trees is proposed by \citet{bhukya14}, along with an experimental evaluation confirming its merit.}
+\citet{bein10} study energy efficiency in wireless networks where all nodes can initiate a message, and the communication takes place in a single all-to-all broadcast tree.
+A heuristic algorithm generating all-to-all broadcast trees is proposed by \citet{bhukya14}, along with an experimental evaluation confirming its merit.
 

### 2018-12-27 smt/sec3-sbtsmt.tex
MI (SMT): Minor changes in several files, mainly in Sect 8 - description of the results.
MI: some typos fixed
MI: Remarks from minor revision incorporated - mostly typos and grammar.

@@
 \label{sec:SBT}
+Like in studies of other problems in combinatorial optimization,
+a theoretical investigation of SMT requires abstraction and simplification of the real-world communication network.
 
@@
 
-Sending a message directly from (to) node $i$ to (from) node $j$ requires a certain \emph{transmission power} $p_{ij}=p_{ji}$.
+Sending a message directly from (to) node $i$ to (from) node $j$ requires a positive \emph{transmission power} $p_{ij}=p_{ji}$.
 Whenever appropriate, $p_{ij}$ is referred to as the \emph{cost} of $\{i,j\}$,
@@
 as the set of nodes requiring no less power at $i$ than $j$ does.
-The notation $G'\subseteq G$ means that $G'=(V_{G'},E_{G'})$ is a subgraph of $G$, i.e., $E_{G'}\subseteq E$ consists of the edges in $E$ with both end nodes in $V_{G'}\subseteq V$.
+The notation $G'\subseteq G$ means that $G'=(V_{G'},E_{G'})$ is a subgraph of $G$, 
+i.e., $E_{G'}\subseteq E$ consists uniquely of edges in $E$ with both end nodes in $V_{G'}\subseteq V$.
 
-Let $y \in \{0,1\}^E$ and $X\in\{0,1\}^{A\times D}$ be binary vectors with components corresponding to
-edges and pairs of arcs and destination nodes, respectively.
+Let $y \in \{0,1\}^E$ and $X\in\{0,1\}^{A\times D}$ be binary vectors with components corresponding to edges,
+and pairs of arcs and destination nodes, respectively.
 The vector $X^s$ then consists of all components of $X$ corresponding to a pair of which $s$ is the destination.
@@
 
-\textcolor{blue}{\subsection{Network model} \label{sec:netmod}}
+\subsection{Network model} \label{sec:netmod}
 
 To illustrate the problem under study, consider the tree depicted in Fig. \ref{fig:objexp}.
-Six destination nodes ($D=\{i_1,i_2,a,b,c,d\}$) are spanned by use of three Steiner nodes ($V\setminus D = \{e,i,i_3\}$).
+Seven destination nodes ($D=\{i_1,i_2,a,b,c,d,e\}$) are spanned by use of two Steiner nodes ($V\setminus D = \{i,i_3\}$).
 Consider node $i$ and its three neighbour nodes $i_1$, $i_2$ and $i_3$, ordered by decreasing power requirement at $i$.
 That is, $p_{ii_1}>p_{ii_2}>p_{ii_3}$.
-If the transmitting node is outside the shaded area ($i_2$, $c$, or $d$), then node $i$ relays the signal to two of its neighbors, \emph{including $i_1$}.
-Exploiting the wireless advantage, node $i$ reaches both neighbors if it is assigned power level $p_{ii_1}$.
+If the transmitting node is outside the shaded area ($i_2$, $c$, $d$, or $e$), then node $i$ relays the signal to two of its neighbors, \emph{including $i_1$}.
+Exploiting the wireless advantage, the signal from node $i$ reaches both neighbors if it is assigned power level $p_{ii_1}$.
 Otherwise, if the signal source is either $i_1$, $a$, or $b$, node $i$ passes the signal on to neighbors $i_2$ and $i_3$.
-This is accomplished by power $p_{ii_2}$.
+This is accomplished by setting the power level of $i$ to $p_{ii_2}$.
 
@@
   \node[fill=none] (i) at (4.5,1.5) {$i$};
-  \node[fill=none] (e) at (8,1.5) {$e$};
+  \node (e) at (8,1.5) {$e$};
  \end{scope}
@@
 
-\textcolor{blue}{
 The cost model outlined above aims for minimization of the transmission energy.
@@
 This is in line with the approach taken in the wide range of literature on optimization models for wireless networks mentioned in Section \ref{sec:literature}.
-For a detailed energy model for wireless sensor networks, the reader is referred to the article by \citet{halgamuge}.}
+For a detailed energy model for wireless sensor networks, the reader is referred to the article by \citet{halgamuge}.
 
@@
 Problem \ref{def:problem} also has close resemblance with the SBT problem, which is the special case of SMT where $D=V$.
-Like many other network design problems, \textsc{Minimum Shared Multicast Tree} is NP-hard.
+Like many other network design problems, SMT is NP-hard.
 This follows directly from the NP-hardness of SBT \citep{Papadimitriou06SBT}.

### 2018-12-27 smt/sec4-milp.tex
MI: Remarks from minor revision incorporated - mostly typos and grammar.

@@
 be the maximal path in $G_{X^s}$ that terminates in $t$.
-Existence and uniqueness of this path is assured since $t$ belongs to an acyclic connected component.
+Existence and uniqueness of this path are assured since $t$ belongs to an acyclic connected component.
 According to (\ref{con:dd:arrowFromNonDestA}), each non-destination in $P$ has an entering arc in $P$,
-while (\ref{con:dd:arrowFromDest}) ensures that the same is true for every destination node, except $s$.
+while constraints (\ref{con:dd:arrowFromDest}) ensure that the same is true for every destination node, except $s$.
 It follows that the first node of $P$ is $s$.
@@
 \begin{align}
-\label{objective:mf} \makebox[0pt][l]{$\displaystyle{} \min \sum\limits_{(i,j) \in A} p_{ij} \sum\limits_{s \in D} \pi^s_{ij} $}\\
+\label{objective:mf} \makebox[0pt][l]{$\displaystyle{} \min \sum\limits_{(i,j) \in A} p_{ij} \sum\limits_{s \in D} \pi^s_{ij}, $}\\
 \notag\text{s.t.}\\
-\notag(\ref{con:dd:arrowFromDest}) &- (\ref{con:dd:vardim})\\
+\notag(\ref{con:dd:arrowFromDest}) &- (\ref{con:dd:vardim}),\\
 \label{con:dd:yvar} X^s_{ij} & \leq \sum_{k\in W_{ij}}\pi^s_{ik} & s\in D, (i,j)\in A,\\
@@
 \begin{proof}
-Follows directly from $p\geq 0$ and Lemma \ref{lem:x0valid}.
+Follows directly from $p>0$ and Lemma \ref{lem:x0valid}.
 \qed
@@
 Their effect is to disallow non-destination leaves.
-Equations (\ref{con:vi:Y1}) say that in arborescence $T^s$, exactly one arc $(s,j)$ leaving the root $s$ is declared to be largest.
+Equations (\ref{con:vi:Y1}) say that in arborescence $T^s$, exactly one arc $(s,j)$ leaving the root $s$ is declared to be the largest.
 Validity of the constraints follows immediately because the root has at least one leaving arc.
@@
 Assume $(F,g)$ satisfies (\ref{con:pf1:xfrel})--(\ref{con:pf1:dim}).
-Then some connected component $T^0\subseteq G_{g}$ is an $s_0$-rooted arborescence spanning $D$,
+Then, some connected component $T^0\subseteq G_{g}$ is an $s_0$-rooted arborescence spanning $D$,
 and some connected component of $G_{F^s}\subseteq T^0$ is a path from $s_0$ to $s\in D_0$.
@@
 Assume $(F,g)$ satisfies (\ref{con:pf1:xfrel})--(\ref{con:pf1:dim}),
-and let $T^s=(V^s,A^s)$ be the arborescence obtained by redirecting the arcs in $T^0$ such that they point away from $s\in D_0$. Then
+and let $T^s=(V^s,A^s)$ be the arborescence obtained by redirecting the arcs in $T^0$ such that they point away from $s\in D_0$. 
+Then,
 $$
@@
 \begin{proof}
-Because $T^0$ is acyclic, it follows from the capacity constraints (\ref{con:pf1:xfrel}) that $g_{ij}+F^s_{ji}\leq 1$.
+Because $T^0$ is acyclic, $g_{ij}+g_{ji}\leq 1$, and it follows from the capacity constraints (\ref{con:pf1:xfrel}) that $g_{ij}+F^s_{ji}\leq 1$.
 The same constraints also imply $g_{ij} - F^s_{ij}\geq 0$.
@@
 Arborescence $T^s$ contains arc $(i,j)$ iff $(i,j)$ is in $A^0$ but not on a path to $s$, or its reverse arc $(j,i)$ is on such a path.
-By Lemma \ref{lem:modelcorrect}, this disjunction is equivalent to $g_{ij}-F_{ij}^s=1$ or $F_{ji}=1$.
+By Lemma \ref{lem:modelcorrect}, this disjunction is equivalent to $g_{ij}-F_{ij}^s=1$ or $F_{ji}^s=1$.
 The result then follows since the two conditions are mutually exclusive.
@@
 \begin{align}
-\notag\min & \sum\limits_{(i,j) \in A} p_{ij} \sum\limits_{s \in D} \pi^s_{ij}\\ 
+\notag\min & \sum\limits_{(i,j) \in A} p_{ij} \sum\limits_{s \in D} \pi^s_{ij},\\ 
 \notag\text{s.t.} && \\
-\notag(\ref{con:pf1:xfrel}) - (\ref{con:pf1:dim})\\
+\notag(\ref{con:pf1:xfrel}) - (\ref{con:pf1:dim}),\\
 \label{con:pf1:yvar} g_{ij} - F^s_{ij}+F^s_{ji} & \leq \sum_{k\in W_{ij}}\pi^s_{ik}  & s\in D_0, (i,j)\in A,\\
@@
 Since $i$ also has a leaving arc in $T^s$, we have $\pi_{ik}^s=1$ for some $k\neq i,s$.
-This proves validity of the constraint $\sum_{j\in V_i}\left(g_{ji} - F^s_{ji}+F^s_{ij}\right)=\sum_{j\in V_i\setminus\{s\}}\pi_{ij}^s$
+This proves validity of the constraint $\sum_{j\in V_i}\left(g_{ji} - F^s_{ji}+F^s_{ij}\right)=\sum_{j\in V_i\setminus\{s\}}\pi_{ij}^s$.
 Application of flow conservation (\ref{con:pf1:flow}) yields the valid constraints
@@
 With $s_0$ equal to the destination node in position (47, 65), the instance given in Remark \ref{rem:x2} also proves independence of the three sets
-(\ref{con:vi:Y1}), (\ref{con:vi:sumYImpSumXTrans})--(\ref{con:pf1:flowX}) of valid inequalities.
+(\ref{con:vi:Y1}), (\ref{con:pf1:flowX})--(\ref{con:vi:sumYImpSumXTrans}) of valid inequalities.
 \end{remark}

### 2018-12-27 smt/sec5-relations.tex
MI: Remarks from minor revision incorporated - mostly typos and grammar.

@@
 \begin{aligned}
- & g_{ij}+g_{ji} \leq 1 & \{i,j\}\in E. \label{eq:lemlpf1c}
+ & g_{ij}+g_{ji} \leq 1 & \text{ for all } \{i,j\}\in E. \label{eq:lemlpf1c}
 \end{aligned}
@@
 For $j\in D_0$ (for $j=s_0$), (\ref{eq:lemlpf2b}) follows directly from (\ref{con:pf1:fitt=xit}) (from (\ref{con:pf1:xi0=0})).
-Consider an $i\in D$ and $j\in V_i$ such that $g_{ij}>F_{ij}^*$.
+Consider an $i\in D$ and $j\in V_i\setminus D$ such that $g_{ij}>F_{ij}^*$.
 Reducing $g_{ij}$ to $F_{ij}^*$, while keeping all other variables unchanged, preserves feasibility and the objective function value.
@@
 Let $\tilde{g}_{j_lj_{l+1}}=g_{j_lj_{l+1}}-\sigma^*$ for $l=0,1,\ldots,p-1$, and let $\tilde{g}_{ij}=g_{ij}$ for all other arcs $(i,j)\in A$.
-Then $(F,\tilde{g},\pi)$ is feasible since, in comparison to $(F,g,\pi)$,  the left hand side of inequality (\ref{con:pf1:flowX}) is
+Then, $(F,\tilde{g},\pi)$ is feasible since, in comparison to $(F,g,\pi)$,  the left hand side of inequality (\ref{con:pf1:flowX}) is
 increased by no more than the current slack at node $i=j_0$, reduced at node $i=j_p$, and unchanged at all other nodes $i\in V\setminus D$.
 For all arcs $(i,j)$ between non-destination nodes, $\sigma_{ij}(F,\tilde{g})\leq\sigma_{ij}(F,g)$, except from arcs for which $i=j_p$.
-But since $(j_p,j)\not\in A[F,g]$, $\sigma_{j_pj}(F,g)>0$.
+But since $(j_p,j)\not\in A[F,g]$, we have that $\sigma_{j_pj}(F,\tilde{g})\leq\tilde{g}_{j_pj}-F^*_{j_pj}=g_{j_pj}-F^*_{j_pj}=0$.
+%$\sigma_{j_pj}(F,g)>0$.
 % and the capacity reduction introduces no new violations of (\ref{eq:lemlpf2a})--(\ref{eq:lemlpf2c}).
@@
 Assume first that $g_{ij}=F_{ij}^*$.
-Then there exists a destination node $t\in D_0\setminus\{i\}$ such that $F_{ij}^t=g_{ij}$.
+Then, there exists a destination node $t\in D_0\setminus\{i\}$ such that $F_{ij}^t=g_{ij}$.
 Because (\ref{eq:lemlpf1a}) gives $F_{ji}^t=0$, flow conservation (\ref{con:pf1:flow}) at node $i$ implies 
@@
 To prove (\ref{con:dd:arrowFromNonDestA}), let $i\in V\setminus D$, $j\in V_i$, and $s\in D$.
-Then transformations (\ref{eq:trangX0})--(\ref{eq:trangX}) yield
+Then, transformations (\ref{eq:trangX0})--(\ref{eq:trangX}) yield
 $X_{ij}^s - \sum_{k\in V_{i}\setminus \{j\}}X_{ki}^s

### 2018-12-27 smt/sec6-strong.tex
MI: Remarks from minor revision incorporated - mostly typos and grammar.

@@
 \begin{align}
-\notag\makebox[0pt][l]{$\displaystyle{} \min \sum\limits_{(i,j) \in A} p_{ij} \sum\limits_{s \in D} \pi^s_{ij} $}\\
+\notag\makebox[0pt][l]{$\displaystyle{} \min \sum\limits_{(i,j) \in A} p_{ij} \sum\limits_{s \in D} \pi^s_{ij}, $}\\
 \text{s.t.} \notag \\
-(\ref{con:dd:arrowFromDest}) - (\ref{con:dd:vardim}) &\text{ and }(\ref{con:dd:yvar}) - (\ref{con:vi:sumYImpSumX}) \notag\\
+(\ref{con:dd:arrowFromDest}) - (\ref{con:dd:vardim}) &\text{ and }(\ref{con:dd:yvar}) - (\ref{con:vi:sumYImpSumX}), \notag\\
 \label{con:mf:flowNormal} \sum\limits_{\substack{ j\in V_i}}x^{st}_{ij}-\sum\limits_{\substack{j\in V_i }}x^{st}_{ji} & = 0 & (s,t)\in A(D), i\in V\setminus\{s,t\},\\
@@
 The flow symmetry equations (\ref{con:mf:fsym}) state that the path from $s$ to $t$ contains $(i,j)$ iff the reverse path contains the reverse arc $(j,i)$.
-Finally, constraints (\ref{con:vi:sumFImpSumY}) reflect that if the path from $s$ to $t$ intersects $(i,j)$,
+Finally, constraints (\ref{con:vi:sumFImpSumY}) reflect the fact that if the path from $s$ to $t$ intersects $(i,j)$,
 then the most expensive arc leaving $i$ in $T^s$ has cost $p_{ij}$ or more.
@@
 Similarly to the extension $\mathcal{X}_3$ of $\mathcal{X}_2$ by $x$-variables, the model $\mathcal{F}_2$ is extended by variables with four node indices.
-Analogously to $A(D)$, let $E(D_0)=E\cap (D_0\times D_0)$ be the set of pairs of \emph{unordered} destination nodes other than $s_0$.
+Analogously to $A(D)$, let $E(D_0)=E\cap (D_0\times D_0)$ be the set of \emph{unordered} pairs of destination nodes other than $s_0$.
 Define variables $f\in\{0,1\}^{A\times E(D)}$, where
@@
 \begin{align}
-\notag\min & \sum\limits_{(i,j) \in A} p_{ij} \sum\limits_{s \in D} \pi^s_{ij}\\ \notag
+\notag\min & \sum\limits_{(i,j) \in A} p_{ij} \sum\limits_{s \in D} \pi^s_{ij},\\ \notag
 \text{s.t.}& && \\
-\notag(\ref{con:pf1:flow})-(\ref{con:vi:sumYImpSumXTrans}) & ~\text{and} ~(\ref{con:vi:Y1}) \\
+\notag(\ref{con:pf1:flow})-(\ref{con:vi:sumYImpSumXTrans}) & ~\text{and} ~(\ref{con:vi:Y1}), \\
 \label{con:pf2:flowHook} \sum\limits_{\substack{ j \in V_i }}\left(f^{st}_{ij}-f^{st}_{ji}\right) & \leq 
@@
 inequalities (\ref{con:pf2:stronger}) impose $f_{ij}^{st}=1$ if $F_{ij}^s=F_{ij}^t=1$.
-By virtue of (\ref{con:pf2:stopInDest}), (\ref{con:pf2:stronger}) replace the weaker capacity constraints (\ref{con:pf1:xfrel}).
+By virtue of (\ref{con:pf2:stopInDest}), constraints (\ref{con:pf2:stronger}) replace the weaker capacity constraints (\ref{con:pf1:xfrel}).
 Finally, (\ref{con:pf2:sumFImpSumY})--(\ref{con:pf2:sumFImpSumY2}) are valid inequalities justified by:
@@
 Assume $(F,f,g,\pi)$ satisfies (\ref{con:pf1:flow})--(\ref{con:vi:sumYImpSumXTrans}), (\ref{con:vi:Y1}), (\ref{con:pf2:flowHook})--(\ref{con:pf2:stronger}), and (\ref{con:pf2:dim}).
-Let $T^0\subseteq G_g$ be the arborescence spanning $D$ (see Lemma \ref{lem:modelcorrect}),
-and let $T^s$ be the arborescence obtained by redirecting the arcs in $T^0$ such that they point away from $s\in D$. Then, for all nodes $i$ and $j$ ($i\neq j$) spanned by $T^0$,
+Let $T^0\subseteq G_g$ be the corresponding arborescence spanning $D$ (see Lemma \ref{lem:modelcorrect}),
+and let $T^s$ be the arborescence obtained by redirecting the arcs in $T^0$ such that they point away from $s\in D$. 
+Then, for all nodes $i$ and $j$ ($i\neq j$) spanned by $T^0$,
 $$
@@
 \begin{algorithm}
-$Q\leftarrow\emptyset$\;
+$Q\leftarrow\emptyset$\\
 \Do {$\Delta Q\neq\emptyset$}{
-	$(X,y,\pi)\leftarrow\text{optimal solution to $\text{LP}(\mathcal{X}_2+Q)$}$\;
+	$(X,y,\pi)\leftarrow\text{optimal solution to $\text{LP}(\mathcal{X}_2+Q)$}$\\
 	\For{$\{s,t\}\in E(D)\setminus Q$} {
-		$v_{st}\leftarrow \text{optimal objective function value in $\mathcal{MF}$}$\;
+		$v_{st}\leftarrow \text{optimal objective function value in $\mathcal{MF}$}$\\
 	}
-	$Q'\leftarrow\left\{\{s,t\}\in E(D)\setminus Q: v_{st}<1\right\}$\;
-	Select some $\Delta Q\subseteq Q'$ such that $\Delta Q\neq\emptyset$ if $Q'\neq\emptyset$\;
-	$Q\leftarrow Q\cup\Delta Q$\;
+	$Q'\leftarrow\left\{\{s,t\}\in E(D)\setminus Q: v_{st}<1\right\}$\\
+	Select some $\Delta Q\subseteq Q'$ such that $\Delta Q\neq\emptyset$ if $Q'\neq\emptyset$\\
+	$Q\leftarrow Q\cup\Delta Q$\\
 }
@@
 we pursue the CG scheme as follows:
-In the graph $(D,Q')$, where edges correspond to pairs $\{s,t\}$ at which $v_{st}<1$, let the edge weights be $1-v_{st}$.
+In the graph $(D,Q')$, where $Q'=\{\{s,t\}\in E(D):v_{st}<1\}$,
+%edges correspond to pairs $\{s,t\}$ at which $v_{st}<1$, 
+let the edge weights be $1-v_{st}$.
 Let $\Delta Q$ be a matching of maximum weight in $(D,Q')$.
-Thus, for each $\{u,w\}\in E(D)\setminus Q$ where (\ref{con:mf:flowNormal})--(\ref{con:vi:sumFImpSumY}) are not satisfiable given the current values of $X$ and $\pi$,
+Thus, for each $\{u,w\}\in E(D)\setminus Q$ where constraints (\ref{con:mf:flowNormal})--(\ref{con:vi:sumFImpSumY}) are not satisfiable given the current values of $X$ and $\pi$,
 $Q$ is extended by an adjacent edge.

### 2018-12-27 smt/sec7-heur.tex
MI: Remarks from minor revision incorporated - mostly typos and grammar.

@@
-\textcolor{blue}{\section{Computing near-optimal solutions} \label{sec:heur}}
+\section{Computing near-optimal solutions} \label{sec:heur}
 
-\textcolor{blue}{
 Instances of size that prohibits computation of the optimal solution should be approached by some fast, but possibly inexact method.
 Imposing a time bound on an IP-solver, and applying it to any of the models introduced in Section \ref{sec:ILP}, readily gives such a method.
-Alternatively, a heuristic method dedicated to SMT can be developed.}
+Alternatively, a heuristic method dedicated to SMT can be developed.
 
-\textcolor{blue}{
 To the best of our knowledge, the scientific literature offers few heuristic methods for SMT besides those studied by \citet{ivanova16isco}.
@@
 If also $|\mathcal{T}|=\gamma$, the most expensive tree $T_{\gamma}$ is removed from $\mathcal{T}$.
-}
 
-\textcolor{blue}{
 Adapting the method by \citet{pajor18} to our problem is straightforward,
@@
 The adapted method is then described as follows, where $\mathcal{T}=\left(T_1,\ldots,T_{|\mathcal{T}|}\right)$
-and $c\left(T_1\right)\leq\dots\leq c\left(T_{|\mathcal{T}|}\right)$:}
+and $c\left(T_1\right)\leq\dots\leq c\left(T_{|\mathcal{T}|}\right)$:
 
 \begin{algorithm}
-\textcolor{blue}{
 \While{the time assigned to the heuristic is not expired} {
   \textbf{for} $\{i,j\}\in E$ \textbf{do} $\hat{p}_{ij}\leftarrow\text{random number uniformly distributed on
-  $\left[\frac{1}{2}p_{ij},\frac{3}{2}p_{ij}\right]$}$\;
-  $T\leftarrow\text{\textit{SMTHeur}$(G,D,\hat{p})$}$\;
-  \textbf{if} $|\mathcal{T}|<\gamma$ \textbf{then} $\mathcal{T}\leftarrow \mathcal{T}\cup \{T\}$\;
+  $\left[\frac{1}{2}p_{ij},\frac{3}{2}p_{ij}\right]$}$\\
+  $T\leftarrow\text{\textit{SMTHeur}$(G,D,\hat{p})$}$\\
+  \textbf{if} $|\mathcal{T}|<\gamma$ \textbf{then} $\mathcal{T}\leftarrow \mathcal{T}\cup \{T\}$\\
   \Else {
-    \textbf{if} $c(T)<c(T_{\gamma})$ \textbf{then} $\mathcal{T}\leftarrow \mathcal{T}\setminus\{T_{\gamma}\}$, $\mathcal{T}\leftarrow \mathcal{T}\cup \{T\}$\;
-    $r\leftarrow\text{random number uniformly distributed on $\{1,\ldots,\gamma\}$}$\;
-    \textbf{for} $\{i,j\}\in E_T\cap E_{T_r}$ \textbf{do} $\tilde{p}_{ij}\leftarrow p_{ij}$\;
+    \textbf{if} $c(T)<c(T_{\gamma})$ \textbf{then} $\mathcal{T}\leftarrow \mathcal{T}\setminus\{T_{\gamma}\}$, $\mathcal{T}\leftarrow \mathcal{T}\cup \{T\}$\\
+    $r\leftarrow\text{random number uniformly distributed on $\{1,\ldots,\gamma\}$}$\\
+    \textbf{for} $\{i,j\}\in E_T\cap E_{T_r}$ \textbf{do} $\tilde{p}_{ij}\leftarrow p_{ij}$\\
     \For{$\{i,j\}\in \left(E_T\cup E_{T_r}\right)\setminus \left(E_T\cap E_{T_r}\right)$}{
-      $\nu\leftarrow\text{random number uniformly distributed on $\{100,\ldots,500\}$}$\;
-      $\tilde{p}_{ij}\leftarrow \nu p_{ij}$\;
+      $\nu\leftarrow\text{random number uniformly distributed on $\{100,\ldots,500\}$}$\\
+      $\tilde{p}_{ij}\leftarrow \nu p_{ij}$\\
     }
-    \textbf{for} $\{i,j\}\in E\setminus E_T\setminus E_{T_r}$ \textbf{do} $\tilde{p}_{ij}\leftarrow 1000p_{ij}$\;
-    $\tilde{T}\leftarrow\text{\textit{SMTHeur}$(G,D,\tilde{p})$}$\;
-    \textbf{if} $c(\tilde{T})<c(T_{\gamma})$ \textbf{then} $\mathcal{T}\leftarrow \mathcal{T}\setminus\{T_{\gamma}\}$, $\mathcal{T}\leftarrow \mathcal{T}\cup \{\tilde{T}\}$\;
+    \textbf{for} $\{i,j\}\in E\setminus E_T\setminus E_{T_r}$ \textbf{do} $\tilde{p}_{ij}\leftarrow 1000p_{ij}$\\
+    $\tilde{T}\leftarrow\text{\textit{SMTHeur}$(G,D,\tilde{p})$}$\\
+    \textbf{if} $c(\tilde{T})<c(T_{\gamma})$ \textbf{then} $\mathcal{T}\leftarrow \mathcal{T}\setminus\{T_{\gamma}\}$, $\mathcal{T}\leftarrow \mathcal{T}\cup \{\tilde{T}\}$\\
   }
@@
 \Return $T_1$
-}
 
-\textcolor{blue}{
 \caption{Outline of the heuristic method}
 \label{alg:meta}
-}
 \end{algorithm}
 
-\textcolor{blue}{
 Following \citet{pajor18}, the perturbed cost $\hat{p}_{ij}$ of edge $\{i,j\}$ is put equal to a random number with expected value $p_{ij}$.
@@
 their cost is magnified by a factor dominating the factors drawn for edges that occur in one of the trees.
-}
 
-\textcolor{blue}{
 A few minor simplification over the approach taken by \citet{pajor18} are made in Alg.\ \ref{alg:meta}:
@@
 This contrasts the method in the cited work, where $\tilde{T}$ is chosen as the best tree found in a sequence of merges.
-} \\

### 2018-12-27 smt/sec8-exp.tex
MI: some typos fixed
MI: Remarks from minor revision incorporated - mostly typos and grammar.

@@
 
-\textcolor{blue}{\subsection{Comparison with alternative methods}}
-\textcolor{blue}{
+\subsection{Comparison with alternative methods}
 We now investigate the ability of IP models $\mathcal{X}_1$ and $\mathcal{F}_1$ to produce good feasible solutions when a time limit is imposed.
-We compare the quality of obtained solutions with those calculated by Alg. \ref{alg:meta}.
+We compare the quality of obtained solutions with those calculated by Alg.\ \ref{alg:meta}.
 The results for time restrictions 5, 10 and 20 CPU-minutes, respectively, are reported in Tab. \ref{tab:metamod}.
 These experiments are carried out on the instance sets that are investigated in the experiments reported in Section \ref{sec:expsmall}.
-}
 
@@
 \renewcommand{\arraystretch}{1.4} % Default value: 1
-\textcolor{blue}{
 \begin{tabular}{rrrrrrrrrrr}
@@
 \end{tabular}
-}
-\caption{\textcolor{blue}{Comparison of the metaheuristic algorithm (Alg.\ \ref{alg:meta}) and interrupted B\&B applied to IP models $\mathcal{X}_1$ and $\mathcal{F}_1$}}
+\caption{Comparison of the metaheuristic algorithm (Alg.\ \ref{alg:meta}) and interrupted B\&B applied to IP models $\mathcal{X}_1$ and $\mathcal{F}_1$}
 \label{tab:metamod}
@@
 
-\textcolor{blue}{
 For a given instance, we define the relative performance of a method as the objective function value obtained by the method relative to the best performing method in that instance.
@@
 Thus, the lowest value in a given instance set indicates the best performing method.
-}
 
-\textcolor{blue}{
 Table \ref{tab:metamod} reveals that Alg.\ \ref{alg:meta} gives the best average results for all combinations of time limits and most instance set.
@@
 Although Alg.\ \ref{alg:meta} fails to prove optimality, it appears to be a better procedure than B\&B for computing near-optimal solutions within a given time bound.
-}
 \newline
@@
 As the instance size increases, exact solution of the SMT problem becomes unrealistic.
-To assess the quality of fast and possible inexact methods, \textcolor{blue}{such as Alg.\ \ref{alg:meta}, 
-or the construction methods by \citep{ivanova16isco}},  access to tightest possible lower bounds is crucial.
+To assess the quality of fast and possible inexact methods, such as Alg.\ \ref{alg:meta}, 
+or the construction methods by \citep{ivanova16isco},  access to tightest possible lower bounds is crucial.
 In the next set of experiments, we compare two different methods for computing the bounds.
@@
 Instances with $|V|=2|D|\in\{24, 26,\ldots, 34\}$ are investigated, with 5 instances for each value of $|V|$.
-The results are illustrated in Fig. \ref{fig:cggr}.
+The results are illustrated in Fig.\ \ref{fig:cggr}.
 Each curve represents the progress of the lower bound within each instance set.
@@
 The vertical axes correspond to the lower bounds as a percentage of $z\left(\text{LP}(\mathcal{F}_1)\right)$, averaged over all instances in the set.
-\textcolor{blue}{
 The tightness of the bounds increases with time, and it can be observed from the graphs that the CG bound becomes tighter at some point.
@@
 After that point, the bound grows rather slowly.
-	In contrast, CG produces a bound that grows faster even after it has reached the value $z\left(\text{LP}(\mathcal{X}_1)\right)$.}
+In contrast, CG produces a bound that grows faster even after it has reached the value $z\left(\text{LP}(\mathcal{X}_1)\right)$.
 
@@
   \caption{Comparison of lower bounds obtained by CG and B\&B on model $\mathcal{F}_1$.
-	\textcolor{blue}{Each figure depicts the progress of lower bounds for instances with varying number of destination and non-destination nodes.
+	Each figure depicts the progress of lower bounds for instances with varying number of destination and non-destination nodes.
 	The lower bounds are obtained by applying the CG scheme to model $\mathcal{F}_1$ (black curves), and by running the B\&B algorithm (grey curves), respectively.
-	}}
+	}
   \label{fig:cggr}

### 2018-12-27 smt/sec9-conclusion.tex
MI: Remarks from minor revision incorporated - mostly typos and grammar.

--- 2018-08-30 smt/sec7-conclusion.tex
@@
 We have presented and analyzed two approaches of modelling the SMT problem as an integer linear program.
-The first formulation, $\mathcal{X}_1$, is based on a formulation of the broadcast version of the problem, while the second one, $\mathcal{F}_1$, is an extension of network flow formulation of the minimum Steiner tree problem.
+The first formulation, $\mathcal{X}_1$, is based on a formulation of the broadcast version of the problem, while the second one,
+$\mathcal{F}_1$, is an extension of network flow formulation of the minimum Steiner tree problem.
 Both formulations are improved by valid inequalities resulting in models $\mathcal{X}_2$ and $\mathcal{F}_2$, respectively. 
@@
 A search for facet-defining valid inequalities, or even convex hull formulations, is an interesting and ambitious direction of future work.
-Owing to the limited size of solvable instances, there is a potential in studying inexact methods, such as approximation algorithms and domain specific heuristics.
+Owing to the limited size of solvable instances, there is a potential in studying inexact methods, such as approximation algorithms and problem specific heuristics.
+In particular, experiments on a simple metaheuristic documented in the current work, 
+suggest that fast and inexact methods have a potential to compute solutions close to optimality in large instances.
 Instance preprocessing combined with variable fixing can possibly increase the size of instances solvable to optimality with the current models.

### 2018-12-28 smt/sec3-sbtsmt.tex
mainly sect 7 rewritten

@@
 For multicasting a message, the power to be assigned to node $i$ is either $p_{ii_1}$ or $p_{ii_2}$, where $i_1$ and $i_2$ are neighbors of $i$ such that
-$p_{ii_1}>p_{ii_2}>p_{ij}$ for all other neighbors $j\in V_T\setminus\{i_1,i_2\}$.
+$p_{ii_1}>p_{ii_2}>p_{ij}$ for all other neighbors $j\neq i_1,i_2$ of $i$ in $V_T$.
 If the path from the message source to $i$ intersects $i_1$, then power $p_{ii_2}$ is sufficient.

### 2018-12-28 smt/sec7-heur.tex
mainly sect 7 rewritten

--- 2018-09-26 smt/sec7-heur.tex
@@
 The main algorithmic components in the most basic heuristic are a construction algorithm, a local improvement algorithm,
-and an algorithm for merging two Steiner trees in order to create a better one.
-Construction and local improvement are applied in order to generate Steiner trees of presumably low cost.
-If the cost is sufficiently low, the tree is stored in an ordered set $\mathcal{T}$ of at most $\gamma$ distinct trees.
-Diversity among the trees is achieved by perturbing the edge costs randomly before construction and local improvement are applied.
-After each tree generation, the new tree is merged with a randomly chosen tree from $\mathcal{T}$, 
-and the resulting tree is kept in $\mathcal{T}$ if its cost is smaller than some other tree already in $\mathcal{T}$.
-If also $|\mathcal{T}|=\gamma$, the most expensive tree $T_{\gamma}$ is removed from $\mathcal{T}$.
-}
+and an algorithm for merging two Steiner trees.
 
-\textcolor{blue}{
-Adapting the method by \citet{pajor18} to our problem is straightforward,
-given the recently published construction and local improvement algorithms for SMT \citep{ivanova16isco}.
-Let \textit{SMTHeur}$(G,D,p)$ denote the output produced when $(G,D,p)$ is input to the construction algorithm,
-and the constructed tree is input to the local improvement method.
-The adapted method is then described as follows, where $\mathcal{T}=\left(T_1,\ldots,T_{|\mathcal{T}|}\right)$
-and $c\left(T_1\right)\leq\dots\leq c\left(T_{|\mathcal{T}|}\right)$:}
+A set $\mathcal{T}=\left(T_1,\ldots,T_{|\mathcal{T}|}\right)$ of at most $\gamma$ distinct Steiner trees,
+ordered by non-decreasing costs, is maintained throughout the heuristic \citep{pajor18}.
+Construction and local improvement are applied in order to generate a new tree $T$ of presumably low cost.
+Diversity in $\mathcal{T}$ is achieved by perturbing the edge costs randomly before construction and local improvement are applied.
+If $\left|\mathcal{T}\right|<\gamma$, $T$ is included in $\mathcal{T}$.
+If $\left|\mathcal{T}\right|=\gamma$, and the cost of $T$ is below the cost of $T_{\gamma}$, $T_{\gamma}$ is replaced by $T$ in $\mathcal{T}$.
+Another Steiner tree, $\tilde{T}$, is generated by merging $T$ with a randomly chosen $T_r\in\mathcal{T}$ ($r=1,\ldots,\gamma$),
+and $\tilde{T}$ replaces $T_{\gamma}$ in $\mathcal{T}$ if its cost is smaller.
 
 \begin{algorithm}
-\textcolor{blue}{
 \While{the time assigned to the heuristic is not expired} {
@@
 \Return $T_1$
-}
 
-\textcolor{blue}{
 \caption{Outline of the heuristic method}
 \label{alg:meta}
-}
 \end{algorithm}
 
-\textcolor{blue}{
+Taking advantage of the recently published construction and local improvement algorithms for SMT \citep{ivanova16isco},
+Alg.\ \ref{alg:meta} shows how the method by \citet{pajor18} is adapted to our problem.
+The procedure \textit{SMTHeur}$(G,D,p)$ produces a multicast tree $T$ by first applying the construction algorithm to the input $(G,D,p)$,
+and next applying local improvement to the constructed tree.
 Following \citet{pajor18}, the perturbed cost $\hat{p}_{ij}$ of edge $\{i,j\}$ is put equal to a random number with expected value $p_{ij}$.
-Further, in the merge process, the adjusted weights $\tilde{p}_{ij}$ are determined in order to 
-promote edges that appear in both $T$ and the randomly selected tree $T_{\gamma}$ in $\mathcal{T}$. 
+Further, in the merge process, adjusted weights $\tilde{p}_{ij}$ are determined in order to 
+promote edges that appear in both $T$ and the randomly selected tree $T_r\in\mathcal{T}$. 
 The cost of edges appearing in exactly one of the trees are thus magnified.
@@
 their cost is magnified by a factor dominating the factors drawn for edges that occur in one of the trees.
-}
 
-\textcolor{blue}{
 A few minor simplification over the approach taken by \citet{pajor18} are made in Alg.\ \ref{alg:meta}:
 Rather than selecting the tree to leave $\mathcal{T}$ randomly, we consistently remove $T_{\gamma}$ when a new tree is added.
-Also, for each new tree $T$ constructed, Alg.\ \ref{alg:meta} makes only one attempt to merge it with some $T_r$ stored in $\mathcal{T}$.
+Also, for each new tree $T$ constructed, Alg.\ \ref{alg:meta} makes only one attempt to merge it with some $T_r\in\mathcal{T}$.
 This contrasts the method in the cited work, where $\tilde{T}$ is chosen as the best tree found in a sequence of merges.
-} \\

### 2018-12-28 smt/sec8-exp.tex
MI: final updates: Explanation of metaheuristic improved, removed semicolons from pseudocodes, some improved text formulations in the description of the experiments.

@@
 Instances of given number of nodes and destinations are generated with random coordinates uniformly distributed on the square with corners at [0, 0] and [100, 100].
-A time limit of 20 CPU-minutes is imposed on each run.
+All edge costs equal the square of the distance between the end nodes.
+A time limit of 20 CPU-minutes is imposed on each run, unless otherwise is stated.
 All experiments are run on an Intel Core 2 Quad CPU at 2.83 GHz and 8 GB RAM.
@@
 The 20 CPU-minutes time restriction manifests itself when applying the strongest LP relaxations to some instances.
-Solving $\text{LP}(\mathcal{F}_k), (k=1,2,3)$ is consistently more time-consuming than solving $\text{LP}(\mathcal{X}_k)$, which demonstrates a trade-off between strength and running time.  
-Moreover, the running times of $\text{LP}(\mathcal{X}_3)$ and $\text{LP}(\mathcal{F}_3)$ are, in instances of the actual size, considerably longer than the time it takes to solve $\mathcal{F}_1$ to optimality using branch-and-bound (B\&B).
+Solving $\text{LP}(\mathcal{F}_k), (k=1,2,3)$ is consistently more time-consuming than solving $\text{LP}(\mathcal{X}_k)$, 
+which demonstrates a trade-off between strength and running time.  
+Moreover, the running times of $\text{LP}(\mathcal{X}_3)$ and $\text{LP}(\mathcal{F}_3)$ are, in instances of the actual size, 
+considerably longer than the time it takes to solve $\mathcal{F}_1$ to optimality (see the last column) using branch-and-bound (B\&B).
 This implies that both $\text{LP}(\mathcal{X}_3)$ and $\text{LP}(\mathcal{F}_3)$ are impractical in their basic form.
@@
 We investigate what is the maximum size of instances that can be solved to optimality within 20 CPU-minutes using models $\mathcal{X}_1$ and $\mathcal{F}_1$.
-The experiments compare sets of instances with different $|V|/|D|$ ratios.
+The experiments compare sets of instances with different values of $|V|/|D|$.
 For instances of the size given in columns 1--2, Table \ref{tab:small_inst} shows the average running time in CPU-seconds (columns 3--4),
@@
 Furthermore, when instances are not solved to optimality, $\mathcal{F}_1$ leaves a tighter optimality gap than does $\mathcal{X}_1$.
-Note that instances with lower $|V|/|D|$ ratio are more difficult to solve than those with higher $|V|/|D|$.
+Note that the lower the value of $|V|/|D|$, the more difficult are the instances to solve.
 This applies to both $\mathcal{F}_1$ and $\mathcal{X}_1$.
@@
 \end{tabular}
-\caption{Comparison of the metaheuristic algorithm (Alg.\ \ref{alg:meta}) and interrupted B\&B applied to IP models $\mathcal{X}_1$ and $\mathcal{F}_1$}
+\caption{Comparison between the metaheuristic algorithm (Alg.\ \ref{alg:meta}) and interrupted B\&B applied to IP models $\mathcal{X}_1$ and $\mathcal{F}_1$}
 \label{tab:metamod}
@@
 we report the relative performances of $\mathcal{X}_1$, $\mathcal{F}_1$, and Alg.\ \ref{alg:meta}, averaged over all instances in the set.
-For example, if all methods find the optimum in every instance, the three corresponding values are all 100.0\%.
+For example, if all methods find an optimum in every instance, the three corresponding values are all 100.0\%.
 Thus, the lowest value in a given instance set indicates the best performing method.
 
-Table \ref{tab:metamod} reveals that Alg.\ \ref{alg:meta} gives the best average results for all combinations of time limits and most instance set.
-Even when the time bound is set beyond what is needed to compute the optimal solution by submitting $\mathcal{X}_1$ or $\mathcal{F}_1$ to B\&B (see Tab.\ \ref{tab:small_inst}),
-this observation applies.
-Although Alg.\ \ref{alg:meta} fails to prove optimality, it appears to be a better procedure than B\&B for computing near-optimal solutions within a given time bound.
+Table \ref{tab:metamod} reveals that Alg.\ \ref{alg:meta} gives the best average results for all combinations of time limit and instance set.
+Even when the time bound is set beyond what is needed to compute an optimal solution by submitting or $\mathcal{F}_1$ to B\&B (see Tab.\ \ref{tab:small_inst}),
+Alg.~\ref{alg:meta} produces no worse solutions, and is thus able to identify an optimal solution in each considered instance.
+Although Alg.\ \ref{alg:meta} fails to prove optimality in general, it appears to be a better procedure than B\&B for computing near-optimal solutions within a given time bound.
 \newline
@@
 To assess the quality of fast and possible inexact methods, such as Alg.\ \ref{alg:meta}, 
-or the construction methods by \citep{ivanova16isco},  access to tightest possible lower bounds is crucial.
-In the next set of experiments, we compare two different methods for computing the bounds.
+or the construction methods by \cite{ivanova16isco},  access to tightest possible lower bounds is crucial.
+In the next set of experiments, 
 Section \ref{sec:expcomplp} indicates that $z\left(\text{LP}(\mathcal{X}_3)\right)$ is a strong bound,
@@
 After that point, the bound grows rather slowly.
-In contrast, CG produces a bound that grows faster even after it has reached the value $z\left(\text{LP}(\mathcal{X}_1)\right)$.
+In contrast, CG produces a bound that grows faster, even after it has reached the value $z\left(\text{LP}(\mathcal{X}_1)\right)$.
 
@@
 While solution of $\text{LP}(\mathcal{F}_1)$ is still in progress, the growth of the lower bound is faster than what is observed for CG (solution of $\text{LP}(\mathcal{F}_1)$). 
-This is particularly apparent for instances of size 28 and larger (Fig. \ref{fig:cggr28-14} - \ref{fig:cggr34-17}). %and can be explained as a consequence of Prop. \ref{prop:f1strx1}.
+This is particularly apparent for instances of size 28 and beyond (Fig. \ref{fig:cggr28-14} - \ref{fig:cggr34-17}). %and can be explained as a consequence of Prop. \ref{prop:f1strx1}.
 Once $\text{LP}(\mathcal{F}_1)$ is solved, the increase slows down substantially, and remains modest until the time limit is reached. 

### 2018-12-28 smt/response2.tex
Response letter to the 2nd revision.

\documentclass[12pt]{article}
\usepackage{a4wide,url}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{color}

\begin{document}
\section*{Response Letter}
We are grateful to the editor and the reviewers for their time, effort, and constructive remarks in the review process.
We are also glad that the reviewers are satisfied with our first revision, and that they recommend acceptance conditioned upon correction of typos.

The manuscript has been revised for a second time, according to the reviewers' suggestions:
We have proofread it carefully, with particular attention to English grammar and spelling, and made a few corrections.
We hope that the editor finds the revision version suitable for publication in \emph{Journal of Combinatorial Optimization}.

\end{document}
